---
title: "Genomic Data Security"
author: "Yongjin Park"
date: "`r format(Sys.time(), '%d %B %Y')`"
classoption: "aspectratio=169"
fontsize: 12pt
bibliography: genomic_security.bib
output:
    beamer_presentation:
        theme: Madrid
        keep_md: true
        keep_tex: true
        latex_engine: pdflatex
        slide_level: 2
header-includes:
  - \usepackage{tikz}
  - \usetikzlibrary{decorations.pathreplacing}
  - \usepackage{cancel}
  - \definecolor{UBCblue}{rgb}{0.04706, 0.13725, 0.26667}
  - \usecolortheme[named=UBCblue]{structure}
  - \setbeamertemplate{frametitle}{\color{UBCblue}\bfseries\insertframetitle\par\vskip-6pt\hrulefill}
  - \setbeamercolor{title in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{section in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{author in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{date in head/foot}{bg=white,fg=gray}
  - \setbeamertemplate{page number in head/foot}{}
  - \setbeamertemplate{frame numbering}[none]
  - \setbeamercolor{alerted text}{bg=yellow}
  - \AtBeginSection[]{\begin{frame}\frametitle{Today's lecture}{\Large\tableofcontents[currentsection]}\end{frame}}
  - |
    \makeatletter
    \def\ps@titlepage{%
      \setbeamertemplate{footline}{}
    }
    \addtobeamertemplate{title page}{\thispagestyle{titlepage}}{}
    \makeatother
---


```{r setup, include=FALSE}
source("Setup.R")
fig.dir <- "../Fig/genomic-data-security/"
dir.create(fig.dir, showWarnings=FALSE)
setup.env(fig.dir)
dir.create("../data", showWarnings=FALSE)

library(data.table)
library(tidyverse)
library(patchwork)
library(matrixTests)

theme_set(theme_minimal())
```

# {.plain}

\centerline{\includegraphics[height=.8\textheight]{img/privacy_opinions.png}}

\vfill
\tiny
\flushright

[https://xkcd.com/1269/](https://xkcd.com/1269/)

# Genetic Data Utility vs. Privacy

##

:::::: {.columns}
::: {.column width=.2}

:::
::: {.column width=.6}

\centerline{\includegraphics[width=\textwidth]{img/nyt_genetic_sleuth-1.png}}

:::
::: {.column width=.2}

:::
::::::

\vfill

With the help of a genomic database...

\vfill
\tiny

[https://www.nytimes.com/2018/10/15/science/gedmatch-genealogy-cold-cases.html](https://www.nytimes.com/2018/10/15/science/gedmatch-genealogy-cold-cases.html)

## Math based on (random) DNA recombination

:::::: {.columns}
::: {.column width=.2}

:::
::: {.column width=.5}

\centerline{\includegraphics[width=\textwidth]{img/nyt_genetic_sleuth-2.png}}

:::
::: {.column width=.2}

:::
::::::

\vfill
\tiny

[https://www.nytimes.com/2018/10/15/science/gedmatch-genealogy-cold-cases.html](https://www.nytimes.com/2018/10/15/science/gedmatch-genealogy-cold-cases.html)

## The Golden State Killer

\large

- **Joseph James DeAngelo** arrested in 2018 after 40+ years

- At least 13 murders, 50+ rapes across California (1970s-80s)

- First major cold case solved using Investigative Genetic Genealogy

## Back story: How the case was solved

\vfill

\only<1>{\centerline{\includegraphics[width=\textwidth]{img/gedmatch.png}}}

\large

\begin{enumerate}
\item<2-> Crime scene DNA uploaded to {\color{teal}\tt\bf GEDmatch.com}
\item<3-> Found matches to {\color{magenta} distant relatives} (not the suspect)
\item<4-> Reconstructed family trees back 5-8 generations
\item<5> Narrowed down the pool of suspects to a handful of living individuals
\item<5> Confirmed by collecting discarded DNA (coffee cup) from DeAngelo
\end{enumerate}

\normalsize

## {.plain}

:::::: {.columns}
::: {.column width=.1}

:::
::: {.column width=.8}

\centerline{\includegraphics[width=\textwidth]{img/nyt-goldenstate-killer-1.png}}

:::
::: {.column width=.1}

:::
::::::

\vfill
\tiny

[https://www.nytimes.com/2018/10/15/science/gedmatch-genealogy-cold-cases.html](https://www.nytimes.com/2018/10/15/science/gedmatch-genealogy-cold-cases.html)

## {.plain}

\begin{tikzpicture}[remember picture,overlay]
\fill[black] (current page.south west) rectangle (current page.north east);
\end{tikzpicture}

\color{white}

\centerline{\includegraphics[width=.95\textwidth]{img/nyt-goldenstate-killer-2.png}}

\vfill
\tiny

\url{https://www.nytimes.com/2019/04/25/us/golden-state-killer-dna.html}

## A tale from the other side of the GSK case

:::::: {.columns}
::: {.column width=.35}

\centerline{\includegraphics[width=\textwidth]{img/erlich-1.pdf}}

:::
::: {.column width=.6}

\large

- 60% success with database covering just 2% of population

- As databases grow, nearly anyone can be identified

- Privacy erosion inevitable?

:::
::::::

\tiny
\vfill

@Erlich2018-es


## Key security breach is to find someone in DB

\centerline{\includegraphics[width=\textwidth]{img/erlich-2.pdf}}

\tiny
\vfill

@Erlich2018-es


## Genome-Wide Association Studies data could...

:::::: {.columns}
::: {.column width=.4}

\centerline{\includegraphics[width=\textwidth]{img/erlich-2b.pdf}}

:::
::: {.column width=.6}

\large

- GWAS summary statistics

- What if we can "look up" (or predict) phenotypes?

- Public genealogical records amplify genetic databases.

:::
::::::

\tiny

@Erlich2018-es

## Narrowing down: from genealogy to individual

:::::: {.columns}
::: {.column width=.3}

\centerline{\includegraphics[width=\textwidth]{img/erlich-3.pdf}}

:::
::: {.column width=.5}

- Demographic identifiers narrow search: genealogy match $\rightarrow$ age/location $\rightarrow$ individual

- We can also narrow down to the most genomically informative locations...

${}$

\onslide<2>{\centerline{\includegraphics[width=\textwidth]{img/idb_wiki.png}}}

\tiny

[https://en.wikipedia.org/wiki/Identity_by_descent](https://en.wikipedia.org/wiki/Identity_by_descent)

:::
::::::

\tiny
\vfill

@Erlich2018-es

## Real Cases: From DNA to Identification

\centerline{\includegraphics[width=\textwidth]{img/erlich-4.pdf}}


\vfill
\tiny

@Erlich2018-es

## {.plain}

\centerline{\includegraphics[width=\textwidth]{img/nyt-23andme-1.png}}


## {.plain}

\centerline{\includegraphics[width=\textwidth]{img/nyt-23andme-2.png}}

- 15 million customers' genetic data $\approx$ \$ 256M?

## {.plain}

\centerline{\includegraphics[width=\textwidth]{img/nyt-23andme-3.png}}


## Privacy Concerns in the midst of corporate dynamics

\large

- Genetic data could be sold as an asset to pay creditors

- Should genetic information have special protections in bankruptcy?

- Commercial partnerships with pharma companies for drug discovery

- Who owns data?

## {.plain}

\centerline{\includegraphics[width=\textwidth]{img/23andme-regeneron.png}}

## How genetic data can compromise privacy

\centerline{\includegraphics[width=\textwidth]{img/gursoy-2.pdf}}

\vfill
\tiny

@Gursoy2022-od


## We may be able to infer genotypes from ...

\centerline{\includegraphics[height=.75\textheight]{img/gursoy-3.pdf}}

\vfill
\tiny
\flushright

@Gursoy2022-od

## We could gain more information than we intended

:::::: {.columns}
::: {.column width=.6}

\centerline{\includegraphics[width=\textwidth]{img/gursoy-4.pdf}}

:::
::: {.column width=.35}

* Single Nucleotide Variants

* Although we didn't want to know SNVs, we would be able to infer them.

* Almost every genomic technology involves "sequencing."

:::
::::::

\tiny

@Gursoy2022-od

## Knowing what the genotypes are ... could open an "unprecedented" possibility?

:::::: {.columns}
::: {.column width=.7}

\centerline{\includegraphics[width=\textwidth]{img/gursoy-5.pdf}}

:::
::: {.column width=.25}

G $\to$ ?

:::
::::::

\tiny

@Gursoy2022-od

## After all this... are we safe?

\centerline{\includegraphics[width=\textwidth]{img/gursoy-6.pdf}}

\tiny
\vfill

@Gursoy2022-od

## Many outstanding questions...

\large

- Should we write "no technology" manifesto like "Unabomber?"

- What if an insurance company attempts to "predict?"

- What if we can link them to social media?

\vfill

1. How genomic data security can be breached

2. How to prevent unwanted information leakage

# Membership inference



## Membership attack in genomic data

:::::: {.columns}
::: {.column width=.1}

:::
::: {.column width=.8}

\centerline{\includegraphics[width=\textwidth]{img/homer_2008_pgen.g001.png}}

:::
::: {.column width=.1}

:::
::::::

\tiny

@Homer2008-mn


## Basic idea 1: Membership inference

\begin{tikzpicture}[>=stealth, font=\normalsize]
% SNP j
\node[left] at (-0.8, 2) {\textbf{SNP $j$}};
\draw[thick] (0,2) -- (9,2);
\foreach \x/\lab in {0/0, 2.25/0.25, 4.5/0.50, 6.75/0.75, 9/1.0}
\draw (\x,1.9) -- (\x,2.1) node[above,font=\scriptsize] {\lab};
\fill[blue!70] (2.25,2) circle (3.5pt) node[below=3pt,font=\scriptsize] {Pop};
\fill[red!70] (6.75,2) circle (3.5pt) node[below=3pt,font=\scriptsize] {$M$};
\fill[black] (9,2) circle (3.5pt) node[below=3pt,font=\scriptsize] {$Y$};
\draw[blue!70,<->,thick] (2.25,1.45) -- node[below,font=\scriptsize] {0.75} (9,1.45);
\draw[red!70,<->,thick] (6.75,1.55) -- node[above,font=\scriptsize] {0.25} (9,1.55);
\node[right] at (9.3, 2) {\color{red!70} closer to $M$?};

% SNP j+1
\node[left] at (-0.8, 0) {\textbf{SNP $j\!+\!1$}};
\draw[thick] (0,0) -- (9,0);
\foreach \x/\lab in {0/0, 2.25/0.25, 4.5/0.50, 6.75/0.75, 9/1.0}
\draw (\x,-0.1) -- (\x,0.1) node[above,font=\scriptsize] {\lab};
\fill[blue!70] (2.25,0) circle (3.5pt) node[below=3pt,font=\scriptsize] {Pop};
\fill[red!70] (6.75,0) circle (3.5pt) node[below=3pt,font=\scriptsize] {$M$};
\fill[black] (4.5,0) circle (3.5pt) node[below=3pt,font=\scriptsize] {$Y$};
\draw[blue!70,<->,thick] (2.25,-0.5) -- node[below,font=\scriptsize] {0.25} (4.5,-0.5);
\draw[red!70,<->,thick] (4.5,-0.45) -- node[above,font=\scriptsize] {0.25} (6.75,-0.45);
\node[right] at (9.3, 0) {\color{gray} equal};

% SNP j+2
\node[left] at (-0.8, -2) {\textbf{SNP $j\!+\!2$}};
\draw[thick] (0,-2) -- (9,-2);
\foreach \x/\lab in {0/0, 2.25/0.25, 4.5/0.50, 6.75/0.75, 9/1.0}
\draw (\x,-2.1) -- (\x,-1.9) node[above,font=\scriptsize] {\lab};
\fill[blue!70] (2.25,-2) circle (3.5pt) node[below=3pt,font=\scriptsize] {Pop};
\fill[red!70] (6.75,-2) circle (3.5pt) node[below=3pt,font=\scriptsize] {$M$};
\fill[black] (0,-2) circle (3.5pt) node[below=3pt,font=\scriptsize] {$Y$};
\draw[blue!70,<->,thick] (0,-2.5) -- node[below,font=\scriptsize] {0.25} (2.25,-2.5);
\draw[red!70,<->,thick] (0,-2.45) -- node[above,font=\scriptsize] {0.75} (6.75,-2.45);
\node[right] at (9.3, -2) {\color{blue!70} closer to Pop?};
\end{tikzpicture}

\medskip
\centering \textbf{Q: Is the person of interest in the mixture pool?}

## Basic idea 2: Membership inference

\begin{tikzpicture}[>=stealth, font=\normalsize]
% Show accumulation across many SNPs
\node[left] at (-0.5, 1.5) {\textbf{Per SNP:}};
\node[right] at (0, 1.5) {$D_j = |Y_j - \text{Pop}_j| - |Y_j - M_j|$};

% Visual bar chart of D_j values
\node[left] at (-0.5, 0.3) {\textbf{Sum:}};
\foreach \x/\h/\col in {0.3/0.6/red!70, 1.1/-0.3/blue!70, 1.9/0.4/red!70, 2.7/0.5/red!70, 3.5/-0.2/blue!70, 4.3/0.3/red!70, 5.1/-0.4/blue!70, 5.9/0.7/red!70, 6.7/0.2/red!70, 7.5/-0.1/blue!70, 8.3/0.5/red!70, 9.1/0.3/red!70}
\fill[\col] (\x,-0.3) rectangle (\x+0.5, -0.3+\h);
\draw[thick] (-0.1,-0.3) -- (9.8,-0.3);
\node[font=\scriptsize] at (4.8, -1) {SNP $1, 2, \ldots, L$};

% Arrow showing net positive
\draw[->,thick,red!70] (10.1, 0.1) -- (10.1, 0.7) node[right,font=\scriptsize] {net $> 0$};
\node[right,font=\scriptsize,red!70] at (10.1, -0.1) {member};

\end{tikzpicture}

Test statistic:

$$T = \displaystyle\sum_{j=1}^{L} D_j \quad\begin{cases} \approx 0 & \text{non-member} \\ > 0 & \text{member}\end{cases}$$

## The test statistic

Deviation of target from population at SNP $j$:
$$d_j = x_j - 2p_j$$

Inner product with the mixture's deviation:
$$D_{\text{mix}} = \sum_{j=1}^{L} \underbrace{(f_j^{\text{mix}} - p_j)}_{\text{mixture shift}} \cdot \underbrace{d_j}_{\text{target deviation}}$$

**Decision rule:** declare "member" if $D_{\text{mix}} > D_{\text{ref}}$

## Why it works

**$H_0$ (not in pool):** $f_j^{\text{mix}}$ and $x_j$ are independent
$$E[D_{\text{mix}}] \approx 0$$

**$H_1$ (in pool of $n$):** target pulls $f_j^{\text{mix}}$ toward $x_j$

$$f_j^{\text{mix}} = \frac{1}{2n}\Big(x_j + \textstyle\sum_{i \neq \text{target}} x_{ij}\Big)$$

$$E[D_{\text{mix}} \mid H_1] \approx \frac{1}{2n}\sum_{j=1}^{L} d_j^2 > 0$$

## Membership Test in Code

```{r membership-test, echo=TRUE, eval=FALSE}
# Homer's membership inference test
                                        # Target genotype: x (0, 1, or 2 for each SNP)
                                        # Mixture allele freq: f_mix
                                        # Population allele freq: p_pop

                                        # Target's deviation from population
d <- x - 2 * p_pop

                                        # Test statistic: inner product
D_mix <- sum((f_mix - p_pop) * d)

                                        # Decision
is_member <- D_mix > threshold  # TRUE if in pool
```

\medskip

**Key insight:** Aggregated statistics leak membership

## Signal-to-Noise Ratio

$$\text{SNR} \sim \frac{\sqrt{L}}{n}$$

| $L$ (SNPs) | $n$ (pool) | SNR | Detectable? |
|---:|---:|---:|:---:|
| 500K | 100 | $\sim 7$ | Yes |
| 500K | 1,000 | $\sim 0.7$ | Marginal |
| 500K | 10,000 | $\sim 0.07$ | No |

With $L = 500\text{K}$ common SNPs, membership detectable in pools up to $n \approx 1000$ [@Sankararaman2009-jt]

## Fairly strong membership prediction

\only<1>{\includegraphics[width=.3\textwidth]{img/sankararaman-1.pdf}}

:::::: {.columns}
::: {.column width=.6}

\only<1>{
\includegraphics[width=\textwidth]{img/sankararaman-2.pdf}
}

\only<2>{
Result is \textbf{not population-specific} (HapMap YRI):

${}$

\begin{itemize}
\item Same pattern in African population
\item $m = 100$: near-random
\item $m = 1{,}000$: moderate power
\item $m = 33{,}138$: near-perfect
\end{itemize}
\vfill
$${}$$
Theory matches empirical across populations
}

:::
::: {.column width=.4}

\only<1>{
\small
\begin{itemize}
\item Pool size $n = 1{,}000$
\item LR outperforms Homer et al.
\item More SNPs ($m$) $\Rightarrow$ higher power
\item Theory $\approx$ simulation
\item Validated on real WTCCC data
\end{itemize}
}

\only<2>{
\includegraphics[width=\textwidth]{img/sankararaman-3.pdf}

\medskip
\scriptsize
HapMap YRI (African) dataset
}


:::
::::::

## The Beacon Attack

:::::: {.columns}
::: {.column width=.35}

\small

A **beacon** answers one query:

> "Does anyone carry allele A at position X?"

Response: **yes** / **no** (1 bit)

\medskip

Even this minimal interface leaks membership via repeated queries + LR test

:::
::: {.column width=.6}

\centerline{\includegraphics[width=\textwidth]{img/beacon-1.pdf}}

\centerline{\includegraphics[width=\textwidth]{img/beacon-2.pdf}}

:::
::::::

\vfill
\tiny

@Shringarpure2015-ut

## {.plain}

\centerline{\includegraphics[width=\textwidth]{img/surname-1.pdf}}

- What's so special about Y chromosome?

- Y-STRs co-segregate with surnames; ~12% of US Caucasian males can be identified

- How does surname frequency affect vulnerability?

- Can rare surnames protect privacy?

\vfill
\tiny

@Gymrek2013-ho

## Why Y-STR? No recombination in the patrilineal line

:::::: {.columns}
::: {.column width=.5}

\includegraphics[width=\textwidth]{img/surname-pedigree-1.pdf}

:::
::: {.column width=.5}

- Y chromosome does not recombine $\Rightarrow$ Y-STR haplotype passed intact father $\to$ son

- Red: all patrilineal males share the same Y-STR profile; Black filled: sequenced individuals

- **Common surname** scenario: large lineage, many carriers in genealogy DB

- Easy to find a Y-STR match, but harder to narrow down to one person

:::
::::::

\vfill
\tiny

@Gymrek2013-ho

## Common surname: Y-STR reaches distant relatives

:::::: {.columns}
::: {.column width=.5}

\includegraphics[width=\textwidth]{img/surname-pedigree-2.pdf}

:::
::: {.column width=.5}

- Arrows: individuals matched via genealogy DB, 2-7 meioses from the sequenced person

- Identical Y-STR links distant relatives who never consented to data sharing

- Large pedigree $\Rightarrow$ many males share the surname, requiring additional metadata to identify

:::
::::::

\vfill
\tiny

@Gymrek2013-ho

## Rare surname: a single match is nearly unique

:::::: {.columns}
::: {.column width=.5}

\includegraphics[width=\textwidth]{img/surname-pedigree-3.pdf}

:::
::: {.column width=.5}

- Small lineage: few males carry this Y-STR haplotype

- Dashed: genealogy DB record holder; Arrow: identified descendant

- One match in DB $\Rightarrow$ probability of random match: 1 in $10^5$ to $5 \times 10^9$

- Rare surnames are the most vulnerable to re-identification

:::
::::::

\vfill
\tiny

@Gymrek2013-ho

## Surname coverage in genealogy databases

:::::: {.columns}
::: {.column width=.4}

\centerline{\includegraphics[width=\textwidth]{img/surname-2.pdf}}

:::
::: {.column width=.5}

- Genealogy DB coverage tracks US census surname frequency

- Common surnames: many DB records $\Rightarrow$ easier to find a Y-STR match, but harder to narrow down to one person

- Rare surnames: fewer DB records, but a single match is highly identifying

- The attack surface grows as hobbyist databases expand

:::
::::::

\vfill
\tiny

@Gymrek2013-ho

## Demographic Triangulation

:::::: {.columns}
::: {.column width=.4}

\centerline{\includegraphics[width=\textwidth]{img/surname-3.pdf}}

:::
::: {.column width=.5}

- Black: surname recovery rate; Gray: false positive rate

- Trade-off: at $\delta = 0.82$, ~12% success with ~5% FPR

- Surname + age + state narrows 60,000 candidates $\rightarrow$ median of 12 individuals

- How much metadata is too much?

- What should be protected beyond HIPAA?

:::
::::::

\vfill
\tiny

@Gymrek2013-ho

## Real-World Re-identification: Craig Venter

:::::: {.columns}
::: {.column width=.4}

\centerline{\includegraphics[width=\textwidth]{img/surname-4.pdf}}

:::
::: {.column width=.5}

- Wait, where is Craig Venter's DNA data?

- Clear match to "Venter" record in genetic genealogy database

- Public genomes in GenBank can be identified by name

:::
::::::

\vfill
\tiny

@Gymrek2013-ho

## How much does Y-STR matching narrow the search?

:::::: {.columns}
::: {.column width=.4}

\centerline{\includegraphics[width=\textwidth]{img/surname-5.pdf}}

:::
::: {.column width=.5}

- Red: candidate pool from census alone (median $\sim 10^5$ males); Black: after Y-STR surname match (median $\sim 10$)

- 99% accuracy from Illumina sequencing; works even at $10\times$ coverage

- Technical noise alone does not prevent the attack

:::
::::::

\vfill
\tiny

@Gymrek2013-ho



# Re-identification attack


## eQTL Linking Attacks

```{r eqtl-sim, include=FALSE}
n <- 500; K <- 50

## Generate realistic MAFs (minor allele frequencies)
set.seed(42)
mafs  <- runif(K, min=0.05, max=0.45)  ## MAF between 5% and 45%

## Generate effect sizes
betas <- runif(K, min=0.8, max=1.5)

sigma <- 0.3  ## strong eQTLs: high signal-to-noise ratio

G <- sapply(mafs, function(p) rbinom(n, 2, p))
E <- sapply(1:K, function(k) betas[k] * G[,k] + rnorm(n, 0, sigma))

pred.geno <- function(e, b) pmin(pmax(round(e / b), 0), 2)
set.seed(73)
target <- sample(1:n, 1)  ## random target individual

## store results for narrowing slides (4 genes)
pred <- sapply(1:K, function(k) pred.geno(E[target, k], betas[k]))
cands <- list(); cs <- 1:n
for(s in 1:K) { cs <- cs[G[cs, s] == pred[s]]; cands[[s]] <- cs }
n.rem <- sapply(cands, length)

## frequency-weighted scores: use MORE genes for clear separation
set.seed(101)
K.score <- 20
mafs.s  <- runif(K.score, 0.05, 0.45)
betas.s <- runif(K.score, 0.8, 1.4)
G.s <- sapply(mafs.s, function(p) rbinom(n, 2, p))
E.s <- sapply(1:K.score, function(k) betas.s[k] * G.s[,k] + rnorm(n, 0, sigma))
pred.s <- sapply(1:K.score, function(k) pred.geno(E.s[target, k], betas.s[k]))

gfreq <- function(g, p) dbinom(g, 2, p)
scores <- sapply(1:n, function(i) {
    sum(sapply(1:K.score, function(k) {
        if(G.s[i, k] == pred.s[k]) log2(1 / gfreq(pred.s[k], mafs.s[k])) else 0
    }))
})

                                        # Select 5 representative genes for visualization (spread across MAF range)
maf.order <- order(mafs)
viz.genes <- maf.order[round(seq(1, K, length.out=5))]

                                        # plot helper: boxplot + expression heatmap
plot.step <- function(step) {
    pool <- if(step == 1) 1:n else cands[[step - 1]]
    k <- step
    matched <- cands[[step]]
    unmatched <- setdiff(pool, matched)

    ## left: boxplot
    d <- data.frame(geno = factor(G[pool, k]),
                    expr = E[pool, k],
                    is.target = pool == target)
    tgt.expr <- E[target, k]
    p1 <- ggplot(d, aes(geno, expr)) +
        geom_boxplot(outlier.shape=NA, fill="grey90", width=.5) +
        ggrastr::rasterise(
                     geom_jitter(aes(color=is.target, size=is.target),
                                 width=.15, stroke=0),
                     dpi=300) +
        scale_color_manual(values=c("grey50","red")) +
        scale_size_manual(values=c(0.6, 1.5)) +
        geom_hline(yintercept=tgt.expr, lty=2, color="red", linewidth=.3) +
        annotate("label", x=3.4, y=tgt.expr,
                 label=paste0("predict g=", pred[k]),
                 size=2.5, color="red", fill="white", hjust=1) +
        labs(x="Genotype (hidden)", y="Expression",
             subtitle=paste0(length(pool), " candidates  >  ",
                             n.rem[step], " remain")) +
        theme_minimal() +
        theme(legend.position="none",
              axis.title=element_text(size=7), axis.text=element_text(size=6),
              plot.subtitle=element_text(size=8, face="bold"))

    ## right: expression heatmap for genes 1..step
    ids <- c(matched, unmatched)
    grp <- c(rep("match", length(matched)), rep("no match", length(unmatched)))
    ## Center expression for better color display
    mat.centered <- scale(E[ids, 1:step, drop=FALSE], center=TRUE, scale=FALSE)
    hd <- expand.grid(row=seq_along(ids), col=1:step)
    hd$expr <- as.vector(mat.centered)
    hd$gene <- factor(paste0("G", hd$col))
    hd$group <- grp[hd$row]
    hd$is.tgt <- ids[hd$row] == target

    p2 <- ggplot(hd, aes(gene, reorder(row, -row), fill=expr)) +
        ggrastr::rasterise(geom_tile(), dpi=300) +
        scale_fill_gradient2(low="steelblue", mid="white", high="firebrick",
                             midpoint=0) +
        facet_grid(group ~ ., scales="free_y", space="free_y") +
        labs(x=NULL, y="Individual", subtitle="Expression matrix") +
        theme_minimal() +
        theme(legend.position="none",
              axis.text.y=element_blank(), axis.ticks.y=element_blank(),
              axis.title=element_text(size=7), axis.text.x=element_text(size=6),
              strip.text=element_text(size=6, face="bold"),
              plot.subtitle=element_text(size=8, face="bold"))

    p1 + p2 + plot_layout(widths=c(3, 2))
}
```

\begin{tikzpicture}[>=stealth, font=\small, every node/.style={align=center}]
\node[draw, rounded corners, fill=green!10, minimum width=2.2cm, minimum height=1cm] (expr) at (0,0) {Expression\\(public)};
\node[draw, rounded corners, fill=yellow!15, minimum width=2.2cm, minimum height=1cm] (eqtl) at (3.5,0) {eQTL catalog\\(public)};
\node[draw, rounded corners, fill=orange!15, minimum width=2.2cm, minimum height=1cm] (pred) at (7,0) {Predicted\\genotype};
\node[draw, rounded corners, fill=red!10, minimum width=2.2cm, minimum height=1cm] (geno) at (10.5,0) {Genotype DB\\\footnotesize(private)};
\draw[->,thick] (expr) -- (eqtl);
\draw[->,thick] (eqtl) -- node[above,font=\scriptsize] {predict} (pred);
\draw[->,thick] (pred) -- node[above,font=\scriptsize] {match} (geno);
\node[font=\footnotesize, red!70!black] at (10.5, -1) {\textbf{Who is the target?}};
\end{tikzpicture}

## What is eQTL?

:::::: {.columns}
::: {.column width=.55}

An **expression quantitative trait locus (eQTL)** is a genomic variant (SNP) whose genotype is associated with the expression level of a gene.

\medskip

\begin{tikzpicture}[>=stealth, font=\small]
% DNA region
\draw[thick, blue!60] (0,0) -- (6,0);
\node[circle, fill=red!70, inner sep=2pt, label={[font=\scriptsize]above:SNP}] (snp) at (1.2,0) {};
\draw[fill=blue!20, draw=blue!60, thick] (3.5,-0.15) rectangle (5.5,0.15);
\node[font=\scriptsize, below=3pt] at (4.5,-0.15) {Gene};
\draw[->, thick, red!60!black, dashed] (snp) to[bend left=25] node[above, font=\scriptsize] {regulates} (4.5, 0);

% Genotype-expression boxplot sketch
\foreach \x/\g/\h in {1/0/0.4, 2.2/1/1.0, 3.4/2/1.6} {
\draw[fill=blue!15] (\x, -2.6) rectangle ++(0.8, \h);
\node[font=\scriptsize] at (\x+0.4, -2.9) {$g{=}\g$};
}
\node[font=\scriptsize, rotate=90] at (0.5, -2.2) {Expression};
\draw[->] (0.7,-2.8) -- (4.5,-2.8);
\draw[->] (0.7,-2.8) -- (0.7,-1.7);

\end{tikzpicture}

:::
::: {.column width=.43}

$$e = \beta \cdot g + \epsilon$$

- $g \in \{0, 1, 2\}$: genotype (minor allele count)
- $e$: expression level
- $\beta$: eQTL effect size

<!-- \medskip -->
<!-- **Key insight for attacks:** -->
<!-- If $\beta$ is known (public eQTL catalog), observing $e$ reveals $g$: -->
<!-- $$\hat{g} = \text{round}(e / \beta)$$ -->

:::
::::::

## {.plain}

${}$

\centerline{\includegraphics[width=\textwidth]{img/harmanci-gerstein-1.pdf}}

${}$

**Input**:

- Expression data
- eQTL summary statistics

**Output**:

- Genotypes

**Implication**: G $\to$ ?

<!-- \vfill -->
<!-- \tiny -->
<!-- [@Harmanci2016-ua] -->


## Genotypes are predictable from expression

:::::: {.columns}
::: {.column width=.1}
:::

::: {.column width=.65}

\centerline{\includegraphics[width=\textwidth]{img/harmanci-gerstein-2.pdf}}

:::
::: {.column width=.25}

- Correlation info links expression to variants

- Predict genotypes from observed expression

- Genotypes $\Rightarrow$ re-identification

- Would involuntarily reveal sensitive information!

$${}$$

\tiny

[@Harmanci2016-ua]

:::
::::::


## Using expression QTL as backdoor...

\centerline{\includegraphics[width=.9\textwidth]{img/harmanci-gerstein-3a.pdf}}

\vfill

- eQTL defines a joint distribution of expression ($E$) and genotype ($V$)

- Given observed expression $e$, slice the joint distribution $\Rightarrow$ conditional $P(V | E = e)$

- Strong eQTL: the slice is peaked at one genotype $\Rightarrow$ high information leakage

\tiny

[@Harmanci2016-ua]

## Tradeoff: predictability vs. identifiability

:::::: {.columns}
::: {.column width=.65}

\begin{tikzpicture}[>=stealth, font=\normalsize, yscale=0.7]
% Common variant
\node[font=\large, anchor=west] at (-0.5, 3.2) {\textbf{Common} (MAF $= 0.4$)};
% Prior bars
\node[font=\small, anchor=east] at (-0.2, 2.5) {Prior:};
\fill[blue!30] (0, 2.0) rectangle (1.1, 2.9);
\fill[blue!30] (1.4, 2.0) rectangle (2.87, 2.9);
\fill[blue!30] (3.2, 2.0) rectangle (3.7, 2.9);
\node[font=\scriptsize] at (0.55, 1.7) {$g{=}0$};
\node[font=\scriptsize] at (2.13, 1.7) {$g{=}1$};
\node[font=\scriptsize] at (3.45, 1.7) {$g{=}2$};
\node[font=\scriptsize] at (0.55, 2.5) {.36};
\node[font=\scriptsize] at (2.13, 2.5) {.48};
\node[font=\scriptsize] at (3.45, 2.5) {.16};
% Posterior bars
\node[font=\small, anchor=east] at (-0.2, 0.8) {$P(V|E{=}e)$:};
\fill[blue!15] (0, 0.3) rectangle (0.35, 1.2);
\fill[red!50] (1.4, 0.3) rectangle (3.1, 1.2);
\fill[blue!15] (3.2, 0.3) rectangle (3.45, 1.2);
\node[font=\scriptsize] at (0.55, 0.0) {$g{=}0$};
\node[font=\scriptsize] at (2.13, 0.0) {$g{=}1$};
\node[font=\scriptsize] at (3.45, 0.0) {$g{=}2$};
\node[font=\scriptsize] at (0.18, 0.75) {.10};
\node[font=\scriptsize, red!70!black] at (2.25, 0.75) {\textbf{.80}};
\node[font=\scriptsize] at (3.33, 0.75) {.10};
\draw[->, thick] (4.0, 0.75) -- (5.0, 0.75) node[right, font=\small, text width=2.2cm] {predict $g{=}1$\\info $= 0.3$ bits};

% Rare variant
\node[font=\large, anchor=west] at (-0.5, -1.0) {\textbf{Rare} (MAF $= 0.05$)};
% Prior bars
\node[font=\small, anchor=east] at (-0.2, -1.7) {Prior:};
\fill[blue!30] (0, -2.2) rectangle (2.7, -1.3);
\fill[blue!30] (3.0, -2.2) rectangle (3.3, -1.3);
\fill[blue!30] (3.6, -2.2) rectangle (3.67, -1.3);
\node[font=\scriptsize] at (1.35, -2.5) {$g{=}0$};
\node[font=\scriptsize] at (3.15, -2.5) {$g{=}1$};
\node[font=\scriptsize] at (3.63, -2.5) {$g{=}2$};
\node[font=\scriptsize] at (1.35, -1.7) {.90};
\node[font=\scriptsize] at (3.15, -1.7) {.10};
\node[font=\scriptsize] at (3.63, -1.7) {$\approx 0$};
% Posterior bars
\node[font=\small, anchor=east] at (-0.2, -3.4) {$P(V|E{=}e)$:};
\fill[blue!15] (0, -3.9) rectangle (0.35, -3.0);
\fill[red!50] (3.0, -3.9) rectangle (4.7, -3.0);
\fill[blue!15] (5.0, -3.9) rectangle (5.07, -3.0);
\node[font=\scriptsize] at (0.18, -4.2) {$g{=}0$};
\node[font=\scriptsize] at (3.85, -4.2) {$g{=}1$};
\node[font=\scriptsize] at (5.03, -4.2) {$g{=}2$};
\node[font=\scriptsize] at (0.18, -3.45) {.10};
\node[font=\scriptsize, red!70!black] at (3.85, -3.45) {\textbf{.80}};
\node[font=\scriptsize] at (5.03, -3.45) {.10};
\draw[->, thick] (5.5, -3.45) -- (5.7, -3.45) node[right, font=\small, text width=2.2cm] {predict $g{=}1$\\info $= 3.4$ bits};
\end{tikzpicture}

:::
::: {.column width=.35}

\small

Conditional self-information:

$$\log_2\!\frac{1}{P(V = g \mid E = e)}$$

\medskip

- Both predict $g{=}1$ with 80% confidence

- Common: $g{=}1$ has prior .48 $\Rightarrow$ low surprise

- Rare: $g{=}1$ has prior .10 $\Rightarrow$ high surprise, highly identifying

:::
::::::

\vfill
\tiny

[@Harmanci2016-ua]


## Three-Step Linking Framework

:::::: {.columns}
::: {.column width=.55}

\centerline{\includegraphics[width=\textwidth]{img/harmanci-gerstein-4.pdf}}

:::
::: {.column width=.43}

\large

1. Select eQTLs (correlation threshold)
2. Predict genotypes from expression
3. Link to database via distance metric

${}$

\normalsize

- Adding gender, population improves accuracy to >95%
- Even genotypes alone achieve ~85% accuracy

:::
::::::


\tiny
\flushright

[@Harmanci2016-ua]


## {.plain}

\vfill

\centerline{\includegraphics[width=\textwidth]{img/harmanci-gerstein-5.pdf}}


\tiny

[@Harmanci2016-ua]


## Simplified attack: extremity-based prediction

${}$

\centerline{\includegraphics[width=.9\textwidth]{img/harmanci-gerstein-6.pdf}}

- Don't need the full joint distribution: just use extreme expression values

- High expression $\Rightarrow$ predict $g = 2$; Low expression $\Rightarrow$ predict $g = 0$

- Robust across eQTL selection criteria; PPV >95% for most thresholds

\tiny
\flushright

[@Harmanci2016-ua]

## A working example on eQTL linkage attack

:::::: {.columns}
::: {.column width=.48}

**Attacker has (public):**

- Target's expression: $\mathbf{e} = (e_1, \ldots, e_K)$
- eQTL catalog: Gene $k$ $\leftrightarrow$ SNP $k$, effect $\beta_k$

\medskip

**Attacker wants:**

- Predicted genotype: $\hat{g}_k = \text{round}(e_k / \beta_k)$
- Match $\hat{\mathbf{g}}$ to a row in the genotype DB

:::
::: {.column width=.5}

```{r eqtl-setup-table, echo=FALSE, results='asis'}
cat("\\footnotesize\n")
cat("\\begin{tabular}{cccc}\n")
cat("\\hline\n")
cat("Gene & Expression & $\\beta$ & Predicted $\\hat{g}$ \\\\\n")
cat("\\hline\n")
for(k in viz.genes) {
    cat(sprintf("%d & %.1f & %.1f & %d \\\\\n",
                k, E[target, k], betas[k], pred[k]))
}
cat("\\hline\n")
cat("\\end{tabular}\n")
cat("\\normalsize\n")
```

\medskip

Match this vector against $n = `r n`$ individuals

$${}$$

* How many unknowns?

* How many pieces of evidence?

:::
::::::

## The Attack in Code

```{r attack-code, echo=TRUE, eval=FALSE}
## eQTL model: expression = beta * genotype + noise
E <- beta * G + rnorm(n, sd = sigma)

## Predict genotype from expression (reverse the model)
pred_genotype <- round(E / beta)
pred_genotype <- pmin(pmax(pred_genotype, 0), 2)  # clip to 0,1,2

## Match predicted genotype to database
## --> could be just nearest neighbour search
matches <- which(G_database == pred_genotype)
```

\medskip

Simple algebra can leak the identity...

<!-- ## Let's see how gene expressions can unlock the identity

```{r eqtl-zoom, echo=FALSE, fig.width=5.5, fig.height=3.2, eval=FALSE}
## Cumulative match against ACTUAL genotypes using target's predicted genotype
cum.match <- matrix(TRUE, n, K)
for(s in 1:K)
    cum.match[,s] <- rowSums(sweep(G[,1:s,drop=FALSE], 2, pred[1:s], "==")) == s

## Order: target first, then by match depth (descending)
n.total <- rowSums(cum.match)
ro <- order(-n.total)
row.ord <- c(target, ro[ro != target])

## Only show top individuals to avoid overcrowding
n.show <- min(30, n)
row.ord.subset <- row.ord[1:n.show]

## Find target's position in subset
target.row <- which(row.ord.subset == target)

## Center expression for better color visualization
E.centered <- scale(E, center = TRUE, scale = FALSE)

## For this cumulative visualization, use first 5 genes sequentially
genes.show <- 1:5

## Build panel data
panels <- do.call(rbind, lapply(genes.show, function(s) {
    data.frame(
        row = rep(1:n.show, length(genes.show)),
        col = rep(1:length(genes.show), each = n.show),
        expr = as.vector(E.centered[row.ord.subset, genes.show]),
        gene = factor(paste0("G", rep(genes.show, each = n.show))),
        active = rep(cum.match[row.ord.subset, s], length(genes.show)),
        is.tgt = rep(1:n.show == 1, length(genes.show)),
        step = paste0("Gene ", s, ": ", sum(cum.match[,s]), " remain"),
        gene.used = rep(genes.show, each = n.show) <= s
    )
}))
panels$step <- factor(panels$step, levels = unique(panels$step))

ggplot(panels, aes(gene, row)) +
    ggrastr::rasterise(
                 geom_tile(aes(fill = ifelse(active & gene.used, expr, NA))),
                 dpi = 300) +
    ggrastr::rasterise(
                 geom_tile(data = panels[!panels$active & panels$gene.used, ],
                           fill = "grey92"),
                 dpi = 300) +
    ggrastr::rasterise(
                 geom_tile(data = panels[!panels$gene.used, ],
                           fill = "grey97"),
                 dpi = 300) +
    scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                         midpoint = 0, na.value = "grey92", guide = "none") +
    ## Mark the target individual with a star on the first gene
    geom_text(data = data.frame(
                  step = factor(levels(panels$step), levels = levels(panels$step)),
                  gene = factor("G1", levels = levels(panels$gene))),
              aes(x = gene, label = "*"), y = target.row,
              size = 5, color = "red", hjust = 0.5) +
    facet_wrap(~ step, nrow = 1) +
    scale_y_reverse() +
    labs(x = NULL, y = "Individuals (top 30)") +
    theme_minimal() +
    theme(axis.text = element_blank(), axis.ticks = element_blank(),
          axis.title.y = element_text(size = 7),
          strip.text = element_text(size = 6, face = "bold"),
          panel.spacing = unit(0.3, "lines"))
```
-->

## Gene 1: predicting genotype from expression

```{r eqtl-step1, echo=FALSE, fig.width=5, fig.height=2.5}
plot.step(1)
```

## Gene 2: narrowing the pool

```{r eqtl-step2, echo=FALSE, fig.width=5, fig.height=2.5}
plot.step(2)
```

## Gene 3: a rare variant

```{r eqtl-step3, echo=FALSE, fig.width=5, fig.height=2.5}
plot.step(3)
```

## Gene 4: identified

```{r eqtl-step4, echo=FALSE, fig.width=5, fig.height=2.5}
plot.step(4)
```

## Genotype vs. expression: the full picture

```{r eqtl-full-view, echo=FALSE, fig.width=5.5, fig.height=2.5}
d.all <- do.call(rbind, lapply(viz.genes, function(k) {
    data.frame(geno = factor(G[,k]),
               expr = E[,k],
               is.target = 1:n == target,
               is.match = 1:n %in% cands[[K]],
               gene = paste0("Gene ", k, " (MAF=", round(mafs[k], 2), ")"),
               pred.g = pred[k])
}))
d.all$gene <- factor(d.all$gene, levels = unique(d.all$gene))

ggplot(d.all, aes(geno, expr)) +
    geom_boxplot(outlier.shape = NA, fill = "grey90", width = .5) +
    ggrastr::rasterise(
                 geom_jitter(color = "grey60", size = 0.4, width = .15, stroke = 0),
                 dpi = 300) +
    ggrastr::rasterise(
                 geom_jitter(data = d.all[d.all$is.target, ],
                             color = "red", size = 1.2, width = .15, stroke = 0),
                 dpi = 300) +
    geom_hline(data = data.frame(
                   gene = factor(unique(d.all$gene), levels = levels(d.all$gene)),
                   yint = E[target, viz.genes]),
               aes(yintercept = yint), lty = 2, color = "red", linewidth = .3) +
    facet_wrap(~ gene, nrow = 1, scales = "free_y") +
    labs(x = "Genotype", y = "Expression") +
    theme_minimal() +
    theme(axis.title = element_text(size = 7),
          axis.text = element_text(size = 6),
          strip.text = element_text(size = 6, face = "bold"),
          panel.spacing = unit(0.4, "lines"))
```


## Not all genes are equal

```{r eqtl-funnel, echo=FALSE, fig.width=5, fig.height=2.5}
before <- c(n, n.rem[1:3])
after  <- n.rem[1:4]
## Find the most informative gene among first 4 (lowest MAF)
most.info.idx <- which.min(mafs[1:4])
most.info.gene <- paste0("Gene ", most.info.idx)
data.frame(
    Gene = factor(paste0("Gene ", 1:4), levels=paste0("Gene ", 1:4)),
    Frac = (before - after) / before,
    Label = paste0(before, " -> ", after),
    MAF = round(mafs[1:4], 2)
) %>%
    ggplot(aes(Gene, Frac)) +
    geom_col(aes(fill = Gene == most.info.gene), width=.6, show.legend=FALSE) +
    scale_fill_manual(values=c("steelblue","firebrick")) +
    geom_text(aes(label=Label), vjust=-.3, size=3) +
    scale_y_continuous(labels=scales::percent, limits=c(0,1)) +
    labs(y="Fraction eliminated") +
    theme_minimal() +
    theme(axis.title=element_text(size=8), axis.text=element_text(size=7))
```

**Q: Why is a rare-variant gene more informative?**

## Weight matches by rarity

Exact matching is fragile: one wrong prediction breaks the chain.

\medskip

**Better approach**: score each match by how informative it is.

\medskip

- If 45% of people have genotype $g=1$, matching on it is **weak evidence**
- If 5% of people have genotype $g=2$, matching on it is **strong evidence**

\medskip

$$S_i = \sum_{k} \mathbb{I}[G_{ik} = \hat{g}_k] \cdot \underbrace{\log_2 (1/f(\hat{g}_k))}_{\text{information content}}$$

\medskip

This is just information theory: rare events carry more information.

## Accumulating evidence across genes

```{r eqtl-scores-cumul, echo=FALSE, fig.width=5.5, fig.height=2.5}
## Cumulative scores gene by gene
cum.scores <- matrix(0, n, K)
for(k in 1:K) {
    contrib <- sapply(1:n, function(i)
        if(G[i,k] == pred[k]) log2(1 / gfreq(pred[k], mafs[k])) else 0)
    cum.scores[,k] <- if(k == 1) contrib else cum.scores[,k-1] + contrib
}

## Build panel data - only for viz.genes
panels <- do.call(rbind, lapply(viz.genes, function(k) {
    ord <- order(cum.scores[,k], decreasing=TRUE)
    data.frame(rank = 1:n, score = cum.scores[ord, k],
               is.target = ord == target,
               step = paste0("+ Gene ", k))
}))
panels$step <- factor(panels$step, levels = paste0("+ Gene ", viz.genes))

## Annotate target rank in each panel
target.ranks <- do.call(rbind, lapply(viz.genes, function(k) {
    ord <- order(cum.scores[,k], decreasing=TRUE)
    r <- which(ord == target)
    data.frame(rank = r, score = cum.scores[target, k],
               step = paste0("+ Gene ", k))
}))
target.ranks$step <- factor(target.ranks$step, levels = paste0("+ Gene ", viz.genes))
target.ranks$label <- paste0("rank ", target.ranks$rank)

ggplot(panels, aes(rank, score, fill=is.target)) +
    ggrastr::rasterise(geom_col(width=.8), dpi=300) +
    scale_fill_manual(values=c("grey70","red")) +
    geom_text(data=target.ranks,
              aes(x=rank, y=score, label=label, fill=NULL),
              vjust=-0.3, hjust=0, size=2.2, color="red", fontface="bold") +
    facet_wrap(~ step, nrow=1) +
    labs(x="Person (ranked by score)", y="Cumulative linking score") +
    theme_minimal() +
    theme(legend.position="none",
          axis.title=element_text(size=7), axis.text=element_text(size=5),
          strip.text=element_text(size=7, face="bold"),
          panel.spacing=unit(0.3, "lines"))
```

Each panel adds one gene. The target (red) moves toward rank 1 as evidence accumulates.



## Auxiliary metadata amplifies the attack

\centerline{\includegraphics[width=\textwidth]{img/harmanci-gerstein-7.pdf}}

- Genotypes alone: ~85% of individuals vulnerable at moderate eQTL threshold

- Adding gender or population pushes vulnerability above 95%


\tiny

[@Harmanci2016-ua]


## {.plain}

${}$

\centerline{\includegraphics[width=\textwidth]{img/walker-1.pdf}}

- Linking attack workflow: scRNA-seq $\rightarrow$ genotype prediction $\rightarrow$ database matching
- How can we share valuable single-cell data while protecting donor identity?

\tiny

@Walker2024-rm


## Linking with Bulk eQTLs

\centerline{\includegraphics[width=\textwidth]{img/walker-2.pdf}}

\tiny

- 93-98% linking accuracy using publicly available tissue eQTLs
- Even "common" cell types leak identity. What counts as rare enough to be safe?

\tiny
@Walker2024-rm


## {.plain}

${}$

\centerline{\includegraphics[width=\textwidth]{img/walker-3.pdf}}

${}$

\tiny

@Walker2024-rm


## The way we called eQTLs can create a backdoor

\centerline{\includegraphics[width=.7\textwidth]{img/walker-4.pdf}}

* **Input**: (1) pseudobulk (2) eQTL catalogue

\tiny

@Walker2024-rm

<!-- - Cell-type-specific eQTLs increase accuracy for rare cell types -->
<!-- - Better biology $\rightarrow$ worse privacy. How do we balance scientific utility vs. risk? -->
<!-- - High accuracy (>80%) even without eQTL information, using variant selection -->
<!-- - Privacy breach possible with just scRNA-seq data. Can we trust "anonymization"? -->

## Well, multiple cell types $\approx$ multiple lines of evidence

:::::: {.columns}
::: {.column width=.4}

\centerline{\includegraphics[width=\textwidth]{img/walker-4.pdf}}

${}$

- Usually, 5-10 cell types in PBMC

- Cell type specific eQTL data

- Do genotypes change with cell types?

:::
::: {.column width=.6}

\centerline{\includegraphics[width=\textwidth]{img/walker-5.pdf}}

:::
::::::

\vfill
\tiny

@Walker2024-rm


## Revealing "donor" information using external DB!

${}$

\centerline{\includegraphics[width=.8\textwidth]{img/walker-6.pdf}}

<!-- \tiny -->
<!-- - Predicted genotypes from scRNA-seq can identify individuals in external databases -->
<!-- - Your data becomes more identifiable as public databases grow. Future-proof privacy? -->

- Rarer cell types $\Rightarrow$ higher linking accuracy (lines trend upward, common $\to$ rare)

- More cells per type $\Rightarrow$ better pseudobulk estimates $\Rightarrow$ better genotype prediction

<!-- - Even with 200 cells, linking accuracy reaches 20-40% -->

\tiny
\flushright

@Walker2024-rm


## What if we have single-cell data and eQTLs?

:::::: {.columns}
::: {.column width=.65}

\centerline{\includegraphics[width=\textwidth]{img/walker-7.pdf}}

:::
::: {.column width=.3}

- Can normalization and PCA remove non-genetic individual-specific information?

- Single-cell count matrices aggregated into cell-type-specific pseudobulk

:::
::::::



## What if we have single-cell data and eQTLs?

:::::: {.columns}
::: {.column width=.3}

\centerline{\includegraphics[width=\textwidth]{img/walker-7.pdf}}

${}$

\large

The power of multiple lines of evidence across different cell types

:::
::: {.column width=.5}

\centerline{\includegraphics[width=\textwidth]{img/walker-8.pdf}}

:::
::::::

\tiny

@Walker2024-rm


## Again, strong prediction performance...

\centerline{\includegraphics[width=\textwidth]{img/walker-9.pdf}}

<!-- - Each approach generates $10^3$-$10^6$ unique predicted genotypes per sample -->
<!-- - More genetic signal extracted $\rightarrow$ easier to identify. Is dimensionality reduction protective? -->

\tiny

@Walker2024-rm


## Do we even need scQTL summary statistics data?


:::::: {.columns}
::: {.column width=.7}
\centerline{\includegraphics[width=\textwidth]{img/walker-10.pdf}}
:::
::: {.column width=.3}

- No external eQTL catalog needed

- AAF = alternative allele frequency

- Split by expression (high/low), compare AAF per locus

- Large $\Delta$AAF $\Rightarrow$ informative variant

- **Cross-cohort**: learn variants from training cohort, attack a different target cohort

:::
::::::

\tiny

@Walker2024-rm


## Genotypes in training data $\to$ predict other cohorts

:::::: {.columns}
::: {.column width=.1}
:::

::: {.column width=.35}
\centerline{\includegraphics[width=\textwidth]{img/walker-11.pdf}}
:::
::: {.column width=.35}
\centerline{\includegraphics[width=\textwidth]{img/walker-12.pdf}}
:::
::: {.column width=.2}
:::
::::::

\tiny

@Walker2024-rm

## Auxiliary metadata amplifies the single-cell attack

:::::: {.columns}
::: {.column width=.6}

\centerline{\includegraphics[width=\textwidth]{img/walker-13.pdf}}

:::
::: {.column width=.4}

- **OneK1K**: genotypes alone reach 97--100 percent; metadata barely helps

- **Lupus**: genotypes alone 60--70 percent; adding sex + population pushes to 85 percent; all metadata reaches near 100 percent

- Linking individuals in one cohort to the other

- Implication?

\tiny

@Walker2024-rm

:::
::::::


# Secured protocols for genomic data analysis 

## {.plain}

\begin{center}\includegraphics[width=0.7\textwidth]{img/gursoy-1} \end{center}

\vfill
\tiny

@Gursoy2022-od

## How can we safely estimate eQTLs?

\begin{center}\includegraphics[width=0.7\textwidth]{img/gursoy-eqtl-1} \end{center}

\vfill

**Goal**: Protect expression profiled individuals from re-identification attacks

\vfill
\tiny

@Gursoy2022-od

## Trusted Execution Environments (TEE) {.plain}

:::::: {.columns}
::: {.column width=.4}

\begin{center}\includegraphics[height=0.8\textheight]{img/gursoy-eqtl-2} \end{center}

:::
::: {.column width=.55}

- Data encrypted in transit and at rest
- Decrypted only inside hardware **enclave** (e.g., Intel SGX)
- Even the OS cannot read enclave memory

\medskip

**Pros:** 

* near-plaintext speed
* no utility loss

**Cons:** 

* side-channel attacks
* trust in hardware vendor
* limited enclave memory

:::
::::::

\tiny\flushright
@Gursoy2022-od

## Homomorphic Encryption (HE) {.plain}

:::::: {.columns}
::: {.column width=.4}

\begin{center}\includegraphics[height=0.8\textheight]{img/gursoy-eqtl-3} \end{center}

:::
::: {.column width=.55}

Arithmetic on **encrypted data** yields correct results when decrypted:

$\text{Dec}(f(\text{Enc}(x))) = f(x)$

- Server runs eQTL caller on encrypted genotypes and expression (never sees the actual data)
- Only the user holds the private key to decrypt the result

\medskip

**Pros:** strong formal guarantee (computational)

**Cons:** $10^3$–$10^6 \times$ slowdown vs.\ plaintext

:::
::::::

\tiny\flushright
@Gursoy2022-od

## Secure Multiparty Computation (MPC) {.plain}

:::::: {.columns}
::: {.column width=.4}

\begin{center}\includegraphics[height=0.8\textheight]{img/gursoy-eqtl-4} \end{center}

:::
::: {.column width=.55}

Multiple hospitals pool data **without sharing it**

- Each site splits data into random **secret shares**
- Third-party servers compute on shares
- No single party can reconstruct individual data

\medskip

**Key property:** **no utility loss**. Exact same result as pooled plaintext.

:::
::::::

\tiny\flushright
@Gursoy2022-od

## Federated Learning (FL) {.plain}

:::::: {.columns}
::: {.column width=.4}

\begin{center}\includegraphics[height=0.8\textheight]{img/gursoy-eqtl-5} \end{center}

:::
::: {.column width=.55}

Raw data **never leaves** each institution

1. Train local models
2. Add noise to model updates
3. Combine noised models centrally
4. Iterate until convergence

\medskip

**Caveat:** gradients can leak training data. FL alone provides only informal privacy guarantees.

:::
::::::

\tiny\flushright
@Gursoy2022-od


# Differential Privacy

## {.plain}

\centerline{\includegraphics[height=.8\textheight]{img/predictive_models.png}}

\vfill
\tiny
\flushright

[https://xkcd.com/2169/](https://xkcd.com/2169/)

## Differential Privacy: Formal Privacy Guarantees

A randomized mechanism $M$ satisfies $\varepsilon$-DP if for all neighboring datasets $D, D'$:

$$P(M(D) \in S) \leq e^\varepsilon \cdot P(M(D') \in S)$$

**Intuition:** whether or not you participate, published results change by at most $e^\varepsilon$.

\medskip

| $\varepsilon$ | $e^\varepsilon$ | Interpretation |
|:---:|:---:|:---|
| 0.1 | 1.11 | Strong privacy |
| 1 | 2.72 | Moderate |
| 10 | $\sim 22{,}000$ | Essentially vacuous |

\vfill
\tiny
[@Uhler2012-kz]

## DP for GWAS: Adding Noise to Protect Individuals

**Private MAF** at a single SNP in $n$ diploid individuals [@Uhler2012-kz; @Yamamoto2021-qw]:

$$\hat{f}_{\text{priv}} = \underbrace{\frac{1}{2n}\sum_{i=1}^{n}x_i}_{f} + \text{Lap}\!\left(\frac{1}{n\varepsilon}\right)$$

- Sensitivity: changing one person shifts $f$ by at most $\Delta f = 1/n$
- For $n = 5{,}000$, $\varepsilon = 1$: noise SD $\approx 0.00028$ vs. sampling error $\sim 0.01$

\medskip

**Tradeoff:** DP adds statistical noise. Most painful for **rare variants** where the signal is already small.

<!-- ## Proof Sketch: Why Laplace Mechanism Satisfies DP -->
<!-- **Goal:** Show $P(M(D) \in S) \leq e^\varepsilon \cdot P(M(D') \in S)$ -->
<!-- \small -->
<!-- **Step 1:** Neighboring databases $D, D'$ differ by one person -->
<!-- $$|f(D) - f(D')| \leq \Delta f = \frac{1}{n}$$ -->
<!-- **Step 2:** Published value: $f_{\text{priv}} = f + Z$ where $Z \sim \text{Lap}(b)$, $b = \frac{\Delta f}{\varepsilon} = \frac{1}{n\varepsilon}$ -->
<!-- **Step 3:** Probability ratio for any output $s$: -->
<!-- $$\frac{P(f_{\text{priv}}(D) = s)}{P(f_{\text{priv}}(D') = s)} = \frac{e^{-|s - f(D)|/b}}{e^{-|s - f(D')|/b}} = e^{(|s - f(D')| - |s - f(D)|)/b}$$ -->
<!-- **Step 4:** By triangle inequality: $|s - f(D')| - |s - f(D)| \leq |f(D) - f(D')| \leq \frac{1}{n}$ -->
<!-- **Step 5:** Therefore: $\frac{P(M(D) = s)}{P(M(D') = s)} \leq e^{\frac{1/n}{1/(n\varepsilon)}} = e^\varepsilon$ $\checkmark$ -->

## DP in Code: Laplace Mechanism

```{r dp-laplace, echo=TRUE, eval=FALSE}
# Add Laplace noise for epsilon-differential privacy
                                        # Sensitivity: max change from adding/removing one person

                                        # Allele frequency with DP
n <- 5000
epsilon <- 1.0
f <- mean(genotypes) / 2  # true MAF

                                        # Add Laplace noise: Lap(sensitivity/epsilon)
sensitivity <- 1/n
noise <- rlaplace(1, scale = sensitivity/epsilon)
f_private <- f + noise

                                        # Publish f_private instead of f
```

\medskip

**Key insight:** Controlled noise gives privacy guarantee

## DP for Association Tests: Protecting $\chi^2$ Statistics

For a GWAS with genotype $\{0,1,2\}$ $\times$ case/control, form a $3 \times 2$ contingency table [@Uhler2012-kz]:

\small

| | Case | Control |
|:---:|:---:|:---:|
| $g=0$ | $n_{00}$ | $n_{01}$ |
| $g=1$ | $n_{10}$ | $n_{11}$ |
| $g=2$ | $n_{20}$ | $n_{21}$ |

\normalsize

**Sensitivity** (Uhler Thm 3.2): Adding or removing one person changes $\chi^2$ by at most
$$\Delta(\chi^2) = \frac{4N}{N+2}$$

For large $N$, this is $\approx 4$, while $\chi^2$ for true associations grows with $N$
$\Rightarrow$ relative noise shrinks as sample size grows.

## Private $\chi^2$ Release

**Private release:**
$$\chi^2_{\text{priv}} = \chi^2 + \text{Lap}\!\left(\frac{1}{\varepsilon}\cdot\frac{4N}{N+2}\right)$$

\medskip

**Example:** $N = 5{,}000$, $\varepsilon = 1$

- Sensitivity: $(4 \times 5000)/5002 \approx 4.0$
- Noise SD: $4.0/1.0 \times \sqrt{2} \approx 5.7$
- A genome-wide significant SNP has $\chi^2 \approx 30$ $\Rightarrow$ signal-to-noise $\approx 5:1$

\medskip

But do we need to release **all** $M'$ test statistics? Noise grows with number released.

<!-- ## Why Raw p-Values Fail [@Uhler2012-kz; @Yamamoto2021-qw]

Raw p-values are **tiny numbers** concentrated near zero:

- Genome-wide significance: $p < 5 \times 10^{-8}$
- Sensitivity of a single raw p-value from a $3\times2$ table: $e^{-2/3}/\varepsilon \approx 0.51/\varepsilon$

\medskip

**Problem:** Laplace noise of scale $\sim 0.5$ overwhelms a signal of $5 \times 10^{-8}$

$$\underbrace{5 \times 10^{-8}}_{\text{signal}} \ll \underbrace{0.51 / \varepsilon}_{\text{noise scale}}$$

\medskip

Uhler (2012, Fig.\ 7) showed that ROC curves for detecting true associations from noisy raw p-values are near the diagonal --- essentially **random guessing**. -->

## Making DP Practical: Use $\log_{10}$ p-Values

**Key insight:** release $-\log_{10}(p)$ instead of $p$ [@Yamamoto2021-qw].

\medskip

**Sensitivity** for a $3 \times 2$ table (Yamamoto Prop.\ 2):
$$\Delta(-\log_{10} p) = \log_{10}(e) \cdot \frac{2N}{N+2}$$

For $N = 100{,}000$: sensitivity $\approx 0.87$ $\Rightarrow$ **manageable noise**

\medskip

**Private release:**
$$(-\log_{10} p)_{\text{priv}} = -\log_{10}(p) + \text{Lap}\!\left(\frac{0.87}{\varepsilon}\right)$$

\medskip

The significance threshold is now **additive**: $-\log_{10}(5 \times 10^{-8}) = 7.3$

$\Rightarrow$ Compare noisy $-\log_{10}(p)$ directly against $7.3$

## {.plain}

```{r dp-log10-aha, echo=FALSE, fig.width=6, fig.height=2.5}
## The "aha moment": why log10 scale saves differential privacy

## Laplace random variable generator
rlaplace <- function(n, scale) {
    u <- runif(n, -0.5, 0.5)
    return(-scale * sign(u) * log(1 - 2 * abs(u)))
}

set.seed(42)
epsilon <- 1.0

## Raw p-value scale
p_signal <- 5e-8  ## genome-wide significant
noise_scale_raw <- 0.51 / epsilon  ## sensitivity on raw scale

## Log10 p-value scale
log10p_signal <- -log10(p_signal)  ## = 7.3
noise_scale_log <- 0.87 / epsilon  ## sensitivity on log scale

## Create comparison data
comparison_data <- data.frame(
    scale = rep(c("Raw p-value", "-log10(p)"), each = 2),
    type = rep(c("Signal", "Noise scale"), 2),
    value = c(p_signal, noise_scale_raw, log10p_signal, noise_scale_log),
    y_pos = rep(c(1, 2), 2)
)

## Panel 1: Raw scale (disaster)
p_raw <- ggplot() +
    ## Noise scale (huge bar)
    geom_segment(aes(x = 0, xend = noise_scale_raw, y = 2, yend = 2),
                 color = "#d62728", linewidth = 6, alpha = 0.7) +
    geom_text(aes(x = noise_scale_raw/2, y = 2,
                  label = sprintf("Noise: %.2f", noise_scale_raw)),
              hjust=0, color = "black", fontface = "bold", size = 3) +
    ## Signal (tiny, invisible)
    geom_segment(aes(x = 0, xend = p_signal, y = 1, yend = 1),
                 color = "#2ca02c", linewidth = 6, alpha = 0.7) +
    geom_text(aes(x = 0.25, y = 1, label = "Signal: 5e-8\n(invisible!)"),
              color = "black", fontface = "bold", size = 3, hjust = 0) +
    scale_x_continuous(expand = c(0, 0), limits = c(0, 0.6)) +
    scale_y_continuous(limits = c(0.5, 2.5)) +
    labs(title = "Raw p-value scale",
         subtitle = "Signal completely buried by noise",
         x = NULL, y = NULL) +
    theme_minimal(base_size = 10) +
    theme(
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 11),
        plot.subtitle = element_text(color = "#d62728", size = 9, face = "bold")
    )

## Panel 2: Log scale (success!)
p_log <- ggplot() +
    ## Signal
    geom_segment(aes(x = 0, xend = log10p_signal, y = 1, yend = 1),
                 color = "#2ca02c", linewidth = 6, alpha = 0.7) +
    geom_text(aes(x = log10p_signal/2, y = 1,
                  label = sprintf("Signal: %.1f", log10p_signal)),
              color = "black", fontface = "bold", size = 3) +
    ## Noise scale
    geom_segment(aes(x = 0, xend = noise_scale_log, y = 2, yend = 2),
                 color = "#ff7f0e", linewidth = 6, alpha = 0.7) +
    geom_text(aes(x = noise_scale_log/2, y = 2,
                  label = sprintf("Noise: %.2f", noise_scale_log)),
              color = "black", fontface = "bold", size = 3) +
    ## Threshold line
    geom_vline(xintercept = 7.3, linetype = "dashed", linewidth = 1, alpha = 0.5) +
    annotate("text", x = 7.4, y = 2.3, label = "Significance\nthreshold",
             hjust = 0, size = 2.5, fontface = "italic") +
    scale_x_continuous(expand = c(0, 0), limits = c(0, 8.5)) +
    scale_y_continuous(limits = c(0.5, 2.5)) +
    labs(title = "-log10(p) scale",
         subtitle = "Signal clearly exceeds noise!",
         x = NULL, y = NULL) +
    theme_minimal(base_size = 10) +
    theme(
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 11),
        plot.subtitle = element_text(color = "#2ca02c", size = 9, face = "bold")
    )

p_raw + p_log
```

## Efficient DP: Release Only Top SNPs, Not All

**Uhler Algorithm 1** [@Uhler2012-kz] (3 steps, $\varepsilon$-DP by Thm 3.7):

Let $S(\chi^2) = 4N/(N+2)$. Given $M'$ total SNPs, release top $M$:

\medskip

1. **Noisy ranking:** Add $\text{Lap}\!\left(\frac{4M}{\varepsilon} \cdot S(\chi^2)\right)$ to all $M'$ statistics

2. **Select top $M$:** Pick the $M$ SNPs with largest noisy $\chi^2$ $\Rightarrow$ set $\mathcal{S}$

3. **Re-noise the winners:** Add *fresh* $\text{Lap}\!\left(\frac{2M}{\varepsilon} \cdot S(\chi^2)\right)$ to the **true** $\chi^2$ values of SNPs in $\mathcal{S}$

\medskip

**Key property:** noise scales with $M$ (number released), **not** $M'$ (total SNPs tested).

## Safe Z-Score Release via Post-Processing

\textbf{Post-processing theorem:} Any function of a DP output is also DP - no additional privacy cost.

\medskip

Since $Z^2 = \chi^2$ for 1-df tests, derive a private Z-score from the private $\chi^2$:

$$Z_{\text{priv}} = \text{sign}(\hat{\beta}) \cdot \sqrt{\max(0,\; \chi^2_{\text{priv}})}$$

\medskip

Alternatively, derive from the private $-\log_{10}(p)$:

$$Z_{\text{priv}} = \Phi^{-1}\!\left(\tfrac{1}{2}\cdot 10^{-(-\log_{10} p)_{\text{priv}}}\right)$$

\medskip

This lets downstream consumers use **familiar Z-scores** for meta-analysis, fine-mapping, etc. All for free under DP composition.

## Code: Uhler Algorithm 1 + Z-Score Derivation

```{r dp-topM-algorithm, echo=TRUE, eval=FALSE}
# Uhler Algorithm 1: Release top M SNPs privately
N <- 5000; M <- 100; M_prime <- 500000; eps <- 1.0

                                        # Sensitivity of chi-squared for a 3x2 table
sens_chi2 <- 4 * N / (N + 2)

                                        # Step 1: Noisy ranking
chi2_noisy <- chi2_all +
    rlaplace(M_prime, scale = 4 * M / eps * sens_chi2)

                                        # Step 2: Pick top M
top_M <- order(chi2_noisy, decreasing = TRUE)[1:M]

                                        # Step 3: Re-noise the winners (fresh noise on true values)
chi2_private <- chi2_all[top_M] +
    rlaplace(M, scale = 2 * M / eps * sens_chi2)

                                        # Post-process to Z-scores (free by DP composition)
Z_private <- sign(beta_hat[top_M]) *
    sqrt(pmax(0, chi2_private))
```

## DP in Action: Privacy vs. Statistical Power

```{r dp-tradeoff, echo=FALSE, fig.width=5.5, fig.height=2.5}
## Simulate power to detect associations vs. epsilon

## Laplace random variable generator
rlaplace <- function(n, scale) {
    u <- runif(n, -0.5, 0.5)
    return(-scale * sign(u) * log(1 - 2 * abs(u)))
}

set.seed(123)
n_sim <- 1000
true_effect_size <- 30  ## chi-squared for true association
N <- 5000

## Range of epsilon values
epsilon_vals <- c(0.01, 0.05, 0.1, 0.5, 1, 2, 5, 10, 100)

## For each epsilon, calculate detection power
power_data <- map_dfr(epsilon_vals, function(eps) {
    sens_chi2 <- 4 * N / (N + 2)
    noise_scale <- sens_chi2 / eps
    noisy_vals <- true_effect_size + rlaplace(n_sim, scale = noise_scale)
    threshold <- qchisq(0.995, df = 1)
    power <- mean(noisy_vals > threshold)
    data.frame(epsilon = eps, power = power)
})

## Single clean plot
ggplot(power_data, aes(x = epsilon, y = power)) +
    geom_line(linewidth = 1.5, color = "#1f77b4") +
    geom_point(size = 3, color = "#1f77b4") +
    geom_vline(xintercept = 1, linetype = "dashed", color = "#2ca02c", linewidth = 1) +
    annotate("rect", xmin = 0.5, xmax = 2, ymin = 0, ymax = 1,
             alpha = 0.15, fill = "#2ca02c") +
    annotate("text", x = 1, y = 0.92, label = "ε = 1\n(practical)",
             color = "#2ca02c", fontface = "bold", size = 2.5, hjust = 0) +
    scale_x_log10(breaks = c(0.01, 0.1, 1, 10, 100)) +
    scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
    labs(x = "Privacy parameter ε", y = "Power") +
    theme_minimal(base_size = 9) +
    theme(
        panel.grid.minor = element_blank(),
        plot.margin = margin(5, 5, 5, 5)
    )
```

**Key insight:** DP doesn't mean "no utility". It means calibrated noise that preserves strong signals while protecting individuals.

## Recall: The eQTL Linking Attack

**The vulnerability:**

1. Published expression data for anonymous samples
2. Public eQTL catalog (genotype $\leftrightarrow$ expression)
3. Reference genotype database

\medskip

**The attack:**

$$\text{Expression} \xrightarrow{\text{eQTL}} \text{Predicted genotype} \xrightarrow{\text{match}} \text{Individual ID}$$

\medskip

**Result:** With 50 genes, individuals can be uniquely identified

\medskip

**Now:** Can differential privacy stop this attack?

## Can DP Protect Against eQTL Attacks?

**Setup:** Realistic scenario (50 genes, 500 individuals)

**Defense:** Add Laplace noise to expression values before publishing

$$e_k^{\text{priv}} = e_k + \text{Lap}(b)$$

**Attack:** Predict genotypes from noisy expression, then match to database

\medskip

**Question:** How much noise is needed to stop the attack?

## Does DP Stop the eQTL Attack?

```{r dp-eqtl, echo=FALSE, fig.width=6, fig.height=2.5}
set.seed(42)
noise.scales <- c(0, 0.2, 0.4, 0.6, 0.8, 1.0, 1.5, 2.0, 3.0)
n.rep <- 200

## Cache file paths
cache.dir <- "../data/cache"
dir.create(cache.dir, showWarnings=FALSE, recursive=TRUE)
dp.results.file <- file.path(cache.dir, "dp_eqtl_results.rds")
corr.results.file <- file.path(cache.dir, "dp_eqtl_corr.rds")

## Compute or load DP attack results
if.needed(dp.results.file, {
    dp.results <- do.call(rbind, lapply(noise.scales, function(b) {
        ranks <- sapply(1:n.rep, function(r) {
            ## Add Laplace noise to expression
            noisy.E <- E + matrix(
                               if(b > 0) sample(c(-1,1), n*K, replace=TRUE) * rexp(n*K, rate=1/b)
                               else rep(0, n*K),
                               nrow=n, ncol=K)
            ## Attacker predicts from noisy expression
            noisy.pred <- sapply(1:K, function(k) pred.geno(noisy.E[target, k], betas[k]))
            ## Compute matching scores
            sc <- sapply(1:n, function(i) {
                sum(sapply(1:K, function(k) {
                    if(G[i,k] == noisy.pred[k]) log2(1 / gfreq(noisy.pred[k], mafs[k])) else 0
                }))
            })
            rank(-(sc + runif(n, 0, 1e-6)))[target]
        })
        data.frame(noise = paste0("b = ", b),
                   rank = ranks,
                   identified = ranks == 1,
                   b = b)
    }))
    saveRDS(dp.results, dp.results.file)
})
dp.results <- readRDS(dp.results.file)
dp.results$noise <- factor(dp.results$noise, levels = paste0("b = ", noise.scales))

## Compute or load correlation results
if.needed(corr.results.file, {
    corr.results <- do.call(rbind, lapply(noise.scales, function(b) {
        correlations <- sapply(1:50, function(r) {
            noisy.E <- E + matrix(
                               if(b > 0) sample(c(-1,1), n*K, replace=TRUE) * rexp(n*K, rate=1/b)
                               else rep(0, n*K),
                               nrow=n, ncol=K)
            cor(as.vector(E), as.vector(noisy.E))
        })
        data.frame(b = b, correlation = mean(correlations))
    }))
    saveRDS(corr.results, corr.results.file)
})
corr.results <- readRDS(corr.results.file)

## Combine rank and correlation data
rank_summary <- dp.results %>%
    group_by(b) %>%
    summarise(mean_rank = mean(rank), .groups="drop")

combined_data <- rank_summary %>%
    left_join(corr.results, by="b")

## Dual y-axis plot
max_rank <- max(combined_data$mean_rank)
scaling_factor <- max_rank / 1.0  ## scale correlation (0-1) to match rank range

ggplot(combined_data, aes(x=b)) +
    ## Privacy: rank (left axis)
    geom_line(aes(y=mean_rank), color="firebrick", linewidth=0.7) +
    geom_point(aes(y=mean_rank), color="firebrick", size=1.5) +
    ## Utility: correlation (scaled to right axis)
    geom_line(aes(y=correlation * scaling_factor), color="steelblue", linewidth=0.7, linetype="dashed") +
    geom_point(aes(y=correlation * scaling_factor), color="steelblue", size=1.5) +
    ## Reference line: rank 1 = uniquely identified
    geom_hline(yintercept=1, linetype="dotted", color="gray40", linewidth=0.4, alpha=0.5) +
    annotate("text", x=max(combined_data$b)*0.95, y=1.8,
             label="Identified as #1", color="gray40", size=2, hjust=1) +
    ## Scales
    scale_y_continuous(
        name="Privacy: Target's position in ranked list",
        sec.axis=sec_axis(~./scaling_factor, name="Utility: Correlation(obs, noisy)")
    ) +
    labs(x="Noise scale (b)") +
    theme_minimal(base_size=9) +
    theme(
        panel.grid.minor=element_blank(),
        axis.title.y.left=element_text(color="firebrick", face="bold", size=8),
        axis.title.y.right=element_text(color="steelblue", face="bold", size=8),
        axis.text.y.left=element_text(color="firebrick"),
        axis.text.y.right=element_text(color="steelblue")
    )
```

## The DP Tradeoff for eQTL Data

**What we learned:**

\medskip

:::::: {.columns}
::: {.column width="50%"}

**Privacy side:**

- No noise ($b = 0$): Target at rank ~1 (identified)
- Small noise ($b < 0.6$): Rank ~5-10 (still narrow)
- Moderate noise ($b \geq 1$): Rank ~20+ (lost in crowd)

:::
::: {.column width="50%"}

**Utility side:**

- No noise: Perfect correlation (r=1.0)
- Moderate noise: Correlation ~0.6-0.7
- High noise: Data quality severely degraded

:::
::::::

\medskip

**Bottom line:** DP can protect individuals, but you pay with statistical power for discovery.

# Summary

## Comparison of approaches

\footnotesize

| | DP | MPC | HE | FL | TEE |
|:---|:---:|:---:|:---:|:---:|:---:|
| Privacy guarantee | Formal ($\varepsilon$) | Info-theoretic | Computational | Informal | Hardware |
| Utility loss | Yes (noise) | None | None | Some | None |
| Computational cost | Low | Moderate | High | Moderate | Low |
| Multi-party native | -- | Yes | Limited | Yes | Limited |

\normalsize

\medskip

In practice these are **combined**: e.g., federated learning + DP noise + MPC aggregation.

## Genomic privacy

1. **Summary statistics are not safe.** Membership inference from pooled allele frequencies.

2. **Expression data leaks genotypes.** eQTL linking attacks on RNA-seq and scRNA-seq.

3. **Defenses exist but involve tradeoffs.** No single mechanism suffices; layered approaches are the norm.


## Nevertheless, information has to be shared

\large

- Why?

- How?

- Ownership?

- Should a government use it?

## References {.allowframebreaks}
