---
title: "Variational Autoencoder in Single-cell Data"
author: "Yongjin Park"
date: "`r format(Sys.time(), '%d %B %Y')`"
classoption: "aspectratio=169"
fontsize: 12pt
bibliography: vae.bib
output:
    beamer_presentation:
        theme: Madrid
        keep_md: true
        keep_tex: true
        latex_engine: xelatex
        slide_level: 2
header-includes:
  - \usepackage{cancel}
  - \usepackage{tikz}
  - \definecolor{UBCblue}{rgb}{0.04706, 0.13725, 0.26667}
  - \usecolortheme[named=UBCblue]{structure}
  - \setbeamertemplate{frametitle}{\color{UBCblue}\bfseries\insertframetitle\par\vskip-6pt\hrulefill}
  - \setbeamercolor{title in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{section in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{author in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{date in head/foot}{bg=white,fg=gray}
  - \setbeamertemplate{page number in head/foot}{}
  - \setbeamertemplate{frame numbering}[none]
  - \setbeamercolor{alerted text}{bg=yellow}
  - \AtBeginSection[]{\begin{frame}\frametitle{Today's lecture}{\Large\tableofcontents[currentsection]}\end{frame}}
  - |
    \makeatletter
    \def\ps@titlepage{%
      \setbeamertemplate{footline}{}
    }
    \addtobeamertemplate{title page}{\thispagestyle{titlepage}}{}
    \makeatother
---


```{r setup, include=FALSE}
source("Setup.R")
fig.dir <- "../Fig/vae-single-cell/"
dir.create(fig.dir, showWarnings=FALSE)
setup.env(fig.dir)
dir.create("../data", showWarnings=FALSE)

library(data.table)
library(tidyverse)
library(patchwork)
library(matrixTests)

theme_set(theme_minimal())
```

## Learning objectives

* Model-based approach to high-dimensional data analysis

* Primer to learning with artificial neural networks


## Overview of single-cell data analysis

\centerline{\includegraphics[width=.8\textwidth]{img/Karchenko_1.pdf}}

\centerline{\includegraphics[width=.8\textwidth]{img/Karchenko_2.pdf}}

\vfill

\flushright
\tiny
[@Kharchenko2021-wr]

## The goal of modelling in high-dimensional space

The goal is to model two types of relationships:

\large

* Relationship between dimensions/features (genes)

\vfill

* Relationship between samples/data points (cells)

\normalsize

\vfill

::: {.incremental}

### Key question

Which cells co-vary? Which genes co-vary?

:::


## What is a latent variable/factor model?

Examples:

\large

* Principal Component Analysis

* Non-negative matrix factorization

* (...)

\normalsize

## What is Unsupervised Learning?

:::::: {.columns}
::: {.column width=.45}
::: {.block}

### Supervised learning

* Input: data $\mathcal{X} = \{\mathbf{x}_{i},\ldots\}$

* Input 2: label $\mathcal{Y} = \{y_{i},\ldots\}$

* Goal: learn $f: \mathcal{X} \to \mathcal{Y}$

:::

X: gene expression samples; Y: disease labels

:::
::: {.column width=.45}

::: {.block}
### Unsupervised learning

* Input: data $\mathcal{X} = \{\mathbf{x}_{i},\ldots\}$

* Goal 1: learn $f: \mathcal{Z} \to \mathcal{X}$

* Goal 2: hidden/latent states, $\mathcal{Z} = \{\mathbf{z}_{i},\ldots\}$

:::

X: gene expression matrices

:::
::::::


# Representation Learning

## 

\Large

How can we obtain a good representation of data?

\normalsize

## Two strategies to learn a "good" representation

:::::: {.columns}
::: {.column width=.45}

::: {.block}

### 1. Feature engineering

Construct useful combinations of variables, feature engineering

- Principal component analysis (matrix factorization)

- Pattern recognition

- Clustering (average patterns $\approx$ combined features/data)

::: 

:::
::: {.column width=.45}

::: {.incremental}

::: {.block}

### 2. Manifold learning

Change the manifold (a coordinate system) of data

- Probabilistic topic models: word frequency $\to$ simplex
	
- Angular space $\to$ Euclidean space

- Observed high-dim. $\to$ interpretable ones

:::

:::

:::
::::::

We use both strategies in practice.

## Feature engineering by combining multiple vars.

:::::: {.columns}
::: {.column width=.5}

```{r echo=FALSE}
.rnorm <- function(d1,d2) matrix(rnorm(d1*d2),d1,d2)
```

```{r echo=FALSE}
U <- .rnorm(100, 2); V <- .rnorm(100, 2)
X <- U %*% t(V) + .rnorm(100, 100)
```

$$\mathbf{x}_{i} = \mathbf{u}_{i} V^{\top} + \mathcal{N}\!\left(\mathbf{0},I\right)$$

```{r echo=FALSE}
X <- .sort.matrix(X)
```

```{r echo=FALSE, fig.width=2.5, fig.height=2, onslide.plot="1-"}
.matshow(X, .axis.lab=0, .lw=0)
```

:::
::: {.column width=.5}

```{r echo=FALSE}
.svd <- svd(X, nu=2, nv=2)
u.hat <- sweep(.svd$u, 2, .svd$d[1:2], `*`)
```

```{r echo=FALSE, fig.width=2.7, fig.height=2.3, only.plot="2"}
p1 <- .matshow(u.hat) +
    ggtitle("features")
p.times <- ggplot() + theme_void() +
    geom_text(aes(0,0,label="X"), size=5)
p2 <- wrap_plots(.matshow(t(.svd$v)),
                 ggplot() + theme_void(),
                 ncol=1, heights=c(1,9)) +
    ggtitle("feature loading")
wrap_plots(list(p1, p.times, p2), nrow=1, widths=c(1,2,8))
```

:::
::::::

## Feature engineering 

:::::: {.columns}
::: {.column width=.5}

$$\mathbf{x}_{i} = \mathbf{u}_{i} V^{\top} + \mathcal{N}\!\left(\mathbf{0},I\right)$$

* We call each column of U and $V$ a factor (row/column)

* One factor corresponds to the activities/patterns/expressions of 100 genes/samples

* We will discuss further later.

:::
::: {.column width=.5}

```{r echo=FALSE, fig.width=2.7, fig.height=2.1}
p1 <- .matshow(u.hat)
p.times <- ggplot() + theme_void() +
    geom_text(aes(0,0,label="X"), size=5)
p2 <- wrap_plots(.matshow(t(.svd$v)),
                 ggplot() + theme_void(),
                 ncol=1, heights=c(1,9))
wrap_plots(list(p1, p.times, p2), nrow=1, widths=c(1,2,8))
```

:::
::::::

## Manifold learning changes the coordinate system

:::::: {.columns}
::: {.column width=.5}

```{r echo=FALSE}
if(!require(sphereplot)){
    install.packages("sphereplot")
}
.dt <- fread("worldcities.csv")
xx <- sphereplot::sph2car(.dt$lng, .dt$lat)
.svd <- svd(xx)
```

```{r echo=FALSE, fig.width=2.5, fig.height=2.3, only.plot="1-2"}
ggplot(as.data.table(xx), aes(x, y)) +
    ggrastr::rasterise(geom_point(stroke=0, size=.5, alpha=.5, colour="gray40"), dpi=300) +
    ggtitle("x and y coordinates")
```


```{r echo=FALSE, fig.width=2.5, fig.height=2.5, only.plot="3-4"}
ggplot(as.data.table(.svd$u), aes(V1, V2)) +
    ggrastr::rasterise(geom_point(stroke=0, size=.5, alpha=.5, colour="gray40"), dpi=300) +
    ggtitle("PC1 and PC2") +
    xlab("PC 1") + ylab("PC 2")
```

:::
::: {.column width=.5}

```{r echo=FALSE, fig.width=2.5, fig.height=2.3, only.plot="2"}
ggplot(as.data.table(xx), aes(x, z)) +
    ggrastr::rasterise(geom_point(stroke=0, size=.5, alpha=.5, colour="gray40"), dpi=300) +
    ggtitle("x and z coordinates")
```

```{r echo=FALSE, fig.width=2.5, fig.height=2.5, only.plot="4"}
ggplot(as.data.table(.svd$u), aes(V2, V3)) +
    ggrastr::rasterise(geom_point(stroke=0, size=.5, alpha=.5, colour="gray40"), dpi=300) +
    ggtitle("PC2 and PC3") +
    xlab("PC 2") + ylab("PC 3")
```

:::
::::::

```{r echo=FALSE, fig.width=4.5, fig.height=2.8, only.plot=5}
yy <- sphereplot::car2sph(xx) %>% as.data.table
ggplot(yy, aes(`long`, `lat`)) +
    ggrastr::rasterise(geom_point(stroke=0, size=.5, alpha=.5, colour="gray40"), dpi=300) +
    ggtitle("(x,y,z) to (long, lat)") +
    xlab("longitude") + ylab("latitude")
```

\tiny
Data from [https://simplemaps.com/data/world-cities](https://simplemaps.com/data/world-cities)


## Representation learning

:::::: {.columns}
::: {.column width=.45}

### Feature engineering

* Combine samples based on similarity to learn common features

* Combine features/genes/proteins to create a new factor

* Matrix factorization to capture factors that explain large variation 

:::
::: {.column width=.45}

### Manifold learning

* Change the coordinate system

* What's not separable in one space can be spread out in another one

* Apply non-linear functions to a set of features

:::
::::::

# MNIST Example

## A visual example: handwritten digits

```{r mnist-prep, include=FALSE}
library(torch)
mnist.all.rds <- "../data/mnist_all.rds"
if.needed(mnist.all.rds, {
    library(dslabs)
    mnist <- read_mnist()
    X.all <- mnist$train$images / 255
    labels <- mnist$train$labels
    saveRDS(list(X = X.all, labels = labels), mnist.all.rds)
})
mnist.all <- readRDS(mnist.all.rds)
```

```{r mnist-show-helper, include=FALSE}
.show.digits <- function(imgs, nrow = 2, ncol = 5, show_grid = FALSE) {
    plots <- lapply(seq_len(nrow * ncol), function(i) {
        m <- matrix(imgs[i, ], 28, 28, byrow = TRUE)
        ## Rotate 90 degrees clockwise: transpose then reverse columns
        m_rot <- t(m)[, 28:1]
        dt <- reshape2::melt(m_rot)
        p <- ggplot(dt, aes(Var1, Var2, fill = value)) +
            geom_raster() +
            scale_fill_gradient(low = "white", high = "black") +
            coord_equal() +
            theme_void() + theme(legend.position = "none")

        if (show_grid) {
            p <- p +
                geom_vline(xintercept = seq(0.5, 28.5, 1), colour = "gray80", linewidth = 0.2) +
                geom_hline(yintercept = seq(0.5, 28.5, 1), colour = "gray80", linewidth = 0.2) +
                scale_x_continuous(expand = c(0, 0)) +
                scale_y_continuous(expand = c(0, 0))
        }
        p
    })
    wrap_plots(plots, nrow = nrow, ncol = ncol)
}

## Random rotation via affine grid
.rotate.batch <- function(x, max.angle = 30) {
    n <- x$shape[1]
    x4d <- x$view(c(n, 1L, 28L, 28L))
    angles <- runif(n, -max.angle, max.angle) * pi / 180
    theta <- torch_zeros(n, 2, 3)
    theta[, 1, 1] <- cos(angles)
    theta[, 1, 2] <- -sin(angles)
    theta[, 2, 1] <- sin(angles)
    theta[, 2, 2] <- cos(angles)
    grid <- nnf_affine_grid(theta, c(n, 1L, 28L, 28L))
    rotated <- nnf_grid_sample(x4d, grid, mode = "bilinear", padding_mode = "zeros")
    rotated$view(c(n, 784L))
}
```

```{r mnist-show-orig, echo=FALSE, fig.width=6, fig.height=2.7}
set.seed(1)
show.idx <- sapply(0:9, function(d) {
    idx <- which(mnist.all$labels == d)
    sample(idx, 1)
})
.show.digits(mnist.all$X[show.idx, ], show_grid = TRUE)
```

* 28 $\times$ 28 = 784 pixels per image — can we compress to $K = 12$?
* We train on all digits (0-9)

## The autoencoder bottleneck $\approx$ PCA

\begin{center}
\begin{tikzpicture}[
    layer/.style={draw, fill=gray!20, minimum width=0.6cm, rounded corners=2pt},
    arr/.style={->, thick, >=stealth}
]
\usetikzlibrary{decorations.pathreplacing}
\node[layer, minimum height=3.5cm] (input) at (0,0) {};
\node[below=2pt] at (input.south) {\scriptsize 784};
\node[layer, minimum height=1cm] (z) at (2,0) {};
\node[below=2pt] at (z.south) {\scriptsize \textbf{12}};
\node[layer, minimum height=3.5cm] (output) at (4,0) {};
\node[below=2pt] at (output.south) {\scriptsize 784};
\draw[arr] (input) -- (z);
\draw[arr] (z) -- (output);
\node[above=8pt] at (input.north) {\small $\mathbf{x}$};
\node[above=8pt] at (z.north) {\small $\mathbf{z}$};
\node[above=8pt] at (output.north) {\small $\hat{\mathbf{x}}$};
\draw[decorate, decoration={brace, mirror, amplitude=5pt}] (-0.5,-2.3) -- (1.7,-2.3)
    node[midway, below=6pt] {\small encoder};
\draw[decorate, decoration={brace, mirror, amplitude=5pt}] (2.3,-2.3) -- (4.5,-2.3)
    node[midway, below=6pt] {\small decoder};
\end{tikzpicture}
\end{center}

* All information about the digit must pass through **12 numbers**

## AE vs. VAE on all digits (K=12)

```{r mnist-ae-train, include=FALSE}
mnist.ae.rds <- "../data/mnist_ae_model.rds"
if.needed(mnist.ae.rds, {
    X.m <- torch_tensor(mnist.all$X)
    n <- nrow(mnist.all$X)
    d <- 784L
    K.m <- 12L

    ## Autoencoder
    set.seed(42)
    encoder <- nn_linear(d, K.m)
    decoder <- nn_linear(K.m, d)
    optimizer <- optim_adam(c(encoder$parameters, decoder$parameters), lr = 1e-3)

    ## Training loop
    for (epoch in seq_len(500)) {
        idx <- sample(n, 128)
        x_batch <- X.m[idx, ]

        # Forward pass
        z <- encoder(x_batch)
        x_recon <- decoder(z) |> torch_sigmoid()

        # Loss: binary cross-entropy
        loss <- nnf_binary_cross_entropy(x_recon, x_batch, reduction = "mean")

        # Backward pass
        optimizer$zero_grad()
        loss$backward()
        optimizer$step()
    }

    ## Get reconstructions
    with_no_grad({
        z_all <- encoder(X.m) |> as.array()
        recon_all <- decoder(encoder(X.m)) |> torch_sigmoid() |> as.array()
    })

    saveRDS(list(z = z_all, recon = recon_all), mnist.ae.rds)
})
mnist.ae <- readRDS(mnist.ae.rds)
```

```{r mnist-vae-train, include=FALSE}
mnist.vae.rds <- "../data/mnist_vae_model.rds"
if.needed(mnist.vae.rds, {
    X.m <- torch_tensor(mnist.all$X)
    n <- nrow(mnist.all$X)
    d <- 784L
    K.m <- 12L

    ## Variational Autoencoder
    set.seed(123)  # Different seed from AE
    encoder_mu <- nn_linear(d, K.m)
    encoder_logvar <- nn_linear(d, K.m)
    decoder <- nn_linear(K.m, d)

    optimizer <- optim_adam(c(encoder_mu$parameters,
                              encoder_logvar$parameters,
                              decoder$parameters), lr = 1e-3)

    ## Training loop
    for (epoch in seq_len(500)) {
        idx <- sample(n, 128)
        x_batch <- X.m[idx, ]
        batch_size <- x_batch$shape[1]

        # Encode
        mu <- encoder_mu(x_batch)
        logvar <- encoder_logvar(x_batch) |> torch_clamp(-8, 8)

        # Reparameterization trick: z = mu + sigma * epsilon
        std <- torch_exp(0.5 * logvar)
        eps <- torch_randn_like(std)
        z <- mu + std * eps

        # Decode
        x_recon <- decoder(z) |> torch_sigmoid()

        # Reconstruction loss (BCE, averaged over batch and features)
        recon_loss <- nnf_binary_cross_entropy(x_recon, x_batch, reduction = "sum")
        recon_loss <- recon_loss / batch_size

        # KL divergence: -0.5 * sum(1 + log(sigma^2) - mu^2 - sigma^2)
        kl_loss <- -0.5 * torch_sum(1 + logvar - mu$pow(2) - logvar$exp())
        kl_loss <- kl_loss / batch_size

        # Total loss
        loss <- recon_loss + kl_loss

        # Backward pass
        optimizer$zero_grad()
        loss$backward()
        optimizer$step()
    }

    ## Get reconstructions (use mean, not sampling)
    with_no_grad({
        mu_all <- encoder_mu(X.m)
        z_all <- mu_all |> as.array()
        recon_all <- decoder(mu_all) |> torch_sigmoid() |> as.array()
    })

    saveRDS(list(z = z_all, recon = recon_all), mnist.vae.rds)
})
mnist.vae <- readRDS(mnist.vae.rds)
```

```{r mnist-models-combine, include=FALSE}
mnist.models <- list(
    z.ae = mnist.ae$z,
    z.vae = mnist.vae$z,
    recon.ae = mnist.ae$recon,
    recon.vae = mnist.vae$recon,
    labels = mnist.all$labels,
    show.idx = show.idx
)
```

:::::: {.columns}
::: {.column width=.33}

\centerline{\textbf{Original}}

```{r mnist-recon-orig, echo=FALSE, fig.width=1.8, fig.height=1.5}
.show.digits(mnist.all$X[show.idx, ], nrow = 2, ncol = 5)
```

:::
::: {.column width=.33}

\centerline{\textbf{AE (K=12)}}

```{r mnist-recon-ae, echo=FALSE, fig.width=1.8, fig.height=1.5}
.show.digits(mnist.models$recon.ae[show.idx, ], nrow = 2, ncol = 5)
```

:::
::: {.column width=.33}

\centerline{\textbf{VAE (K=12)}}

```{r mnist-recon-vae, echo=FALSE, fig.width=1.8, fig.height=1.5}
.show.digits(mnist.models$recon.vae[show.idx, ], nrow = 2, ncol = 5)
```

:::
::::::

* Trained on all digits (0-9)

## Latent representations (heatmaps sorted by digit)

```{r}
## Sort by digit label
digit.order <- order(mnist.models$labels)
z.ae <- mnist.models$z.ae[digit.order, ]
z.vae <- mnist.models$z.vae[digit.order, ]

## Sample for visualization (100 per digit = 1000 total)
set.seed(42)
z.ae.list <- lapply(0:9, function(d) {
    idx <- which(mnist.models$labels[digit.order] == d)
    sub.idx <- sample(idx, min(100, length(idx)))
    dt <- as.data.table(z.ae[sub.idx, ])
    dt$digit <- factor(d)
    dt$sample <- 1:nrow(dt)
    dt
})
z.ae.dt <- rbindlist(z.ae.list)

z.vae.list <- lapply(0:9, function(d) {
    idx <- which(mnist.models$labels[digit.order] == d)
    sub.idx <- sample(idx, min(100, length(idx)))
    dt <- as.data.table(z.vae[sub.idx, ])
    dt$digit <- factor(d)
    dt$sample <- 1:nrow(dt)
    dt
})
z.vae.dt <- rbindlist(z.vae.list)
```

```{r mnist-latent-ae, echo=FALSE, fig.width=5.5, fig.height=2.5, only.plot=1}
z.ae.long <- melt(z.ae.dt, id.vars = c("sample", "digit"),
                   variable.name = "dim", value.name = "value")
ggplot(z.ae.long, aes(x = sample, y = dim, fill = value)) +
    ggrastr::rasterise(geom_tile(), dpi = 300) +
    facet_grid(. ~ digit, scales = "free_x", space = "free_x") +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
    ggtitle("AE") + xlab("") + ylab("latent dims") +
    theme_minimal() + theme(axis.text.x = element_blank())
```

```{r mnist-latent-vae, echo=FALSE, fig.width=5.5, fig.height=2.5, only.plot=2}
z.vae.long <- melt(z.vae.dt, id.vars = c("sample", "digit"),
                    variable.name = "dim", value.name = "value")
ggplot(z.vae.long, aes(x = sample, y = dim, fill = value)) +
    ggrastr::rasterise(geom_tile(), dpi = 300) +
    facet_grid(. ~ digit, scales = "free_x", space = "free_x") +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
    ggtitle("VAE") + xlab("") + ylab("latent dims") +
    theme_minimal() + theme(axis.text.x = element_blank())
```

* What differences do you observe between AE and VAE?
* How are the digits distributed in latent space?

## Why non-linearity?

* **Linear layers alone** cannot learn complex patterns

* Stacking linear layers: $f(g(\mathbf{x})) = \mathbf{W}_2(\mathbf{W}_1\mathbf{x}) = (\mathbf{W}_2\mathbf{W}_1)\mathbf{x}$

* Result is still linear — just a single matrix multiplication

## ReLU activation

* **ReLU**: $\text{ReLU}(z) = \max(0, z)$

* Introduces non-linearity between layers

* Simple, computationally efficient, avoids vanishing gradients

## Two-layer VAE architecture

* $\mathbf{x} \xrightarrow{\text{Linear}} \mathbf{h} \xrightarrow{\text{ReLU}} \mathbf{h}' \xrightarrow{\text{Linear}} \mathbf{z}$

* Hidden layer $\mathbf{h}$ (128 dimensions) learns intermediate features

* Batch normalization stabilizes training

## Nonlinear layers VAE results

```{r mnist-mlp-train, include=FALSE}
mnist.mlp.rds <- "../data/mnist_mlp_models.rds"
if.needed(mnist.mlp.rds, {
    X.m <- torch_tensor(mnist.all$X)
    n <- nrow(mnist.all$X)
    d <- 784L
    K.m <- 12L
    h <- 128L  ## hidden dimension

    ## Nonlinear layers VAE with batch normalization
    set.seed(456)  # Different from linear models
    encoder_hidden <- nn_linear(d, h)
    encoder_bn <- nn_batch_norm1d(h)
    encoder_mu <- nn_linear(h, K.m)
    encoder_logvar <- nn_linear(h, K.m)

    decoder_hidden <- nn_linear(K.m, h)
    decoder_bn <- nn_batch_norm1d(h)
    decoder_out <- nn_linear(h, d)

    optimizer <- optim_adam(c(encoder_hidden$parameters, encoder_bn$parameters,
                              encoder_mu$parameters, encoder_logvar$parameters,
                              decoder_hidden$parameters, decoder_bn$parameters,
                              decoder_out$parameters), lr = 1e-3)

    ## Training loop
    for (epoch in seq_len(500)) {
        idx <- sample(n, 128)
        x_batch <- X.m[idx, ]
        batch_size <- x_batch$shape[1]

        ## Encoder: x -> hidden -> mu, logvar
        h_enc <- encoder_hidden(x_batch) |> encoder_bn() |> nnf_relu()
        mu <- encoder_mu(h_enc)
        logvar <- encoder_logvar(h_enc) |> torch_clamp(-8, 8)

        ## Reparameterization: z = mu + std * epsilon
        std <- torch_exp(0.5 * logvar)
        eps <- torch_randn_like(std)
        z <- mu + std * eps

        ## Decoder: z -> hidden -> x
        h_dec <- decoder_hidden(z) |> decoder_bn() |> nnf_relu()
        x_recon <- decoder_out(h_dec) |> torch_sigmoid()

        ## Reconstruction loss (sum then normalise)
        recon_loss <- nnf_binary_cross_entropy(x_recon, x_batch, reduction = "sum")
        recon_loss <- recon_loss / batch_size

        ## KL divergence (sum then normalise)
        kl_loss <- -0.5 * torch_sum(1 + logvar - mu$pow(2) - logvar$exp())
        kl_loss <- kl_loss / batch_size

        ## Total loss
        loss <- recon_loss + kl_loss

        ## Backward pass
        optimizer$zero_grad()
        loss$backward()
        optimizer$step()
    }

    ## Get reconstructions (use mean, not sampling)
    with_no_grad({
        h_enc <- encoder_hidden(X.m) |> encoder_bn() |> nnf_relu()
        mu_all <- encoder_mu(h_enc)
        h_dec <- decoder_hidden(mu_all) |> decoder_bn() |> nnf_relu()
        recon_all <- decoder_out(h_dec) |> torch_sigmoid() |> as.array()
        z_all <- mu_all |> as.array()
    })

    saveRDS(list(recon = recon_all, z = z_all, labels = mnist.all$labels,
                 show.idx = show.idx), mnist.mlp.rds)
})
mnist.mlp <- readRDS(mnist.mlp.rds)
```

:::::: {.columns}
::: {.column width=.5}

\centerline{\textbf{Linear VAE}}

```{r mnist-mlp-vae-linear, echo=FALSE, fig.width=2.5, fig.height=2}
.show.digits(mnist.models$recon.vae[show.idx, ], nrow = 2, ncol = 5)
```

:::
::: {.column width=.5}

\centerline{\textbf{Nonlinear layers VAE}}

```{r mnist-mlp-vae-mlp, echo=FALSE, fig.width=2.5, fig.height=2}
.show.digits(mnist.mlp$recon[show.idx, ], nrow = 2, ncol = 5)
```

:::
::::::

* 2 encoder layers: 784 → 128 → 12 (with batch norm + ReLU)
* 2 decoder layers: 12 → 128 → 784 (with batch norm + ReLU)

## Nonlinear layers VAE architecture

\begin{center}
\begin{tikzpicture}[
    layer/.style={draw, fill=gray!20, minimum width=0.6cm, rounded corners=2pt},
    arr/.style={->, thick, >=stealth}
]

%% Encoder
\node[layer, minimum height=3.5cm] (x) at (0,0) {\small $\mathbf{x}$};
\node[below=2pt] at (x.south) {\scriptsize 784};

\node[layer, minimum height=1.8cm] (h1) at (1.5,0) {};
\node[below=2pt] at (h1.south) {\scriptsize 128};
\draw[arr] (x) -- (h1);

\node[layer, minimum height=0.8cm, fill=blue!10] (mu) at (3,0.5) {\scriptsize $\boldsymbol{\mu}$};
\node[below=2pt] at (mu.south) {\scriptsize 12};
\draw[arr] (h1) -- (mu);

\node[layer, minimum height=0.8cm, fill=blue!10] (logvar) at (3,-0.5) {\scriptsize $\log\boldsymbol{\sigma}^2$};
\node[below=4pt] at (logvar.south) {\scriptsize 12};
\draw[arr] (h1) -- (logvar);

%% Reparameterization
\node[layer, minimum height=0.8cm, fill=green!10] (eps) at (4.5,-0.5) {\scriptsize $\boldsymbol{\epsilon}$};

\node[layer, minimum height=0.8cm, fill=orange!20] (z) at (5.5,0) {\small $\mathbf{z}$};
\node[below=4pt] at (z.south) {\tiny $\boldsymbol{\epsilon} \sim \mathcal{N}(0,\mathbf{I})$};
\draw[arr] (mu) -- (z);
\draw[arr] (logvar) -- (z);
\draw[arr] (eps) -- (z);

%% Decoder
\node[layer, minimum height=1.8cm] (h2) at (7,0) {};
\node[below=2pt] at (h2.south) {\scriptsize 128};
\draw[arr] (z) -- (h2);

\node[layer, minimum height=3.5cm] (xhat) at (8.5,0) {\small $\hat{\mathbf{x}}$};
\node[below=2pt] at (xhat.south) {\scriptsize 784};
\draw[arr] (h2) -- (xhat);

\end{tikzpicture}
\end{center}

* **Reparameterization trick**: $\mathbf{z} = \boldsymbol{\mu} + \boldsymbol{\sigma} \odot \boldsymbol{\epsilon}$ where $\boldsymbol{\epsilon} \sim \mathcal{N}(0, \mathbf{I})$

## t-SNE visualization of latent space

```{r mnist-tsne, include=FALSE}
library(Rtsne)

mnist.tsne.rds <- "../data/mnist_tsne.rds"
if.needed(mnist.tsne.rds, {
    set.seed(42)

    ## Linear VAE latent space
    tsne_linear <- Rtsne(mnist.vae$z, dims = 2, perplexity = 30, verbose = FALSE)
    tsne_linear_dt <- data.table(
        x = tsne_linear$Y[, 1],
        y = tsne_linear$Y[, 2],
        digit = factor(mnist.all$labels)
    )

    ## Nonlinear layers VAE latent space
    tsne_mlp <- Rtsne(mnist.mlp$z, dims = 2, perplexity = 30, verbose = FALSE)
    tsne_mlp_dt <- data.table(
        x = tsne_mlp$Y[, 1],
        y = tsne_mlp$Y[, 2],
        digit = factor(mnist.all$labels)
    )

    saveRDS(list(linear = tsne_linear_dt, mlp = tsne_mlp_dt), mnist.tsne.rds)
})
mnist.tsne <- readRDS(mnist.tsne.rds)
tsne_linear_dt <- mnist.tsne$linear
tsne_mlp_dt <- mnist.tsne$mlp
```

:::::: {.columns}
::: {.column width=.5}

\centerline{\textbf{Linear VAE}}

```{r tsne-linear, echo=FALSE, fig.width=2.8, fig.height=2.5}
ggplot(tsne_linear_dt, aes(x, y, colour = digit)) +
    geom_point(size = 0.3, alpha = 0.6, stroke = 0) +
    scale_colour_brewer(palette = "Paired") +
    theme_minimal() +
    theme(legend.position = "right", legend.key.size = unit(0.3, "lines")) +
    labs(x = "t-SNE 1", y = "t-SNE 2")
```

:::
::: {.column width=.5}

\centerline{\textbf{Nonlinear layers VAE}}

```{r tsne-mlp, echo=FALSE, fig.width=2.8, fig.height=2.5}
ggplot(tsne_mlp_dt, aes(x, y, colour = digit)) +
    geom_point(size = 0.3, alpha = 0.6, stroke = 0) +
    scale_colour_brewer(palette = "Paired") +
    theme_minimal() +
    theme(legend.position = "right", legend.key.size = unit(0.3, "lines")) +
    labs(x = "t-SNE 1", y = "t-SNE 2")
```

:::
::::::

* t-SNE reduces K=12 dimensions to 2D for visualization

## How do we learn nonlinear encoding-decoding?

\Large

* Encoder: $\mathbf{z} \gets f(\cdots f(\mathbf{x}))$

* Decoder: $\hat{\mathbf{x}} \gets g(\cdots g(\mathbf{z}))$

* Loss: $\mathcal{L} = L(\mathbf{x}, \hat{\mathbf{x}})$

* Gradient (differentiation): $\nabla \mathcal{L}$

\normalsize

# Building Autoencoders on Single-Cell Data

## The PDAC data

```{r load-seurat, message=FALSE, warning=FALSE}
library(Seurat)
```

```{r load-pdac, include=FALSE}
pdac.rds <- "../data/GSE155698_PDAC_Steele2020/pdac_tissues123_processed.rds"
if.needed(pdac.rds, {
    ## Load all three tissues
    data.dirs <- paste0("../data/GSE155698_PDAC_Steele2020/PDAC_TISSUE_", 1:3, "/")

    pdac.list <- lapply(1:3, function(i) {
        counts <- Read10X(data.dir = data.dirs[i])
        obj <- CreateSeuratObject(counts = counts,
                                  project = paste0("PDAC_TISSUE_", i),
                                  min.cells = 3,
                                  min.features = 200)
        obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = "^MT-")
        obj <- subset(obj,
                      nFeature_RNA > 200 &
                      nFeature_RNA < 6000 &
                      percent.mt < 20)
        obj
    })

    ## Merge tissues
    pdac <- merge(pdac.list[[1]],
                  y = c(pdac.list[[2]], pdac.list[[3]]),
                  add.cell.ids = c("T1", "T2", "T3"))

    ## Join layers (required for v5 assay)
    pdac <- JoinLayers(pdac)

    ## Process combined data (NO batch correction - we'll use scVI for that!)
    pdac <- pdac |>
        NormalizeData() |>
        FindVariableFeatures(nfeatures = 2000) |>
        ScaleData() |>
        RunPCA(npcs = 30) |>
        FindNeighbors(dims = 1:20) |>
        FindClusters(resolution = 0.5) |>
        RunUMAP(dims = 1:20)

    saveRDS(pdac, pdac.rds)
})
pdac <- readRDS(pdac.rds)
```

```{r show-pdac-pipeline, eval=FALSE, echo=TRUE}
## Load and merge tissues 1-3
pdac.list <- lapply(1:3, function(i) {
    Read10X(data.dir = paste0("PDAC_TISSUE_", i, "/")) |>
        CreateSeuratObject(project = paste0("PDAC_T", i))
})
pdac <- merge(pdac.list[[1]], y = c(pdac.list[[2]], pdac.list[[3]]))

## Process
pdac <- pdac |>
    NormalizeData() |>
    FindVariableFeatures(nfeatures = 2000) |>
    ScaleData() |>
    RunPCA(npcs = 30) |>
    FindNeighbors(dims = 1:20) |>
    FindClusters(resolution = 0.5) |>
    RunUMAP(dims = 1:20)
```

* `r num.int(nrow(pdac))` genes/features/rows and `r num.int(ncol(pdac))` cells/columns

## PDAC: `r ncol(pdac)` cells from 3 tissues

```{r show-pdac-umap, echo=FALSE, fig.width=4.5, fig.height=3.5}
DimPlot(pdac, reduction = "umap", label = TRUE, label.size = 3) +
    theme(legend.position = "right")
```

* Seurat clustering identifies `r length(unique(pdac$seurat_clusters))` clusters


```{r load-torch, include=FALSE}
library(torch)
```

```{r prep-torch-data, include=FALSE}
torch.data.rds <- "../data/GSE155698_PDAC_Steele2020/pdac_torch_data.rds"
if.needed(torch.data.rds, {
    ## Use RAW COUNTS (not normalized/batch-corrected)
    ## VAE/scVI models need raw counts to learn distributions
    counts <- GetAssayData(pdac, layer = "counts")
    hvg <- VariableFeatures(pdac)
    X <- t(as.matrix(counts[hvg, ]))
    saveRDS(list(X = X, hvg = hvg), torch.data.rds)
})
torch.data <- readRDS(torch.data.rds)
X.tensor <- torch_tensor(torch.data$X)
n.cells <- nrow(torch.data$X)
n.genes <- ncol(torch.data$X)
```

```{r prep-batch-data, include=FALSE}
batch.data.rds <- "../data/GSE155698_PDAC_Steele2020/batch_data.rds"
if.needed(batch.data.rds, {
    ## Extract batch labels (orig.ident contains tissue info)
    batch.labels <- pdac$orig.ident
    ## Convert to numeric: PDAC_TISSUE_1 -> 0, PDAC_TISSUE_2 -> 1, PDAC_TISSUE_3 -> 2
    batch.idx <- as.integer(factor(batch.labels,
                                    levels = c("PDAC_TISSUE_1", "PDAC_TISSUE_2", "PDAC_TISSUE_3"))) - 1
    ## Extract library sizes (total counts per cell)
    lib.sizes <- Matrix::colSums(GetAssayData(pdac, layer = "counts"))
    saveRDS(list(batch.idx = batch.idx, lib.sizes = lib.sizes), batch.data.rds)
})
batch.data <- readRDS(batch.data.rds)
batch.tensor <- torch_tensor(batch.data$batch.idx, dtype = torch_long())
lib.size.tensor <- torch_tensor(batch.data$lib.sizes)
n.batches <- length(unique(batch.data$batch.idx))
```

## Prepare the data for `torch`

```{r show-prep-data, echo=TRUE, eval=FALSE}
counts <- GetAssayData(pdac, layer = "counts")
hvg <- VariableFeatures(pdac)
X <- t(as.matrix(counts[hvg, ]))
```

* Extract **raw counts** (not normalized/batch-corrected)
* Rows = cells, columns = genes (2000 highly variable genes)

```{r show-dim, echo=TRUE}
dim(torch.data$X)
```

## What does the data look like?

```{r show-data-corner, echo=TRUE}
torch.data$X[1:5, 1:6]
```

* Sparse count matrix -- most entries are zero

```{r show-data-summary, echo=TRUE}
mean(torch.data$X == 0)
```

## Convert to a `torch` tensor

```{r show-tensor, echo=TRUE}
X.tensor <- torch_tensor(torch.data$X)
X.tensor$shape
```

* This is our input: each row $\mathbf{x}_i$ is one cell's expression profile

## What is an autoencoder?

\large

* Compress high-dimensional data into a low-dimensional **code**

* Then reconstruct the original data from that code

\normalsize

\vfill

$$\mathbf{x} \xrightarrow{\text{encoder}} \mathbf{z} \xrightarrow{\text{decoder}} \hat{\mathbf{x}}$$

\vfill

* The **bottleneck** $\mathbf{z}$ forces the model to learn a compact summary

## A linear encoder

$$\mathbf{z}_{i} = W \, \mathbf{x}_{i} + \mathbf{b}$$

* Input: $\mathbf{x}_i \in \mathbb{R}^{p}$ (2000 genes)
* Output: $\mathbf{z}_i \in \mathbb{R}^{K}$ (e.g., $K = 10$ latent dimensions)

```{r def-encoder, echo=TRUE}
K <- 10

linear.encoder <- nn_module(
    initialize = function(d.in, K) {
        self$linear <- nn_linear(d.in, K)
    },
    forward = function(x) {
        x |> torch_log1p() |> self$linear()
    }
)
```

## A linear decoder

$$\hat{\mathbf{x}}_{i} = W' \mathbf{z}_{i} + \mathbf{b}'$$

* Input: $\mathbf{z}_i \in \mathbb{R}^{K}$
* Output: $\hat{\mathbf{x}}_i \in \mathbb{R}^{p}$ (reconstructed expression)

```{r def-decoder, echo=TRUE}
linear.decoder <- nn_module(
    initialize = function(K, d.out) {
        self$linear <- nn_linear(K, d.out)
    },
    forward = function(z) {
        self$linear(z)
    }
)
```

## Putting them together

```{r def-autoencoder, echo=TRUE}
autoencoder <- nn_module(
    initialize = function(d.in, K) {
        self$enc <- linear.encoder(d.in, K)
        self$dec <- linear.decoder(K, d.in)
    },
    forward = function(x) {
        z <- self$enc(x)
        x.hat <- self$dec(z)
        list(z = z, x.hat = x.hat)
    }
)
```

* Encoder compresses $p \to K$, decoder expands $K \to p$

## Loss function: mean squared error

$$\mathcal{L} = \frac{1}{n}\sum_{i} \| \mathbf{x}_{i} - \hat{\mathbf{x}}_{i} \|^{2}$$

* How well can we reconstruct the original data?

```{r def-mse-loss, echo=TRUE}
recon.loss <- function(x, x.hat) {
    torch_mean(torch_sum((x - x.hat)^2, dim = 2))
}
```

## Training: set up the model and optimizer

```{r train-ae-setup, echo=TRUE}
set.seed(42)
model.ae <- autoencoder(n.genes, K)
optimizer <- optim_adam(model.ae$parameters, lr = 1e-3)
```

```{r train-ae-setup-show, echo=TRUE}
model.ae
```

## First, let's sort the observed data for visualization

```{r prep-obs-heatmap, include=FALSE}
obs.rds <- "../data/GSE155698_PDAC_Steele2020/obs_sorted.rds"
if.needed(obs.rds, {
    set.seed(1)
    show.idx <- sample(n.cells, 200)
    X.show <- torch.data$X[show.idx, ]
    ## pick top 200 genes by variance for clearer visualization
    gene.var <- apply(X.show, 2, var)
    top.genes <- order(gene.var, decreasing = TRUE)[1:200]
    X.show <- X.show[, top.genes]
    X.log <- log1p(X.show)
    X.scaled <- scale(X.log)
    col.center <- attr(X.scaled, "scaled:center")
    col.scale <- attr(X.scaled, "scaled:scale")
    ## get sort order via order.pair
    .op <- as.matrix(X.scaled) %>%
        reshape2::melt() %>%
        rename(row = Var1, col = Var2, weight = value) %>%
        order.pair()
    row.perm <- match(.op$rows, rownames(X.scaled))
    col.perm <- match(.op$cols, colnames(X.scaled))
    X.sorted <- X.scaled[row.perm, col.perm]
    saveRDS(list(idx = show.idx, top.genes = top.genes,
                 col.center = col.center, col.scale = col.scale,
                 row.perm = row.perm, col.perm = col.perm,
                 X.sorted = X.sorted), obs.rds)
})
obs.data <- readRDS(obs.rds)
.clim <- c(-3, 3)
```

```{r plot-obs-heatmap, echo=FALSE, fig.width=5, fig.height=2.8}
.matshow(obs.data$X.sorted, .axis.lab = 0, .limits = .clim) +
    ggtitle("Observed (log1p, scaled) — 200 cells x 200 genes")
```

## Training loop

```{r train-ae-show, echo=TRUE, eval=FALSE}
n.epoch <- 2000
batch.size <- 256
for (epoch in seq_len(n.epoch)) {
    idx <- sample(n.cells, batch.size)
    x.batch <- X.tensor[idx, ]
    out <- model.ae(x.batch)
    loss <- recon.loss(torch_log1p(x.batch), out$x.hat)
    optimizer$zero_grad()
    loss$backward()
    optimizer$step()
}
```

* Sample a minibatch $\to$ forward pass $\to$ compute loss $\to$ backprop $\to$ update weights

```{r train-ae, include=FALSE}
ae.rds <- "../data/GSE155698_PDAC_Steele2020/ae_trained.rds"
if.needed(ae.rds, {
    n.epoch <- 2000
    batch.size <- 256
    loss.history <- c()
    snapshots <- list()
    record.at <- c(1, 100, 500, 2000)
    for (epoch in seq_len(n.epoch)) {
        idx <- sample(n.cells, batch.size)
        x.batch <- X.tensor[idx, ]
        out <- model.ae(x.batch)
        loss <- recon.loss(torch_log1p(x.batch), out$x.hat)
        optimizer$zero_grad()
        loss$backward()
        optimizer$step()
        loss.history <- c(loss.history, loss$item())
        if (epoch %in% record.at) {
            with_no_grad({
                out.all <- model.ae(X.tensor[obs.data$idx, ])
                snapshots[[as.character(epoch)]] <-
                    out.all$x.hat |> as.array()
            })
        }
    }
    with_no_grad({
        z.all <- model.ae$enc(X.tensor) |> as.array()
    })
    saveRDS(list(loss = loss.history,
                 z = z.all,
                 snapshots = snapshots), ae.rds)
})
ae.result <- readRDS(ae.rds)
```

## Training loss

```{r plot-ae-loss, echo=FALSE, fig.width=4.5, fig.height=2.5}
data.table(epoch = seq_along(ae.result$loss),
           loss = ae.result$loss) |>
    ggplot(aes(epoch, loss)) +
    geom_point(size = 0.3, alpha = 0.5) +
    geom_smooth(se = FALSE, colour = "firebrick") +
    ylab("reconstruction loss") +
    ggtitle("Autoencoder training")
```

## Reconstruction across epochs

```{r plot-ae-snapshots-helper, include=FALSE}
.recon.heatmap <- function(recon.mat, title = "") {
    recon.sub <- recon.mat[, obs.data$top.genes]
    recon.scaled <- sweep(recon.sub, 2, obs.data$col.center, `-`)
    recon.scaled <- sweep(recon.scaled, 2, obs.data$col.scale, `/`)
    recon.scaled[is.na(recon.scaled)] <- 0
    .mat <- recon.scaled[obs.data$row.perm, obs.data$col.perm]
    .matshow(.mat, .axis.lab = 0, .limits = .clim) + ggtitle(title)
}
```

```{r}
p.obs <- .matshow(obs.data$X.sorted, .axis.lab = 0, .limits = .clim) +
    ggtitle("Observed")
p.ep1 <- .recon.heatmap(ae.result$snapshots[["1"]], "epoch 1")
p.ep100 <- .recon.heatmap(ae.result$snapshots[["100"]], "epoch 100")
p.ep500 <- .recon.heatmap(ae.result$snapshots[["500"]], "epoch 500")
```


```{r plot-ae-snapshots-0, echo=FALSE, fig.width=5.8, fig.height=2.9, only.plot="1"}
p.obs | p.ep1
```

```{r plot-ae-snapshots-1, echo=FALSE, fig.width=5.8, fig.height=2.9, only.plot="2"}
p.obs | p.ep100
```

```{r plot-ae-snapshots-2, echo=FALSE, fig.width=5.8, fig.height=2.9, only.plot="3"}
p.obs | p.ep500
```



## Observed vs. reconstructed (epoch 2000)

```{r plot-ae-compare, echo=FALSE, fig.width=5.8, fig.height=2.9}
p.obs <- .matshow(obs.data$X.sorted, .axis.lab = 0, .limits = .clim) +
    ggtitle("Observed")
p.recon <- .recon.heatmap(ae.result$snapshots[["2000"]], "Reconstructed (epoch 2000)")
p.obs + p.recon
```

## Latent space of the autoencoder

```{r plot-ae-latent, echo=FALSE, fig.width=5.5, fig.height=2.8}
z.dt <- as.data.table(ae.result$z)
colnames(z.dt) <- paste0("z", 1:K)
p1 <- ggplot(z.dt, aes(z1, z2)) +
    ggrastr::rasterise(geom_point(size = 0.2, alpha = 0.3), dpi = 300) +
    ggtitle("z1 vs z2")
p2 <- ggplot(z.dt, aes(z3, z4)) +
    ggrastr::rasterise(geom_point(size = 0.2, alpha = 0.3), dpi = 300) +
    ggtitle("z3 vs z4")
p1 + p2
```

* Each dot is a cell; the latent space has no particular structure yet


# Variational Autoencoder on single cell data

## From autoencoder to VAE: one key change

\large

Autoencoder encoder outputs a **point**:

$$\mathbf{z}_{i} = W \mathbf{x}_{i} + \mathbf{b}$$

\vfill

VAE encoder outputs a **distribution**:

$$\boldsymbol{\mu}_{i} = W_{\mu} \mathbf{x}_{i} + \mathbf{b}_{\mu}, \quad \log \boldsymbol{\sigma}^{2}_{i} = W_{\sigma} \mathbf{x}_{i} + \mathbf{b}_{\sigma}$$

$$\mathbf{z}_{i} \sim \mathcal{N}(\boldsymbol{\mu}_{i},\, \text{diag}(\boldsymbol{\sigma}^{2}_{i}))$$

## The reparameterization trick

We can't backpropagate through a random sample. But we can rewrite:

$$\mathbf{z}_{i} = \boldsymbol{\mu}_{i} + \boldsymbol{\sigma}_{i} \odot \boldsymbol{\epsilon}, \quad \boldsymbol{\epsilon} \sim \mathcal{N}(\mathbf{0}, I)$$

\vfill

* The randomness is in $\boldsymbol{\epsilon}$ (external), not in the parameters
* Gradients flow through $\boldsymbol{\mu}$ and $\boldsymbol{\sigma}$ as usual

## VAE encoder in `torch`

```{r def-vae-encoder, echo=TRUE}
vae.encoder <- nn_module(
    initialize = function(d.in, K, h = 128) {
        self$hidden <- nn_linear(d.in, h)
        self$bn <- nn_batch_norm1d(h)
        self$z.mean <- nn_linear(h, K)
        self$z.logvar <- nn_linear(h, K)
    },
    forward = function(x) {
        x <- torch_log1p(x)
        h <- self$hidden(x) |> self$bn() |> nnf_relu()
        mu <- self$z.mean(h)
        lv <- self$z.logvar(h) |> torch_clamp(-8, 8)
        ## reparameterization trick
        z <- mu + torch_exp(lv / 2) * torch_randn_like(lv)
        list(z = z, mu = mu, logvar = lv)
    }
)
```

## What changed? Only the encoder

:::::: {.columns}
::: {.column width=.4}

### Autoencoder encoder

```r
forward = function(x) {
  x |>
    torch_log1p() |>
    self$linear()
}
```

Returns a **point** $\mathbf{z}$

:::
::: {.column width=.55}

### VAE encoder

```r
forward = function(x) {
  x <- torch_log1p(x)
  h <- self$hidden(x) |>
       self$bn() |> nnf_relu()
  mu <- self$z.mean(h)
  lv <- self$z.logvar(h) |>
       torch_clamp(-8, 8)
  z <- mu + torch_exp(lv/2) *
       torch_randn_like(lv)
  list(z=z, mu=mu, logvar=lv)
}
```

Returns $\mathbf{z}$ **sampled**

:::
::::::

## KL divergence: regularize the latent space

$$D_{\text{KL}}\!\left(q(\mathbf{z}|\mathbf{x}) \,\|\, p(\mathbf{z})\right) = -\frac{1}{2}\sum_{k}\left(1 + \log\sigma^{2}_{k} - \mu^{2}_{k} - \sigma^{2}_{k}\right)$$

* Pushes each $q(\mathbf{z}|\mathbf{x})$ toward the prior $\mathcal{N}(\mathbf{0}, I)$
* Without it, the encoder would set $\sigma \to 0$ (killing the noise)

```{r def-kl-loss, echo=TRUE}
kl.loss <- function(mu, logvar) {
    -0.5 * torch_mean(
        torch_sum(1 + logvar - mu^2 - torch_exp(logvar), dim = 2)
    )
}
```

## VAE = encoder + decoder + new loss

```{r def-vae, echo=TRUE}
vae <- nn_module(
    initialize = function(d.in, K) {
        self$enc <- vae.encoder(d.in, K)
        self$dec <- linear.decoder(K, d.in)
    },
    forward = function(x) {
        q <- self$enc(x)
        x.hat <- self$dec(q$z)
        list(x.hat = x.hat, mu = q$mu,
             logvar = q$logvar, z = q$z)
    }
)
```

* Same linear decoder as before
* Loss = reconstruction + KL divergence

## Train the VAE

```{r train-vae-show, echo=TRUE, eval=FALSE}
model.vae <- vae(n.genes, K)
optimizer <- optim_adam(model.vae$parameters, lr = 1e-3)

for (epoch in seq_len(n.epoch)) {
    idx <- sample(n.cells, batch.size)
    x.batch <- X.tensor[idx, ]
    out <- model.vae(x.batch)
    loss <- recon.loss(torch_log1p(x.batch), out$x.hat) +
            kl.loss(out$mu, out$logvar)
    optimizer$zero_grad()
    loss$backward()
    optimizer$step()
}
```

* One extra term in the loss: that's it

```{r train-vae, include=FALSE}
vae.rds <- "../data/GSE155698_PDAC_Steele2020/vae_trained.rds"
if.needed(vae.rds, {
    set.seed(42)
    model.vae <- vae(n.genes, K)
    optimizer <- optim_adam(model.vae$parameters, lr = 1e-3)
    n.epoch <- 2000
    batch.size <- 256
    loss.history <- c()
    snapshots <- list()
    record.at <- c(1, 100, 500, 2000)
    for (epoch in seq_len(n.epoch)) {
        idx <- sample(n.cells, batch.size)
        x.batch <- X.tensor[idx, ]
        out <- model.vae(x.batch)
        loss <- recon.loss(torch_log1p(x.batch), out$x.hat) +
                kl.loss(out$mu, out$logvar)
        optimizer$zero_grad()
        loss$backward()
        optimizer$step()
        loss.history <- c(loss.history, loss$item())
        if (epoch %in% record.at) {
            with_no_grad({
                out.all <- model.vae(X.tensor[obs.data$idx, ])
                snapshots[[as.character(epoch)]] <-
                    out.all$x.hat |> as.array()
            })
        }
    }
    with_no_grad({
        out.all <- model.vae(X.tensor)
        z.all <- out.all$mu |> as.array()
    })
    saveRDS(list(loss = loss.history,
                 z = z.all,
                 snapshots = snapshots), vae.rds)
})
vae.result <- readRDS(vae.rds)
```

## VAE training loss

```{r plot-vae-loss, echo=FALSE, fig.width=4.5, fig.height=2.5}
data.table(epoch = seq_along(vae.result$loss),
           loss = vae.result$loss) |>
    ggplot(aes(epoch, loss)) +
    geom_point(size = 0.3, alpha = 0.5) +
    geom_smooth(se = FALSE, colour = "firebrick") +
    ylab("reconstruction + KL loss") +
    ggtitle("VAE training")
```

## VAE reconstruction across epochs

```{r plot-vae-snapshots-prep, include=FALSE}
p.obs.v <- .matshow(obs.data$X.sorted, .axis.lab = 0, .limits = .clim) +
    ggtitle("Observed")
p.vep1 <- .recon.heatmap(vae.result$snapshots[["1"]], "epoch 1")
p.vep100 <- .recon.heatmap(vae.result$snapshots[["100"]], "epoch 100")
p.vep500 <- .recon.heatmap(vae.result$snapshots[["500"]], "epoch 500")
```

```{r plot-vae-snap-0, echo=FALSE, fig.width=5.8, fig.height=2.9, only.plot="1"}
p.obs.v | p.vep1
```

```{r plot-vae-snap-1, echo=FALSE, fig.width=5.8, fig.height=2.9, only.plot="2"}
p.obs.v | p.vep100
```

```{r plot-vae-snap-2, echo=FALSE, fig.width=5.8, fig.height=2.9, only.plot="3"}
p.obs.v | p.vep500
```

## VAE: observed vs. reconstructed (epoch 2000)

```{r plot-vae-compare, echo=FALSE, fig.width=5.8, fig.height=2.9}
p.obs.v2 <- .matshow(obs.data$X.sorted, .axis.lab = 0, .limits = .clim) +
    ggtitle("Observed")
p.vae.recon <- .recon.heatmap(vae.result$snapshots[["2000"]], "VAE (epoch 2000)")
p.obs.v2 | p.vae.recon
```

## Latent space: autoencoder vs. VAE

```{r plot-compare-latent-prep, include=FALSE}
z.ae <- as.data.table(ae.result$z)
colnames(z.ae) <- paste0("z", 1:K)
z.vae <- as.data.table(vae.result$z)
colnames(z.vae) <- paste0("z", 1:K)
```

```{r plot-compare-z12, echo=FALSE, fig.width=5.5, fig.height=2.8, only.plot="1"}
p1 <- ggplot(z.ae, aes(z1, z2)) +
    ggrastr::rasterise(geom_point(size = 0.2, alpha = 0.3), dpi = 300) +
    ggtitle("Autoencoder: z1 vs z2")
p2 <- ggplot(z.vae, aes(z1, z2)) +
    ggrastr::rasterise(geom_point(size = 0.2, alpha = 0.3), dpi = 300) +
    ggtitle("VAE: z1 vs z2")
p1 | p2
```

```{r plot-compare-z34, echo=FALSE, fig.width=5.5, fig.height=2.8, only.plot="2"}
p1 <- ggplot(z.ae, aes(z3, z4)) +
    ggrastr::rasterise(geom_point(size = 0.2, alpha = 0.3), dpi = 300) +
    ggtitle("Autoencoder: z3 vs z4")
p2 <- ggplot(z.vae, aes(z3, z4)) +
    ggrastr::rasterise(geom_point(size = 0.2, alpha = 0.3), dpi = 300) +
    ggtitle("VAE: z3 vs z4")
p1 | p2
```

```{r plot-compare-z56, echo=FALSE, fig.width=5.5, fig.height=2.8, only.plot="3"}
p1 <- ggplot(z.ae, aes(z5, z6)) +
    ggrastr::rasterise(geom_point(size = 0.2, alpha = 0.3), dpi = 300) +
    ggtitle("Autoencoder: z5 vs z6")
p2 <- ggplot(z.vae, aes(z5, z6)) +
    ggrastr::rasterise(geom_point(size = 0.2, alpha = 0.3), dpi = 300) +
    ggtitle("VAE: z5 vs z6")
p1 | p2
```

* Each dot is a cell; the latent space has no particular structure yet
* VAE latent space is more structured (KL pushes toward $\mathcal{N}(0,I)$)

# Multinomial likelihood

## The problem with MSE loss

\large

Our data is **counts**, not continuous values

\normalsize

* MSE on $\log(1+x)$ ignores the count nature of the data

* A cell with 10,000 total reads and a cell with 1,000 reads are treated the same

\vfill

**Better**: model each cell's expression as a **multinomial** draw

$$\mathbf{x}_{i} \sim \text{Multinomial}(N_{i},\, \boldsymbol{\pi}_{i})$$

* $N_i = \sum_j x_{ij}$ is the library size
* $\boldsymbol{\pi}_{i}$ is the gene probability vector

## Multinomial decoder

Each cell $i$ has **topic proportions** $\boldsymbol{\theta}_i$ and each topic $k$ has **gene probabilities** $\boldsymbol{\phi}_k$:

$$\boldsymbol{\theta}_{i} = \text{softmax}(\mathbf{z}_{i}), \quad \boldsymbol{\phi}_{k} = \text{softmax}(\boldsymbol{\beta}_{k})$$

$$\boldsymbol{\pi}_{i} = \boldsymbol{\theta}_{i}^{\top} \boldsymbol{\Phi} = \sum_{k} \theta_{ik} \, \boldsymbol{\phi}_{k}$$

\vfill

* This is exactly a **topic model**: cells are documents, genes are words

## Multinomial decoder in `torch`

```{r def-multinom-decoder, echo=TRUE}
multinom.decoder <- nn_module(
    initialize = function(K, d.out) {
        self$logit.beta <- nn_parameter(
            torch_randn(K, d.out)
        )
    },
    forward = function(z) {
        log.theta <- nnf_log_softmax(z, dim = 2)
        log.phi <- nnf_log_softmax(self$logit.beta, dim = 2)
        ## log(theta %*% phi)
        log.pi <- torch_logsumexp(
            log.theta$unsqueeze(3) + log.phi$unsqueeze(1),
            dim = 2
        )
        log.pi
    },
    get_weight = function() {
        self$logit.beta
    }
)
```

## Multinomial log-likelihood

$$\log p(\mathbf{x}_i \mid \boldsymbol{\pi}_i) = \sum_{j} x_{ij} \log \pi_{ij} + \text{const.}$$

```{r def-multinom-loss, echo=TRUE}
multinom.llik <- function(x, log.pi) {
    torch_mean(torch_sum(x * log.pi, dim = 2))
}
```

* We maximize the log-likelihood (minimize negative log-likelihood)
* No need for $\log(1+x)$ transform — we work directly with counts

## Autoencoder with multinomial likelihood

```{r def-ae-multinom, echo=TRUE}
ae.multinom <- nn_module(
    initialize = function(d.in, K) {
        self$enc <- linear.encoder(d.in, K)
        self$dec <- multinom.decoder(K, d.in)
    },
    forward = function(x) {
        z <- self$enc(x)
        log.pi <- self$dec(z)
        list(z = z, log.pi = log.pi)
    }
)
```

## VAE with multinomial likelihood

```{r def-vae-multinom, echo=TRUE}
vae.multinom <- nn_module(
    initialize = function(d.in, K) {
        self$enc <- vae.encoder(d.in, K)
        self$dec <- multinom.decoder(K, d.in)
    },
    forward = function(x) {
        q <- self$enc(x)
        log.pi <- self$dec(q$z)
        list(log.pi = log.pi, mu = q$mu,
             logvar = q$logvar, z = q$z)
    }
)
```

* Same encoder changes as before (point $\to$ distribution)
* Now the decoder is shared: multinomial in both cases

## Train both models

```{r train-multinom, include=FALSE}
mn.rds <- "../data/GSE155698_PDAC_Steele2020/multinom_trained.rds"
if.needed(mn.rds, {
    set.seed(42)
    ## AE multinomial
    model.ae.mn <- ae.multinom(n.genes, K)
    opt.ae <- optim_adam(model.ae.mn$parameters, lr = 1e-3)

    ## VAE multinomial
    model.vae.mn <- vae.multinom(n.genes, K)
    opt.vae <- optim_adam(model.vae.mn$parameters, lr = 1e-3)

    n.epoch <- 2000
    batch.size <- 256
    ae.loss.hist <- c()
    vae.loss.hist <- c()
    ae.snaps <- list()
    vae.snaps <- list()
    record.at <- c(1, 100, 500, 2000)

    for (epoch in seq_len(n.epoch)) {
        ## AE step
        idx <- sample(n.cells, batch.size)
        x.batch <- X.tensor[idx, ]
        out <- model.ae.mn(x.batch)
        loss <- -multinom.llik(x.batch, out$log.pi)
        opt.ae$zero_grad()
        loss$backward()
        opt.ae$step()
        ae.loss.hist <- c(ae.loss.hist, loss$item())

        ## VAE step
        idx <- sample(n.cells, batch.size)
        x.batch <- X.tensor[idx, ]
        out <- model.vae.mn(x.batch)
        loss <- -multinom.llik(x.batch, out$log.pi) +
                kl.loss(out$mu, out$logvar)
        opt.vae$zero_grad()
        loss$backward()
        opt.vae$step()
        vae.loss.hist <- c(vae.loss.hist, loss$item())

        if (epoch %in% record.at) {
            with_no_grad({
                out.ae <- model.ae.mn(X.tensor[obs.data$idx, ])
                ae.snaps[[as.character(epoch)]] <-
                    out.ae$log.pi |> torch_exp() |> as.array()
                out.vae <- model.vae.mn(X.tensor[obs.data$idx, ])
                vae.snaps[[as.character(epoch)]] <-
                    out.vae$log.pi |> torch_exp() |> as.array()
            })
        }
    }

    with_no_grad({
        z.ae <- model.ae.mn$enc(X.tensor) |> as.array()
        z.vae <- (model.vae.mn$enc(X.tensor))$mu |> as.array()
    })

    saveRDS(list(ae.loss = ae.loss.hist, vae.loss = vae.loss.hist,
                 ae.z = z.ae, vae.z = z.vae,
                 ae.snaps = ae.snaps, vae.snaps = vae.snaps,
                 model.vae = model.vae.mn),
            mn.rds)
})
mn.result <- readRDS(mn.rds)
```

```{r train-multinom-show, echo=TRUE, eval=FALSE}
## AE: minimize negative log-likelihood
loss <- -multinom.llik(x.batch, out$log.pi)

## VAE: add KL divergence
loss <- -multinom.llik(x.batch, out$log.pi) +
        kl.loss(out$mu, out$logvar)
```

* The only change from MSE: swap the reconstruction loss

## Multinomial training loss

```{r plot-mn-loss, echo=FALSE, fig.width=5.5, fig.height=2.5}
dt.loss <- rbind(
    data.table(epoch = seq_along(mn.result$ae.loss),
               loss = mn.result$ae.loss, model = "AE"),
    data.table(epoch = seq_along(mn.result$vae.loss),
               loss = mn.result$vae.loss, model = "VAE")
)
ggplot(dt.loss, aes(epoch, loss, colour = model)) +
    geom_point(size = 0.2, alpha = 0.3) +
    geom_smooth(se = FALSE) +
    ylab("negative log-likelihood") +
    ggtitle("Multinomial AE vs. VAE")
```

## Multinomial reconstruction (AE vs. VAE, epoch 2000)

```{r plot-mn-recon-helper, include=FALSE}
## for multinomial, the decoder outputs probabilities (pi)
## scale by library size then use same visualization
.recon.heatmap.mn <- function(pi.mat, title = "") {
    ## pi is probability; convert to expected log counts
    lib.sizes <- rowSums(torch.data$X[obs.data$idx, ])
    expected <- pi.mat * lib.sizes
    recon.log <- log1p(expected[, obs.data$top.genes])
    recon.scaled <- sweep(recon.log, 2, obs.data$col.center, `-`)
    recon.scaled <- sweep(recon.scaled, 2, obs.data$col.scale, `/`)
    recon.scaled[is.na(recon.scaled)] <- 0
    .mat <- recon.scaled[obs.data$row.perm, obs.data$col.perm]
    .matshow(.mat, .axis.lab = 0, .limits = .clim) + ggtitle(title)
}
```

```{r plot-mn-recon, echo=FALSE, fig.width=5.5, fig.height=2.9}
p.obs.mn <- .matshow(obs.data$X.sorted, .axis.lab = 0, .limits = .clim) +
    ggtitle("Observed")
p.ae.mn <- .recon.heatmap.mn(mn.result$ae.snaps[["2000"]], "AE (multinomial)")
p.vae.mn <- .recon.heatmap.mn(mn.result$vae.snaps[["2000"]], "VAE (multinomial)")
p.obs.mn | p.ae.mn | p.vae.mn
```

## Latent space: multinomial AE vs. VAE

```{r plot-mn-latent-prep, include=FALSE}
z.ae.mn <- as.data.table(mn.result$ae.z)
colnames(z.ae.mn) <- paste0("z", 1:K)
z.vae.mn <- as.data.table(mn.result$vae.z)
colnames(z.vae.mn) <- paste0("z", 1:K)
```

```{r plot-mn-z12, echo=FALSE, fig.width=5.5, fig.height=2.8, only.plot="1"}
p1 <- ggplot(z.ae.mn, aes(z1, z2)) +
    ggrastr::rasterise(geom_point(size = 0.2, alpha = 0.3), dpi = 300) +
    ggtitle("AE (multinomial): z1 vs z2")
p2 <- ggplot(z.vae.mn, aes(z1, z2)) +
    ggrastr::rasterise(geom_point(size = 0.2, alpha = 0.3), dpi = 300) +
    ggtitle("VAE (multinomial): z1 vs z2")
p1 | p2
```

```{r plot-mn-z34, echo=FALSE, fig.width=5.5, fig.height=2.8, only.plot="2"}
p1 <- ggplot(z.ae.mn, aes(z3, z4)) +
    ggrastr::rasterise(geom_point(size = 0.2, alpha = 0.3), dpi = 300) +
    ggtitle("AE (multinomial): z3 vs z4")
p2 <- ggplot(z.vae.mn, aes(z3, z4)) +
    ggrastr::rasterise(geom_point(size = 0.2, alpha = 0.3), dpi = 300) +
    ggtitle("VAE (multinomial): z3 vs z4")
p1 | p2
```

```{r plot-mn-z56, echo=FALSE, fig.width=5.5, fig.height=2.8, only.plot="3"}
p1 <- ggplot(z.ae.mn, aes(z5, z6)) +
    ggrastr::rasterise(geom_point(size = 0.2, alpha = 0.3), dpi = 300) +
    ggtitle("AE (multinomial): z5 vs z6")
p2 <- ggplot(z.vae.mn, aes(z5, z6)) +
    ggrastr::rasterise(geom_point(size = 0.2, alpha = 0.3), dpi = 300) +
    ggtitle("VAE (multinomial): z5 vs z6")
p1 | p2
```

## Decoder structure: top genes per factor

\large

Without the KL term, the encoder can "cheat":

\normalsize

* Set $\sigma \to 0$ — no noise, just a point estimate
* The latent space becomes **irregular** — no structure
* Some dimensions may **collapse** (carry no information)

\vfill

**$\beta$-VAE** [@Higgins2016-ml]: weight the KL term to control the tradeoff

$$\mathcal{L} = -\log p(\mathbf{x}|\mathbf{z}) + \alert{\beta} \cdot D_{\text{KL}}(q(\mathbf{z}|\mathbf{x}) \| p(\mathbf{z}))$$

* $\beta = 0$: pure autoencoder (no regularization)
* $\beta = 1$: standard VAE
* $\beta > 1$: stronger push toward $\mathcal{N}(0, I)$

## Training with different $\beta$ values

```{r train-beta-sweep, include=FALSE}
beta.rds <- "../data/GSE155698_PDAC_Steele2020/beta_sweep.rds"
if.needed(beta.rds, {
    betas <- c(0, 0.1, 0.5, 1, 2, 5)
    results <- list()

    for (beta in betas) {
        set.seed(42)
        model <- vae.multinom(n.genes, K)
        opt <- optim_adam(model$parameters, lr = 1e-3)

        n.epoch <- 2000
        batch.size <- 256
        kl.hist <- c()
        llik.hist <- c()

        for (epoch in seq_len(n.epoch)) {
            idx <- sample(n.cells, batch.size)
            x.batch <- X.tensor[idx, ]
            out <- model(x.batch)

            llik.val <- multinom.llik(x.batch, out$log.pi)
            kl.val <- kl.loss(out$mu, out$logvar)
            loss <- -llik.val + beta * kl.val

            opt$zero_grad()
            loss$backward()
            opt$step()

            kl.hist <- c(kl.hist, kl.val$item())
            llik.hist <- c(llik.hist, llik.val$item())
        }

        with_no_grad({
            out.all <- model(X.tensor)
            z.all <- out.all$mu |> as.array()
            z.var <- apply(z.all, 2, var)
            out.sub <- model(X.tensor[obs.data$idx, ])
            pi.mat <- out.sub$log.pi |> torch_exp() |> as.array()
        })

        results[[as.character(beta)]] <- list(
            kl = kl.hist, llik = llik.hist,
            z = z.all, z.var = z.var, pi = pi.mat
        )
    }

    saveRDS(results, beta.rds)
})
beta.results <- readRDS(beta.rds)
```

```{r train-beta-show, echo=TRUE, eval=FALSE}
betas <- c(0, 0.1, 0.5, 1, 2, 5)

for (beta in betas) {
    ## ... same training loop, but:
    loss <- -multinom.llik(x, out$log.pi) +
            beta * kl.loss(out$mu, out$logvar)
}
```

* We track KL and log-likelihood **separately** at each epoch

## Loss traces across epochs

```{r plot-beta-traces-prep, include=FALSE}
show.betas <- c("0", "0.1", "1", "5")
.beta.trace.plot <- function(b) {
    n <- length(beta.results[[b]]$llik)
    kl.vals <- beta.results[[b]]$kl
    kl.upper <- quantile(kl.vals, 0.99)
    dt <- rbind(
        data.table(epoch = seq_len(n),
                   value = beta.results[[b]]$llik,
                   metric = "log-likelihood"),
        data.table(epoch = seq_len(n),
                   value = kl.vals,
                   metric = "KL divergence")
    )
    dt <- dt[!(metric == "KL divergence" & value > kl.upper)]
    ggplot(dt, aes(epoch, value)) +
        geom_point(size = 0.1, alpha = 0.15) +
        geom_smooth(se = FALSE, linewidth = 0.7) +
        facet_wrap(~ metric, ncol = 1, scales = "free_y") +
        ylab("") +
        ggtitle(bquote(beta == .(as.numeric(b)))) +
        theme(strip.text = element_text(size = 7))
}
```

:::::: {.columns}
::: {.column width=.6}

```{r plot-beta-trace-0, echo=FALSE, fig.width=4, fig.height=3, only.plot="1"}
.beta.trace.plot("0")
```

```{r plot-beta-trace-01, echo=FALSE, fig.width=4, fig.height=3, only.plot="2"}
.beta.trace.plot("0.1")
```

```{r plot-beta-trace-1, echo=FALSE, fig.width=4, fig.height=3, only.plot="3"}
.beta.trace.plot("1")
```

```{r plot-beta-trace-5, echo=FALSE, fig.width=4, fig.height=3, only.plot="4"}
.beta.trace.plot("5")
```

:::
::: {.column width=.35}

* Higher $\beta$: lower KL (more Gaussian) but worse reconstruction

:::
::::::




## Latent space shape at different $\beta$

```{r plot-beta-latent-shape-prep, include=FALSE}
beta.labels <- c("AE~(no~sampling)", paste0("beta == ", show.betas))
.make.z.scatter <- function(dims, xlab, ylab) {
    dt <- rbindlist(c(
        list(data.table(zx = mn.result$ae.z[, dims[1]],
                        zy = mn.result$ae.z[, dims[2]],
                        beta = beta.labels[1])),
        lapply(seq_along(show.betas), function(i) {
            z <- beta.results[[show.betas[i]]]$z
            data.table(zx = z[, dims[1]], zy = z[, dims[2]],
                       beta = beta.labels[i + 1])
        })
    ))
    dt[, beta := factor(beta, levels = beta.labels)]
    ggplot(dt, aes(zx, zy)) +
        ggrastr::rasterise(geom_point(size = 0.1, alpha = 0.2), dpi = 300) +
        facet_wrap(~ beta, nrow = 1, scales = "free",
                   labeller = label_parsed) +
        xlab(xlab) + ylab(ylab) +
        theme(strip.text = element_text(size = 7))
}
```

```{r plot-beta-latent-z12, echo=FALSE, fig.width=6.1, fig.height=2.8, only.plot="1"}
.make.z.scatter(c(1, 2), "z1", "z2")
```

```{r plot-beta-latent-z34, echo=FALSE, fig.width=6.1, fig.height=2.8, only.plot="2"}
.make.z.scatter(c(3, 4), "z3", "z4")
```

```{r plot-beta-latent-z56, echo=FALSE, fig.width=6.1, fig.height=2.8, only.plot="3"}
.make.z.scatter(c(5, 6), "z5", "z6")
```

```{r plot-beta-latent-z78, echo=FALSE, fig.width=6.1, fig.height=2.8, only.plot="4"}
.make.z.scatter(c(7, 8), "z7", "z8")
```

```{r plot-beta-latent-z910, echo=FALSE, fig.width=6.1, fig.height=2.8, only.plot="5"}
.make.z.scatter(c(9, 10), "z9", "z10")
```

<!-- * AE (deterministic): most irregular — no sampling to regularize -->
<!-- * Even $\beta = 0$ (sampling, no KL) is smoother than AE -->

## Topic proportions $\boldsymbol{\theta} = \text{softmax}(\mathbf{z})$

```{r plot-beta-softmax, echo=FALSE, fig.width=5.5, fig.height=2.8}
.softmax <- function(z) {
    ez <- exp(z - apply(z, 1, max))
    ez / rowSums(ez)
}

dt.sm <- rbindlist(c(
    list(data.table(theta = as.vector(.softmax(mn.result$ae.z)),
                    beta = beta.labels[1])),
    lapply(seq_along(show.betas), function(i) {
        data.table(theta = as.vector(.softmax(beta.results[[show.betas[i]]]$z)),
                   beta = beta.labels[i + 1])
    })
))
dt.sm[, beta := factor(beta, levels = beta.labels)]

ggplot(dt.sm, aes(theta)) +
    geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
    scale_y_log10() +
    facet_wrap(~ beta, nrow = 1, scales = "free_y",
               labeller = label_parsed) +
    xlab(expression(theta[k])) + ylab("count") +
    theme(strip.text = element_text(size = 7),
          axis.text = element_text(size = 5),
          axis.title = element_text(size = 6))
```

* AE: spikes near 0 and 1 — one topic dominates per cell
* Higher $\beta$: more uniform $\boldsymbol{\theta}$ — cells use multiple topics


## Latent distribution: is $\mathbf{z}$ Gaussian?

```{r plot-beta-zhist, echo=FALSE, fig.width=5.5, fig.height=2.8}
dt.z <- rbindlist(lapply(show.betas, function(b) {
    z <- beta.results[[b]]$z
    data.table(value = as.vector(z),
               beta = paste0("beta == ", b))
}))

ggplot(dt.z, aes(value)) +
    geom_histogram(aes(y = after_stat(density)),
                   bins = 50, fill = "steelblue", alpha = 0.6) +
    facet_wrap(~ beta, nrow = 1, labeller = label_parsed, scales = "free") +
    xlab("z value (all dimensions pooled)") + ylab("density") +
    theme(strip.text = element_text(size = 8))
```

* Higher $\beta$ $\Rightarrow$ closer to the standard normal prior

## Reconstruction at different $\beta$

```{r plot-beta-recon, echo=FALSE, fig.width=5.5, fig.height=2.9}
p.obs.b <- .matshow(obs.data$X.sorted, .axis.lab = 0, .limits = .clim) +
    ggtitle("Observed")
p.b0 <- .recon.heatmap.mn(beta.results[["0"]]$pi, expression(beta == 0))
p.b1 <- .recon.heatmap.mn(beta.results[["1"]]$pi, expression(beta == 1))
p.b5 <- .recon.heatmap.mn(beta.results[["5"]]$pi, expression(beta == 5))
p.obs.b | p.b0 | p.b1 | p.b5
```

* $\beta = 0$ reconstructs best but latent space is unstructured
* $\beta = 5$ has smooth latent space but blurry reconstruction

## Latent space structure at different $\beta$

```{r beta-tsne-prep, include=FALSE}
library(Rtsne)

beta.tsne.rds <- "../data/GSE155698_PDAC_Steele2020/beta_tsne.rds"
if.needed(beta.tsne.rds, {
    set.seed(42)
    beta.tsne.list <- lapply(show.betas, function(b) {
        z <- beta.results[[b]]$z
        tsne.out <- Rtsne(z, dims = 2, perplexity = 30, verbose = FALSE)
        data.table(
            x = tsne.out$Y[, 1],
            y = tsne.out$Y[, 2],
            beta = b
        )
    })
    saveRDS(beta.tsne.list, beta.tsne.rds)
})
beta.tsne.list <- readRDS(beta.tsne.rds)
```

```{r beta-tsne-0, echo=FALSE, fig.width=3.5, fig.height=2.8, only.plot="1"}
ggplot(beta.tsne.list[[1]], aes(x, y)) +
    geom_point(size = 0.3, alpha = 0.5, colour = "steelblue") +
    ggtitle(expression(beta == 0)) +
    labs(x = "t-SNE 1", y = "t-SNE 2") +
    theme_minimal()
```

```{r beta-tsne-01, echo=FALSE, fig.width=3.5, fig.height=2.8, only.plot="2"}
ggplot(beta.tsne.list[[2]], aes(x, y)) +
    geom_point(size = 0.3, alpha = 0.5, colour = "steelblue") +
    ggtitle(expression(beta == 0.1)) +
    labs(x = "t-SNE 1", y = "t-SNE 2") +
    theme_minimal()
```

```{r beta-tsne-1, echo=FALSE, fig.width=3.5, fig.height=2.8, only.plot="3"}
ggplot(beta.tsne.list[[3]], aes(x, y)) +
    geom_point(size = 0.3, alpha = 0.5, colour = "steelblue") +
    ggtitle(expression(beta == 1)) +
    labs(x = "t-SNE 1", y = "t-SNE 2") +
    theme_minimal()
```

```{r beta-tsne-5, echo=FALSE, fig.width=3.5, fig.height=2.8, only.plot="4"}
ggplot(beta.tsne.list[[4]], aes(x, y)) +
    geom_point(size = 0.3, alpha = 0.5, colour = "steelblue") +
    ggtitle(expression(beta == 5)) +
    labs(x = "t-SNE 1", y = "t-SNE 2") +
    theme_minimal()
```

* t-SNE visualization of latent space $\mathbf{z}$
* Higher $\beta$ $\Rightarrow$ more structured, continuous latent space

## Connection to topic models

\centerline{\includegraphics[width=.8\linewidth]{img/BleiTopicLDA_0.pdf}}

\vfill
\tiny
\flushright
Slide credit: David Blei

## Word frequencies define topics

\centerline{\includegraphics[width=.8\linewidth]{img/BleiTopicLDA_1.pdf}}

\vfill
\tiny
\flushright
Slide credit: David Blei

## This is LDVAE (Svensson et al., 2020)

\large

What we just built:

* Linear encoder + multinomial decoder + VAE

\normalsize

\vfill

This is exactly **Linearly Decoded VAE** (LDVAE):

* The decoder weight matrix $\boldsymbol{\beta}$ is directly interpretable
* Each row = a "topic" = a gene program
* Equivalent to probabilistic PCA with count likelihood

\vfill

**scVI** (Lopez et al., 2018) extends this with:

* Deep (nonlinear) encoder and decoder
* ZINB likelihood (handles overdispersion + dropout)
* Batch effect correction via conditional decoder


## scVI: VAE with more flexible encoder, decoder

\centerline{\includegraphics[height=.75\textheight]{img/scVI.pdf}}

\vfill
\tiny

[@Lopez2018-lb]

## Batch correction with scVI

**Problem:** PDAC data has 3 tissues (3 batches)

* Technical variation between batches can dominate biological signal
* PCA/UMAP may cluster by batch instead of cell type

\vfill

**scVI solution:** Conditional decoder $p(x \mid z, s)$

* Latent code $z$: biological variation (shared across batches)
* Batch indicator $s \in \{T1, T2, T3\}$: technical variation
* Decoder learns to generate batch-specific expression from shared $z$

\vfill

**Result:** Latent space $z$ is batch-corrected

* Cells group by biology, not batch
* Can integrate multiple datasets/experiments

## Implementing scVI in R `torch`

**Strategy:** Reuse VAE encoder, upgrade decoder to ZINB with batch conditioning

* Same encoder as before (outputs $\mu$, $\log\sigma^2$, $z$)
* New decoder: outputs ZINB parameters ($\mu$, $\theta$, $\pi$) conditioned on batch
* Loss = $-\log p(x|z,s) + D_{\text{KL}}(q(z|x) \| p(z))$

\vfill

**Why ZINB likelihood?**

* Negative binomial handles overdispersion (variance > mean)
* Zero-inflation handles dropout (technical zeros)
* Better fit than multinomial for scRNA-seq counts

## ZINB likelihood math

**Negative Binomial distribution:**

$$\text{NB}(x|\mu, \theta) = \frac{\Gamma(x + \theta)}{\Gamma(\theta) \, \Gamma(x+1)} \left(\frac{\theta}{\theta + \mu}\right)^\theta \left(\frac{\mu}{\theta + \mu}\right)^x$$

* $\mu$: mean expression
* $\theta$: dispersion (inverse overdispersion)

\vfill

**Zero-Inflated NB:**

$$\text{ZINB}(x|\mu, \theta, \pi) = \begin{cases}
\pi + (1-\pi) \cdot \text{NB}(0|\mu,\theta) & \text{if } x = 0\\
(1-\pi) \cdot \text{NB}(x|\mu,\theta) & \text{if } x > 0
\end{cases}$$

* $\pi$: dropout probability (probability of technical zero)

## Why Zero-Inflated Negative Binomial (ZINB)?

**Problem with multinomial/Poisson:** Cannot handle

1. **Overdispersion**: variance $\gg$ mean in real scRNA-seq data
2. **Excess zeros**: both biological (gene off) and technical (dropout)

\vfill

**Solution: ZINB** = mixture model with two sources of counts

\vfill

For gene $j$ in cell $i$:

$$
x_{ij} \sim \begin{cases}
0 & \text{with probability } \pi_{ij} \text{ (dropout)}\\
\text{NB}(\mu_{ij}, \theta_j) & \text{with probability } 1 - \pi_{ij}
\end{cases}
$$

\vfill

* $\mu_{ij}$: mean expression (from decoder)
* $\theta_j$: dispersion parameter (gene-specific overdispersion)
* $\pi_{ij}$: dropout probability (zero-inflation)

## ZINB likelihood in `torch` - 1

```{r zinb-utils, echo=TRUE}
## Negative Binomial log-probability
nb.log.prob <- function(x, mu, theta) {
    eps <- 1e-6
    mu <- torch_clamp(mu, min = eps, max = 1e6)
    theta <- torch_clamp(theta, min = eps, max = 1e6)

    log_nb <- (torch_lgamma(x + theta) - torch_lgamma(theta) -
               torch_lgamma(x + 1) +
               theta * torch_log(theta / (theta + mu)) +
               x * torch_log(mu / (theta + mu)))
    log_nb
}
```

$$\text{NB}(x|\mu, \theta) = \frac{\Gamma(x + \theta)}{\Gamma(\theta) \, \Gamma(x+1)} \left(\frac{\theta}{\theta + \mu}\right)^\theta \left(\frac{\mu}{\theta + \mu}\right)^x$$

* $\mu$: mean expression
* $\theta$: dispersion (inverse overdispersion)

## ZINB likelihood in `torch` - 2

```{r echo=T}
## Zero-Inflated NB log-probability
zinb.log.prob <- function(x, mu, theta, pi) {
    eps <- 1e-6
    pi <- torch_clamp(pi, min = eps, max = 1 - eps)

    ## For x = 0: log(π + (1-π)·NB(0))
    ## For x > 0: log((1-π)·NB(x))
    nb_logprob <- nb.log.prob(x, mu, theta)
    zero_nb <- nb.log.prob(torch_zeros_like(x), mu, theta)

    ## Use logaddexp for numerical stability
    log_prob_zero <- torch_logaddexp(
        torch_log(pi + eps),
        torch_log(1 - pi + eps) + zero_nb
    )
    log_prob_nonzero <- torch_log(1 - pi + eps) + nb_logprob

    ## Select based on whether x is zero
    zero_mask <- (x < eps)
    torch_where(zero_mask, log_prob_zero, log_prob_nonzero)
}
```

$$
x_{ij} \sim \begin{cases}
0 & \text{with probability } \pi_{ij} \text{ (dropout)}\\
\text{NB}(\mu_{ij}, \theta_j) & \text{with probability } 1 - \pi_{ij}
\end{cases}
$$

## Putting them all together

```{r echo=T}
## Full ZINB log-likelihood (sum over genes, mean over cells)
zinb.llik <- function(x, mu, theta, pi) {
    log_probs <- zinb.log.prob(x, mu, theta, pi)
    ## Clamp to avoid extreme values
    log_probs <- torch_clamp(log_probs, min = -1e6, max = 1e6)
    torch_mean(torch_sum(log_probs, dim = 2))
}
```

## Batch-conditional ZINB decoder

```{r zinb-decoder, include=FALSE}
zinb.decoder <- nn_module(
    initialize = function(K, d.out, n.batches, h = 128) {

        self$batch.embed <- nn_embedding(n.batches, 8) # Batch embedding
       
        ## Hidden layer
        self$hidden <- nn_linear(K + 8, h)  
        ## Normalization
        self$bn <- nn_batch_norm1d(h)       

        ## Output for ZINB parameters
        self$mu.head <- nn_linear(h, d.out) 
        self$pi.head <- nn_linear(h, d.out)

        ## Gene-specific dispersion
        self$log.theta <- nn_parameter(torch_ones(d.out) * log(10))
    },
    forward = function(z, batch, lib.size) {
        ## Concatenate latent code with batch embedding
        batch.emb <- self$batch.embed(batch + 1L)  # +1 for 1-indexing
        z.batch <- torch_cat(list(z, batch.emb), dim = 2)

        ## Hidden layer with batch norm + ReLU
        h <- self$hidden(z.batch) |> self$bn() |> nnf_relu()

        ## Output ZINB parameters
        ## μ: mean expression (scale by library size)
        mu.raw <- nnf_softplus(self$mu.head(h))
        ## Normalize library sizes to mean 1 for stability
        lib.scale <- (lib.size / torch_mean(lib.size))$unsqueeze(2)
        nb.mu <- mu.raw * lib.scale

        ## π: dropout probability
        nb.pi <- torch_sigmoid(self$pi.head(h))

        ## θ: dispersion (shared across cells, gene-specific)
        nb.theta <- torch_exp(self$log.theta)$unsqueeze(1)$expand_as(nb.mu)

        list(nb.mu = nb.mu, nb.theta = nb.theta, nb.pi = nb.pi)
    }
)
```

:::::: {.columns}
::: {.column width=.65}

```{r zinb-decoder-init, eval=FALSE, echo=T}
zinb.decoder <- nn_module(
    initialize = function(K, d.out, n.batches, h = 128) {

        ## Batch embedding
        self$batch.embed <- nn_embedding(n.batches, 8) 
       
        ## Hidden layer
        self$hidden <- nn_linear(K + 8, h)  
        ## Normalization
        self$bn <- nn_batch_norm1d(h)       

        ## Output for ZINB parameters
        self$mu.head <- nn_linear(h, d.out)
        self$pi.head <- nn_linear(h, d.out)

        ## Gene-specific dispersion
        self$log.theta <- nn_parameter(torch_ones(d.out) * log(10))
    },
    forward = function(z, batch, lib.size) {
        ## (...)
    }
)
```

:::
::: {.column width=.3}


* Batch embedding: 3 batches → 8-dim vectors
* Concatenate with latent code $z$ before decoder
* Three output heads: $\mu$, $\pi$ (cell+gene-specific), $\theta$ (gene-specific)


:::
::::::

## Batch-conditional ZINB decoder (cont'd)


:::::: {.columns}
::: {.column width=.65}

```{r zinb-decoder-forward, eval=FALSE, echo=T}
zinb.decoder <- nn_module(
    ## (...)
    forward = function(z, batch, lib.size) {
        batch.emb <- self$batch.embed(batch + 1L)  # 1-indexing
        ## Concatenate latent code with batch embedding
        z.batch <- torch_cat(list(z, batch.emb), dim = 2)
        ## Hidden layer with batch norm + ReLU
        h <- self$hidden(z.batch) |> self$bn() |> nnf_relu()
        ## Output ZINB parameters
        ## μ: mean expression (scale by library size)
        mu.raw <- nnf_softplus(self$mu.head(h))
        ## Normalize library sizes to mean 1 for stability
        lib.scale <- (lib.size / torch_mean(lib.size))$unsqueeze(2)
        nb.mu <- mu.raw * lib.scale
        ## π: dropout probability
        nb.pi <- torch_sigmoid(self$pi.head(h))
        ## θ: dispersion (shared across cells, gene-specific)
        nb.theta <- torch_exp(self$log.theta)$unsqueeze(1)$expand_as(nb.mu)
        list(nb.mu = nb.mu, nb.theta = nb.theta, nb.pi = nb.pi)
    }
)
```

:::
::: {.column width=.3}

* Batch embedding: 3 batches → 8-dim vectors
* Concatenate with latent code $z$ before decoder
* Three output heads: $\mu$, $\pi$ (cell+gene-specific), $\theta$ (gene-specific)

:::
::::::

## Complete `scVI` model

```{r vae-zinb, echo=TRUE}
vae.zinb <- nn_module(
    initialize = function(d.in, K, n.batches) {
        self$enc <- vae.encoder(d.in, K)         ## REUSE existing VAE encoder
        self$dec <- zinb.decoder(K, d.in, n.batches)         ## NEW: ZINB decoder
    },
    forward = function(x, batch, lib.size) {
        ## Encode
        q <- self$enc(x)  # Returns: z, mu, logvar
        ## Decode (batch-conditional)
        params <- self$dec(q$z, batch, lib.size)
        ## Return everything
        list(z = q$z, mu = q$mu, logvar = q$logvar,
             nb.mu = params$nb.mu,
             nb.theta = params$nb.theta,
             nb.pi = params$nb.pi)
    }
)
```

* Only decoder changes – encoder is identical to previous VAE!
* Decoder now takes batch labels and library sizes as input

## Train scVI with ZINB loss

```{r train-scvi-zinb, include=FALSE, cache=TRUE}
scvi.trained.rds <- "../data/GSE155698_PDAC_Steele2020/scvi_zinb_trained.rds"
if.needed(scvi.trained.rds, {
    set.seed(42)

    ## Initialize model
    model.scvi <- vae.zinb(n.genes, K, n.batches)
    optimizer <- optim_adam(model.scvi$parameters, lr = 1e-3)

    ## Training parameters
    n.epoch <- 2000
    batch.size <- 256

    ## Track losses
    losses <- data.frame(
        epoch = integer(),
        total = numeric(),
        nll = numeric(),
        kl = numeric()
    )

    ## Training loop
    for (epoch in seq_len(n.epoch)) {
        ## Sample mini-batch
        idx <- sample(n.cells, batch.size)
        x.batch <- X.tensor[idx, ]
        batch.batch <- batch.tensor[idx]
        lib.batch <- lib.size.tensor[idx]

        ## Forward pass
        out <- model.scvi(x.batch, batch.batch, lib.batch)

        ## Loss = -ZINB log-likelihood + KL divergence
        llik <- zinb.llik(x.batch, out$nb.mu, out$nb.theta, out$nb.pi)
        kl <- kl.loss(out$mu, out$logvar)  # REUSE existing KL function!
        loss <- -llik + kl

        ## Check for NaN
        if (is.nan(as.numeric(loss$item()))) {
            warning("NaN loss at epoch ", epoch, ", skipping")
            next
        }

        ## Backward pass
        optimizer$zero_grad()
        loss$backward()
        ## Gradient clipping for stability
        nn_utils_clip_grad_norm_(model.scvi$parameters, max_norm = 5.0)
        optimizer$step()

        ## Track progress
        if (epoch %% 50 == 0) {
            losses <- rbind(losses, data.frame(
                epoch = epoch,
                total = as.numeric(loss$item()),
                nll = -as.numeric(llik$item()),
                kl = as.numeric(kl$item())
            ))
        }
    }

    ## Extract latent representation
    model.scvi$eval()
    with_no_grad({
        ## Use encoder mean (mu) instead of sampled z for stable embedding
        out.full <- model.scvi(X.tensor, batch.tensor, lib.size.tensor)
        z.scvi <- as.matrix(out.full$mu$cpu())  # Use mean, not sample
    })

    ## Verify no NaN values
    if (any(is.na(z.scvi))) {
        warning("NaN values in latent representation, falling back to safe extraction")
        model.scvi$eval()
        with_no_grad({
            ## Extract just encoder output
            enc.out <- model.scvi$enc(X.tensor)
            z.scvi <- as.matrix(enc.out$mu$cpu())
        })
    }

    saveRDS(list(model = model.scvi, losses = losses, z = z.scvi), scvi.trained.rds)
})
scvi.results <- readRDS(scvi.trained.rds)
model.scvi <- scvi.results$model
scvi.losses <- scvi.results$losses
z.scvi <- scvi.results$z
```

```{r show-train-scvi, echo=TRUE, eval=FALSE}
## Initialize
model.scvi <- vae.zinb(n.genes, K, n.batches)
optimizer <- optim_adam(model.scvi$parameters, lr = 1e-3)

## Training loop
for (epoch in seq_len(2000)) {
    idx <- sample(n.cells, 256)
    x.batch <- X.tensor[idx, ]
    batch.batch <- batch.tensor[idx]
    lib.batch <- lib.size.tensor[idx]

    out <- model.scvi(x.batch, batch.batch, lib.batch)

    ## Loss = -ZINB log-likelihood + KL
    llik <- zinb.llik(x.batch, out$nb.mu, out$nb.theta, out$nb.pi)
    kl <- kl.loss(out$mu, out$logvar)
    loss <- -llik + kl

    optimizer$zero_grad()
    loss$backward()
    optimizer$step()
}
```

## scVI training converges smoothly

```{r plot-scvi-losses, echo=FALSE, fig.width=6, fig.height=3}
# Calculate scaling factor to fit both on same plot
kl_range <- range(scvi.losses$kl)
llik_range <- range(-scvi.losses$nll)
scale_factor <- diff(llik_range) / diff(kl_range)
offset <- llik_range[1] - kl_range[1] * scale_factor

ggplot(scvi.losses, aes(x = epoch)) +
    geom_line(aes(y = -nll), linewidth = 0.8, color = "#E41A1C") +
    geom_line(aes(y = kl * scale_factor + offset), linewidth = 0.8, color = "#377EB8") +
    scale_y_continuous(
        name = "Log-likelihood",
        sec.axis = sec_axis(~ (. - offset) / scale_factor, name = "KL divergence")
    ) +
    labs(x = "Epoch") +
    theme_minimal() +
    theme(
        axis.title.y.left = element_text(color = "#E41A1C"),
        axis.text.y.left = element_text(color = "#E41A1C"),
        axis.title.y.right = element_text(color = "#377EB8"),
        axis.text.y.right = element_text(color = "#377EB8")
    )
```

## Batch correction comparison

```{r compute-scvi-umap, include=FALSE}
scvi.umap.rds <- "../data/GSE155698_PDAC_Steele2020/scvi_umap.rds"
if.needed(scvi.umap.rds, {
    ## Check for NaN values in scVI latent representation
    if (any(is.na(z.scvi))) {
        stop("NaN values found in scVI latent representation. Check training stability.")
    }

    ## Compute UMAP on scVI latent space
    library(uwot)
    umap.scvi <- umap(z.scvi, n_neighbors = 15, min_dist = 0.3, metric = "euclidean")

    ## Also get PCA UMAP for comparison
    pca.embed <- Embeddings(pdac, "pca")[, 1:10]
    umap.pca <- umap(pca.embed, n_neighbors = 15, min_dist = 0.3, metric = "euclidean")

    saveRDS(list(scvi = umap.scvi, pca = umap.pca), scvi.umap.rds)
})
scvi.umap.results <- readRDS(scvi.umap.rds)
umap.scvi <- scvi.umap.results$scvi
umap.pca <- scvi.umap.results$pca

## Create data frames for plotting
umap.df <- data.frame(
    UMAP1 = c(umap.pca[, 1], umap.scvi[, 1]),
    UMAP2 = c(umap.pca[, 2], umap.scvi[, 2]),
    batch = rep(batch.data$batch.idx, 2),
    cluster = rep(as.integer(pdac$seurat_clusters), 2),
    method = rep(c("PCA", "scVI"), each = n.cells)
)
umap.df$batch.label <- factor(umap.df$batch, levels = 0:2, labels = c("T1", "T2", "T3"))
```

:::::: {.columns}
::: {.column width=.48}

**PCA (batch-confounded)**

```{r plot-pca-batch, echo=FALSE, fig.width=3, fig.height=3}
umap.df |>
    filter(method == "PCA") |>
    ggplot(aes(x = UMAP1, y = UMAP2, color = batch.label)) +
    geom_point(size = 0.3, alpha = 0.6, stroke = 0) +
    scale_color_manual(values = c("#E41A1C", "#377EB8", "#4DAF4A")) +
    labs(x = "UMAP 1", y = "UMAP 2") +
    theme_minimal() +
    theme(legend.position = "none")
```

* Clusters by tissue batch

:::
::: {.column width=.48}

**scVI (batch-corrected)**

```{r plot-scvi-batch, echo=FALSE, fig.width=3, fig.height=3}
umap.df |>
    filter(method == "scVI") |>
    ggplot(aes(x = UMAP1, y = UMAP2, color = batch.label)) +
    geom_point(size = 0.3, alpha = 0.6, stroke = 0) +
    scale_color_manual(values = c("#E41A1C", "#377EB8", "#4DAF4A")) +
    labs(x = "UMAP 1", y = "UMAP 2") +
    theme_minimal() +
    theme(legend.position = "none")
```

* Batches mix together

:::
::::::

## `Seurat` PCA clustering $\approx$ `scVI`

:::::: {.columns}
::: {.column width=.48}

`Seurat`

```{r plot-pca-clusters, echo=FALSE, fig.width=3, fig.height=2.8}
umap.df |>
    filter(method == "PCA") |>
    ggplot(aes(x = UMAP1, y = UMAP2, color = factor(cluster))) +
    geom_point(size = 0.3, alpha = 0.6, stroke = 0) +
    scale_color_manual(values = scales::hue_pal()(length(unique(umap.df$cluster)))) +
    labs(x = "UMAP 1", y = "UMAP 2") +
    theme_minimal() +
    theme(legend.position = "none")
```

:::
::: {.column width=.48}

`scVI`

```{r plot-scvi-clusters, echo=FALSE, fig.width=3, fig.height=2.8}
umap.df |>
    filter(method == "scVI") |>
    ggplot(aes(x = UMAP1, y = UMAP2, color = factor(cluster))) +
    geom_point(size = 0.3, alpha = 0.6, stroke = 0) +
    scale_color_manual(values = scales::hue_pal()(length(unique(umap.df$cluster)))) +
    labs(x = "UMAP 1", y = "UMAP 2") +
    theme_minimal() +
    theme(legend.position = "none")
```

:::
::::::

* Cell type structure (Seurat clusters) preserved after batch correction
* scVI removes technical variation while keeping biological signal

## Quantifying batch mixing

```{r compute-batch-entropy, include=FALSE}
## Compute batch entropy for each cell's k-NN
library(FNN)

compute.batch.entropy <- function(embedding, batch.labels, k = 30) {
    nn.idx <- get.knn(embedding, k = k)$nn.index

    entropies <- sapply(1:nrow(nn.idx), function(i) {
        nn.batches <- batch.labels[nn.idx[i, ]]
        batch.freqs <- table(nn.batches) / k
        -sum(batch.freqs * log(batch.freqs + 1e-10))
    })

    mean(entropies)
}

pca.embed <- Embeddings(pdac, "pca")[, 1:10]
entropy.pca <- compute.batch.entropy(pca.embed, batch.data$batch.idx)
entropy.scvi <- compute.batch.entropy(z.scvi, batch.data$batch.idx)
```

**Batch entropy metric:**

* For each cell, find $k=30$ nearest neighbors
* Compute entropy of batch labels among neighbors
* Higher entropy = better batch mixing (tissues)

\vfill

```{r show-batch-entropy, echo=FALSE}
entropy.df <- data.frame(
    Method = c("PCA", "scVI"),
    `Batch Entropy` = c(entropy.pca, entropy.scvi)
)
knitr::kable(entropy.df, digits = 3, col.names = c("Method", "Batch Entropy"))
```

\vfill

* scVI increases batch mixing by `r round((entropy.scvi - entropy.pca) / entropy.pca * 100, 1)`%

## What did we learn?

**scVI in R torch shows:**

1. **Modular design:** Reuse VAE encoder, only change decoder + likelihood
2. **ZINB likelihood:** Three parameters ($\mu$, $\theta$, $\pi$) handle overdispersion + dropout
3. **Batch conditioning:** Simple embedding layer enables integration
4. **Interpretable:** See exactly what each component does

\vfill

**Key takeaways:**

* From multinomial → ZINB: better fit to scRNA-seq count data
* Conditional decoder: batch correction without destroying biology
* Build-it-yourself: understand what's "under the hood" of scvi-tools

## Running scVI with Python `scvi-tools`

```{python setup-scvi, eval=FALSE, echo=TRUE}
import scvi
import scanpy as sc
import numpy as np

# Load data (convert from Seurat to AnnData)
# In practice: use sceasy or manual conversion
adata = sc.read_h5ad("pdac_tissues123.h5ad")
# Setup scVI model
scvi.model.SCVI.setup_anndata(
    adata,
    layer="counts",           # raw counts
    batch_key="orig.ident"    # tissue batch (T1, T2, T3)
)
# Create and train model
model = scvi.model.SCVI(
    adata,
    n_latent=10,              # K = 10 latent dimensions
    gene_likelihood="zinb"    # zero-inflated negative binomial
)

model.train(max_epochs=100)
```

## Extract latent representation

```{python get-latent, eval=FALSE, echo=TRUE}
# Get latent representation (batch-corrected)
adata.obsm["X_scVI"] = model.get_latent_representation()

# Compute UMAP on scVI latent space
sc.pp.neighbors(adata, use_rep="X_scVI")
sc.tl.umap(adata)

# Compare: PCA vs scVI
sc.pl.umap(adata, color=["orig.ident", "seurat_clusters"])
```

**Key difference from LDVAE:**

* Deep encoder/decoder (nonlinear)
* ZINB likelihood handles dropout
* Batch correction via conditional generation

##

<!----------->
<!-- blank -->
<!----------->

## References {.allowframebreaks}

