---
title: "RNA-seq Data Analysis"
author: "Yongjin Park\n(slide credit: Keegan Korthauer)"
date: "20 January 2026"
classoption: "aspectratio=169"
fontsize: 12pt
output:
    beamer_presentation:
        theme: Madrid
        keep_md: true
        keep_tex: true
        latex_engine: xelatex
        slide_level: 2
bibliography: rna-seq.bib
header-includes:
  - \usepackage{cancel}
  - \usepackage{tikz}
  - \usetikzlibrary{positioning,calc,arrows.meta}
  - \tikzset{>={Stealth[length=3mm,width=2mm]}}
  - \definecolor{UBCblue}{rgb}{0.04706, 0.13725, 0.26667}
  - \usecolortheme[named=UBCblue]{structure}
  - \setbeamertemplate{frametitle}{\color{UBCblue}\bfseries\insertframetitle\par\vskip-6pt\hrulefill}
  - \setbeamercolor{title in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{section in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{author in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{date in head/foot}{bg=white,fg=gray}
  - \setbeamertemplate{page number in head/foot}{}
  - \setbeamertemplate{frame numbering}[none]
  - \setbeamercolor{alerted text}{bg=yellow}
  - \AtBeginSection[]{\begin{frame}\frametitle{Today's lecture}{\Large\tableofcontents[currentsection]}\end{frame}}
  - |
    \makeatletter
    \def\ps@titlepage{%
      \setbeamertemplate{footline}{}
    }
    \addtobeamertemplate{title page}{\thispagestyle{titlepage}}{}
    \makeatother
    \include{toc}
---

```{r setup, include=FALSE}
source("Setup.R")
fig.dir <- "../Fig/rna-seq/"
dir.create(fig.dir, showWarnings=FALSE)
setup.env(fig.dir)
dir.create("../data", showWarnings=FALSE)

library(recount3)
library(data.table)
library(tidyverse)
library(patchwork)
library(matrixTests)
library(ggpubr)
library(ggrepel)

theme_set(theme_minimal())
```


## Roadmap of STAT/BIOF/GSAT540 2026

_data generation_ followed by _inference_ or validation

\vfill

\large

:::::: {.columns}
::: {.column width=.45}

3. Formulating biological questions

4. Hypothesis testing

5. \textbf{\color{magenta}RNA-seq data} (Today)

6. \textbf{\color{teal}Inference focusing on RNA-seq} (Thursday)

7. \textbf{\color{olive} Experimental design} (next week)

:::
::: {.column width=.45}

8. Single-cell genomics

9. Matrix factorization

10. Genome-wide association studies

11. Multiple hypothesis testing

:::
::::::

\normalsize

## Learning objectives of RNA-seq analysis

:::::: {.columns}
::: {.column width=.45}

**How do we get sequencing data?**

\vspace{0.5cm}

* High-throughput sequencing data
  - data generation scheme
  - quantification

* Data normalization
  - intrinsic biases
  - gene-level adjustment
  - gene and sample adjustment

* Over-dispersed nature
  - count data
  - transformation

:::
::: {.column width=.45}

**How can we fit a model to sequencing data?** 

* Linear modelling
  - a Gaussian linear model
  - a generalized linear model

* Multivariate linear models
  - a column space view

* Model-based inference
  - maximum likelihood learning
  - parametric inference

* Revisit RNA-seq
  - \textcolor{blue}{\texttt{\textbf{DESeq2}}} vs. \textcolor{blue}{\texttt{\textbf{limma-voom}}}
  - modelling and estimation

:::
::::::


# RNA-Seq data generation scheme

## RNA-seq: pooling mRNA reads with poly-A tails

\centerline{\includegraphics[height=.7\textheight]{img/RNA_seq_step1_WangGersteinSnyder.pdf}}

\vfill\tiny
@Wang2009-tg

## RNA-seq: mapping short reads to reference genome

\centerline{\includegraphics[height=.7\textheight]{img/RNA_seq_step2_WangGersteinSnyder.pdf}}

\vfill\tiny
@Wang2009-tg

## Before RNA-seq: The Pre-2008 Era

:::::: {.columns}
::: {.column width=.45}

**EST sequencing** (1990s)

- Low throughput Sanger sequencing of cDNA clones
- Expensive and labor-intensive

\vspace{.5cm}

**SAGE** (1995)

- Short sequence tags from cDNA
- Limited dynamic range

:::
::: {.column width=.45}

**Microarrays** (late 1990s-2000s)

- Required **predefined probes** for known transcripts
- High background noise, cross-hybridization
- Limited dynamic range (~3 orders of magnitude)
- Poor at detecting low-abundance transcripts
- Species-specific probe design required

:::
::::::

\vfill\tiny
@Wang2009-tg

## Microarrays had been revolutionary, too

:::::: {.columns}
::: {.column width=.55}

\centerline{\includegraphics[height=.7\textheight]{img/Brown_Botstein_1999.pdf}}

:::
::: {.column width=.4}

> Thousands of genes are being discovered for the first time by sequencing the genomes of model organisms, **an exhilarating reminder** that much of the natural world remains to be explored at the molecular level. 

:::
::::::

\vfill\tiny
@Brown1999-sz

## RNA-seq: mapping short reads to reference genome

\centerline{\includegraphics[height=.7\textheight]{img/RNA_seq_step2_WangGersteinSnyder.pdf}}

\vfill\tiny
@Wang2009-tg

## Why sequencing-based high-throughput assays?

:::::: {.columns}
::: {.column width=.45}

**Key improvements**

- Much wider dynamic range (~5 orders of magnitude)
- No prior annotation needed
- Single-nucleotide resolution
- Digital counts vs. analog signals
- Works for any organism

:::
::: {.column width=.45}

**New capabilities**

- Discovery of novel transcripts and genes
- Alternative splicing detection
- Allele-specific expression
- Study of non-model organisms
- Precise isoform quantification

:::
::::::

\vfill

\begin{center}
From \textbf{hypothesis-driven} (design probes; fixed arrays)

$\Downarrow$

\textbf{\color{magenta}discovery-driven} (sequence everything, analyze later)
\end{center}

\vfill\tiny
@Wang2009-tg

## Foundational experimental work (2008)

:::::: {.columns}
::: {.column width=.45}

@Nagalakshmi2008-wq

- First complete transcriptome map using deep sequencing
- Yeast genome transcriptional landscape

:::
::: {.column width=.45}

@Mortazavi2008-es

- First comprehensive RNA-seq in mouse tissues
- Introduced **RPKM** normalization
- Demonstrated superiority over microarrays

:::
::::::


## Sequencing Technologies

:::::: {.columns}
::: {.column width=.45}

**Early Platforms (2005-2010)**

\vspace{.5cm}

- **454 Pyrosequencing**: First NGS platform, longer reads but lower throughput
- **Illumina/Solexa**: Became dominant platform, higher throughput, shorter reads
- **SOLiD**: Two-base encoding sequencing

:::
::: {.column width=.45}

**Modern Era (2011-present)**

\vspace{.5cm}

- **Pacific Biosciences (PacBio)**: Long-read sequencing, full-length transcripts without assembly
- **Oxford Nanopore**: Portable, long-read, direct RNA sequencing without reverse transcription

:::
::::::

## RNA-seq: mapping short reads to reference genome

\centerline{\includegraphics[height=.7\textheight]{img/RNA_seq_step2_WangGersteinSnyder.pdf}}

\vfill\tiny
@Wang2009-tg


## The isoform quantification problem

\begin{center}
\begin{tikzpicture}[scale=0.75]
  % Gene structure label
  \node[anchor=east, font=\scriptsize\bfseries] at (-0.5, 3.5) {Gene};

  % Exons (boxes)
  \fill[blue!60] (0,3.2) rectangle (1.2,3.8);
  \node[font=\tiny, white] at (0.6, 3.5) {E1};
  \fill[green!60] (2.2,3.2) rectangle (3.4,3.8);
  \node[font=\tiny, white] at (2.8, 3.5) {E2};
  \fill[orange!60] (4.4,3.2) rectangle (5.6,3.8);
  \node[font=\tiny, white] at (5.0, 3.5) {E3};
  \fill[red!60] (6.6,3.2) rectangle (7.8,3.8);
  \node[font=\tiny, white] at (7.2, 3.5) {E4};

  % Introns (lines)
  \draw (1.2, 3.5) -- (2.2, 3.5);
  \draw (3.4, 3.5) -- (4.4, 3.5);
  \draw (5.6, 3.5) -- (6.6, 3.5);

  % Isoform A (all exons)
  \node[anchor=east, font=\scriptsize\bfseries] at (-0.5, 2.0) {Isoform A};
  \fill[blue!60] (0,1.7) rectangle (1.2,2.3);
  \fill[green!60] (1.3,1.7) rectangle (2.5,2.3);
  \fill[orange!60] (2.6,1.7) rectangle (3.8,2.3);
  \fill[red!60] (3.9,1.7) rectangle (5.1,2.3);

  % Short reads on Isoform A
  \fill[gray!70] (0.1,2.4) rectangle (0.7,2.55);
  \fill[gray!70] (1.6,2.4) rectangle (2.2,2.55);
  \fill[gray!70] (2.9,2.4) rectangle (3.5,2.55);
  \fill[gray!70] (4.2,2.4) rectangle (4.8,2.55);
  % Junction-spanning read (E1-E2) - unique to A
  \fill[cyan!70] (0.9,2.4) rectangle (1.5,2.55);

  % Isoform B (skip E2)
  \node[anchor=east, font=\scriptsize\bfseries] at (-0.5, 0.5) {Isoform B};
  \fill[blue!60] (0,0.2) rectangle (1.2,0.8);
  \fill[orange!60] (1.3,0.2) rectangle (2.5,0.8);
  \fill[red!60] (2.6,0.2) rectangle (3.8,0.8);

  % Short reads on Isoform B
  \fill[gray!70] (0.3,-0.05) rectangle (0.9,0.1);
  \fill[gray!70] (1.6,-0.05) rectangle (2.2,0.1);
  \fill[gray!70] (2.9,-0.05) rectangle (3.5,0.1);
  % Junction-spanning read (E1-E3) - unique to B
  \fill[magenta!70] (0.9,-0.05) rectangle (1.5,0.1);

  % Annotations
  \node[anchor=west, font=\tiny] at (5.3, 2.0) {E1-E2-E3-E4};
  \node[anchor=west, font=\tiny] at (4.0, 0.5) {E1-E3-E4 (E2 skipped)};

  % Legend box
  \draw[rounded corners, gray] (9, 3.5) rectangle (14, 0.5);

  \fill[gray!70] (9.3, 3.1) rectangle (9.9, 3.3);
  \node[anchor=west, font=\tiny] at (10.1, 3.2) {Ambiguous (E1, E3, E4)};

  \fill[cyan!70] (9.3, 2.5) rectangle (9.9, 2.7);
  \node[anchor=west, font=\tiny] at (10.1, 2.6) {Unique to A (E1-E2 junction)};

  \fill[magenta!70] (9.3, 1.9) rectangle (9.9, 2.1);
  \node[anchor=west, font=\tiny] at (10.1, 2.0) {Unique to B (E1-E3 junction)};

  \node[anchor=north west, font=\tiny, align=left, text width=4.5cm] at (9.3, 1.6) {
    Ambiguous $\rightarrow$ EM assigns\\
    fractional counts to isoforms
  };

\end{tikzpicture}
\end{center}

Most reads are **ambiguous** $\rightarrow$ require **statistical inference** (EM) to estimate isoform abundance

## Long-read RNA-seq: Full-length transcripts

:::::: {.columns}
::: {.column width=.45}

**The isoform problem**

- Short reads (~150bp) cannot span full transcripts
- Transcript assembly is computationally challenging
- Isoform quantification relies on statistical inference

\centerline{$\Downarrow$}

**Long-read solution**: Sequence entire transcripts (1-10+ kb) without fragmentation

:::
::: {.column width=.45}

**Key advantages**

- Direct isoform identification without assembly
- Novel splice variant discovery
- Fusion transcript detection
- Allele-specific (break the tie)

\vspace{.3cm}

**Trade-offs**

- Lower throughput than short-read
- Higher error rates
- Higher cost per base

:::
::::::

\vfill\tiny
@Sharon2013-dr

## RNA-seq: mapping short reads to reference genome

\centerline{\includegraphics[height=.7\textheight]{img/RNA_seq_step2_WangGersteinSnyder.pdf}}

\vfill\tiny
@Wang2009-tg


## Gene-level quantification methods (CS-oriented)

:::::: {.columns}
::: {.column width=.45}

**Burrow-Wheeler (2009-2012)**

- \textcolor{blue}{\texttt{\textbf{bowtie}}} [@Langmead2009-us]: Ultrafast "dictionary" based mapping

- \textcolor{blue}{\texttt{\textbf{TopHat}}} [@Trapnell2009-lq]: Parsing out the `bowtie` results

- \textcolor{blue}{\texttt{\textbf{cufflinks}}} [@Trapnell2010-cv]: Exact transcript assembly model (first suggested FPKM)

- \textcolor{blue}{\texttt{\textbf{RSEM}}} [@Li2011-sj]: Expectation Maximization-based splice isoform quantification 

:::
::: {.column width=.45}

**Isoform-aware mapping (2012-2014)**

- \textcolor{blue}{\texttt{\textbf{STAR}}} [@Dobin2013-cv]: longer reads and splice-aware

\vspace{.5cm}

**Pseudo-alignment (2015-2017)**

- \textcolor{blue}{\texttt{\textbf{kallisto}}}: Ultra-fast pseudoalignment, orders of magnitude faster [@Bray2016-cn]

- \textcolor{blue}{\texttt{\textbf{salmon}}}: Fast, bias-aware quantification [@Patro2017-ch]

:::
::::::

## Hands-on session in `Seminar 06`

\Large

RNA-seq: from beginning to end – BAM to count data

\normalsize

## Aside: Allele-specific analysis separately quantifies diploid transcriptome

\begin{center}
\begin{tikzpicture}[scale=0.9]
  % Maternal chromosome
  \node[font=\small\bfseries] at (-2, 4.5) {Maternal};

  % Gene on maternal chromosome
  \draw[thick, fill=blue!30] (0, 4) rectangle (4, 4.5);
  \node[font=\scriptsize] at (1, 4.25) {Gene};

  % SNP marker on maternal
  \fill[red] (2, 4.25) circle (0.1);
  \node[above] at (2, 4.6) {SNP: A};

  % RNA from maternal (many)
  \foreach \x in {0.5, 1, 1.5, 2, 2.5, 3, 3.5} {
    \draw[->, thick, blue!60] (\x, 3.9) -- (\x, 3.2);
  }
  \node[font=\scriptsize, blue!60] at (5, 3.5) {High expression};

  % Sequencing reads from maternal (15 reads)
  % Row 1
  \draw[fill=blue!20] (0.2, 2.8) rectangle (0.9, 3.0);
  \node[font=\tiny] at (0.55, 2.9) {A};
  \draw[fill=blue!20] (1.0, 2.8) rectangle (1.7, 3.0);
  \node[font=\tiny] at (1.35, 2.9) {A};
  \draw[fill=blue!20] (1.8, 2.8) rectangle (2.5, 3.0);
  \node[font=\tiny] at (2.15, 2.9) {A};
  \draw[fill=blue!20] (2.6, 2.8) rectangle (3.3, 3.0);
  \node[font=\tiny] at (2.95, 2.9) {A};
  \draw[fill=blue!20] (3.4, 2.8) rectangle (4.1, 3.0);
  \node[font=\tiny] at (3.75, 2.9) {A};
  % Row 2
  \draw[fill=blue!20] (0.2, 2.55) rectangle (0.9, 2.75);
  \node[font=\tiny] at (0.55, 2.65) {A};
  \draw[fill=blue!20] (1.0, 2.55) rectangle (1.7, 2.75);
  \node[font=\tiny] at (1.35, 2.65) {A};
  \draw[fill=blue!20] (1.8, 2.55) rectangle (2.5, 2.75);
  \node[font=\tiny] at (2.15, 2.65) {A};
  \draw[fill=blue!20] (2.6, 2.55) rectangle (3.3, 2.75);
  \node[font=\tiny] at (2.95, 2.65) {A};
  \draw[fill=blue!20] (3.4, 2.55) rectangle (4.1, 2.75);
  \node[font=\tiny] at (3.75, 2.65) {A};
  % Row 3
  \draw[fill=blue!20] (0.2, 2.3) rectangle (0.9, 2.5);
  \node[font=\tiny] at (0.55, 2.4) {A};
  \draw[fill=blue!20] (1.0, 2.3) rectangle (1.7, 2.5);
  \node[font=\tiny] at (1.35, 2.4) {A};
  \draw[fill=blue!20] (1.8, 2.3) rectangle (2.5, 2.5);
  \node[font=\tiny] at (2.15, 2.4) {A};
  \draw[fill=blue!20] (2.6, 2.3) rectangle (3.3, 2.5);
  \node[font=\tiny] at (2.95, 2.4) {A};
  \draw[fill=blue!20] (3.4, 2.3) rectangle (4.1, 2.5);
  \node[font=\tiny] at (3.75, 2.4) {A};

  % Paternal chromosome
  \node[font=\small\bfseries] at (-2, 1.3) {Paternal};

  % Gene on paternal chromosome
  \draw[thick, fill=orange!30] (0, 0.8) rectangle (4, 1.3);
  \node[font=\scriptsize] at (1, 1.05) {Gene};

  % SNP marker on paternal
  \fill[blue] (2, 1.05) circle (0.1);
  \node[above] at (2, 1.4) {SNP: G};

  % RNA from paternal (fewer)
  \foreach \x in {1, 2, 3} {
    \draw[->, thick, orange!60] (\x, 0.7) -- (\x, 0.0);
  }
  \node[font=\scriptsize, orange!60] at (5, 0.3) {Low expression};

  % Sequencing reads from paternal (3 reads)
  \draw[fill=orange!20] (0.5, -0.4) rectangle (1.3, -0.2);
  \node[font=\tiny] at (0.9, -0.3) {G};
  \draw[fill=orange!20] (1.5, -0.4) rectangle (2.3, -0.2);
  \node[font=\tiny] at (1.9, -0.3) {G};
  \draw[fill=orange!20] (2.5, -0.4) rectangle (3.3, -0.2);
  \node[font=\tiny] at (2.9, -0.3) {G};

  % Result box (smaller and moved right)
  \node[font=\small\bfseries] at (8.25, 4.3) {RNA-seq};

  \node[font=\scriptsize, align=left] at (8.25, 3.4) {
    Read counts:\\
    Allele A: 15 reads\\
    Allele G: 3 reads
  };

  \node[font=\scriptsize, align=left] at (8.25, 2.4) {
    $\Rightarrow$ Allelic imbalance\\
    5:1 ratio
  };

  \node[font=\scriptsize, align=left] at (8.25, 1.0) {
    \textbf{Biological insights:}\\
    \tiny
    - Cis-regulatory\\
    - Imprinting\\
    - X-inactivation
  };

\end{tikzpicture}
\end{center}

\vfill\small

[@Montgomery2010-th; @Pickrell2010-ie] Let's do this in `lect12`

# RNA-seq data normalization

## Why normalize RNA-seq data?

:::::: {.columns}
::: {.column width=.45}

**Technical biases**

- Sequencing depth varies across samples
- Gene length affects read counts
- GC-content bias
- Batch effects

:::
::: {.column width=.45}

**Goal**

- Make expression levels comparable across genes and samples
- Remove technical variation while preserving biological signal

:::
::::::

## Normalization methods

:::::: {.columns}
::: {.column width=.45}

**Within-sample normalization**

- **RPKM/FPKM**: Reads per kilobase per million mapped reads [@Mortazavi2008-es]
- **TPM**: Transcripts per million (improved RPKM)

\vspace{.3cm}

*Adjusts for sequencing depth and gene length*

:::
::: {.column width=.45}

**Between-sample normalization**

- **TMM**: Trimmed mean of M-values [@Robinson2010-mk]
- **DESeq size factors**: Median of ratios [@Anders2010-cf]
- **RUVSeq**: Removing unwanted variation [@Risso2014-sa]

\vspace{.3cm}

*Addresses composition bias*

:::
::::::

## Continue to work on TCGA data

```{r}
data.dir <- "../data/recount3"
dir.create(data.dir, recursive = TRUE, showWarnings = FALSE)
.file <- file.path(data.dir, "tcga_brca_rse.rds")

if.needed(.file, {
    tcga.proj <- subset(available_projects(),
                        file_source == "tcga")
    brca.info <- tcga.proj[tcga.proj$project == "BRCA", ]
    rse.brca <- create_rse(brca.info)
    saveRDS(rse.brca, .file)
})
rse <- readRDS(.file)
```

## After alignment: `SummarizedExperiment`

```
                        samples (columns)
                     +---------------------+
                     |  S1   S2   S3  ...  |
               +-----|---------------------+
     genes     | G1  |  10   25   18  ...  |  <- assays (counts)
     (rows)    | G2  |   5   12    8  ...  |
               | ... | ...  ...  ...  ...  |
               +-----+---------------------+

  rowData: genomic coordinates (chr, start, end, strand)
  colData: sample metadata (tumor/normal, clinical info)
```

## Learning objectives: identify biases and address them

\Large

1. Gene bias (rows)

2. Sequencing depth (samples/columns)

3. Composition bias (corollary to sequencing depth)

\normalsize

## Get the data and compute the statistics

Here is gene expression count data:

```{r}
raw.counts <- assay(rse, "raw_counts")
```

Here is gene/row annotations

```{r size="scriptsize"}
rowData(rse) %>% head(3)
```

## Bias 1. Gene length vs. gene expression

```{r gene-length-bias-avg, echo=F, fig.width=3, fig.height=2.8, message=F}
## Calculate average log1p expression per gene
avg.expr <- rowMeans(raw.counts)

## Get gene length from rowData
bp.length <- rowData(rse)$bp_length

## Create data frame for plotting
.dt <- data.table(bp.length, avg.expr) %>%
    filter(avg.expr > 0)

ggplot(.dt, aes(x = log10(bp.length), y = log10(avg.expr))) +
    ggrastr::rasterize(geom_point(alpha=.5,size=0.5,colour="gray60", stroke=0)) +
    geom_density_2d(colour = "blue", linewidth=0.5) +
    geom_smooth(method = "lm", se = F, color = "red", lty=2, linewidth=.5) +
    stat_regline_equation(size = 4) +
    labs(x = "log10 gene length (bp)", y="log10 average expression")
```

## How did the gene length bias occur?

Original transcript:

```
5'--[5'UTR]--[=======CDS=======]--[3'UTR]--AAAAAAA 3'
                                           (polyA)
```

**RNA is fragmented before sequencing**

After fragmentation:

```
   ===    ===   ===    ===   ===    ===   ===
     ===    ===   ===    ===   ===    ===
```

## Longer genes produce more fragments

```
Short gene (1 kb):
|----------|
   ==  ==  ==
     ==  ==
   ~5 reads
```

```
Long gene (4 kb):
|------------------------------------------|
   ==  ==  ==  ==  ==  ==  ==  ==  ==  ==
     ==  ==  ==  ==  ==  ==  ==  ==  ==
   ~20 reads
```

Same expression $\implies$ 4 $\times$ more counts for the 4 $\times$ longer

## RPKM (Reads Per Kilobase per Million mapped)

$$\text{RPKM}_{gi} = \frac{Y_{gi}}{L_{g} / 1000 \times N_{i} / 10^6} = \frac{Y_{gi} \times 10^9}{L_{g} \times N_{i}}$$

Where:

- $Y_{gi}$ = raw count for gene $g$ in sample $i$
- $L_{g}$ = gene length of gene $g$ in bp
- $N_{i}$ = total mapped reads on a sample $i$ (library size)

*Order:* Normalize by library size first, then gene length.

\vfill\tiny
@Mortazavi2008-es

## TPM (Transcripts Per Million)

$$\text{TPM}_{gi} 
= \frac{Y_{gi} / L_{g}}{\sum_{g'} (Y_{g'i} / L_{g'})} \, 10^6
= \frac{\text{FPKM}_{gi}}{\sum_{g'}\text{FPKM}_{g'i} / 10^{6}}$$

*Order:* Normalize by gene length first, then scale to per million.

**Key difference:** TPM values sum to 1 million across genes within each sample, making samples directly comparable.

\vfill\tiny
@Li2011-sj, @Wagner2012-tpm

## Gene length vs. expression after RPKM adjustment

```{r gene-length-rpkm, fig.width=5, fig.height=2.8, message=F, echo=F, only.plot="1"}
## Calculate RPKM: RPKM = Y_g × 10^9 / (L_g × N)
## Y_g = raw count, L_g = gene length, N = total mapped reads
lib.sizes <- colSums(raw.counts)

rpkm <- sweep(sweep(raw.counts, 2, lib.sizes, `/`), 1, bp.length, `/`) * 1e9
      
## rpkm <- t(t(raw.counts) / lib.sizes) * 1e9 / bp.length

## Average RPKM per gene
avg.rpkm <- rowMeans(rpkm)
avg.expr <- rowMeans(raw.counts)

## Create data frame for plotting
.dt <-
    rbind(data.table(bp.length, y=avg.expr, type="count"),
          data.table(bp.length, y=avg.rpkm, type="RPKM")) %>% 
    filter(y > 0, y < 1e7)

ggplot(.dt, aes(x = log10(bp.length), y = y)) +
    facet_wrap(~ `type`, nrow=1, scales="free") +
    ggrastr::rasterize(geom_point(alpha=.5, size=0.5, colour="gray60", stroke=0)) +
    geom_density_2d(colour = "blue", linewidth=0.5) +
    geom_smooth(method = "lm", se = F,color = "red", lty=2, linewidth=.5) +
    stat_regline_equation(size = 4) +
    labs(x = "log10 gene length (bp)", y="average ...")
```

```{r gene-length-rpkm-log10, fig.width=5, fig.height=2.8, message=F, echo=F, only.plot="2"}
ggplot(.dt, aes(x = log10(bp.length), y = log10(y))) +
    facet_wrap(~ `type`, nrow=1, scales="free") +
    ggrastr::rasterize(geom_point(alpha=.5, size=0.5, colour="gray60", stroke=0)) +
    geom_density_2d(colour = "blue", linewidth=0.5) +
    geom_smooth(method = "lm", se = F,color = "red", lty=2, linewidth=.5) +
    stat_regline_equation(size = 4) +
    labs(x = "log10 gene length (bp)", y="log10 of average ...")
```

## Gene length vs. expression after TPM adjustment

```{r gene-length-tpm, fig.width=5, fig.height=2.8, echo=F, message=F, only.plot="1"}
## Calculate TPM: TPM = (Y_g / L_g) / sum(Y_g' / L_g') × 10^6
## First normalize by gene length, then scale to per million
reads.per.bp <- sweep(raw.counts, 1, bp.length, `/`)
tpm <- sweep(reads.per.bp, 2, colSums(reads.per.bp), `/`) * 1e6

## t(t(reads.per.bp) / colSums(reads.per.bp)) * 1e6

## Average TPM per gene
avg.tpm <- rowMeans(tpm)
avg.expr <- rowMeans(raw.counts)

## Create data frame for plotting
.dt <-
    rbind(data.table(bp.length, y=avg.expr, type="count"),
          data.table(bp.length, y=avg.tpm, type="TPM")) %>% 
    filter(y > 0, y < 1e7)

ggplot(.dt, aes(x = log10(bp.length), y = y)) +
    facet_wrap(~ `type`, nrow=1, scales="free") +
    ggrastr::rasterize(geom_point(alpha=.5, size=0.5, colour="gray60", stroke=0)) +
    geom_density_2d(colour = "blue", linewidth=0.5) +
    geom_smooth(method = "lm", se = F,color = "red", lty=2, linewidth=.5) +
    stat_regline_equation(size = 4) +
    labs(x = "log10 gene length (bp)", y="average ...")
```

```{r gene-length-tpm-log10, fig.width=5, fig.height=2.8, message=F, echo=F, only.plot="2"}
ggplot(.dt, aes(x = log10(bp.length), y = log10(y))) +
    facet_wrap(~ `type`, nrow=1, scales="free") +
    ggrastr::rasterize(geom_point(alpha=.5, size=0.5, colour="gray60", stroke=0)) +
    geom_density_2d(colour = "blue", linewidth=0.5) +
    geom_smooth(method = "lm", se = F,color = "red", lty=2, linewidth=.5) +
    stat_regline_equation(size = 4) +
    labs(x = "log10 gene length (bp)", y="log10 of average ...")
```

## Bias 2. Sequencing depth and library size

* Different samples are sequenced to different depths

* Different total read counts even when biological content is identical.

\vfill

```{r eval=F, echo=T, size="Large"}
lib.sizes <- colSums(raw.counts)
```

## Bias 2. Sequencing depth and library size

```{r library-size-variation, echo=F, fig.width=5, fig.height=2.5, message=F}
## Calculate library sizes (total reads per sample)
lib.sizes <- colSums(raw.counts)

## Visualize library size variation
.dt <- data.table(sample = 1:length(lib.sizes),
                  lib_size = sort(lib.sizes)) %>%
    mutate(lib_size_M = lib_size / 1e6,
           label = case_when(
               sample == 1 ~ sprintf("min: %.1fM", lib_size_M),
               sample == n() ~ sprintf("max: %.1fM", lib_size_M),
               TRUE ~ NA_character_
           ))

ggplot(.dt, aes(x = sample, y = lib_size_M)) +
    geom_point(alpha = 0.5) +
    geom_label_repel(aes(label = label), na.rm = TRUE,
                     nudge_y = 5, size = 3) +
    labs(x = "Sample (sorted)", y = "Library Size (millions of reads)",
         title = "Variation in sequencing depth across TCGA BRCA samples")
```

## Bias 3. RNA composition change by normalization

- Two samples, each with 1 million total reads
- Sample A: 1000 genes each expressed equally (1000 reads/gene)
- Sample B: One gene upregulated 100×, now takes 500,000 reads
  - The remaining 999 genes now share only 500,000 reads (~500 reads/gene)
  - These 999 genes appear "downregulated" even though their absolute expression is unchanged!

## RNA composition bias due to outlier genes

![](img/SequenceSpace.pdf)

\vfill

What would happen if normalized by library size?

## What would happen with naive normalization

\begin{center}
\begin{tikzpicture}[scale=0.8]
  % Sample 1 box
  \node[font=\small\bfseries] at (1.5, 4.2) {Sample 1};
  \node[font=\scriptsize] at (1.5, 3.7) {(10 transcripts)};

  % Sample 1 normalized values
  \node[anchor=east, font=\scriptsize] at (0.3, 2.8) {\textcolor{red}{Red:}};
  \node[anchor=west, font=\scriptsize] at (0.5, 2.8) {$3/10 = 0.30$};

  \node[anchor=east, font=\scriptsize] at (0.3, 2.2) {\textcolor{blue}{Blue:}};
  \node[anchor=west, font=\scriptsize] at (0.5, 2.2) {$3/10 = 0.30$};

  \node[anchor=east, font=\scriptsize] at (0.3, 1.6) {\textcolor{green!60!black}{Green:}};
  \node[anchor=west, font=\scriptsize] at (0.5, 1.6) {$4/10 = 0.40$};

  % Arrow and fold change (between samples)
  \draw[->, thick] (2.8, 2.8) -- (4.2, 2.8);
  \node[font=\tiny] at (3.5, 3.2) {0.45x};

  \draw[->, thick] (2.8, 2.2) -- (4.2, 2.2);
  \node[font=\tiny] at (3.5, 2.6) {0.45x};

  \draw[->, thick] (2.8, 1.6) -- (4.2, 1.6);
  \node[font=\tiny] at (3.5, 2.0) {1.8x};

  % Sample 2 box
  \node[font=\small\bfseries] at (6.8, 4.2) {Sample 2};
  \node[font=\scriptsize] at (6.8, 3.7) {(22 transcripts)};

  % Sample 2 normalized values
  \node[anchor=east, font=\scriptsize] at (5.8, 2.8) {\textcolor{red}{Red:}};
  \node[anchor=west, font=\scriptsize] at (6.0, 2.8) {$3/22 = 0.14$};

  \node[anchor=east, font=\scriptsize] at (5.8, 2.2) {\textcolor{blue}{Blue:}};
  \node[anchor=west, font=\scriptsize] at (6.0, 2.2) {$3/22 = 0.14$};

  \node[anchor=east, font=\scriptsize] at (5.8, 1.6) {\textcolor{green!60!black}{Green:}};
  \node[anchor=west, font=\scriptsize] at (6.0, 1.6) {$16/22 = 0.73$};

  % Conclusion box
  \draw[rounded corners, red!50, thick] (8.5, 3.5) rectangle (13.8, 1.2);
  \node[font=\scriptsize\bfseries, red!70!black] at (11.1, 3.2) {False conclusion:};
  \node[font=\scriptsize, align=left] at (11.1, 2.5) {Red \& Blue appear};
  \node[font=\scriptsize, align=left] at (11.1, 2.0) {\textbf{downregulated} (0.45x)};
  \node[font=\scriptsize, align=left] at (11.1, 1.5) {but actual counts unchanged!};

\end{tikzpicture}
\end{center}

\vfill

\small
**Problem**: Green gene's high expression "steals" proportional representation from others

## TMM (Trimmed Mean of M-values) by @Robinson2010-mk

- "A scaling normalization method for differential expression analysis of RNA-seq data"

- Implemented in **edgeR** Bioconductor package

- Calculate scaling factors based on weighted trimmed mean of log-fold-changes between samples.

- Normalization tailored for differential expression analysis (Thursday)

## TMM algorithm

1. Choose a reference sample (typically sample with median library size)
2. For each sample vs. reference, calculate:
   - M-values (log fold-change): 
   
$$M_g = \log_2\left(\frac{\text{count}_g^{\text{sample}}}{\text{count}_g^{\text{ref}}}\right)$$

3. Trim extreme M-values (default: 30%)

4. Calculate weighted mean of remaining M-values

5. Convert to scaling factor

**Assumption:** Most genes are not differentially expressed between samples.

## TMM implemented in \textcolor{blue}{\texttt{\textbf{edgeR}}}

```{r edger_setup}
library(edgeR)

## Create DGEList object
dge <- DGEList(counts = raw.counts)
```

```{r tmm_normalization, size="large"}
## Calculate TMM normalization factors
dge <- calcNormFactors(dge, method = "TMM")
```

## Another idea implemented in \textcolor{blue}{\texttt{\textbf{DESeq2}}}

```{r deseq2_setup}
library(DESeq2)
coldata <- data.frame(sample = colnames(raw.counts))

dds <- DESeqDataSetFromMatrix(countData = raw.counts,
                              colData = coldata,
                              design = ~ 1)
```

```{r deseq_normalization, size="large"}
## Calculate size factors
dds <- estimateSizeFactors(dds)

## Get normalized counts
deseq.norm <- counts(dds, normalized = TRUE)
```

## Summary: Three main sources of bias

\large

**1. Gene length bias** (RPKM/FPKM/TPM)

**2. Sequencing depth / Library size bias**

**3. RNA composition bias** (TMM)

## Pop quiz!

\large

1. For differential expression analysis, namely gene-by-gene comparison, which normalization methods should be applied?

1. For gene set enrichment analysis, which normalization methods should be applied?

\normalsize

# Understanding the nature of RNA-seq data

## Sequencing generates count data

\begin{center}
\begin{tikzpicture}[scale=0.9]
  % Microarray side
  \node[font=\large\bfseries, UBCblue] at (2, 5) {Microarray};

  % Continuous spectrum
  \draw[thick, ->] (0, 4) -- (4, 4);
  \node[below, font=\scriptsize] at (0, 4) {0};
  \node[below, font=\scriptsize] at (4, 4) {$\infty$};
  \node[above, font=\small] at (2, 4) {Continuous intensity};

  % Example values on spectrum
  \fill[blue!60] (1.2, 4) circle (0.08);
  \fill[red!60] (2.8, 4) circle (0.08);

  % Normal distribution (bell curve)
  \draw[thick, blue, smooth] plot[domain=0.2:3.8, samples=50]
    (\x, {2.2 + 1.3*exp(-(\x-2)^2/0.5)});
  \node[font=\scriptsize] at (2, 1.75) {$Y \sim \mathcal{N}(\mu, \sigma^2)$};

  % Methods
  \node[font=\scriptsize, align=center] at (2, 1.6) {t-test, ANOVA\\linear models};

  % Vertical divider
  \draw[thick, gray!50] (5, 0.5) -- (5, 5.5);

  % RNA-seq side
  \node[font=\large\bfseries, UBCblue] at (8, 5) {RNA-seq};

  % Discrete counts
  \draw[thick, ->] (6, 4) -- (10, 4);
  \foreach \x in {6,6.5,7,7.5,8,8.5,9,9.5,10} {
    \draw (\x, 3.9) -- (\x, 4.1);
  }
  \node[below, font=\scriptsize] at (6, 4) {0};
  \node[below, font=\scriptsize] at (7, 4) {1};
  \node[below, font=\scriptsize] at (8, 4) {2};
  \node[below, font=\scriptsize] at (9, 4) {3};
  \node[above, font=\small] at (8, 4.15) {Discrete counts};

  % Example values
  \fill[blue!60] (7, 4) circle (0.08);
  \fill[red!60] (8.5, 4) circle (0.08);

  % Discrete distribution
  \foreach \x/\h in {6.5/0.15, 7/0.35, 7.5/0.45, 8/0.50, 8.5/0.45, 9/0.35, 9.5/0.20} {
    \draw[thick, fill=orange!60] (\x, 2.8) rectangle (\x+0.3, 2.8+\h);
  }
  \node[font=\scriptsize] at (8, 2.4) {$Y \sim \text{NB}(\mu, \phi)$};

  % Methods
  \node[font=\scriptsize, align=center] at (8, 1.6) {GLM (log-link)\\DESeq2, edgeR};

\end{tikzpicture}
\end{center}

:::::: {.columns}
::: {.column width=.45}

**Microarray**

- Continuous intensity values
- Normal distribution assumptions

:::
::: {.column width=.45}

**RNA-seq**

- Discrete count data
- Poisson/Negative binomial models

:::
::::::

## RNA-seq: mapping short reads to reference genome

\centerline{\includegraphics[height=.7\textheight]{img/RNA_seq_step2_WangGersteinSnyder.pdf}}

\vfill\tiny
@Wang2009-tg


## Learning objectives

\Large

1. Understand mean-variance relationship in RNA-seq

2. Why Negative Binomial distribution for RNA-seq

3. Identify challenges in NB modelling

\normalsize

## Recap: the distributions for discrete data

```{r fig.width=6, fig.height=2.5, echo=F}
# Set up parameters for different discrete distributions
x_range <- 0:20

## Binomial
binom_n <- 20
binom_p <- 0.3
p1 <- data.frame(x = 0:binom_n,
                 pmf = dbinom(0:binom_n, size = binom_n, prob = binom_p)) %>%
    ggplot(aes(x = x, y = pmf)) +
    geom_col(fill = "coral", width = 0.7) +
    labs(title = paste0("Binomial(n=", binom_n, ", p=", binom_p, ")"),
         x = "x", y = "P(X = x)") +
    theme_minimal() +
    theme(plot.title = element_text(size = 8), axis.text = element_text(size=6))

## Poisson
pois_lambda <- 6
p2 <- data.frame(x = x_range,
                 pmf = dpois(x_range, lambda = pois_lambda)) %>%
    ggplot(aes(x = x, y = pmf)) +
    geom_col(fill = "darkgreen", width = 0.7) +
    labs(title = paste0("Poisson(λ=", pois_lambda, ")"),
         x = "x", y = "P(X = x)") +
    theme_minimal() +
    theme(plot.title = element_text(size = 8), axis.text = element_text(size=6))

## Negative Binomial
nb_r <- 5
nb_p <- 0.5
p3 <- data.frame(x = x_range,
                 pmf = dnbinom(x_range, size = nb_r, prob = nb_p)) %>%
    ggplot(aes(x = x, y = pmf)) +
    geom_col(fill = "orange", width = 0.7) +
    labs(title = paste0("Negative Binomial\n(r=", nb_r, ", p=", nb_p, ")"),
         x = "x", y = "P(X = x)") +
    theme_minimal() +
    theme(plot.title = element_text(size = 8), axis.text = element_text(size=6))

p1 | p2 | p3
```

\vfill
Which one best captures RNA-seq count data?

## Let's take a look at gene-level statistics

```{r mean-var-raw, size="large"}
## Calculate mean and variance per gene
gene.mean <- rowMeans(raw.counts)
gene.var <- rowVars(raw.counts)
```

## Mean-variance relationships in RNA-seq data

```{r rna_seq_mean-vs-var, fig.width=4.5, fig.height=2.5, only.plot="1", echo=F}
.dt <- data.table(mean = gene.mean, var = gene.var) %>%
    filter(`mean` > 0, `var` > 0, `var` < 1e7)

ggplot(.dt, aes(`mean`, `var`)) +
    ggrastr::rasterize(geom_point(alpha=.5, size=0.5, colour="gray60", stroke=0)) +
    geom_density_2d(colour = "blue", linewidth=0.5) +
    labs(x = "gene-level mean", y = "gene-level variance",
         title = "Mean-Variance relationship in raw counts")
```

```{r rna_seq_log10mean-vs-log10var, fig.width=4.5, fig.height=2.5, only.plot="2", echo=F}
.dt <- data.table(mean = gene.mean, var = gene.var) %>%
    filter(`mean` > 0, `var` > 0)

ggplot(.dt, aes(log10(`mean`), log10(`var`))) +
    ggrastr::rasterize(geom_point(alpha=.5, size=0.5, colour="gray60", stroke=0)) +
    geom_density_2d(colour = "blue", linewidth=0.5) +
    geom_abline(slope = 1, intercept = 0, color = "red", lty=2, linewidth=.5) +
    annotate("text", x = 0, y = 0, label = "mean = var",
             color = "red", hjust = 0, size = 4) +
    labs(x = "log10 gene-level mean", y = "log10 gene-level variance",
         title = "Mean-Variance relationship in raw counts")
```

\vfill

What is this called? \only<2>{\textbf{overdispersion}}


## Does TMM normalization fix overdispersion?

```{r tmm-mean-var, fig.width=4.5, fig.height=2.5, echo=F}
## Get TMM-normalized counts (CPM with TMM factors)
tmm.counts <- edgeR::cpm(dge)

## mean and variance per gene after TMM
tmm.mean <- rowMeans(tmm.counts)
tmm.var <- rowVars(as.matrix(tmm.counts))

.dt <- data.table(mean = tmm.mean, var = tmm.var) %>%
    filter(`mean` > 0, `var` > 0)

ggplot(.dt, aes(log10(`mean`), log10(`var`))) +
    ggrastr::rasterize(geom_point(alpha=.5, size=0.5, colour="gray60", stroke=0)) +
    geom_density_2d(colour = "blue", linewidth=0.5) +
    geom_abline(slope = 1, intercept = 0, color = "red", lty=2, linewidth=.5) +
    annotate("text", x = 0, y = 0.5, label = "mean = var",
             color = "red", hjust = 0, size = 4) +
    labs(x = "log10 gene-level mean (TMM)", y = "log10 gene-level variance (TMM)",
         title = "Mean-Variance relationship after TMM normalization")
```

\vfill

TMM addresses **composition bias**, not **overdispersion**.

## What about column-wise mean-variance?

```{r size="large"}
sample.mean <- colMeans(raw.counts)
sample.var <- colVars(raw.counts)
```

## Mean and variance per sample (column-wise)

```{r sample-mean-var, fig.width=4.5, fig.height=2.5, echo=F, only.plot="1"}
.dt <- data.table(mean = sample.mean, var = sample.var) %>%
    filter(`mean` > 0, `var` > 0)

ggplot(.dt, aes(`mean`, `var`)) +
    ggrastr::rasterize(geom_point(alpha=.5, size=1, colour="darkblue", stroke=0)) +
    geom_density_2d(colour = "blue", linewidth=0.5) +
    labs(x = "sample-level mean", y = "sample-level variance",
         title = "Sample-wise Mean-Variance relationship")
```

```{r sample-mean-var-log10, fig.width=4.5, fig.height=2.5, echo=F, only.plot="2"}
ggplot(.dt, aes(log10(`mean`), log10(`var`))) +
    ggrastr::rasterize(geom_point(alpha=.5, size=1, colour="darkblue", stroke=0)) +
    geom_density_2d(colour = "blue", linewidth=0.5) +
    labs(x = "log10 sample-level mean", y = "log10 sample-level variance",
         title = "Sample-wise Mean-Variance relationship")
```

\vfill

\onslide<2>{
similar mean-variance relationship?
}

## What about log2-transformed gene-level variance?

```{r log2-tmm-mean-var, fig.width=4.5, fig.height=2.5, echo=F}
## Log2 transform TMM-normalized counts
log2.counts <- log2(tmm.counts + 1)

## Mean and variance per gene after log2 transformation
log2.mean <- rowMeans(log2.counts)
log2.var <- rowVars(as.matrix(log2.counts))

.dt <- data.table(mean = log2.mean, var = log2.var) %>%
    filter(`mean` > 0, `var` > 0)

ggplot(.dt, aes(`mean`, `var`)) +
    ggrastr::rasterize(geom_point(alpha=.5, size=0.5, colour="gray60", stroke=0)) +
    geom_density_2d(colour = "blue", linewidth=0.5) +
    geom_hline(yintercept = 1, color = "red", lty=2, linewidth=.5) +
    labs(x = "gene-level mean (log2 TMM + 1)", y = "gene-level variance (TMM)",
         title = "Mean-Variance after log2 transformation")
```

\vfill

Log transformation helps, but **low-expression genes still have high variance**




## What does "over" dispersion of a random variable mean?

* Is there "under" dispersion?

* If there is neither "over" nor "under" dispersion:

$$\mathbb{E}\!\left[X\right] = \mathbb{V}\!\left[X\right]$$

\vfill

**Which one for RNA-seq data?**

:::::: {.columns}
::: {.column width=.45}

$$\mathbb{E}\!\left[X\right] < \mathbb{V}\!\left[X\right]$$

:::
::: {.column width=.45}

$$\mathbb{E}\!\left[X\right] > \mathbb{V}\!\left[X\right]$$

:::
::::::

## Why not Poisson?

:::::: {.columns}
::: {.column width=.45}

**Poisson assumption**

$$\text{Var}(Y) = \mu$$

- Mean equals variance
- Too restrictive for biological data

:::
::: {.column width=.45}

**Biological reality**

$$\text{Var}(Y) > \mu$$

- Overdispersion is common
- Variability exceeds expectation

:::
::::::

## Negative Binomial model

\only<1>{

$$Y \sim \text{NB}(\mu, \phi)$$

}

- $\mu$: mean expression
- $\phi$: dispersion parameter (captures overdispersion)
- $\phi > 0$

\only<2>{

$$P(Y = y) = \binom{y + r - 1}{y} p^r (1-p)^y$$

where $r = 1/\phi$ and $p = 1/(1 + \phi\mu)$

}

$$\mathbb{V}\!\left[Y\right] = \mu + \phi \mu^2$$

\only<3>{

$$P(Y = y) = \frac{\Gamma(y + 1/\phi)}{\Gamma(1/\phi) \, \Gamma(y + 1)} \left(\frac{1}{1 + \phi\mu}\right)^{1/\phi} \left(\frac{\phi\mu}{1 + \phi\mu}\right)^y$$

where $\Gamma(\cdot)$ is Euler's gamma function.

}

## Some intuition of NB distribution

:::::: {.columns}
::: {.column width=.45}

**Classical interpretation**

NB models the number of **failures** before achieving $r$ **successes** in a sequence of independent Bernoulli trials.

\vspace{.3cm}

- $r$: target number of successes
- $p$: probability of success in each trial
- $Y$: number of failures before $r$ successes

:::
::: {.column width=.45}

**Special cases**

\vspace{.3cm}

$r = 1$ $\Rightarrow$ **Geometric distribution**

- Number of failures before the first success

\vspace{.3cm}

$r \to \infty$ (or $\phi \to 0$) $\Rightarrow$ **Poisson**

- Overdispersion vanishes
- Mean = Variance

\vspace{.3cm}

**For RNA-seq:**

- NB is a flexible count distribution
- Accommodates overdispersion through $\phi$ (or $r$)

:::
::::::

\vfill\tiny
Source: \url{https://en.wikipedia.org/wiki/Negative_binomial_distribution}


## Let's take a look at overdispersion $\phi$ (or $1/\phi$)

$$P(Y = y) = \frac{\Gamma(y + 1/\phi)}{\Gamma(1/\phi) \, \Gamma(y + 1)} {\color{teal} \left(\frac{1}{1 + \phi\mu}\right) }^{1/\phi} {\color{magenta} \left(\frac{\phi\mu}{1 + \phi\mu}\right) }^y$$

where $\Gamma(\cdot)$ is Euler's gamma function.

$$\text{\color{magenta}success:} \qquad p = \frac{\phi \mu}{1 + \phi \mu} = \frac{ \mu}{1/\phi + \mu}$$

$$\text{\color{teal}failure:} \qquad 1-p = \frac{1}{1 + \phi \mu} = \frac{1/\phi}{1/\phi + \mu}$$

## What do overdispersion parameters do?

```{r nb-overdispersion-comparison, fig.width=6, fig.height=2.5, echo=F}
## Compare two negative binomial distributions with different overdispersion

## Set parameters
mu <- 10  # mean expression level
phi_low <- 0.1   # low overdispersion (closer to Poisson)
phi_high <- 1.0  # high overdispersion

## Convert to (size, prob) parameterization used by dnbinom
## phi = 1/size, so size = 1/phi
## prob = 1/(1 + phi*mu)
size_low <- 1/phi_low
prob_low <- 1/(1 + phi_low * mu)

size_high <- 1/phi_high
prob_high <- 1/(1 + phi_high * mu)

## Generate x values
x_range <- 0:30

## Create data
plot_data <- rbind(
  data.table(
    x = x_range,
    pmf = dnbinom(x_range, size = size_low, prob = prob_low),
    distribution = sprintf("Low overdispersion (φ=%.1f)", phi_low),
    variance = mu + phi_low * mu^2
  ),
  data.table(
    x = x_range,
    pmf = dnbinom(x_range, size = size_high, prob = prob_high),
    distribution = sprintf("High overdispersion (φ=%.1f)", phi_high),
    variance = mu + phi_high * mu^2
  )
)

## Create plot
## Create named vector for colors
color_vals <- c("darkgreen", "orange")
names(color_vals) <- c(sprintf("Low overdispersion (φ=%.1f)", phi_low),
                       sprintf("High overdispersion (φ=%.1f)", phi_high))

## Extract specific values for annotation at y=1 and y=10
highlight_vals <- plot_data[x %in% c(1, 10)]

## Calculate log-probability gaps
gap_low <- highlight_vals[x == 10 & grepl("Low", distribution), log(pmf)] -
           highlight_vals[x == 1 & grepl("Low", distribution), log(pmf)]
gap_high <- highlight_vals[x == 10 & grepl("High", distribution), log(pmf)] -
            highlight_vals[x == 1 & grepl("High", distribution), log(pmf)]

## Prepare annotation data with positions for dodged bars
dodge_width <- 0.8
highlight_vals[, x_pos := x + c(-0.5, 0.5) * dodge_width/2, by = x]

## Add log probabilities to plot_data
plot_data[, log_pmf := log(pmf)]
highlight_vals[, log_pmf := log(pmf)]

## Create vertical bar data showing the gap at x=25
gap_bars <- data.table(
    distribution = c(sprintf("Low overdispersion (φ=%.1f)", phi_low),
                     sprintf("High overdispersion (φ=%.1f)", phi_high)),
    x = 10,
    y_min = c(highlight_vals[x == 1 & grepl("Low", distribution), log_pmf],
              highlight_vals[x == 1 & grepl("High", distribution), log_pmf]),
    y_max = c(highlight_vals[x == 10 & grepl("Low", distribution), log_pmf],
              highlight_vals[x == 10 & grepl("High", distribution), log_pmf]),
    gap = c(abs(gap_low), abs(gap_high))
)

ggplot(plot_data, aes(x = x, y = log_pmf, fill = distribution)) +
    facet_grid(. ~ `distribution`) +
    geom_col(position = position_dodge(width = dodge_width), width = 0.7, alpha = 0.8) +
    scale_fill_manual(values = color_vals) +
    ## Add vertical lines at x=1 and x=10
    geom_vline(xintercept = c(1, 10), linetype = "dashed", color = "gray40", linewidth = 0.3) +
    ## Add horizontal lines showing log probability levels
    geom_hline(data = highlight_vals,
               aes(yintercept = log_pmf, color = distribution),
               linetype = 2, linewidth = 0.5, alpha = 0.7) +
    scale_color_manual(values = color_vals, guide = "none") +
    ## Add vertical bars showing the gap at x=25
    geom_segment(data = gap_bars,
                 aes(x = x, xend = x, y = y_min, yend = y_max),
                 color = "red", linewidth = .5, inherit.aes = FALSE,
                 arrow = arrow(ends = "both", length = unit(0.05, "inches"))) +
    ## Add gap value labels
    geom_text(data = gap_bars,
              aes(x = x, y = (y_min + y_max)/2, label = sprintf("%.2f", gap)),
              size = 3.5, fontface = "bold", color = "red",
              hjust = -0.3, inherit.aes = FALSE) +
    ## Annotate x=1 and x=10
    annotate("text", x = 1, y = min(plot_data$log_pmf) * 1.05,
             label = "y=1", size = 2, fontface = "bold", color = "gray20") +
    annotate("text", x = 10, y = min(plot_data$log_pmf) * 1.05,
             label = "y=10", size = 2, fontface = "bold", color = "gray20") +
    labs(
        title = sprintf("Negative Binomial distributions (μ=%d)", mu),
        x = "Count (y)",
        y = "log P(Y = y| μ, φ)",
        fill = "Distribution"
    ) +
    theme(legend.position = "none")
```

\vfill

With higher $\phi$, low count, e.g., $y=1$, matters more or less?

## Poisson vs. Negative Binomial in terms of $\mu,\phi$

\begin{center}
\begin{tikzpicture}[scale=1.0]
  % Poisson model (left)
  \node[font=\large\bfseries, UBCblue] at (2, 5) {Poisson};

  % Lambda (fixed parameter)
  \node[draw, circle, minimum size=1cm, fill=white] (lambda1) at (2, 3.5) {$\lambda$};
  \node[font=\tiny, above] at (2, 4.05) {fixed rate};

  % Y (observed)
  \node[draw, circle, minimum size=1cm, fill=blue!20, thick] (Y1) at (2, 2) {$Y$};
  \node[font=\tiny, below] at (2, 1.45) {count};

  % Arrow
  \draw[->, thick] (lambda1) -- (Y1);

  % Model equation
  \node[font=\scriptsize, align=center] at (2, 0.5) {$Y \sim \text{Poisson}(\lambda)$\\[0.1cm]$\mathbb{E}[Y] = \mathbb{V}[Y] = \lambda$};

  % Vertical divider
  \draw[thick, gray!50] (4.5, 0) -- (4.5, 5.5);

  % Negative Binomial model (right)
  \node[font=\large\bfseries, UBCblue] at (8, 5.3) {Negative Binomial};

  % Hyperparameters (fixed)
  \node[draw, rectangle, minimum width=0.8cm, minimum height=0.6cm, fill=gray!10] (mu) at (7, 4.5) {$\mu$};
  \node[draw, rectangle, minimum width=0.8cm, minimum height=0.6cm, fill=gray!10] (phi) at (9, 4.5) {$\phi$};

  % Lambda (random variable - latent)
  \node[draw, circle, minimum size=1cm, fill=white, dashed, thick] (lambda2) at (8, 3.5) {$\lambda$};
  \node[font=\tiny, above right] at (8.5, 3.0) {latent};

  % Y (observed)
  \node[draw, circle, minimum size=1cm, fill=blue!20, thick] (Y2) at (8, 2) {$Y$};
  \node[font=\tiny, below] at (8, 1.45) {count};

  % Arrows
  \draw[->, thick] (mu) -- (lambda2);
  \draw[->, thick] (phi) -- (lambda2);
  \draw[->, thick] (lambda2) -- (Y2);

  % Model equations
  \node[font=\scriptsize, align=center] at (8, 0.5) {$\lambda \sim \text{Gamma}(\mu, \phi)$\\[0.05cm]$Y | \lambda \sim \text{Poisson}(\lambda)$\\[0.1cm]$\mathbb{E}[Y] = \mu,\; \mathbb{V}[Y] = \mu + \phi\mu^2$};

  % Legend
  \draw[rounded corners, gray] (10.5, 5.0) rectangle (13.5, 3.4);

  \node[draw, circle, minimum size=0.45cm, fill=blue!20, thick] at (11, 4.6) {};
  \node[font=\tiny, right] at (11.3, 4.6) {observed};

  \node[draw, circle, minimum size=0.45cm, fill=white, dashed, thick] at (11, 4.1) {};
  \node[font=\tiny, right] at (11.3, 4.1) {latent};

  \node[draw, rectangle, minimum width=0.4cm, minimum height=0.3cm, fill=gray!10] at (11, 3.7) {};
  \node[font=\tiny, right] at (11.3, 3.7) {fixed param};

\end{tikzpicture}
\end{center}


## Poisson vs. Negative Binomial 

\begin{center}
\begin{tikzpicture}[scale=1.0]
  % Poisson model (left)
  \node[font=\large\bfseries, UBCblue] at (2, 5) {Poisson};

  % Lambda (fixed parameter)
  \node[draw, circle, minimum size=1cm, fill=white] (lambda1) at (2, 3.5) {$\lambda$};
  \node[font=\tiny, above] at (2, 4.05) {fixed rate};

  % Y (observed)
  \node[draw, circle, minimum size=1cm, fill=blue!20, thick] (Y1) at (2, 2) {$Y$};
  \node[font=\tiny, below] at (2, 1.45) {count};

  % Arrow
  \draw[->, thick] (lambda1) -- (Y1);

  % Model equation
  \node[font=\scriptsize, align=center] at (2, 0.5) {$Y \sim \text{Poisson}(\lambda)$\\[0.1cm]$\mathbb{E}[Y] = \mathbb{V}[Y] = \lambda$};

  % Vertical divider
  \draw[thick, gray!50] (4.5, 0) -- (4.5, 5.5);

  % Negative Binomial model (right)
  \node[font=\large\bfseries, UBCblue] at (8, 5.3) {Negative Binomial};

  % Hyperparameters (fixed)
  \node[draw, rectangle, minimum width=0.8cm, minimum height=0.6cm, fill=gray!10] (r) at (7, 4.5) {$r$};
  \node[draw, rectangle, minimum width=0.8cm, minimum height=0.6cm, fill=gray!10] (p) at (9, 4.5) {$p$};

  % Lambda (random variable - latent)
  \node[draw, circle, minimum size=1cm, fill=white, dashed, thick] (lambda2) at (8, 3.5) {$\lambda$};
  \node[font=\tiny, above right] at (8.5, 3.0) {latent};

  % Y (observed)
  \node[draw, circle, minimum size=1cm, fill=blue!20, thick] (Y2) at (8, 2) {$Y$};
  \node[font=\tiny, below] at (8, 1.45) {count};

  % Arrows
  \draw[->, thick] (r) -- (lambda2);
  \draw[->, thick] (p) -- (lambda2);
  \draw[->, thick] (lambda2) -- (Y2);

  % Model equations
  \node[font=\scriptsize, align=center] at (8, 0.5) {$\lambda \sim \text{Gamma}(r, p)$\\[0.05cm]$Y | \lambda \sim \text{Poisson}(\lambda)$\\[0.1cm]$\mathbb{E}[Y] = \mu,\; \mathbb{V}[Y] = \mu + \phi\mu^2$};

  % Legend
  \draw[rounded corners, gray] (10.5, 5.0) rectangle (13.5, 3.4);

  \node[draw, circle, minimum size=0.45cm, fill=blue!20, thick] at (11, 4.6) {};
  \node[font=\tiny, right] at (11.3, 4.6) {observed};

  \node[draw, circle, minimum size=0.45cm, fill=white, dashed, thick] at (11, 4.1) {};
  \node[font=\tiny, right] at (11.3, 4.1) {latent};

  \node[draw, rectangle, minimum width=0.4cm, minimum height=0.3cm, fill=gray!10] at (11, 3.7) {};
  \node[font=\tiny, right] at (11.3, 3.7) {fixed param};

\end{tikzpicture}
\end{center}


## Optional: different parameterizations of NB

:::::: {.columns}
::: {.column width=.45}

**$(r, p)$ parameterization**

\vspace{.3cm}

$$Y \sim \text{NB}(r, p)$$

\vspace{.2cm}

$$\mathbb{E}[Y] = \frac{r(1-p)}{p}$$

\vspace{.2cm}

$$\mathbb{V}[Y] = \frac{r(1-p)}{p^2}$$

:::
::: {.column width=.45}

**$(\mu, \phi)$ parameterization**

\vspace{.3cm}

$$Y \sim \text{NB}(\mu, \phi)$$

\vspace{.2cm}

$$\mathbb{E}[Y] = \mu$$

\vspace{.2cm}

$$\mathbb{V}[Y] = \mu + \phi\mu^2$$

:::
::::::

\vspace{.5cm}

## Optional: different parameterizations of NB


$$\mu = \frac{r(1-p)}{p} \qquad \phi = \frac{1}{r}$$

$$r = \frac{1}{\phi} \qquad p = \frac{1}{1 + \phi\mu}$$

\vfill

As $\phi \to 0$ (or $r \to \infty$): NB $\to$ Poisson with mean $\mu$

## Aside: Why Gamma-Poisson = Negative Binomial?

**Hierarchical model:**

$$\lambda|r,\beta \sim \text{Gamma}(r, \beta), \qquad Y | \lambda \sim \text{Poisson}(\lambda)$$

**Marginal distribution** (integrate out $\lambda$):

\begin{eqnarray*}
P(Y = y | r, \beta) &=& \int_0^\infty P(Y = y | \lambda) \, P(\lambda) \, d\lambda \\
\only<1>{
&=& \int_0^\infty \frac{\lambda^y e^{-\lambda}}{y!} \cdot \frac{\beta^r}{\Gamma(r)} \lambda^{r-1} e^{-\beta\lambda} \, d\lambda \\
&=& \frac{\beta^r}{y! \, \Gamma(r)} \int_0^\infty \lambda^{y+r-1} e^{-(1+\beta)\lambda} \, d\lambda\\
}
\only<2>{
&=& \frac{\beta^r}{y! \, \Gamma(r)} \cdot \frac{\Gamma(y+r)}{(1+\beta)^{y+r}} = \binom{y+r-1}{y} \left(\frac{\beta}{1+\beta}\right)^r \left(\frac{1}{1+\beta}\right)^y
}
\end{eqnarray*}
\only<2>{
$\implies$ negative binomial PMF with $p = \beta/ (1+\beta)$
}


## Challenge of using NB distribution

:::::: {.columns}
::: {.column width=.5}

\begin{center}
\begin{tikzpicture}[scale=1.0]
  % Parameters (open circles)
  \node[draw, circle, minimum size=1cm, fill=white] (mu) at (3, 5) {$\mu_{gi}$};
  \node[font=\tiny, above] at (3, 5.6) {mean};

  \node[draw, circle, minimum size=1cm, fill=white] (phi) at (6, 5) {$\phi_g$};
  \node[font=\tiny, above] at (6, 5.6) {dispersion};

  % Lambda (latent)
  \node[draw, circle, minimum size=1cm, fill=white, dashed, thick] (lambda) at (4.5, 3.5) {$\lambda_{gi}$};
  \node[font=\tiny, right] at (5.1, 3.5) {latent};

  % Y (observed)
  \node[draw, circle, minimum size=1cm, fill=blue!20, thick] (Y) at (4.5, 1.5) {$Y_{gi}$};
  \node[font=\tiny, below] at (4.5, 0.95) {count};

  % Arrows
  \draw[->, thick] (mu) -- (lambda);
  \draw[->, thick] (phi) -- (lambda);
  \draw[->, thick] (lambda) -- (Y);

  % Model equation
  \node[font=\scriptsize, align=center] at (4.5, 0.0) {$\lambda_{gi} \sim \text{Gamma}(\mu_{gi}, \phi_g)$\\[0.05cm]$Y_{gi} | \lambda_{gi} \sim \text{Poisson}(\lambda_{gi})$};

\end{tikzpicture}
\end{center}

:::
::: {.column width=.4}

\large

* $\mu_{gi}$: gene $g$, sample $i$

* $\phi_g$: gene $g$ only

* $Y_{gi}$: observed count


\vspace{.5cm}

**How many $\phi_{g}$ values?**

* Just one: $\phi_{g} = \phi_{g'}$ for all $g \neq g'$

* For each gene: $\phi_{g} \neq \phi_{g'}$ if $g \neq g'$

\normalsize

:::
::::::






## Discussion

\large

1. Can I ignore overdispersion if I don't have deep sequencing?

1. I forgot to perform TMM normalization, but can I adjust library depth by incorporating overdispersion parameters?

\normalsize


## Statistical methods for estimating overdispersion $\phi_{g}$

\large

* \textcolor{blue}{\texttt{\textbf{edgeR}}}: @Robinson2008-pd

* \textcolor{blue}{\texttt{\textbf{DESeq}}}: @Anders2010-cf

* \textcolor{blue}{\texttt{\textbf{DESeq2}}}: @Love2014-dl

\normalsize

## Next lecture 

\large

1. How to estimate overdispersion $\phi_{g}$

2. How to incorporate our working model into $\mu_{gi}$

3. Practical solutions using \textcolor{blue}{\texttt{\textbf{DESeq2}}} and \textcolor{blue}{\texttt{\textbf{limma-voom}}}

\normalsize

## Summary

\large

* RNA-seq count data generation process

* Potential Biases in RNA-seq data (row \& column)

* Properties in RNA-seq-derived count data

\normalsize


## Reference {.allowframebreaks}
