---
title: "How can we formulate our research questions effectively?"
author: "Yongjin Park"
date: "13 January 2026"
classoption: "aspectratio=169"
fontsize: 12pt
output:
    beamer_presentation:
        theme: Madrid
        keep_md: true
        keep_tex: true
        latex_engine: xelatex
        slide_level: 2
header-includes:
  - \usepackage{cancel}
  - \usepackage{tikz}
  - \usetikzlibrary{positioning,calc,arrows.meta}
  - \tikzset{>={Stealth[length=3mm,width=2mm]}}
  - \definecolor{UBCblue}{rgb}{0.04706, 0.13725, 0.26667}
  - \usecolortheme[named=UBCblue]{structure}
  - \setbeamertemplate{frametitle}{\color{UBCblue}\bfseries\insertframetitle\par\vskip-6pt\hrulefill}
  - \setbeamercolor{title in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{section in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{author in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{date in head/foot}{bg=white,fg=gray}
  - \setbeamertemplate{page number in head/foot}{}
  - \setbeamertemplate{frame numbering}[none]
  - \setbeamercolor{alerted text}{bg=yellow}
  - \AtBeginSection[]{\begin{frame}\frametitle{Today's lecture}{\Large\tableofcontents[currentsection]}\end{frame}}
  - |
    \makeatletter
    \def\ps@titlepage{%
      \setbeamertemplate{footline}{}
    }
    \addtobeamertemplate{title page}{\thispagestyle{titlepage}}{}
    \makeatother
    \include{toc}
---


```{r setup, include=FALSE}
setwd("~/work/teaching/stat540/lectures/lect03-defining-problems/")
library(data.table)
library(tidyverse)
library(patchwork)
library(matrixTests)
source("Setup.R")
fig.dir <- "../Fig/lect03-problem-formulation/"
dir.create(fig.dir, showWarnings=FALSE)
setup.env(fig.dir)
dir.create("../data", showWarnings=FALSE)
```

## Course Logistics

:::::: {.columns}
::: {.column width=.2}

:::
::: {.column width=.7}

\Large

**Please fill out the survey**

of your background

:::
::::::


## Advertisement

\Large

* [ASDA](https://asda.stat.ubc.ca/)

\normalsize

* https://asda.stat.ubc.ca/2025/11/10/free-stats-help.html

* If you and your team are interested in the following ASDA projects, let me know. 

* I will send you one-page project proposal document. 

## ASDA \#1: Blood RNAs for Febrile Illness Diagnosis

* Scoping review identified 286 RNA signatures from literature

* Signatures diagnose or predict outcomes of febrile illnesses (sepsis, TB, infections)

* Goal: Characterize and understand potential sources of biases

* Meta-analysis across many different studies

## ASDA \#2: `Breath print` data for Lung Cancer Detection

* Breath samples from 500+ participants (lung cancer vs. healthy controls)

* Volatile organic compounds (VOCs) analyzed via mass spectrometry

* High-dimensional data: ~500-1500 compounds per sample

* Goal: Build analysis pipeline for biomarker discovery and classification

## Roadmap of STAT/BIOF/GSAT540 2026

_data generation_ followed by _inference_ or validation

\vfill

\large

:::::: {.columns}
::: {.column width=.45}

2. _Statistics and matrix algebra_

3. \textbf{\color{magenta}Formulating biological questions} (Today)

4. \textbf{\color{teal} Hypothesis testing} (Thursday)

5. RNA-seq data analysis

6. Statistical inference

:::
::: {.column width=.45}

8. Single-cell genomics

9. Matrix factorization

10. Genome-wide association studies

11. Multiple hypothesis testing

:::
::::::

\normalsize

##

\huge

A biological problem $\implies$ 

\flushright 

a computational problem

\normalsize

## Guide 1: Crystallize abstract ideas into a concrete computational problem by writing a paper

:::::: {.columns}
::: {.column width=.45}

::: {.block}
### Research Model \#1

1. Idea
2. Do research
3. Write a paper

:::

:::
::: {.column width=.45}



::: {.block}
### Research Model \#2

* Idea
* **Write a paper**
* Do research

:::

:::
::::::

Which model would work better for you? Why?

\only<2>{
${}$
\centerline{\textbf{\color{teal} Discussion with your peers around you (5 min).}}
${}$
}

\vfill

\normalsize

\begin{enumerate}
\item<3-> Model \#2 forces us to be clear, focused
\item<3-> Crystallizes what we don’t understand yet (and what we don't need to know)
\item<3-> Opens the way to dialogue with others: reality check and collaboration
\end{enumerate}

\tiny

Simon Peyton Jones, [How to write a great research paper](https://www.microsoft.com/en-us/research/wp-content/uploads/2016/07/How-to-write-a-great-research-paper.pdf)

\normalsize

##

\huge
\begin{center}
${}$ Model \#1 or \#2?
\end{center}

## Guide 2: Ten simple rules ...

:::::: {.columns}
::: {.column width=.3}

:::
::: {.column width=.66}

\large

William Noble,

[**Ten simple rules for defining a computational biology project.**](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1010786)

*PLoS Comput Biol* (2023)

\normalsize

:::
::::::


## Summary: Ten Simple Rules

\large

:::::: {.columns}
::: {.column width=.43}

1. \textbf{\color{teal} Write it down}
2. \textbf{\color{teal} Define the problem}
3. Define potential impact
4. Summarize related work
5. \textbf{\color{teal} State central hypothesis}

:::
::: {.column width=.47}

6. Sketch your approach
7. \textbf{\color{teal} Identify available data}
8. Choose validation measures
9. Set up version control
10. Share with colleagues

:::
::::::

\normalsize

$${}$$

\textbf{\color{blue}\large Discussion: How many items have you followed in the past?}

## Rule 1: Write it down

\Large

Document your idea in "writing"

\normalsize

\vfill

* Clarifies your thinking and the underlying problem

* Exposes half-baked concepts early

* Forces you to articulate assumptions

> "The act of writing your idea down will inevitably clarify it for you."

\vfill

**Even a rough draft is better than no draft**

## Rule 2: Define the problem

\Large

Clearly delineate the scope and nature of the problem

\normalsize

\vfill

* What exactly are you trying to solve?

* What is **in scope** vs. **out of scope**?

* Be specific, not vague


  * ma\vfill

Which sounds more specific?

:::::: {.columns}
::: {.column width=.45}

* Understand gene expression programs in myeloid cancer

:::
::: {.column width=.45}

* Predict enhancer-promoter interaction in K562 cells

:::
::::::


<!-------------------->
<!-- skip this item -->
<!-------------------->

<!------------------------------------------------------>
<!-- ## Rule 3: Define potential impact				  -->
<!-- 												  -->
<!-- \Large											  -->
<!-- 												  -->
<!-- Identify who benefits and how					  -->
<!-- 												  -->
<!-- \normalsize									  -->
<!-- 												  -->
<!-- \vfill											  -->
<!-- 												  -->
<!-- * Who is your audience?						  -->
<!-- * What concrete benefits will your work provide? -->
<!-- * How does it advance the field?				  -->
<!-- 												  -->
<!-- \vfill											  -->
<!-- 												  -->
<!-- This helps:									  -->
<!-- 												  -->
<!-- 1. Motivate stakeholders						  -->
<!-- 2. Write compelling grant applications			  -->
<!-- 3. Frame your paper's contribution				  -->
<!------------------------------------------------------>

## Rule 4: Summarize related work

\Large

Conduct literature review early

\normalsize

\vfill

> "A terrible trap to fall into is carrying out an entire research
> project only to find that the same work was already published before
> you began."

* Learn from others' approaches

* Identify gaps in current knowledge

* **Summarize** methods in a chronologically ordered table

## Rule 5: State central hypothesis

\Large

Reduce your idea to a single sentence

\large

\vfill

* Focus on the one, single, **overarching** theme

* Make it **testable**

* Keep it **simple and clear**

\vfill

\normalsize

* It takes practice and time.

<!---------------------------------------------------------------->
<!-- ## Rule 6: Sketch your approach						    -->
<!-- 														    -->
<!-- \Large													    -->
<!-- 														    -->
<!-- Describe the computational method in plain English		    -->
<!-- 														    -->
<!-- \normalsize											    -->
<!-- 														    -->
<!-- \vfill													    -->
<!-- 														    -->
<!-- * What is your strategy?								    -->
<!-- * Highlight strengths and weaknesses					    -->
<!-- * Compare to alternative approaches					    -->
<!-- * Don't dive into code yet								    -->
<!-- 														    -->
<!-- \vfill													    -->
<!-- 														    -->
<!-- **Goal:** Communicate the *idea* before the implementation -->
<!---------------------------------------------------------------->

## Rule 7: Identify available data

\Large

Ensure you have both inputs and validation data

\normalsize

\vfill

* What are the input?

* How do we know that our discovery is correct?

* Simulation studies are okay.

<!------------------------------------------------------------->
<!-- ## Rule 8: Choose validation measures					 -->
<!-- 														 -->
<!-- \Large													 -->
<!-- 														 -->
<!-- Design quantitative metrics aligned with end-user needs -->
<!-- 														 -->
<!-- \normalsize											 -->
<!-- 														 -->
<!-- \vfill													 -->
<!-- 														 -->
<!-- * Don't rely only on textbook metrics					 -->
<!-- * Consider what matters to biologists					 -->
<!-- * Define success criteria upfront						 -->
<!-- 														 -->
<!-- \vfill													 -->
<!-- 														 -->
<!-- Examples:												 -->
<!-- 														 -->
<!-- * Classification: AUC, precision, recall				 -->
<!-- * Prediction: correlation, RMSE						 -->
<!-- * Discovery: enrichment, biological plausibility		 -->
<!------------------------------------------------------------->

## Rule 9: Set up version control

\Large

Use `git` or a similar tool from day one

\normalsize

\vfill

* Track code evolution from project inception

    * Keep the dates

* Enable collaboration

* Maintain reproducibility

* Document your progress


## Rule 10: Share with colleagues

\Large

Seek feedback on feasibility and scientific importance

\normalsize

\vfill

* Prepare to "Think Again" (don't be embarrassed or offended by being wrong)

* Refine your ideas through discussion

* Find potential collaborators

## Ten Simple Rules (items) $\to$ A paper

:::::: {.columns}
::: {.column width=.43}

1. \textbf{\color{teal} Write it down}
2. \textbf{\color{teal} Define the problem}
3. Define potential impact
4. Summarize related work
5. \textbf{\color{teal} State central hypothesis}
6. Sketch your approach
7. \textbf{\color{teal} Identify available data}
8. Choose validation measures
9. Set up version control
10. Share with colleagues

:::
::: {.column width=.47}

* Abstract
  * Rule \#1, \#2, \#5

* Introduction
  * Rule \#2, \#3, \#4, \#5

* Results
  * Rule \#5, \#7, \#10

* Discussion
  * Rule \#10

* Methods
  * Rule \#6, \#7, \#8, \#9, \#10

:::
::::::


## Bottom line

\Large

${}$ \textbf{It's so critical to describe your problem clearly.}

\begin{itemize}
\item <2-> Recognize the underlying computational problem(s)
\item <3-> Make an actionable plan to solve them.
\item <4> Evaluate what/how it worked and didn't.
\end{itemize}

## Is defining a problem worth my time?

\centerline{\includegraphics[height=.72\textheight]{img/automation_2x.png}}

\vfill

\tiny
[https://xkcd.com/1319/](https://xkcd.com/1319/)
\normalsize

## Two flavors

:::::: {.columns}
::: {.column width=.45}

* \#1: Optimization approach

	* \textbf{\color{magenta} The art of abstraction}

	* Define **input**, **output**, and **metric** (objective function)

	* Find the same or similar existing problems in the literature

:::
::: {.column width=.45}

* \#2: Probabilistic Modelling

	* \textbf{ Think of a potential {\color{teal} generative model}}

	* Describe the problem in terms of random variables

	* Structural Equation Model (in epidemiology)

	* Use an existing inference algorithm

:::
::::::

# Define a problem as an optimization problem

## The art of abstraction: Same data, different questions

\Large

Single-cell RNA-seq data: a matrix of **genes $\times$ cells**

\normalsize

\vfill

Biological questions $\rightarrow$ computational problems


## Exercise: Eleven Grand Challenges in Single-Cell

\footnotesize

Lähnemann et al., *Genome Biology* (2020)

\normalsize

:::::: {.columns}
::: {.column width=.48}

**Transcriptomics**

1. Handling sparsity (dropout)
2. \textbf{\color{magenta} Defining cell identity}
3. Tracking cellular dynamics
4. Mapping to reference atlases

**Genomics**

5. Finding variants (scDNA-seq)
6. Identifying copy-number changes
7. Inferring regulatory networks

:::
::: {.column width=.48}

**Phylogenetics**

8. Understanding tumor evolution
9. Calling structural variants

**Integration**

10. Multi-modal integration
11. Scaling to millions of cells

:::
::::::

Let's work on one of these examples

## Example: Defining cell identity (Challenge 2)

\Large

**Biological question:** What cell types are in my sample?

\vfill


* What are the variables of interest?

* Is it important?

\normalsize


## How should we think of cells in this context?

\Large

Data: $m\times{n}$ expression matrix $\mathbf{\color{magenta} X}$ ($m$ genes $\times$ $n$ cells)

\vfill

\only<1>{

$$
\mathbf{X} = \begin{pmatrix}
x_{11} & x_{12} & \cdots & x_{1n} \\
x_{21} & x_{22} & \cdots & x_{2n} \\
\vdots & \vdots & \ddots & \vdots \\
x_{m1} & x_{m2} & \cdots & x_{mn}
\end{pmatrix}
=
\left(
\mathbf{x}_{1}, \mathbf{x}_{2}, \ldots, \mathbf{x}_{n}
\right)$$
where
$m$ corresponds to 19k genes.
}

\vfill

\begin{itemize}
\item<2> Is a cell a vector of gene expressions over 19k genes?
\item<2> Is a cell a vector of gene expressions over several genes?
\item<2> Is a cell a point in a "cell type landscape?"
\item<2> ...
\end{itemize}

\normalsize

## Let's start writing out the problem

\Large

**Biological question:** What cell types are in my sample?

\normalsize

\vfill

:::::: {.columns}
::: {.column width=.45}

::: {.block}
### Input
Gene expression $X$ ($m$ genes $\times$ $n$ cells)
:::

:::
::: {.column width=.45}

::: {.block}
### Output
Assign each cell vector $\mathbf{x}_{j}$ to label $c_i \in \{1, \ldots, K\}$
:::

:::
::::::

## Brainstorming ...

* How did the cell types emerge in multicellular organisms?

* How did the cell express genes?

* What do we measure with a cell-level gene expression matrix?

* How do we know cells belong to a cell type A, not B?


## Example: Let's write it down with several sentences

1. We can consider each cell as a point along a differentiation trajectory.

2. We can represent the relationships between cells as a graph (connecting points).

3. Cells that share similar expression profiles are closely located in a graph space.

\vfill

Can we do better? Is it specific enough?


## How do we measure the distance between cells?

Multiple levels of distance measures between two cell vectors:

1. **Use all genes**: Compute distance using all $m$ features (e.g., Euclidean, cosine)

2. **Focus on selected genes**: Use only marker genes or highly variable genes

3. **Learn new coordinates**: Project cells into a lower-dimensional space (e.g., PCA) and compute distance there


## A toy example: 100 genes and 200 cells/samples

```{r simulate_data, include=F}
library(MASS)
set.seed(42)

## Simulate 3 cell states with branching trajectory
## State 1 (root) -> State 2 (branch A) or State 3 (branch B)
n_cells <- c(80, 60, 60)  # cells per state
n_genes <- 100
n_total <- sum(n_cells)

## Generate cells along trajectories in latent space
simulate_trajectory <- function(n, mu_start, mu_end, noise = 0.3) {
    t <- seq(0, 1, length.out = n)  # pseudotime
    X <- matrix(0, n, 2)
    for (i in 1:n) {
        mu_t <- (1 - t[i]) * mu_start + t[i] * mu_end
        X[i, ] <- mvrnorm(1, mu = mu_t, Sigma = diag(2) * noise)
    }
    list(X = X, t = t)
}

## Define trajectory endpoints in 2D latent space
mu_root <- c(-3, 0)
mu_mid <- c(0, 0)
mu_branch_A <- c(3, 2)
mu_branch_B <- c(3, -2)

## Generate cells along each trajectory segment
cells_1 <- simulate_trajectory(n_cells[1], mu_root, mu_mid, noise = 0.2)
cells_2 <- simulate_trajectory(n_cells[2], mu_mid, mu_branch_A, noise = 0.25)
cells_3 <- simulate_trajectory(n_cells[3], mu_mid, mu_branch_B, noise = 0.25)

## Combine latent coordinates
Z <- rbind(cells_1$X, cells_2$X, cells_3$X)
true_state <- factor(c(rep("Root", n_cells[1]),
                       rep("Branch A", n_cells[2]),
                       rep("Branch B", n_cells[3])))
pseudotime <- c(cells_1$t, 1 + cells_2$t, 1 + cells_3$t)

## Generate gene expression from latent space
## First 10 genes are trajectory-informative, rest are noise
W <- matrix(rnorm(n_genes * 2, sd = 0.5), nrow = 2, ncol = n_genes)
W[, 1:5] <- W[, 1:5] * 3   # PC1-associated genes
W[, 6:10] <- W[, 6:10] * 2  # PC2-associated genes

## Expression = latent * loadings + noise
X <- Z %*% W + matrix(rnorm(n_total * n_genes, sd = 0.5), nrow = n_total)
X <- t(X)  # genes x cells format
colnames(X) <- paste0("cell_", 1:n_total)
rownames(X) <- paste0("gene_", 1:n_genes)

## Run PCA for visualization
pca_result <- prcomp(t(X), scale. = TRUE)
```

```{r fig.width=5, fig.height=3, echo=F}
.matshow(X) + ylab("genes") + xlab("cells")
```

## Compute cell-cell distance in Top 5 PCA space

```{r include=F}
x.pca <- predict(pca_result)[, 1:5]
D <- as.matrix(dist(x.pca))
## Extract upper triangle (unique pairwise distances)
d.vec <- D[upper.tri(D)]
```

:::::: {.columns}
::: {.column width=.2}

* row = cells; column = cells

${}$

Note: _Don't worry about what PCA is yet (`lect-09`)_

${}$

Instead of dealing with 100 (genes) dimensions, we only need to deal with 5 dimensions.

:::
::: {.column width=.4}

```{r fig.width=2.4, fig.height=2.4, echo=F, onslide.plot="1-"}
.matshow(D)
```

\tiny
blue (small); red (large)
\normalsize

:::
::: {.column width=.3}

```{r fig.width=2, fig.height=2.8, echo=F, onslide.plot="2"}
ggplot(data.frame(d = d.vec), aes(x = d)) +
    geom_histogram(bins = 40, fill = "steelblue", alpha = 0.7) +
    theme_minimal() +
    theme(axis.title=element_text(size=8)) +
    theme(axis.text=element_text(size=8)) +
    labs(x = "Distance", y = "Count")
```

:::
::::::


## Visualize cells in PC space

```{r include=F}
library(igraph)
## Build MST using Prim's algorithm (via igraph)
g.pca <- graph_from_adjacency_matrix(D, mode="undirected", weighted=TRUE)
mst <- mst(g.pca)
## Extract MST edges and weights
mst.edges <- as_edgelist(mst)
mst.weights <- E(mst)$weight
```

```{r include=F}
## Prepare data for ggplot
pca.df <- data.frame(
    PC1 = pca_result$x[, 1],
    PC2 = pca_result$x[, 2],
    state = true_state
)
```

```{r fig.width=4, fig.height=3, echo=F}
ggplot(pca.df, aes(x = PC1, y = PC2)) +
    geom_point(size = 1.5, alpha = 0.8) +
    theme_minimal()
```

## Reminder: Our goal is to build a graph of cells

\large

* Graph = connecting dots (dot = cell)

* Tree = a special case of graph

* Minimum Spanning Tree = one way to connect the dots

* What is our assumption?

\normalsize


## Building Minimum Spanning Tree

```{r include=F}
## Prepare edge data for ggplot
## Match vertex names to row indices
idx1 <- match(mst.edges[, 1], colnames(X))
idx2 <- match(mst.edges[, 2], colnames(X))
edge.df <- data.frame(
    x = pca.df$PC1[idx1],
    y = pca.df$PC2[idx1],
    xend = pca.df$PC1[idx2],
    yend = pca.df$PC2[idx2],
    weight = mst.weights
)
```

```{r include=F}
## Function to plot with first k edges
plot_mst_step <- function(k) {
    p <- ggplot(pca.df, aes(x = PC1, y = PC2))
    if (k > 0) {
        p <- p + geom_segment(data = edge.df[1:k, , drop = FALSE],
                              aes(x = x, y = y, xend = xend, yend = yend,
                                  linewidth = weight),
                              color = "grey50", alpha = 0.6) +
            scale_linewidth_continuous(range = c(0.3, 1.5), guide = "none")
    }
    p + geom_point(aes(color = state), size = 1.5, alpha = 0.8) +
        theme_minimal() +
        labs(color = "Cell State") +
        ggtitle(paste0("Edges: ", k))
}
```

```{r mst_animation, fig.width=4, fig.height=3, echo=F, results='asis'}
for (k in 1:10) {
    cat(sprintf("\\only<%d>{\n", k))
    print(plot_mst_step(k * 10))
    cat("}\n")
}
```

## Build minimum spanning tree (MST)

```{r fig.width=4.5, fig.height=3, echo=F}
ggplot(pca.df, aes(x = PC1, y = PC2)) +
    geom_segment(data = edge.df, aes(x = x, y = y, xend = xend, yend = yend,
                                     linewidth = weight),
                 color = "grey50", alpha = 0.5) +
    scale_linewidth_continuous(range = c(0.2, 1.5)) +
    geom_point(aes(color = state), size = 1.5, alpha = 0.8) +
    theme_minimal() +
    labs(color = "Cell State", linewidth = "Distance")
```

## Graph-based 2D layout show the MST more explicitly

```{r include=F}
## Compute graph-based layouts from MST
## layout.fr <- layout_with_fr(mst)  # Fruchterman-Reingold
layout.kk <- layout_with_kk(mst)  # Kamada-Kawai

## Create data frames for plotting
pca.df.kk <- data.frame(
    X1 = layout.kk[, 1],
    X2 = layout.kk[, 2],
    state = true_state
)
edge.df.kk <- data.frame(
    x = layout.kk[idx1, 1], y = layout.kk[idx1, 2],
    xend = layout.kk[idx2, 1], yend = layout.kk[idx2, 2],
    weight = mst.weights
)
```

:::::: {.columns}
::: {.column width=.48}

**PCA layout**

```{r fig.width=2.8, fig.height=2.5, echo=F}
ggplot(pca.df, aes(x = PC1, y = PC2)) +
    geom_segment(data = edge.df, aes(x = x, y = y, xend = xend, yend = yend,
                                     linewidth = weight),
                 color = "grey50", alpha = 0.5) +
    scale_linewidth_continuous(range = c(0.2, 1.5), guide = "none") +
    geom_point(aes(color = state), size = 1, alpha = 0.8) +
    theme_minimal() +
    theme(legend.position = "none")
```

:::
::: {.column width=.48}

**Graph-based layout**

```{r fig.width=2.8, fig.height=2.5, echo=F}
ggplot(pca.df.kk, aes(x = X1, y = X2)) +
    geom_segment(data = edge.df.kk, aes(x = x, y = y, xend = xend, yend = yend,
                                        linewidth = weight),
                 color = "grey50", alpha = 0.5) +
    scale_linewidth_continuous(range = c(0.2, 1.5), guide = "none") +
    geom_point(aes(color = state), size = 1, alpha = 0.8) +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(x = "", y = "")
```

:::
::::::

## What if we build Maximum Spanning Tree?

```{r include=F}
## Build Maximum Spanning Tree by inverting weights
D.inv <- max(D) - D + 1  # Invert distances
g.inv <- graph_from_adjacency_matrix(D.inv, mode = "undirected", weighted = TRUE)
mst.max <- mst(g.inv)

## Extract edges and get original weights
mst.max.edges <- as_edgelist(mst.max)
mst.max.idx1 <- match(mst.max.edges[, 1], colnames(X))
mst.max.idx2 <- match(mst.max.edges[, 2], colnames(X))
mst.max.weights <- D[cbind(mst.max.idx1, mst.max.idx2)]

## Graph-based layouts for Min and Max ST
layout.mst.min <- layout_with_kk(mst)
layout.mst.max <- layout_with_kk(mst.max)

## Data frames for min spanning tree layout
pca.df.mst <- data.frame(X1 = layout.mst.min[, 1], X2 = layout.mst.min[, 2], state = true_state)
edge.df.mst <- data.frame(
    x = layout.mst.min[idx1, 1], y = layout.mst.min[idx1, 2],
    xend = layout.mst.min[idx2, 1], yend = layout.mst.min[idx2, 2],
    weight = mst.weights
)

## Data frames for max spanning tree layout
pca.df.max <- data.frame(X1 = layout.mst.max[, 1], X2 = layout.mst.max[, 2], state = true_state)
edge.df.max <- data.frame(
    x = layout.mst.max[mst.max.idx1, 1], y = layout.mst.max[mst.max.idx1, 2],
    xend = layout.mst.max[mst.max.idx2, 1], yend = layout.mst.max[mst.max.idx2, 2],
    weight = mst.max.weights
)
```

:::::: {.columns}
::: {.column width=.48}

**Minimum Spanning Tree**

```{r fig.width=2.8, fig.height=2.5, echo=F}
ggplot(pca.df.mst, aes(x = X1, y = X2)) +
    geom_segment(data = edge.df.mst, aes(x = x, y = y, xend = xend, yend = yend,
                                         linewidth = weight),
                 color = "grey50", alpha = 0.5) +
    scale_linewidth_continuous(range = c(0.2, 1.5), guide = "none") +
    geom_point(aes(color = state), size = 1, alpha = 0.8) +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(x = "", y = "")
```

:::
::: {.column width=.48}

**Maximum Spanning Tree**

```{r fig.width=2.8, fig.height=2.5, echo=F}
ggplot(pca.df.max, aes(x = X1, y = X2)) +
    geom_segment(data = edge.df.max, aes(x = x, y = y, xend = xend, yend = yend,
                                         linewidth = weight),
                 color = "grey50", alpha = 0.5) +
    scale_linewidth_continuous(range = c(0.2, 1.5), guide = "none") +
    geom_point(aes(color = state), size = 1, alpha = 0.8) +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(x = "", y = "")
```

:::
::::::


## Alternative: k-Nearest Neighbor (kNN) Graph

```{r include=F}
## Build kNN graph (k=5 nearest neighbors)
k <- 5
knn.adj <- matrix(0, n_total, n_total)
for (i in 1:n_total) {
    ## Find k nearest neighbors (excluding self)
    neighbors <- order(D[i, ])[2:(k + 1)]
    knn.adj[i, neighbors] <- 1
}
## Make symmetric (undirected)
knn.adj <- pmax(knn.adj, t(knn.adj))
colnames(knn.adj) <- rownames(knn.adj) <- colnames(X)

## Create igraph object
g.knn <- graph_from_adjacency_matrix(knn.adj, mode = "undirected")

## Get edge list and weights (distances)
knn.edges <- as_edgelist(g.knn)
knn.idx1 <- match(knn.edges[, 1], colnames(X))
knn.idx2 <- match(knn.edges[, 2], colnames(X))
knn.weights <- D[cbind(knn.idx1, knn.idx2)]

## Edge data frame for PCA layout
edge.df.knn <- data.frame(
    x = pca.df$PC1[knn.idx1], y = pca.df$PC2[knn.idx1],
    xend = pca.df$PC1[knn.idx2], yend = pca.df$PC2[knn.idx2],
    weight = knn.weights
)

## Compute force-directed layout for kNN graph
layout.knn <- layout_with_fr(g.knn)
pca.df.knn <- data.frame(X1 = layout.knn[, 1], X2 = layout.knn[, 2], state = true_state)
edge.df.knn.fr <- data.frame(
    x = layout.knn[knn.idx1, 1], y = layout.knn[knn.idx1, 2],
    xend = layout.knn[knn.idx2, 1], yend = layout.knn[knn.idx2, 2],
    weight = knn.weights
)
```

:::::: {.columns}
::: {.column width=.48}

**PCA layout**

```{r fig.width=2.8, fig.height=2.5, echo=F}
ggplot(pca.df, aes(x = PC1, y = PC2)) +
    geom_segment(data = edge.df.knn, aes(x = x, y = y, xend = xend, yend = yend,
                                         linewidth = weight),
                 color = "grey50", alpha = 0.3) +
    scale_linewidth_continuous(range = c(0.1, 1), guide = "none") +
    geom_point(aes(color = state), size = 1, alpha = 0.8) +
    theme_minimal() +
    theme(legend.position = "none")
```

:::
::: {.column width=.48}

**Force-directed layout**

```{r fig.width=2.8, fig.height=2.5, echo=F}
ggplot(pca.df.knn, aes(x = X1, y = X2)) +
    geom_segment(data = edge.df.knn.fr, aes(x = x, y = y, xend = xend, yend = yend,
                                            linewidth = weight),
                 color = "grey50", alpha = 0.3) +
    scale_linewidth_continuous(range = c(0.1, 1), guide = "none") +
    geom_point(aes(color = state), size = 1, alpha = 0.8) +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(x = "", y = "")
```

:::
::::::



## Let's refine our paragraph on the problem definition.

1. We consider each cell as a point along a differentiation trajectory. We assume that \textbf{\color{teal} most genes are affected by the differentiation process}, so cell-to-cell distances are best captured in "principal" component analysis.

2. We represent the relationships between cells as a graph, \textbf{\color{magenta} more precisely, a tree structure reflecting their developmental lineage.}

3. Since gene expressions are \textbf{\color{olive} gradually changed} over the differentiation process, each \textbf{\color{olive} branch of the underlying tree structure} represents distinctive cell types.

4. We annotate cell type labels based on prior knowledge (more work needed).

## More formal definition of the problem

:::::: {.columns}
::: {.column width=.45}

::: {.block}
### Input
* Gene expression $X$ ($m$ genes $\times$ $n$ cells)
* Assumption: a differentiation process (minimum spanning tree)
:::

:::
::: {.column width=.45}

::: {.block}
### Output
* Cell-to-cell MST graph
* Assign each cell vector $\mathbf{x}_{j}$ to label $c_i \in \{1, \ldots, K\}$
:::

:::
::::::

## More specific problem definition of MST

:::::: {.columns}
::: {.column width=.45}

::: {.block}
### Input
Distance graph $G = (V, E, D)$
where

* $V$ = vertices (cells)

* $E$ = edges between cells

* $D$ = distance matrix

:::

:::
::: {.column width=.45}

::: {.block}
### Output
Tree $T$ connecting all cells
:::

:::
::::::

\vfill

:::::: {.columns}
::: {.column width=.45}

::: {.block}
### Objective
Minimize total edge weight (MST)
:::

:::
::: {.column width=.45}

$$
\min_{T} \sum_{(i,j) \in E(T)} D_{ij}
$$

:::
::::::

## Optional: more on identifying branches in the tree

:::::: {.columns}
::: {.column width=.5}

```{r fig.width=3, fig.height=2.5, echo=F}
ggplot(pca.df.kk, aes(x = X1, y = X2)) +
    geom_segment(data = edge.df.kk, aes(x = x, y = y, xend = xend, yend = yend,
                                        linewidth = weight),
                 color = "grey50", alpha = 0.5) +
    scale_linewidth_continuous(range = c(0.2, 1.5), guide = "none") +
    geom_point(aes(color = state), size = 1, alpha = 0.8) +
    theme_void()
```

:::
::: {.column width=.4}

* True "state" is hidden (need to estimate)

* We will revisit how we can do clustering on graphs

* But the following three slides are there for the sake of completeness

:::
::::::


## Optional: Spectral clustering on MST

```{r include=F}
## Compute graph Laplacian for MST
L <- laplacian_matrix(mst, normalized = FALSE)

## Compute eigenvectors (Fiedler vector is 2nd smallest)
eig <- eigen(as.matrix(L), symmetric = TRUE)
## Eigenvalues are in decreasing order, so reverse
eigvals <- rev(eig$values)
eigvecs <- eig$vectors[, ncol(eig$vectors):1]

## Use first k eigenvectors for k clusters
spectral_cluster <- function(k) {
    if (k == 1) return(rep(1, n_total))
    ## Use eigenvectors 2 to k for clustering
    V <- eigvecs[, 2:k, drop = FALSE]
    kmeans(V, centers = k, nstart = 20)$cluster
}

## Data frame with spectral embedding (first 2 non-trivial eigenvectors)
spectral.df <- data.frame(
    V1 = eigvecs[, 2],
    V2 = eigvecs[, 3],
    state = true_state
)
```

:::::: {.columns}
::: {.column width=.48}

**Eigenvalues of Laplacian**

```{r fig.width=2.6, fig.height=2.2, echo=F}
ggplot(data.frame(k = 1:20, eigval = eigvals[1:20]), aes(x = k, y = eigval)) +
    geom_point() + geom_line() +
    theme_minimal() +
    labs(x = "Index", y = "Eigenvalue") +
    geom_vline(xintercept = 3, linetype = "dashed", color = "red")
```

\tiny
Gap at k=3 suggests 3 clusters
\normalsize

:::
::: {.column width=.48}

**Spectral embedding**

```{r fig.width=2.6, fig.height=2.2, echo=F}
ggplot(spectral.df, aes(x = V1, y = V2, color = state)) +
    geom_point(size = 1.5, alpha = 0.8) +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(x = "Eigenvector 2", y = "Eigenvector 3")
```

\tiny
Fiedler vector separates branches
\normalsize

:::
::::::

## Optional: Spectral cut results

```{r include=F}
## Get spectral clusters for k=3
spectral_labels <- factor(spectral_cluster(3))

## Use graph layout
layout.spectral <- layout_with_kk(mst)
pca.df.spectral <- data.frame(
    X1 = layout.spectral[, 1],
    X2 = layout.spectral[, 2],
    true_state = true_state,
    spectral_state = spectral_labels
)
edge.df.spectral <- data.frame(
    x = layout.spectral[idx1, 1], y = layout.spectral[idx1, 2],
    xend = layout.spectral[idx2, 1], yend = layout.spectral[idx2, 2],
    weight = mst.weights
)
```

:::::: {.columns}
::: {.column width=.48}

**True cell states**

```{r fig.width=2.8, fig.height=2.5, echo=F}
ggplot(pca.df.spectral, aes(x = X1, y = X2)) +
    geom_segment(data = edge.df.spectral,
                 aes(x = x, y = y, xend = xend, yend = yend),
                 color = "grey70", alpha = 0.5) +
    geom_point(aes(color = true_state), size = 1.5, alpha = 0.8) +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(x = "PC1", y = "PC2")
```

:::
::: {.column width=.48}

**Spectral clustering (k=3)**

```{r fig.width=2.8, fig.height=2.5, echo=F}
ggplot(pca.df.spectral, aes(x = X1, y = X2)) +
    geom_segment(data = edge.df.spectral,
                 aes(x = x, y = y, xend = xend, yend = yend),
                 color = "grey70", alpha = 0.5) +
    geom_point(aes(color = spectral_state), size = 1.5, alpha = 0.8) +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(x = "PC1", y = "PC2")
```

:::
::::::


# Use probabilities to describe data generation process

## Why Probabilistic (Graphical) Model?

PGM forces us to be explicit about:

\vfill

1. **What we observe** (data)

2. **What we want to learn** (parameters, latent variables)

3. **What assumptions we make** (model structure)

4. **What we don't know** (uncertainty)

\vfill

> "All models are wrong, but some are useful." -- George Box

## Describe your science by stitching together random variables

\Large

Represent joint probability distributions using graphs

\normalsize

\vfill

:::::: {.columns}
::: {.column width=.45}

**Nodes** = Random variables

* Observed (shaded)
* Latent/hidden (open)
* Parameters (dots or small nodes)

:::
::: {.column width=.45}

**Edges** = Dependencies

* Arrow $\rightarrow$ = causal/generative direction
* No arrow = correlation/association

:::
::::::

## `lect-02`: What is statistics? 


\centerline{\includegraphics[width=.55\textwidth]{img/probability-inference.pdf}}

\tiny 

Wasserman, _All of Statistics_ (2003)

\vfill

\normalsize

As long as you can write out the probability...

## 


## Central question: How did we get our data?

\Large

1. **Write down your data generating model**

2. Inference: Fit observed data to the model

3. Validation: How well? How significant?

## Example: the number of mutations in HIV (`lect02`)

:::::: {.columns}
::: {.column width=.55}

Let's graphically represent

$$X \mid \lambda \sim \text{Poisson}(\lambda)$$

where

* $\lambda$: unknown mutation rate parameter (open)

* $X$: observed number of mutations (shaded)

:::
::: {.column width=.35}

\begin{center}
\begin{tikzpicture}[
    node distance=1.5cm,
    observed/.style={circle, draw, fill=gray!30, minimum size=1cm},
    latent/.style={circle, draw, minimum size=1cm},
    param/.style={circle, draw, fill=black, minimum size=0.2cm}
]
\node[latent] (lambda) {$\lambda$};
\node[observed, below=of lambda] (x) {$X$};

\draw[->, thick] (lambda) -- (x);
\end{tikzpicture}
\end{center}

:::
::::::


## Revisit the same question: cell type identification

**Data**: RNA-seq of **genes $\times$ cells**

\large

Let $X_{gi}$ be the expression value of a gene $g$ in a cell $i$.

$$
\mathbf{X} = \begin{pmatrix}
X_{11} & X_{12} & \cdots & X_{1n} \\
X_{21} & X_{22} & \cdots & X_{2n} \\
\vdots & \vdots & \ddots & \vdots \\
X_{m1} & X_{m2} & \cdots & X_{mn}
\end{pmatrix}
=
\left(
\mathbf{x}_{1}, \mathbf{x}_{2}, \ldots, \mathbf{x}_{n}
\right)$$
where
$m$ corresponds to 19k genes.


## The number of mRNA counts in the same way?

:::::: {.columns}
::: {.column width=.55}

Let's graphically represent

$$X_{gi} \mid \lambda \sim \text{Poisson}(\lambda_{gi})$$

where

* $\lambda_{gi}$: unknown transcription rate parameter of a gene $g$ within a cell $i$ (open; latent; parameter)

* $X_{gi}$: the number of mRNA reads of a gene $g$ in a cell $i$ (shaded; observed; data)

:::
::: {.column width=.35}

\begin{center}
\begin{tikzpicture}[
    node distance=1.5cm,
    observed/.style={circle, draw, fill=gray!30, minimum size=1cm},
    latent/.style={circle, draw, minimum size=1cm},
    param/.style={circle, draw, fill=black, minimum size=0.2cm}
]
\node[latent] (lambda) {$\lambda_{gi}$};
\node[observed, below=of lambda] (x) {$X_{gi}$};

\draw[->, thick] (lambda) -- (x);
\end{tikzpicture}
\end{center}

:::
::::::

## Can we elaborate more on the $\lambda_{gi}$ (gene $\times$ cell)?

```{r include = F}
X.pois <- exp(apply(X, 2, scale))
```

```{r fig.width=5, fig.height=2.5, echo=F}
.matshow(pmin(log1p(X.pois), 3))
```

\vfill

What's wrong with having $\lambda_{gi}$ for each $X_{gi}$?

## Let's think of the "generation" of $X$ (or $\lambda$)

HINT (just sorted rows and columns):

```{r include=F}
xx.sorted <- .sort.matrix(pmin(log1p(X.pois), 3))
```

```{r fig.width=5, fig.height=2.5, echo=F}
.matshow(xx.sorted)
```

## Write out data generating scheme

:::::: {.columns}
::: {.column width=.45}

\large

\onslide<1->{Of the $K$ cell types:

$$Z_{ik} = 1 \iff \text{cell $i$ in type $k$}$$

}

\onslide<2->{
For each cell type $k$, a gene $g$ is expressed with the rate

$$\lambda_{gk}$$

*Note*: do we have more $\lambda_{gi}$'s or $\lambda_{gk}$'s?
}

\normalsize

:::
::: {.column width=.45}

\large

\onslide<3->{
Given the cell type assignment $Z_{ik}$ and type specific expression rate parameters $\lambda$, we have

$$X_{gi} | Z_{ik}=1 \sim \mathsf{Poisson}(\lambda_{gk})$$
}

\normalsize

:::
::::::

## Or we could simply draw it out

\large

\begin{center}
\begin{tikzpicture}[
    node distance=1.2cm,
    observed/.style={circle, draw, fill=gray!30, minimum size=0.9cm},
    latent/.style={circle, draw, minimum size=0.9cm}
]
\node[latent] (z) {$Z_{ik}$};
\node[latent, right=1.5cm of z] (lambda) {$\lambda_{gk}$};
\node[observed, below=1.2cm of $(z)!0.5!(lambda)$] (x) {$X_{gi}$};

\draw[->, thick] (z) -- (x);
\draw[->, thick] (lambda) -- (x);
\end{tikzpicture}
\end{center}

What's next?

## Where do these $Z$ and $\lambda$ variables correspond to?



:::::: {.columns}
::: {.column width=.1}

:::
::: {.column width=.85}

\onslide<2>{\large $Z_{ik} \to \to \to$}

:::
::::::

:::::: {.columns}
::: {.column width=.1}

\onslide<2>{\large
\begin{align*}
\lambda_{gk} \\
\downarrow \\
\downarrow
\end{align*}
}

:::
::: {.column width=.85}

```{r fig.width=5, fig.height=2.5, echo=F}
.matshow(xx.sorted)
```

:::
::::::



## In many cases we have assignment $Z$ observed/given

\large

\begin{center}
\begin{tikzpicture}[
    node distance=1.2cm,
    observed/.style={circle, draw, fill=gray!30, minimum size=0.9cm},
    latent/.style={circle, draw, minimum size=0.9cm}
]
\node[observed] (z) {$Z_{ik}$};
\node[latent, right=1.5cm of z] (lambda) {$\lambda_{gk}$};
\node[observed, below=1.2cm of $(z)!0.5!(lambda)$] (x) {$X_{gi}$};

\draw[->, thick] (z) -- (x);
\draw[->, thick] (lambda) -- (x);
\end{tikzpicture}
\end{center}

What's next (`lect04`)? 

$$\lambda_{g1} = \lambda_{g2} \quad 
\text{or} \quad \lambda_{g1} \neq \lambda_{g2}$$

## As long as we have PGM...

\large

* Estimate the unknown parameter, such as $\lambda$

* Check how "well" generated data, such as $X$, can explain observed $X$

## Summary

\large

**Two approaches to formulate computational problems:**

\normalsize

:::::: {.columns}
::: {.column width=.45}

1. **Optimization approach**
   * Define input, output, metric
   * The art of abstraction
   * Example: MST for cell trajectories

:::
::: {.column width=.45}

2. **Probabilistic modeling**
   * Write down generative model
   * Use graphical models (PGM)
   * Example: Poisson model for gene expression

:::
::::::

\vfill

\large

**Key takeaway:** Write down your problem clearly before coding!


