---
title: "Biological Networks"
author: "Yongjin Park"
date: "`r format(Sys.time(), '%d %B %Y')`"
classoption: "aspectratio=169"
fontsize: 12pt
bibliography: network.bib
output:
    beamer_presentation:
        theme: Madrid
        keep_md: true
        keep_tex: true
        latex_engine: pdflatex
        slide_level: 2
header-includes:
  - \usepackage{cancel}
  - \usepackage{tikz}
  - \usetikzlibrary{shapes.geometric,arrows,positioning}
  - \definecolor{UBCblue}{rgb}{0.04706, 0.13725, 0.26667}
  - \usecolortheme[named=UBCblue]{structure}
  - \setbeamertemplate{frametitle}{\color{UBCblue}\bfseries\insertframetitle\par\vskip-6pt\hrulefill}
  - \setbeamercolor{title in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{section in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{author in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{date in head/foot}{bg=white,fg=gray}
  - \setbeamertemplate{page number in head/foot}{}
  - \setbeamertemplate{frame numbering}[none]
  - \setbeamercolor{alerted text}{bg=yellow}
  - \AtBeginSection[]{\begin{frame}\frametitle{Today's lecture}{\Large\tableofcontents[currentsection]}\end{frame}}
  - |
    \makeatletter
    \def\ps@titlepage{%
      \setbeamertemplate{footline}{}
    }
    \addtobeamertemplate{title page}{\thispagestyle{titlepage}}{}
    \makeatother
---


```{r setup, include=FALSE}
source("Setup.R")
fig.dir <- "../Fig/biological-networks/"
dir.create(fig.dir, showWarnings=FALSE)
setup.env(fig.dir)
dir.create("../data", showWarnings=FALSE)

library(data.table)
library(tidyverse)
library(patchwork)
library(matrixTests)
library(igraph)

theme_set(theme_minimal())

## Helper functions for network visualization
find.nice.layout <- function(G, layout.file) {
    if(!file.exists(layout.file)){
        layout.dt <- layout_nicely(G) %>% as.data.table()
        saveRDS(layout.dt, layout.file)
    }
    readRDS(layout.file)
}

apply.layout.dt <- function(G, .name){
    .out.file <- str_c(fig.dir,"/",.name,".rds")
    .out <- find.nice.layout(G, .out.file) %>%
        dplyr::mutate(v = 1:dplyr::n())
    as_long_data_frame(G) %>%
        as.data.table %>%
        merge(.out, by.x= "from", by.y="v") %>%
        merge(.out, by.x= "to", by.y="v", suffixes=c(".from",".to"))
}

plot.layout.dt <- function(.dt, .palette="Paired", .palette.e = "Set2",
                           .esize = .5, .vsize = 5, .add.lab=TRUE){
    if(!("k.from" %in% colnames(.dt))) .dt[, k.from := 1]
    if(!("k.to" %in% colnames(.dt))) .dt[, k.to := 1]
    .rename <- function(xx) { colnames(xx) <- c("v","V1","V2","k"); xx }
    .dt.v <- rbind(.dt[, .(`from`, `V1.from`, `V2.from`, `k.from`)] %>% .rename(),
                   .dt[, .(`to`, `V1.to`, `V2.to`, `k.to`)] %>% .rename()) %>%
        dplyr::mutate(k = as.factor(k)) %>% unique
    .aes.e <- aes(V1.from, V2.from, xend=V1.to, yend=V2.to)
    if("k.edge" %in% colnames(.dt)) {
        .aes.e <- aes(V1.from, V2.from, xend=V1.to, yend=V2.to, colour=as.factor(k.edge))
    }
    plt <- ggplot() + theme_void()
    if("k.edge" %in% colnames(.dt)) {
        plt <- plt + geom_segment(.aes.e, data = .dt, size = .esize) +
            scale_colour_brewer(palette=.palette.e, na.value="grey", guide="none")
    } else {
        plt <- plt + geom_segment(.aes.e, data = .dt, size = .esize, colour="gray")
    }
    plt <- plt +
        geom_point(aes(V1, V2, fill=k), data = .dt.v, pch=21, stroke=.1, size = .vsize) +
        scale_fill_brewer(palette = .palette, guide="none")
    if(.add.lab){ plt <- plt + geom_text(aes(V1, V2, label=v), data = .dt.v, size = 3) }
    return(plt)
}
```

## Why do we study biological networks?

- Genes and proteins function through **interactions** with each other
- Understanding interactions helps us:
  - Map cellular pathways and processes
  - Identify disease mechanisms
  - Predict gene function
  - Design targeted therapies

## Why networks matter in cancer biology

Challenge in cancer therapy:

- Tumour cells develop resistance to single-target drugs
- Genetic redundancy provides backup pathways
- Need to hit multiple targets simultaneously

Network solution:

- Synthetic lethality: exploit genetic interactions
- Target pairs of genes that are lethal together
- Example: BRCA1/2 mutations + PARP inhibitors

## Networks of oncogene and inhibition intertwined

:::::: {.columns}
::: {.column width=.35}

\includegraphics[width=\textwidth]{img/1_how_to_build/cancer_genetic_interaction_review-1.pdf}

* Oncogenes may regulate other genes

* Which one should be suppressed?

* Which one can be?

:::
::: {.column width=.55}

\includegraphics[width=\textwidth]{img/1_how_to_build/cancer_genetic_interaction_review-2.pdf}

:::
::::::

\tiny
\flushright

[@Kaelin2005-sl]


## Cancer treatment may need to deal with multiple genes...

\centering
\includegraphics[width=0.8\textwidth]{img/1_how_to_build/cancer_genetic_interaction_review-3.pdf}

\tiny
\flushright

[@Kaelin2005-sl]



# How do we measure protein/gene interaction networks?

## Y2H to capture protein-protein interactions

:::::: {.columns}
::: {.column width=.35}

\centering
\includegraphics[height=0.85\textheight]{img/1_how_to_build/ppi_fields_song.pdf}

:::
::: {.column width=.6}

* Yeast Two Hybrid

* GAL4-lacZ reporter system

* Transcription factors have two separable domains: DNA-binding (BD) and activation (AD)

* Fuse BD to "bait" protein, AD to "prey" protein

* If they interact $\implies$ reconstitute functional TF $\implies$ reporter gene ON

$${}$$

\footnotesize
[@Fields1989-yh]

:::
::::::


## Synthesize TF binding sites with reporter genes

\centering
\begin{tikzpicture}[scale=1.4, every node/.style={font=\footnotesize}]

% DNA alone
\draw[thick] (0, 0) -- (8, 0);
\draw[thick] (0, 0.4) -- (8, 0.4);
\draw[fill=green!20, rounded corners] (1, 0) rectangle (2, 0.4);
\node[anchor=north, font=\scriptsize] at (1.5, -0.1) {Binding site};
\draw[->, thick] (2.5, 0.2) -- (3.5, 0.2);
\draw[fill=yellow!40, rounded corners] (4, -0.2) rectangle (7, 0.6);
\node at (5.5, 0.2) {Reporter gene};

% Separated domains - floating away - rectangles
\draw[fill=blue!30, draw, rounded corners] (0.9, 1.2) rectangle (2.1, 1.9);
\node[font=\scriptsize] at (1.5, 1.65) {DNA-binding};
\node[font=\scriptsize] at (1.5, 1.45) {domain};

\draw[fill=red!30, draw, rounded corners] (5.4, 1.5) rectangle (6.6, 2.2);
\node[font=\scriptsize] at (6, 1.95) {Activation};
\node[font=\scriptsize] at (6, 1.75) {domain};

\end{tikzpicture}

## Create fusion proteins

\centering
\begin{tikzpicture}[scale=1.2, every node/.style={font=\footnotesize}]

% Fusion protein 1 - Bait on top of DNA-binding
\node[anchor=south] at (2, 3.2) {Fusion protein 1};
\node[ellipse, fill=green!50, draw, minimum width=2.5cm, minimum height=1cm] at (2, 2.3) {Bait protein};
\draw[fill=blue!30, draw, rounded corners] (1.3, 1.3) rectangle (2.7, 1.9);
\node[font=\scriptsize] at (2, 1.7) {DNA-binding};
\node[font=\scriptsize] at (2, 1.5) {domain};

% Fusion protein 2 - Prey on top of Activation
\node[anchor=south] at (7, 3.2) {Fusion protein 2};
\node[ellipse, fill=orange!50, draw, minimum width=2.5cm, minimum height=1cm] at (7, 2.3) {Prey protein};
\draw[fill=red!30, draw, rounded corners] (6.3, 1.3) rectangle (7.7, 1.9);
\node[font=\scriptsize] at (7, 1.7) {Activation};
\node[font=\scriptsize] at (7, 1.5) {domain};

% DNA with binding site and reporter
\draw[thick] (0.5, 0.2) -- (9, 0.2);
\draw[thick] (0.5, 0.6) -- (9, 0.6);
\draw[fill=green!20, rounded corners] (1, 0.2) rectangle (2, 0.6);
\node[anchor=south, font=\scriptsize] at (1.5, 0.7) {Binding site};
\draw[->, thick] (2.5, 0.4) -- (3.5, 0.4);
\node[anchor=south, font=\scriptsize] at (6, 0.7) {Reporter gene};
\draw[fill=yellow!40, rounded corners] (4.5, 0) rectangle (7.5, 0.8);
\node at (6, 0.4) {Reporter};

\end{tikzpicture}

## Negative: No interaction between bait and prey

\centering
\begin{tikzpicture}[scale=1.2, every node/.style={font=\footnotesize}]

% DNA with binding site and reporter
\draw[thick] (0, 0) -- (7, 0);
\draw[thick] (0, 0.4) -- (7, 0.4);
\draw[fill=green!20, rounded corners] (0.5, 0) rectangle (1.5, 0.4);
\draw[->, thick] (2, 0.2) -- (3, 0.2);
\draw[fill=yellow!40, rounded corners] (3.5, -0.2) rectangle (6.5, 0.6);
\node at (5, 0.2) {Reporter gene};

% Fusion 1 bound to DNA - Bait on top of DNA-binding
\node[ellipse, fill=green!50, draw, minimum width=2.2cm, minimum height=0.9cm] at (1.5, 1.6) {Bait protein};
\draw[fill=blue!30, draw, rounded corners] (0.8, 0.5) rectangle (2.2, 1.1);
\node[font=\scriptsize] at (1.5, 0.9) {DNA-binding};
\node[font=\scriptsize] at (1.5, 0.7) {domain};

% Fusion 2 floating away - Prey on top of Activation
\node[ellipse, fill=orange!50, draw, minimum width=2.2cm, minimum height=0.9cm] at (9, 1.8) {Prey protein};
\draw[fill=red!30, draw, rounded corners] (8.3, 0.9) rectangle (9.7, 1.5);
\node[font=\scriptsize] at (9, 1.3) {Activation};
\node[font=\scriptsize] at (9, 1.1) {domain};

% No interaction
\node[red, font=\normalsize] at (9, 0) {No interaction};
\draw[->, red, very thick] (7.5, -0.5) -- (8.5, -0.5);
\node[red, font=\normalsize] at (10, -0.5) {Reporter OFF};

\end{tikzpicture}

## Positive: Bait-prey interaction reconstitutes TF

\centering
\begin{tikzpicture}[scale=1.2, every node/.style={font=\footnotesize}]

% DNA with binding site and reporter
\draw[thick] (0, 0) -- (7, 0);
\draw[thick] (0, 0.4) -- (7, 0.4);
\draw[fill=green!20, rounded corners] (0.5, 0) rectangle (1.5, 0.4);
\draw[->, thick] (2, 0.2) -- (3, 0.2);
\draw[fill=yellow!40, rounded corners] (3.5, -0.2) rectangle (6.5, 0.6);
\node at (5, 0.2) {Reporter gene};

% Fusion 1 bound to DNA - Bait on top of DNA-binding
\node[ellipse, fill=green!50, draw, minimum width=2.2cm, minimum height=0.9cm] at (1.5, 1.6) {Bait protein};
\draw[fill=blue!30, draw, rounded corners] (0.8, 0.5) rectangle (2.2, 1.1);
\node[font=\scriptsize] at (1.5, 0.9) {DNA-binding};
\node[font=\scriptsize] at (1.5, 0.7) {domain};

% Fusion 2 bound to bait - Prey on top of Activation
\node[ellipse, fill=orange!50, draw, minimum width=2.2cm, minimum height=0.9cm] at (3.5, 1.6) {Prey protein};
\draw[fill=red!30, draw, rounded corners] (2.8, 0.5) rectangle (4.2, 1.1);
\node[font=\scriptsize] at (3.5, 0.9) {Activation};
\node[font=\scriptsize] at (3.5, 0.7) {domain};

% Interaction indicator between Bait and Prey
\draw[<->, very thick, dashed, green!70!black] (2.4, 1.6) -- (2.6, 1.6);
\node[anchor=south, green!70!black, font=\small] at (2.5, 2.2) {Bait-Prey interaction!};

% Transcription happening
\draw[->, ultra thick, green!70!black] (5, -0.5) -- (5, -1.2);
\node[anchor=north, green!70!black] at (5, -1.3) {Transcription};

% Result
\draw[->, ultra thick, green!70!black] (8, 0.6) -- (9.5, 0.6);
\node[fill=green!20, rounded corners, font=\normalsize] at (11, 0.6) {Reporter ON};

\end{tikzpicture}


## First genome-wide Y2H screen

:::::: {.columns}
::: {.column width=.55}

\only<1>{
\includegraphics[width=\textwidth]{img/1_how_to_build/ppi_uetz_1.pdf}
}

\only<2>{
\centering
\includegraphics[width=\textwidth]{img/1_how_to_build/ppi_uetz_2.pdf}
}

:::
::: {.column width=.4}

* Array of ~6000 prey proteins (AD fusions)

* Each bait (BD fusion) screened against entire array

* 192 bait proteins tested

* Growth on selective media = positive interaction

:::
::::::

\footnotesize

[@Uetz2000-ppi]

## Affinity purification–mass spectrometry (AP-MS)

<!-- - Y2H finds pairwise interactions -->
<!-- - But proteins often work in larger complexes -->
<!-- - Tag a protein → purify it + everything bound to it -->
<!-- - Mass spec identifies co-purified proteins -->
<!-- - Reveals multi-protein assemblies and stoichiometry -->
<!-- - BioPlex: large-scale human AP-MS interactome -->

\centerline{
\includegraphics[width=.6\textwidth]{img/1_how_to_build/ppi_apms_gavin_1.pdf}
}

\vspace{0.3cm}

TAP tag (Tandem Affinity Purification):

* CBP (Calmodulin Binding Peptide) + Protein A (binds IgG).

* Two-step purification $\to$ reduce background

\footnotesize
\flushright

[@Gavin2002-apms]


## AP-MS workflow [@Gavin2002-apms]

\centering
\includegraphics[height=0.8\textheight]{img/1_how_to_build/ppi_apms_gavin-2.pdf}

## AP-MS network analysis

:::::: {.columns}
::: {.column width=.45}

**Purification**

\includegraphics[width=\textwidth]{img/1_how_to_build/ppi_apms_gavin-3.pdf}

:::
::: {.column width=.45}

**Mass Spec result:**

\includegraphics[width=\textwidth]{img/1_how_to_build/ppi_apms_gavin-4.pdf}

:::
::::::

\footnotesize

[@Gavin2002-apms]

## Y2H vs. AP-MS: True or False?

**Question 1:** Y2H detects only direct binary protein-protein interactions, while AP-MS can identify both direct and indirect interactions within protein complexes.

\pause

\textcolor{teal}{TRUE} - Y2H requires direct physical contact between bait and prey to reconstitute the transcription factor. AP-MS co-purifies everything bound to the tagged protein, including indirect interactions mediated by other proteins in the complex.

## Y2H vs. AP-MS: True or False?

**Question 2:** Y2H is an in vivo method (interactions measured in living cells), while AP-MS is an in vitro method (interactions measured after cell lysis).

\pause

\textcolor{teal}{TRUE} - Y2H detects interactions in living yeast cells (in vivo, though in artificial nuclear context). AP-MS requires cell lysis before purification and mass spec analysis (in vitro).

## AP-MS Challenge: True or False?

**Question 3:** In AP-MS, any protein detected by mass spectrometry is definitely part of the protein complex.

\pause

\textcolor{red}{FALSE} - Need to distinguish true interactors from contaminants. Solutions: (1) Compare to negative controls, (2) Quantitative mass spec (abundance ratios), (3) Reciprocal tagging experiments, (4) Statistical scoring (e.g., hypergeometric p-value).


## Genetic interaction networks — synthetic lethality

<!-- - Y2H and AP-MS measure physical interactions -->
<!--  -->
<!-- - Synthetic lethality: double mutant lethal, singles viable -->
<!-- - Indicates genes buffer each other (same or parallel pathways) -->
<!-- -  (~23M combinations in yeast) -->
<!-- - Basis for cancer synthetic lethality therapies -->

:::::: {.columns}
::: {.column width=.45}

${}$

\includegraphics[width=\textwidth]{img/1_how_to_build/SL_tong-1.pdf}

- Genetic interactions reveal functional relationships

- Most salient phenotype in yeast?

- How many cells within each strain?

${}$

\footnotesize

[@Tong2001-mz]


:::
::: {.column width=.4}

\includegraphics[height=.8\textheight]{img/1_how_to_build/SL_tong-2.pdf}

:::
::::::

## Synthetic Gene Array creates double mutants

:::::: {.columns}
::: {.column width=.45}

\includegraphics[width=\textwidth]{img/1_how_to_build/SL_tong-3-a.pdf}

\begin{itemize}
\item<2-> Query mutation (bni1$\Delta$::natR) crossed against deletion array.
\item<2-> Smaller colonies = synthetic lethal/sick
\end{itemize}

:::
::: {.column width=.45}

\onslide<2->{
\includegraphics[width=\textwidth]{img/1_how_to_build/SL_tong-3-b.pdf}
}

\begin{itemize}
\item<3> Tetrad analysis confirmation.
\item<3> T=tetrad, NPD=non-parental ditype, PD=parental ditype.
\item<3> Yellow arrows = double mutants with reduced growth.
\end{itemize}

:::
::::::

\vfill

\footnotesize

[@Tong2001-mz]

## SGA reveals functional hub-and-spoke patterns

:::::: {.columns}
::: {.column width=.6}

\includegraphics[height=.8\textheight]{img/1_how_to_build/SL_tong-4.pdf}

:::
::: {.column width=.4}

$${}$$

* Hub-and-spoke architecture

* Genes in same pathway cluster together

* Hubs (e.g., BNI1, ARP2, RAD27) are functionally critical

* Predict function of unknown genes by their neighbours

$${}$$

\footnotesize

[@Tong2001-mz]

:::
::::::

## How should we think of synthetic lethality?

:::::: {.columns}
::: {.column width=.35}

\onslide<1->{
\includegraphics[width=\textwidth]{img/1_how_to_build/ye_bader_synthetic_lethal-1.pdf}
}

:::
::: {.column width=.55}

\onslide<2>{
\includegraphics[width=\textwidth]{img/1_how_to_build/ye_bader_synthetic_lethal-2.pdf}
}

:::
::::::

\footnotesize

[@Ye2005-ps]


# How can we use network data?

## Network is a data set

Graph $G = (V, E)$:

* **Vertices** (nodes): $V = \{v_1, v_2, \ldots, v_n\}$ — genes/proteins
* **Edges**: $E = \{(v_i, v_j)\}$ — interactions between genes
* **Adjacency matrix**: $A_{ij} = 1$ if edge exists, 0 otherwise
* **Degree**: $d_v = \sum_u A_{uv}$ — number of neighbours

Example:

* BioGRID: $n \approx 18{,}000$ genes, $m \approx 500{,}000$ interactions
* Sparse: most genes connect to few others (not all-to-all)
* Scale-free: few hubs with many connections, most with few

## We can use `igraph` in `R`

```{r size = "large", echo = T}
library(igraph)
```

* We can use `igraph` in `R` or `Python` or `C++`.

## Let's take a look at some classic examples

Download some from Mark Newman's website:

\vspace{10pt}

```{r size = "tiny", echo=TRUE}
#' @param .src www location
#' @param .dir downloading local directory
#' @param .zip zip file name
#' @param .tgt target gml file name
.read.online <- function(.src,
                         .dir="../data/network/",
                         .zip = str_c(.dir,basename(.src)),
                         .tgt = str_replace(.zip, "zip$", "gml")){
    if.needed(.tgt, {
        .gml <- basename(.tgt)                     # remove the dir name
        download.file(.src, .zip)                  # download from the web
        unzip(.zip, files=.gml, exdir = .dir)      # unzip
    })
    return(igraph::read_graph(.tgt, format="gml")) # read them into memory
}
```

```{r size = "scriptsize", eval=TRUE, echo=TRUE}
karate <- .read.online("http://www-personal.umich.edu/~mejn/netdata/karate.zip")
polblog <- .read.online("http://www-personal.umich.edu/~mejn/netdata/polblogs.zip")
netsci <- .read.online("http://www-personal.umich.edu/~mejn/netdata/netscience.zip")
football <- .read.online("http://www-personal.umich.edu/~mejn/netdata/football.zip")
condmat <- .read.online("http://www-personal.umich.edu/~mejn/netdata/cond-mat-2005.zip")
```

## Handy functions for visualizing a network

We want to match graph vertices with the layout coordinates found by
`igraph::layout_nicely` and make it easy to show in `ggplot` or something else.

:::::: {.columns}
::: {.column width=.5}

```{r echo=TRUE, eval=FALSE}
#' @param G igraph object
#' @param layout.file local file
find.nice.layout <-
    function(G, layout.file) {

    if(!file.exists(layout.file)){
        layout.dt <-
            layout_nicely(G) %>%
            as.data.table()
        saveRDS(layout.dt, layout.file)
    }
    readRDS(layout.file)
}
```

:::
::: {.column width=.5}

```{r echo=TRUE, eval=FALSE}
#' @param G igraph object
#' @param .name file header
apply.layout.dt <- function(G, .name){
  .out.file <-
    str_c(fig.dir,"/",.name,".rds")

  .out <-
    find.nice.layout(G, .out.file) %>%
    dplyr::mutate(v = 1:dplyr::n()) # index

  as_long_data_frame(G) %>%
    as.data.table %>%
    merge(.out, by.x= "from", by.y="v") %>%
    merge(.out, by.x= "to", by.y="v",
        suffixes=c(".from",".to"))
}
```

:::
::::::

## Plotting a graph with points and segments in `ggplot`

```{r echo=TRUE, size = "tiny", eval=FALSE}
plot.layout.dt <- function(.dt, .palette="Paired", .palette.e = "Set2", .esize = .5, .vsize = 5, .add.lab=TRUE){
  if(!("k.from" %in% colnames(.dt))) .dt[, k.from := 1]                    # add membership
  if(!("k.to" %in% colnames(.dt))) .dt[, k.to := 1]                        # add membership
  .rename <- function(xx) { colnames(xx) <- c("v","V1","V2","k"); xx }     # Find a consolidated
  .dt.v <-                                                                 # set of vertices
    rbind(.dt[, .(`from`, `V1.from`, `V2.from`, `k.from`)] %>% .rename(),
        .dt[, .(`to`, `V1.to`, `V2.to`, `k.to`)] %>% .rename()) %>%
    dplyr::mutate(k = as.factor(k)) %>% unique

  .aes.e <- aes(V1.from, V2.from, xend=V1.to, yend=V2.to)   # edge colours
  if("k.edge" %in% colnames(.dt)) {
    .aes.e <- aes(V1.from, V2.from, xend=V1.to, yend=V2.to, colour=as.factor(k.edge))
  }
  plt <- ggplot() + theme_void()                                           # show no axis
  if("k.edge" %in% colnames(.dt)) {
    plt <- plt +
      geom_segment(.aes.e, data = .dt, size = .esize) +                    # draw edges
      scale_colour_brewer(palette=.palette.e, na.value="grey", guide="none")
  } else {
    plt <- plt +
      geom_segment(.aes.e, data = .dt, size = .esize, colour="gray")       # draw edges
  }
  plt <- plt +
    geom_point(aes(V1, V2, fill=k), data = .dt.v, pch=21, stroke=.1, size = .vsize) + # vertices
    scale_fill_brewer(palette = .palette, guide="none")                    # vertex colouring

  if(.add.lab){ plt <- plt + geom_text(aes(V1, V2, label=v), data = .dt.v, size = 3) }
  return(plt)
}
```

## Social network examples

:::::: {.columns}
::: {.column width=.5}

```{r echo = FALSE, fig.width=2, fig.height=1.75}
.vv <- V(polblog)[degree(polblog) >= 3]
polblog.sub <- induced_subgraph(polblog, .vv)
apply.layout.dt(polblog.sub, "polblog.sub") %>%
    plot.layout.dt(.add.lab=FALSE, .vsize=2, .esize=.2)
```

[@Adamic2005-pb]

:::
::: {.column width=.5}

```{r echo = FALSE, fig.width=2, fig.height=1.75}
.comp <- components(netsci)
.vv <- V(netsci)[.comp$membership == which.max(.comp$csize)]
netsci.sub <- induced_subgraph(netsci, .vv)

apply.layout.dt(netsci.sub, "netsci.sub") %>%
    plot.layout.dt(.add.lab=FALSE, .vsize=2, .esize=.2)
```

* Coauthorship network of scientists working on network theory and experiment

* Showing only the largest component

[@Newman2006-ev]

:::
::::::

## More examples

:::::: {.columns}
::: {.column width=.5}

```{r echo = FALSE, fig.width=2, fig.height=1.6}
.vv <- V(football)[degree(football) >= 3]
football.sub <- induced_subgraph(football, .vv)
apply.layout.dt(football.sub, "football.sub") %>%
    plot.layout.dt(.add.lab=FALSE, .vsize=2, .esize=.2)
```

* Network of American football games between Division IA colleges during regular season Fall 2000.

[@Girvan2002-cs]

:::
::: {.column width=.5}

```{r echo = FALSE, fig.width=2, fig.height=1.8}
G <- condmat

apply.degree.cutoff <- function(G, .cutoff = 5) {
    .sub <- G
    .rr <- sum(degree(.sub) < .cutoff)
    while(.rr > 0){
        vv <- V(.sub)
        .vv <- vv[degree(.sub) >= .cutoff]
        .sub <- induced_subgraph(.sub, .vv)
        .rr <- sum(degree(.sub) < .cutoff)
    }
    return(.sub)
}

condmat.sub <- apply.degree.cutoff(G)
apply.layout.dt(condmat.sub, "condmat.sub") %>%
    plot.layout.dt(.add.lab=FALSE, .vsize=.3)
```

[@Newman2001-sc]

* Showing densely-connected part (iterative degree cutoff >= 5)

:::
::::::

```{r include=FALSE}
db.url <- "http://www.interactome-atlas.org"
ppi.net.file <- "../data/HI-union.tsv.gz"
.ppi.rds <- "../data/HI-union.G.rds"
if(!file.exists(ppi.net.file)){
    .file <- str_remove(ppi.net.file, "[.]gz$")
    .url <- str_c(db.url,"/data/HI-union.tsv")
    download.file(.url, destfile = .file)
    R.utils::gzip(.file)
}
if.needed(.ppi.rds, {
    ppi.pairs <- fread(ppi.net.file, header=FALSE, col.names=c("v1","v2")) %>%
        filter(v1 != v2) %>% as.matrix
    saveRDS(graph_from_edgelist(ppi.pairs, directed = FALSE), .ppi.rds)
})
G.ppi <- readRDS(.ppi.rds)
n <- length(V(G.ppi))
m <- length(E(G.ppi))
```

## Human Reference Protein-Protein interaction map

:::::: {.columns}
::: {.column width=.5}

* Download the data from `interactome-atlas.org`

```{r echo=TRUE, eval=FALSE}
db.url <- "http://www.interactome-atlas.org"
ppi.net.file <- "../data/HI-union.tsv.gz"
if.needed(ppi.net.file,{
    .file <- str_remove(ppi.net.file, "[.]gz$")
    .url <- str_c(db.url,"/data/HI-union.tsv")
    download.file(.url, destfile = .file)
    R.utils::gzip(.file)
})
```

* Build a network from a list of interacting pairs

```{r echo=TRUE, eval=FALSE}
ppi.pairs <-
  fread(ppi.net.file, header=FALSE, col.names=c("v1","v2")) %>%
  filter(v1 != v2) %>% as.matrix
G.ppi <- graph_from_edgelist(ppi.pairs, directed = FALSE)
```

:::
::: {.column width=.45}

* Vertices: genes (or the resulting proteins)

* Edges: two or multiple genes work together (could have many different meanings)

```{r echo=TRUE, eval=FALSE}
n <- length(V(G.ppi))
m <- length(E(G.ppi))
```

* n = `r n` vertices

* m = `r m` edges

:::
::::::

## Human Reference Protein-Protein interaction map

:::::: {.columns}
::: {.column width=.5}

```{r echo=FALSE, fig.width=3, fig.height=3}
apply.layout.dt(G.ppi, "ppi") %>% plot.layout.dt(.add.lab=FALSE, .vsize=1, .esize=.1)
```

:::
::: {.column width=.45}

* Vertices: genes (or the resulting proteins)

* Edges: two or multiple genes work together (could have many different meanings)

```{r}
n <- length(V(G.ppi))
m <- length(E(G.ppi))
```

* n = `r n` vertices

* m = `r m` edges

Sources of biological networks:

* High-throughput experiments

* Co-expression analysis

* Literature-based curation

[@Luck2020-hu; @Rual2005-ls]

:::
::::::

## Degree distribution (How many friends?)

::: {.block}

### Vertex degree

For each vertex $v \in V(G)$: $d_{v} = \sum_{u} A_{uv}$

:::

* It is simple to calculate in `igraph`:

```{r}
deg.dt <- degree(G.ppi) %>%
    (function(.d)
        data.table(d=.d, v = names(.d))
    )
```

* Degree distributions may teach us the underlying generative scheme

* Power-law (almost linear slope in a log-log plot) may indicate a "rich-get-richer" phenomenon.

* The majority with $< 10$ neighbours vs. hub nodes with $> 100$ neighbours

* Poisson distribution if edges are drawn uniformly at random

## Degree distribution (How many friends?)

A handy one-liner:

```{r}
.dt <- deg.dt[, .(count=.N), by=.(d)]
```

Log-log plot:

```{r fig.width=3, fig.height=2, echo=FALSE}
.gg.plot(.dt, aes(x=`d`, y=`count`)) +
    geom_point(stroke=0) +
    scale_x_log10() +
    scale_y_log10()
```

## Connected components and a giant component

:::::: {.columns}
::: {.column width=.6}

```{r echo=FALSE, fig.width=3, fig.height=3}
apply.layout.dt(G.ppi, "ppi") %>% plot.layout.dt(.add.lab=FALSE, .vsize=1, .esize=.1)
```

:::
::: {.column width=.4}

A (connected) component = a set of vertices reachable by hopping
through edges.

```{r}
.comp <- components(G.ppi)
```

Found `r .comp$no` connected components

```{r echo = FALSE, results="asis"}
.dt <- data.table(n=.comp$csize);
knitr::kable(.dt[, .(count=.N), by=.(`n`)])
```

* Can we take the giant one?

```{r}
k <- which.max(.comp$csize)
.sub <- names(.comp$membership)[
    .comp$membership==k]
G.ppi.sub <- induced_subgraph(G.ppi, .sub)
```

:::
::::::

## Only the giant component in PPI

:::::: {.columns}
::: {.column width=.64}

```{r echo=FALSE, fig.width=4, fig.height=3}
apply.layout.dt(G.ppi.sub, "ppi-giant") %>% plot.layout.dt(.add.lab=FALSE, .vsize=1, .esize=.1)
```

:::
::: {.column width=.35}

* $n(G)$: `r length(V(G.ppi.sub))` (vs. `r length(V(G.ppi))`)

* $m(G)$: `r length(E(G.ppi.sub))` (vs. `r length(E(G.ppi))`)

* Degree distribution:

```{r echo=FALSE, fig.width=2, fig.height=1.75}
.dt <- data.table(d=degree(G.ppi.sub))
.dt <- .dt[, .(count = .N), by = .(d)]
.gg.plot(.dt, aes(x=d, y=count)) + geom_point(stroke=0) + scale_x_log10() + scale_y_log10()
```

:::
::::::

## Iterative degree cutoff to remove dangling vertices

:::::: {.columns}
::: {.column width=.5}

We may consider this as a QC step:

```{r}
degree.cutoff <- function(G, .cutoff = 3) {
    G.sub <- G
    n.remove <- sum(degree(G.sub) < .cutoff)
    while(n.remove > 0){
        vv <- V(G.sub)
        .retain <- vv[degree(G.sub) >= .cutoff]
        G.sub <- induced_subgraph(G.sub, .retain)
        n.remove <- sum(degree(G.sub) < .cutoff)
    }
    return(G.sub)
}
G.dc <- degree.cutoff(G.ppi.sub, 3)
```

Comparison with the raw PPI data:

* $n(G)$: `r length(V(G.dc))` (vs. `r length(V(G.ppi))`)

* $m(G)$: `r length(E(G.dc))` (vs. `r length(E(G.ppi))`)

:::
::: {.column width=.5}

Will it look nicer?

```{r echo=FALSE, fig.width=2, fig.height=1.5}
apply.layout.dt(G.dc, "ppi-dc") %>% plot.layout.dt(.add.lab=FALSE, .vsize=1, .esize=.1)
```

:::
::::::


## Iterative degree cutoff to remove dangling vertices

:::::: {.columns}
::: {.column width=.5}

Degree distribution:

```{r echo=FALSE, fig.width=2, fig.height=1.75}
.dt <- data.table(d=degree(G.dc))
.dt <- .dt[, .(count = .N), by = .(d)]
.gg.plot(.dt, aes(x=d, y=count)) + geom_point(stroke=0) + scale_x_log10() + scale_y_log10()
```

* $n(G)$: `r length(V(G.dc))` (vs. `r length(V(G.ppi))`)

* $m(G)$: `r length(E(G.dc))` (vs. `r length(E(G.ppi))`)

* Note: $d$ starts from `3`

:::
::: {.column width=.5}

Will it look nicer?

```{r echo=FALSE, fig.width=2, fig.height=1.5}
apply.layout.dt(G.dc, "ppi-dc") %>% plot.layout.dt(.add.lab=FALSE, .vsize=1, .esize=.1)
```

* Rationale: no genes act alone. If some genes have significantly fewer edges connected, high-throughput experiments may not have covered them sufficiently.

:::
::::::



## Random walks in network amplify associations

\includegraphics[width=\textwidth]{img/2_network_application/app_network_propagation-1.pdf}

\vfill

Square "D" is the original hit $\implies$ found circled "D"

\footnotesize

[@Cowen2017-ka]

## Propagation amplifies disease associations

\includegraphics[width=\textwidth]{img/2_network_application/app_network_propagation-2.pdf}

Network propagation strengthens weak genetic signals by spreading information across interacting genes.

\footnotesize

[@Cowen2017-ka]


## Network Propagation as natural data integration

:::::: {.columns}
::: {.column width=.6}

\includegraphics[height=.6\textheight]{img/2_network_application/app_network_propagation-3.pdf}

:::
::: {.column width=.4}

* Different data types (mutations, expression, pathways) mapped onto network

* Network propagation combines signals across data sources

* Unified mathematical framework for genetic discovery

:::
::::::

\footnotesize

[@Cowen2017-ka]


## Example: Gene prioritisation in Multiple Sclerosis

Challenge:

* GWAS identifies ~100 MS-associated genes
* But these represent only a fraction of disease biology
* Many relevant genes may not show strong GWAS signal
* How to find additional disease-relevant genes?

Solution: Network propagation

* Start with MS GWAS genes as seeds
* Propagate signal through protein-protein interaction network
* Genes frequently "visited" by random walks are prioritised

## Random walk on PPI network

Mathematical framework:

$$\mathbf{p}^{(t)} \gets \gamma W \mathbf{p}^{(t-1)} + (1- \gamma) \mathbf{p}^{(0)}$$

* $\mathbf{p}^{(0)}$: initial probability (1 for MS genes, 0 otherwise)
* $W$: transition matrix from adjacency matrix
* $\gamma$: restart probability (typically 0.85)
* Iterate until convergence

Result:

* Genes close to MS seeds in network receive high scores
* Captures functional relationships beyond direct interactions

## Step 1: Download MS GWAS genes

```{r include=FALSE}
.file <- "gwas-catalog-download-associations-alt-full.tsv"
if.needed(file.path("../data", .file), {
    .url <- "https://ftp.ebi.ac.uk/pub/databases/gwas/releases/2026/02/17/gwas-catalog-associations_ontology-annotated-full.zip"
    .zip <- "../data/gwas_catalog.zip"
    download.file(.url, destfile = .zip, mode = "wb")
    unzip(.zip, exdir = "../data/")
    file.remove(.zip)
})

.gwas.dt <- fread(file.path("../data",.file), sep="\t", quote="")
is.ms <- function(x) str_detect(x, "[Mm]ultiple sclerosis")
ms.genes <- .gwas.dt[is.ms(`DISEASE/TRAIT`),] %>%
    filter(str_length(`MAPPED_GENE`) > 2) %>%
    select(`MAPPED_GENE`) %>% na.omit %>% unlist %>%
    str_split(pattern="[ ,;]+") %>% unlist %>% unique
```

* Download GWAS Catalog from NHGRI-EBI
* Extract genes associated with "Multiple Sclerosis"
* **Result:** `r length(ms.genes)` MS-associated genes

## Step 2: Load human PPI network

```{r include=FALSE}
.biogrid.file <- list.files("../data", pattern="BIOGRID-ALL.*tab3.txt$",
                            full.names=TRUE)
if(length(.biogrid.file) == 0) {
    .url <- "https://downloads.thebiogrid.org/Download/BioGRID/Latest-Release/BIOGRID-ALL-LATEST.tab3.zip"
    .zip <- "../data/biogrid-all.zip"
    download.file(.url, destfile = .zip, mode = "wb")
    unzip(.zip, exdir = "../data/")
    file.remove(.zip)
    .biogrid.file <- list.files("../data", pattern="BIOGRID-ALL.*tab3.txt$",
                                full.names=TRUE)
}
.biogrid.rds <- "../data/biogrid-human.G.rds"
if.needed(.biogrid.rds, {
    biogrid <- fread(.biogrid.file[1], sep="\t", quote="") %>%
        filter(`Organism ID Interactor A` == 9606,
               `Organism ID Interactor B` == 9606) %>%
        select(`Official Symbol Interactor A`, `Official Symbol Interactor B`) %>%
        filter(`Official Symbol Interactor A` != `Official Symbol Interactor B`)
    saveRDS(graph_from_edgelist(biogrid %>% as.matrix, directed = FALSE),
            .biogrid.rds)
})
G.ppi <- readRDS(.biogrid.rds)
```

* BioGRID human protein-protein interactions
* Uses gene symbols (not Ensembl IDs)
* **Result:** `r num.int(length(V(G.ppi)))` genes, `r num.int(length(E(G.ppi)))` interactions

## Step 3: Map MS genes to network

```{r include=FALSE}
# Match MS genes to network (both use gene symbols)
vtot <- V(G.ppi)
gene.names <- names(vtot)
ms.idx <- which(gene.names %in% ms.genes)
```

* BioGRID uses gene symbols - direct matching with MS genes
* **Result:** `r length(ms.idx)` of `r length(ms.genes)` MS genes found in network
* **`r round(100*length(ms.idx)/length(V(G.ppi)), 1)`%** of network genes
* Need to expand beyond direct hits via propagation

## Step 4: Random walk with restart

```{r include=FALSE}
if.needed(str_c(fig.dir, "/ms_propagation.rds"), {
    A <- as_adjacency_matrix(G.ppi)
    W <- sweep(A, 2, degree(G.ppi), `/`)
    n <- length(V(G.ppi))
    pr <- pr.0 <- matrix(0, n, 1)
    pr.0[ms.idx, 1] <- 1 / length(ms.idx)
    gamma <- 0.85

    # Store intermediate steps
    P <- matrix(NA, n, 101)
    P[, 1] <- as.numeric(pr.0)
    pr <- pr.0

    for(t in 1:100) {
        pr <- gamma * W %*% pr + (1 - gamma) * pr.0
        pr <- pr / sum(pr)
        P[, t + 1] <- as.numeric(pr)
    }
    saveRDS(list(P=P, pr.0=pr.0, idx=ms.idx), str_c(fig.dir, "/ms_propagation.rds"))
})
.result <- readRDS(str_c(fig.dir, "/ms_propagation.rds"))
P <- .result$P
pr.0 <- .result$pr.0
```

* Initialize: MS genes = 1, others = 0
* Iterate: $\mathbf{p}^{(t)} \gets 0.85 \times W \mathbf{p}^{(t-1)} + 0.15 \times \mathbf{p}^{(0)}$
* Run 100 iterations until convergence
* **Result:** All genes ranked by propagation score

## Probability spreads through network over iterations

```{r echo=FALSE, fig.width=6, fig.height=1.5, only.plot="1"}
.dt <- data.table(v=as.integer(V(G.ppi)), pr=P[,1])
.gg.plot(.dt, aes(v,pr)) + geom_line() + ylab("p(0)") + xlab("vertices")
```

```{r echo=FALSE, fig.width=6, fig.height=1.5, only.plot="2"}
.dt <- data.table(v=as.integer(V(G.ppi)), pr=P[,2])
.gg.plot(.dt, aes(v,pr)) + geom_line() + ylab("p(1)") + xlab("vertices")
```

```{r echo=FALSE, fig.width=6, fig.height=1.5, only.plot="3"}
.dt <- data.table(v=as.integer(V(G.ppi)), pr=P[,11])
.gg.plot(.dt, aes(v,pr)) + geom_line() + ylab("p(10)") + xlab("vertices")
```

```{r echo=FALSE, fig.width=6, fig.height=1.5, only.plot="4"}
.dt <- data.table(v=as.integer(V(G.ppi)), pr=P[,51])
.gg.plot(.dt, aes(v,pr)) + geom_line() + ylab("p(50)") + xlab("vertices")
```

```{r echo=FALSE, fig.width=6, fig.height=1.5, only.plot="5"}
.dt <- data.table(v=as.integer(V(G.ppi)), pr=P[,101])
.gg.plot(.dt, aes(v,pr)) + geom_line() + ylab("p(100)") + xlab("vertices")
```

## Top 50 propagation genes + all MS genes in network

```{r echo=FALSE, fig.width=4, fig.height=3}
vtot <- V(G.ppi)
gene.names <- names(vtot)

# Top 50 by propagation score
top.idx <- order(P[,101], decreasing=TRUE)[1:50]

# All MS genes in network
ms.in.net <- which(gene.names %in% ms.genes)

# Union of both sets
all.idx <- union(top.idx, ms.in.net)
v.all <- vtot[all.idx]
G.sub <- induced_subgraph(G.ppi, v.all)

# Three categories:
# 1 = MS gene recaptured in top 50
# 2 = MS gene NOT in top 50
# 3 = top 50 gene, not MS (discovered by propagation)
# Use Fruchterman-Reingold with more spread
.layout.file <- str_c(fig.dir, "/ppi_ms_full.rds")
if(!file.exists(.layout.file)){
    layout.dt <- layout_with_fr(G.sub, niter=1000) %>%
        as.data.table()
    saveRDS(layout.dt, .layout.file)
}
.out <- readRDS(.layout.file) %>% dplyr::mutate(v = 1:dplyr::n())
.dt <- as_long_data_frame(G.sub) %>%
    as.data.table %>%
    merge(.out, by.x="from", by.y="v") %>%
    merge(.out, by.x="to", by.y="v", suffixes=c(".from",".to"))
.sub.names <- names(v.all)
.cat <- ifelse(.sub.names %in% ms.genes & (all.idx %in% top.idx), 1,
        ifelse(.sub.names %in% ms.genes, 2, 3))

for(i in seq_along(.cat)) {
    .dt[`from` == i, k.from := .cat[i]]
    .dt[`to` == i, k.to := .cat[i]]
}

# Build vertex table with categories
.rename <- function(xx) { colnames(xx) <- c("v","V1","V2","k"); xx }
.dt.v <- rbind(.dt[, .(`from`, `V1.from`, `V2.from`, `k.from`)] %>% .rename(),
               .dt[, .(`to`, `V1.to`, `V2.to`, `k.to`)] %>% .rename()) %>%
    dplyr::mutate(k = as.integer(k)) %>% unique

# Split into layers: propagation (3) bottom, MS (1,2) on top
.dt.prop <- .dt.v %>% filter(k == 3)
.dt.ms.miss <- .dt.v %>% filter(k == 2)
.dt.ms.hit <- .dt.v %>% filter(k == 1)

.aes.e <- aes(V1.from, V2.from, xend=V1.to, yend=V2.to)
ggplot() + theme_void() +
    geom_segment(.aes.e, data = .dt, size = .3, colour="gray80") +
    geom_point(aes(V1, V2), data = .dt.prop,
               pch=21, fill=alpha("#4DAF4A", .5), stroke=.1, size=2) +
    geom_point(aes(V1, V2), data = .dt.ms.miss,
               pch=21, fill=alpha("#377EB8", .5), stroke=.2, size=3) +
    geom_point(aes(V1, V2), data = .dt.ms.hit,
               pch=21, fill=alpha("#E41A1C", .5), stroke=.2, size=3)
```

Red = MS gene in top 50, Blue = MS gene not in top 50, Green = discovered by propagation

## Are we just finding high-degree vertices?

```{r echo=FALSE, fig.width=3, fig.height=2.5}
.dt <- data.table(v=as.integer(V(G.ppi)),
                  d=degree(G.ppi),
                  pr=P[,101])

.dt.top <- .dt %>% arrange(desc(pr)) %>% head(100)

.gg.plot(.dt, aes(d,pr)) +
    geom_hex(aes(fill = stat(log10(count)))) +
    geom_point(data = .dt.top, colour="red", pch=21) +
    geom_smooth(method="lm", se=FALSE, size=.5, colour="black", lty=2) +
    scale_fill_distiller("count", direction = 1, labels=function(x) round(10^x)) +
    theme(legend.key.width=unit(.2, "lines")) +
    theme(legend.key.height=unit(.5, "lines")) +
    scale_x_log10() +
    scale_y_log10() +
    ylab("p(100)") +
    xlab("degree")
```

Red dots: top 100 genes

## Network propagation summary

- Start with seed genes (e.g., from GWAS, mutations)
- Propagate signal through the network
- Neighbors of seed genes receive scores based on network distance
- Discover related genes not in original list
- Aggregates weak signals across pathways
- Applications: disease gene prioritization, drug target discovery


## Integrate PPI with tumour expression profiles

\centerline{\includegraphics[width=.8\textwidth]{img/2_network_application/app_network_brca_classifier-1.pdf}}

\footnotesize

@Chuang2007-ti

## Subnetwork activity aggregates coordinated gene expression

\includegraphics[width=\textwidth]{img/2_network_application/app_network_brca_classifier-2.pdf}

\footnotesize

@Chuang2007-ti


## Search for subnetworks maximally discriminative of phenotype

\includegraphics[width=\textwidth]{img/2_network_application/app_network_brca_classifier-3.pdf}

\footnotesize

@Chuang2007-ti

## Significant subnetworks $\implies$ a feature matrix for classification of many relevant genes

\centerline{\includegraphics[width=.6\textwidth]{img/2_network_application/app_network_brca_classifier-4.pdf}}

\footnotesize

@Chuang2007-ti

## Subnetworks centre on TP53 and ERBB2 hubs

\includegraphics[width=\textwidth]{img/2_network_application/app_network_brca_classifier-5.pdf}

\footnotesize

@Chuang2007-ti


## Subnetworks recover more known susceptibility genes than single-gene markers

\includegraphics[width=\textwidth]{img/2_network_application/app_network_brca_classifier-6.pdf}

\footnotesize

@Chuang2007-ti



## Most cancer genes are rarely mutated

\onslide<1->{\includegraphics[height=.5\textheight]{img/2_network_application/app_pancancer-1.pdf}}
\onslide<2>{~\includegraphics[height=.5\textheight]{img/2_network_application/app_pancancer-2.pdf}}

\footnotesize

@Leiserson2015-sw


## HotNet2: heat diffusion through PPI finds connected altered subnetworks

\onslide<1->{\includegraphics[height=.5\textheight]{img/2_network_application/app_pancancer-3.pdf}}
\onslide<2>{~\includegraphics[height=.5\textheight]{img/2_network_application/app_pancancer-4.pdf}}

\footnotesize

@Leiserson2015-sw

## Subnetworks map to known pathways and show co-occurrence patterns across cancer types

\centerline{\includegraphics[height=.65\textheight]{img/2_network_application/app_pancancer-5.pdf}}

\footnotesize

@Leiserson2015-sw

## SWI/SNF chromatin remodelling complex altered in 16.8\% of pan-cancer samples

\centerline{\includegraphics[height=.7\textheight]{img/2_network_application/app_pancancer-6.pdf}}

\footnotesize

@Leiserson2015-sw



## Pan-cancer network analysis

- Cancer is driven by pathways, not just individual genes
- Challenge: different patients have mutations in different genes
- Solution: project mutations onto PPI network, find mutated subnetworks
- HotNet2 identifies altered pathways and protein complexes
- Reveals cancer driver pathways missed by single-gene analysis

# New frontiers: cell-cell communication and regulatory perturbation

## Cells communicate with one another

:::::: {.columns}
::: {.column width=.6}

\centerline{\includegraphics[height=.7\textheight]{img/3_cell_cell_network/cell_cell_review-1.pdf}}

:::
::: {.column width=.4}

**Autocrine**: cell signals to itself

**Paracrine**: signals to nearby cells

**Endocrine**: signals via bloodstream to distant cells

**Juxtacrine**: signals through direct membrane contact

:::
::::::

\footnotesize

[@Armingol2021-tu]

## Cell-cell communication shapes development, tissue homeostasis, and disease

\centerline{\includegraphics[height=.6\textheight]{img/3_cell_cell_network/cell_cell_review-2.pdf}}

\footnotesize

[@Armingol2021-tu]

## Maternal-fetal interface $\iff$ cell-cell interactions

:::::: {.columns}
::: {.column width=.35}

\centerline{\includegraphics[height=.7\textheight]{img/3_cell_cell_network/cell_cell_communication_fetal-2.pdf}}

\tiny
[@Vento-Tormo2018-wn]

:::
::: {.column width=.6}

\centerline{\includegraphics[height=.6\textheight]{img/3_cell_cell_network/cell_cell_communication_fetal-5.pdf}}

Decidual^[pregnant uterine lining] NK cells coordinate trophoblast recruitment and tissue remodelling

:::
::::::


## Single-cell atlas reveals diverse cell types at the maternal-fetal interface

\centerline{\includegraphics[height=.6\textheight]{img/3_cell_cell_network/cell_cell_communication_fetal-1.pdf}}

\footnotesize

[@Vento-Tormo2018-wn]

<!-- ## Ligand-receptor connectome -->

<!-- - Cells communicate via secreted ligands binding to receptors -->
<!-- - FANTOM5: catalogue of 1,894 ligand-receptor pairs across 144 human cell types -->
<!-- - Defines potential communication channels -->
<!-- - Tissue context determines which interactions are active -->

## CellPhoneDB scores ligand-receptor pairs using mean expression and permutation testing

:::::: {.columns}
::: {.column width=.4}

\centerline{\includegraphics[height=.6\textheight]{img/3_cell_cell_network/cell_cell_communication_fetal-6.pdf}}

:::
::: {.column width=.6}

\centerline{\includegraphics[height=.6\textheight]{img/3_cell_cell_network/cell_cell_communication_fetal-7.pdf}}

:::
::::::

\footnotesize

[@Vento-Tormo2018-wn]

## Inferring cell-cell communication from scRNA-seq

\centerline{\includegraphics[height=.7\textheight]{img/3_cell_cell_network/cell_cell_review-3.pdf}}

\footnotesize

[@Armingol2021-tu]

## Visualizing cell-cell communication scores across ligand-receptor pairs and cell types

\centerline{\includegraphics[height=.6\textheight]{img/3_cell_cell_network/cell_cell_review-4.pdf}}

\footnotesize

[@Armingol2021-tu]

## Decidual NK subsets modulate immune tolerance via HLA-C/KIR and checkpoint interactions

\centerline{\includegraphics[height=.6\textheight]{img/3_cell_cell_network/cell_cell_communication_fetal-3.pdf}}

\footnotesize

[@Vento-Tormo2018-wn]

## Cell-cell communication in tissues

- Single-cell atlas of maternal-fetal interface (early pregnancy)
- Mapped ligand-receptor interactions between immune, stromal, and trophoblast cells
- Revealed communication networks maintaining pregnancy
- Showed how immune tolerance is established at the maternal-fetal boundary

## CellOracle: from multi-omics to GRN to in silico TF perturbation

\centerline{\includegraphics[height=.6\textheight]{img/3_cell_cell_network/cell_perturb_kamimoto-1.pdf}}

\footnotesize

[@Kamimoto2023-yf]

## Chromatin accessibility defines the regulatory wiring

- scATAC-seq open chromatin defines $\mathcal{R}(j)$: which TFs can regulate gene $j$

$$\hat{x}_j = \sum_{i \in \mathcal{R}(j)} \beta_{ij}\, x_i$$

- KO of TF $k$: set $x_k = 0$ $\to$ propagates through GRN $\to$ genome-wide $\Delta\mathbf{x}$
- Cell fate shift: cosine similarity between $\Delta\mathbf{x}_i$ and $(\mathbf{x}_j - \mathbf{x}_i)$ for each neighbor $j$

\footnotesize

[@Kamimoto2023-yf]

## Simulated differentiation flow recapitulates hematopoietic lineage trajectories

\centerline{\includegraphics[height=.6\textheight]{img/3_cell_cell_network/cell_perturb_kamimoto-2.pdf}}

\footnotesize

[@Kamimoto2023-yf]

## In silico TF knockdown shifts cells toward erythroid/megakaryocyte fate

\centerline{\includegraphics[height=.7\textheight]{img/3_cell_cell_network/cell_perturb_kamimoto-3.pdf}}

\tiny

[@Kamimoto2023-yf]

## Opposing TF knockdown promotes granulocyte-monocyte differentiation

\centerline{\includegraphics[height=.7\textheight]{img/3_cell_cell_network/cell_perturb_kamimoto-4.pdf}}

\tiny

[@Kamimoto2023-yf]

## Spi1 KO: loss of a myeloid driver redirects cells toward erythroid fate

:::::: {.columns}
::: {.column width=.55}

\centerline{\includegraphics[height=.65\textheight]{img/3_cell_cell_network/cell_perturb_kamimoto-5.pdf}}

:::
::: {.column width=.4}

- Spi1 (PU.1) normally promotes GM/myeloid differentiation
- KO: GM lineage inhibited (magenta), MEP differentiation promoted (green)
- PS = perturbation score: signed measure of fate shift

:::
::::::

\footnotesize

[@Kamimoto2023-yf]

## Gata1 KO has the opposite effect: ME inhibited, GMP promoted

:::::: {.columns}
::: {.column width=.55}

\centerline{\includegraphics[height=.65\textheight]{img/3_cell_cell_network/cell_perturb_kamimoto-6.pdf}}

:::
::: {.column width=.4}

- Gata1 normally promotes erythroid/ME differentiation
- KO: ME lineage inhibited (magenta), GMP differentiation promoted (green)
- Spi1 and Gata1 are mutual antagonists at the MEP/GMP branch point (Fig. 1g)

:::
::::::

\footnotesize

[@Kamimoto2023-yf]

<!-- ## In silico gene perturbation (CellOracle) -->
<!-- - GRN nodes = genes; edges = TF $\to$ target, weighted by $\beta_{ij}$ from regression -->
<!-- - scATAC-seq constrains which TFs can regulate each gene (motif accessibility) -->
<!-- - In silico KO: set TF $\to$ 0, propagate $\Delta x_j$ through GRN -->
<!-- - Project $\Delta \mathbf{x}$ into embedding $\to$ cell fate transition vector -->
<!-- - Not a cell-cell network: operates entirely within individual cell states -->

## Networks $\implies$ context for interpreting genomic data

**Measuring networks:**

- Physical interactions:  Y2H, AP-MS
- Genetic interactions: synthetic lethality

**Network analysis:**

- Propagation: amplify signals across network
- Modules: disease genes cluster together

**Cell-cell communication** (CellPhoneDB):

<!-- - Nodes = cell types; edges = significant ligand-receptor pairs -->
- scRNA-seq + curated L-R database → who signals to whom

**Intracellular GRN + perturbation** (CellOracle):

<!-- - Nodes = genes; edges = TF $\to$ target (constrained by scATAC-seq) -->
- In silico TF KO $\to$ predict cell fate shift

##

<!-- intentional blank -->

## References {.allowframebreaks}
