---
title: "Statistical Methods for single cell data analysis 3"
author: "Yongjin Park, UBC Path + Stat, BC Cancer"
date: "`r format(Sys.time(), '%d %B %Y')`"
mainfont: Roboto
classoption: "aspectratio=169"
fontsize: 12pt
output:
    beamer_presentation:
        theme: Madrid
        keep_md: true
        keep_tex: true
        latex_engine: lualatex
        slide_level: 2
header-includes:
  - \usepackage{cancel}
  - \usepackage{fontspec}
  - \definecolor{UBCblue}{rgb}{0.04706, 0.13725, 0.26667}
  - \usecolortheme[named=UBCblue]{structure}
  - \setbeamertemplate{frametitle}{\color{UBCblue}\bfseries\insertframetitle\par\vskip-6pt\hrulefill}
  - \setbeamercolor{title in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{section in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{author in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{date in head/foot}{bg=white,fg=gray}
  - \setbeamertemplate{page number in head/foot}{}
  - \setbeamertemplate{frame numbering}[none]
  - \setbeamercolor{alerted text}{bg=yellow}
  - \AtBeginSection[]{\begin{frame}\frametitle{Today's lecture}{\Large\tableofcontents[currentsection]}\end{frame}}
  - |
    \makeatletter
    \def\ps@titlepage{%
      \setbeamertemplate{footline}{}
    }
    \addtobeamertemplate{title page}{\thispagestyle{titlepage}}{}
    \makeatother
    \include{toc}
---


```{r setup, include=FALSE}
setwd('~/work/teaching/stat540/lectures/lect18-scRNAseq-ML/')
library(data.table)
library(tidyverse)
library(patchwork)
library(matrixTests)
library(mmutilR)

source("Setup.R")
source("Util.R")
fig.dir <- "../Fig/scRNA/vae/"
dir.create(fig.dir, showWarnings=FALSE)
setup.env(fig.dir)
dir.create("../data", showWarnings=FALSE)
library(extrafont)
library(xkcd)
extrafont::font_import(pattern="[X/x]kcd", prompt=F)
extrafont::loadfonts()
theme_set(theme_xkcd() +
          theme(title = element_text(size=10),
                legend.background = element_blank())
          )
```

##

\large

Source code available:

`https://github.com/stat540-UBC/lectures`

## Overview of single-cell data analysis

\centerline{\includegraphics[width=\textwidth]{img/scRNA_overview/Karchenko_1.pdf}}

\vfill

\flushright
Karchenko, \emph{Nature Methods} (2021)

## Overview of single-cell data analysis cont'd

\centerline{\includegraphics[width=\textwidth]{img/scRNA_overview/Karchenko_2.pdf}}

\vfill

\flushright
Karchenko, \emph{Nature Methods} (2021)

## The goal of modelling in high-dimensional space

The goal is to model two types of relationships:

\large

* Relationship between dimensions/features (genes)

\vfill

* Relationship between samples/data points (cells)

\normalsize

\vfill

* One of the most fundamental relationships is co-variation.

## Example: scRNA-seq data of human pancreatic cells

```{r}
raw.data <- fileset.list("../data/GSE85241/GSE85241")
.info <- rcpp_mmutil_info(raw.data$mtx)
```

:::::: {.columns}
::: {.column width=.45}

\centerline{\includegraphics[height=.7\textheight]{img/Muraro_GSE85241.png}}

:::
::: {.column width=.45}

\large

* `r num.int(.info$max.row)` genes/features/rows

* `r num.int(.info$max.col)` cells/columns

* `r num.int(.info$max.elem)` non-zero elements

* $\approx$ `r round(100 * .info$max.elem / .info$max.col / .info$max.row)` % non-zero elements

:::
::::::

\tiny
Muraro \textit{et al.} \emph{Cell Systems} (2016)


## What is a latent variable/factor model?

Examples:

\large

* Principal Component Analysis

* Non-negative matrix factorization

* (...)

\normalsize

## What is Unsupervised Learning?

:::::: {.columns}
::: {.column width=.45}
::: {.block}

### Supervised learning

* Input: data $\mathcal{X} = \{\mathbf{x}_{i},\ldots\}$

* Input 2: label $\mathcal{Y} = \{y_{i},\ldots\}$

* Goal: learn $f: \mathcal{X} \to \mathcal{Y}$

:::

X: gene expression samples; Y: disease labels

:::
::: {.column width=.45}

::: {.block}
### Unsupervised learning

* Input: data $\mathcal{X} = \{\mathbf{x}_{i},\ldots\}$

* Goal 1: learn $f: \mathcal{Z} \to \mathcal{X}$

* Goal 2: hidden/latent states, $\mathcal{Z} = \{\mathbf{z}_{i},\ldots\}$

:::

X: gene expression matrices

:::
::::::

# Representation learning

## 

\Large

How can we obtain a good representation of data?

\normalsize

## Two strategies to learn a "good" representation

:::::: {.columns}
::: {.column width=.45}

::: {.block}

### Feature engineering

Construct useful combinations of variables, feature engineering

- Principal component analysis (matrix factorization)

- Pattern recognition

- Clustering (average patterns $\approx$ combined features/data)

::: 

:::
::: {.column width=.45}

::: {.block}

### Manifold learning

Change the manifold (a coordinate system) of data

- Probabilistic topic models: word frequency $\to$ simplex
	
- Angular space $\to$ Euclidean space

- Observed high-dim. $\to$ interpretable ones

:::

:::
::::::

We use both strategies in practice.

## Feature engineering by combining multiple vars.

:::::: {.columns}
::: {.column width=.5}

```{r echo=FALSE}
.rnorm <- function(d1,d2) matrix(rnorm(d1*d2),d1,d2)
```

```{r echo=FALSE}
U <- .rnorm(100, 2); V <- .rnorm(100, 2)
X <- U %*% t(V) + .rnorm(100, 100)
```

$$\mathbf{x}_{i} = \mathbf{u}_{i} V^{\top} + \mathcal{N}\!\left(\mathbf{0},I\right)$$

```{r echo=FALSE}
X <- .sort.matrix(X)
```

```{r echo=FALSE, fig.width=2.5, fig.height=2, onslide.plot="1-"}
.matshow(X, .axis.lab=0, .lw=0)
```

:::
::: {.column width=.5}

```{r echo=FALSE}
.svd <- svd(X, nu=2, nv=2)
u.hat <- sweep(.svd$u, 2, .svd$d[1:2], `*`)
```

```{r echo=FALSE, fig.width=2.7, fig.height=2.3, only.plot="2"}
p1 <- .matshow(u.hat) +
    ggtitle("features")
p.times <- ggplot() + theme_void() +
    geom_text(aes(0,0,label="X"), size=5)
p2 <- wrap_plots(.matshow(t(.svd$v)),
                 ggplot() + theme_void(),
                 ncol=1, heights=c(1,9)) +
    ggtitle("feature loading")
wrap_plots(list(p1, p.times, p2), nrow=1, widths=c(1,2,8))
```

:::
::::::

## Feature engineering 

:::::: {.columns}
::: {.column width=.5}

$$\mathbf{x}_{i} = \mathbf{u}_{i} V^{\top} + \mathcal{N}\!\left(\mathbf{0},I\right)$$

* We call each column of U and $V$ a factor (row/column)

* One factor corresponds to the activities/patterns/expressions of 100 genes/samples

* We will discuss further later.

:::
::: {.column width=.5}

```{r echo=FALSE, fig.width=2.7, fig.height=2.1}
p1 <- .matshow(u.hat)
p.times <- ggplot() + theme_void() +
    geom_text(aes(0,0,label="X"), size=5)
p2 <- wrap_plots(.matshow(t(.svd$v)),
                 ggplot() + theme_void(),
                 ncol=1, heights=c(1,9))
wrap_plots(list(p1, p.times, p2), nrow=1, widths=c(1,2,8))
```

:::
::::::

## Manifold learning changes the coordinate system of data

:::::: {.columns}
::: {.column width=.5}

```{r echo=FALSE}
if(!require(sphereplot)){
    install.packages("sphereplot")
}
.dt <- fread("worldcities.csv")
xx <- sphereplot::sph2car(.dt$lng, .dt$lat)
.svd <- svd(xx)
```

```{r echo=FALSE, fig.width=2.5, fig.height=2.3, only.plot="1-2"}
ggplot(as.data.table(xx), aes(x, y)) +
    ggrastr::rasterise(geom_point(stroke=0, size=.5, alpha=.5, colour="gray40"), dpi=300) +
    ggtitle("x and y coordinates")
```


```{r echo=FALSE, fig.width=2.5, fig.height=2.5, only.plot="3-4"}
ggplot(as.data.table(.svd$u), aes(V1, V2)) +
    ggrastr::rasterise(geom_point(stroke=0, size=.5, alpha=.5, colour="gray40"), dpi=300) +
    ggtitle("PC1 and PC2") +
    xlab("PC 1") + ylab("PC 2")
```

:::
::: {.column width=.5}

```{r echo=FALSE, fig.width=2.5, fig.height=2.3, only.plot="2"}
ggplot(as.data.table(xx), aes(x, z)) +
    ggrastr::rasterise(geom_point(stroke=0, size=.5, alpha=.5, colour="gray40"), dpi=300) +
    ggtitle("x and z coordinates")
```

```{r echo=FALSE, fig.width=2.5, fig.height=2.5, only.plot="4"}
ggplot(as.data.table(.svd$u), aes(V2, V3)) +
    ggrastr::rasterise(geom_point(stroke=0, size=.5, alpha=.5, colour="gray40"), dpi=300) +
    ggtitle("PC2 and PC3") +
    xlab("PC 2") + ylab("PC 3")
```

:::
::::::

```{r echo=FALSE, fig.width=4.5, fig.height=2.8, only.plot=5}
yy <- sphereplot::car2sph(xx) %>% as.data.table
ggplot(yy, aes(`long`, `lat`)) +
    ggrastr::rasterise(geom_point(stroke=0, size=.5, alpha=.5, colour="gray40"), dpi=300) +
    ggtitle("(x,y,z) to (long, lat)") +
    xlab("longitude") + ylab("latitude")
```

\tiny
Data from [https://simplemaps.com/data/world-cities](https://simplemaps.com/data/world-cities)


## Representation learning

:::::: {.columns}
::: {.column width=.45}

### Feature engineering

* Combine samples based on similarity to learn common features

* Combine features/genes/proteins to create a new factor

* Matrix factorization to capture factors that explain large variation 

:::
::: {.column width=.45}

### Manifold learning

* Change the coordinate system

* What's not separable in one space can be spread out in another one

* Apply non-linear functions to a set of features

:::
::::::




# Clustering as a simplest possible unsupervised learning

```{r}
library(torch)
torch_set_num_threads(8)
```

## The goal of this clustering section 

* Model-based unsupervised learning

    - Latent variable model

* Expectation Maximization

* Hands-on experience with clustering algorithm

```{r}
#' @param nn sample size
#' @param D dimensionality
#' @param K number of clusters
simulate.data <- function(nn, D, K, sig = .2) {
  require(torch)
  torch_manual_seed(1447)
 .mu <- torch_randn(K, D)
  kk <- torch_tensor(sample(K,nn,TRUE),
                     dtype=torch_long())
 .z <- nnf_one_hot(kk)$to(dtype=torch_float())
 .eps <- torch_randn(nn, D)
 list(x = torch_mm(.z, .mu) + .eps * sig,
      k = as.integer(kk), z = .z, mu = .mu)
}
``` 

## Clustering: Find commonality across rows and cols

```{r echo=FALSE, fig.width=4.5, fig.height=2.5, only.plot="1"}
sim.out <- simulate.data(300, 200, 3, sig = 3)
oo <- order(sim.out$k)
xx <- sim.out$x %r% oo
.matshow(xx) + ggtitle("sample x gene")
```

```{r echo=FALSE, fig.width=4.5, fig.height=2.5, only.plot="2-"}
xx <- .sort.matrix(sim.out$x)
.matshow(xx) + ggtitle("sample x gene")
```


## What do we want to know from data?

```{r fig.width=4, fig.height=2.5, echo = FALSE}
X <- as.matrix(sim.out$x)
.svd <- rsvd::rsvd(X,2)

.dt <- as.dt(X) %>%
    cbind(group = as.factor(sim.out$k))

.dt.mu <- as.dt(sim.out$mu)

plt.data <- 
    ggplot(.dt, aes(V1, V2, colour=group)) +
    geom_point(size=1.5) +
    geom_point(data = .dt.mu, colour = "red", size=5, pch = "+") +
    scale_colour_brewer(palette="Paired") +
    theme(axis.title = element_blank())
print(plt.data)
```

**Two goals**: recover (1) group membership and (2) the centroids (red marks)

## A chicken-and-egg problem: guessing latent membership vs. parametric inference

* If we knew the membership of all the points, we can simply estimate the centre (e.g., taking sample mean within each cluster)

* If we knew the centre coordinates, we would be able to assign points to most probable groups easily based on distance from the centre points.

* Statistical answer: Solve the underlying inference problem.

* To a parameter estimator, the membership assignments are *hidden* (latent).

<!-- Make your dimensions right... -->
<!-- * $D$: each data point's dimension -->
<!-- * $K$: desired number of clusters (can be relaxed or tuned) -->

```{r}
N <- nrow(X)
D <- ncol(X)
K <- 4
```

<!-- Random initialization of model parameters: -->
<!-- * $\mu_{k} \sim \mathcal{N}\!\left(0, 1/10 I\right)$ -->
<!-- * $\log\sigma_{k} \gets 0$ for all $k$ -->
<!-- Alternatively, we can start with random assignment of samples to the clusters. -->

```{r}
set.seed(3)
.mu <- scale(matrix(rnorm(K*D), K, D)) * .1
mu <- torch_tensor(.mu, requires_grad = TRUE)
ln.sig <- torch_zeros(K, D, requires_grad=TRUE)
```

## Let's think about the data-generating process to "reverse" it

:::::: {.columns}
::: {.column width=.5}

```{r fig.width=2.5, fig.height=2}
print(plt.data)
```

:::
::: {.column width=.45}

::: {.block}

### Latent Variable Model

What could have been done to generate observe data?

* Latent membership: $Z_{ik}$

* Model parameters: $\mu_{k}$ and $\sigma_{k}$

:::

:::
::::::

## Gaussian Mixture Model (k-means)

:::::: {.columns}
::: {.column width=.4}

<!-- ```{r results="asis", out.width=".7\\linewidth"} -->
<!-- knitr::include_graphics("./Vis/unsupervised/GMM_graphical_model.pdf") -->
<!-- ``` -->

:::
::: {.column width=.5}
::: {.block}
### GMM data generating process

1. Initialize $\mu_{k}$ (the centre of each group) and $\sigma_{k}$ (the spread within each group)

2. Randomly assign group membership, $Z_{ik} = 1$ iff a point $i$ belongs to a group $k$.

3. Generate: $\mathbf{x}_{i} | Z_{ik} = 1, \boldsymbol{\mu}_{k} \sim \mathcal{N}\!\left(\boldsymbol{\mu}_{k}, \sigma^{2}I\right)$

:::

:::
::::::

How do we infer $Z$ and $\mu,\sigma$?


## Expectation Maximization for GMM MLE

\begin{eqnarray*}
\onslide<1->{
	J &\equiv& \log \prod_{i=1}^{n} p(\mathbf{x}_{i}|\mu,\sigma) \\
}
\onslide<2>{
	&=& \log \prod_{i=1}^{n} \sum_{Z} p(\mathbf{x}_{i}|Z,\mu,\sigma)p(Z)
}
\end{eqnarray*}

\onslide<2->{
It might be difficult to enumerate all the $Z$'s... 
}

\onslide<3>{
so let's introduce some other distributions that will help our "guessing" work, which we call it $q(Z)$
}

## EM algorithm: What is the best way to guess latent variables?

\begin{eqnarray*}
\onslide<1->{
  \sum_{i} \log p(\mathbf{x}_{i}|\mu,\sigma)
  &=& \sum_{i=1}^{n} {\color{blue} \log} \sum_{Z_{i}} {\color{red} \frac{q(Z_{i})}{q(Z_{i})} } p(\mathbf{x}_{i}|Z_{i},\mu,\sigma)p(Z_{i}) \\
  }
\onslide<2->{
  &\underset{\textsf{\color{blue} Jensen}}{\ge}&
   \sum_{i=1}^{n} \sum_{Z_{i}} q(Z_{i}) {\color{blue}\log} \frac{p(\mathbf{x}_{i}|Z_{i},\mu,\sigma)p(Z_{i})}{q(Z_{i})} \\ }
\end{eqnarray*}

\onslide<3->{ What is the best $q(Z)$? }

## EM: optimal E-step is to take the posterior probability

\onslide<1->{ If $q(Z)=p(Z|\mathbf{x}_{i},\mu,\sigma)$ (by Bayes rule),  }

\begin{eqnarray*}
\onslide<1->{
&=&
\sum_{i}^{n}\sum_{Z_{i}} p(Z_{i}|\mathbf{x}_{i},\mu,\sigma)
\log \frac{p(\mathbf{x}_{i}|Z_{i},\mu,\sigma)p(Z_{i})}{p(Z_{i}|\mathbf{x}_{i}\mu,\sigma)}\\
}
\onslide<2->{
&=&
\sum_{i}^{n}\sum_{Z_{i}} p(Z_{i}|\mathbf{x}_{i},\mu,\sigma)
\log \frac{p(\mathbf{x}_{i},Z_{i}|\mu,\sigma) p(\mathbf{x}_{i}|\mu,\sigma)}{p(\mathbf{x}_{i}, Z_{i}|\mu,\sigma)} \\
}
\onslide<3>{
&=&
\sum_{i=1}^{n} \underbrace{\left[ \sum_{Z_{i}} p(Z_{i}|\mathbf{x}_{i},\mu,\sigma) \right]}_{\color{red} = 1} \log p(\mathbf{x}_{i}|\mu,\sigma) 
\\
}
\end{eqnarray*}

\onslide<3>{The inequality becomes equality.}

## Expectation Maximization algorithm = expected MLE

\begin{eqnarray}
  \log p(X|\mu,\sigma) &=& \log \sum_{Z} p(X,Z|\mu,\sigma) \\
  &\underset{\textsf{\color{blue} Jensen}}{\ge}&
  \sum_{Z} \underbrace{p(Z|X,\mu,\sigma)}_{\color{red}\textsf{E-step}} \underbrace{\log p(X|Z,\mu,\sigma)}_{\color{red}\textsf{M-step}} \\  
 &=& \underbrace{\mathbb{E}_{p(Z|X,\mu,\sigma)}}_{\color{red}\textsf{E-step}} \left[ \underbrace{\log p(X|Z,\mu,\sigma)}_{\color{red}\textsf{M-step}} \right]
\end{eqnarray}

($Z$ is discrete, e.g., a membership indicator)

**Solution**: **Maximize** the lower bound by taking **the expectation** over the posterior probability.

## EM algorithm of GMM: E-step

Log-likelihood under some group ($\mu_{k}$ and $\sigma_{k}$):

$\log p(\mathbf{x}_{i}|\mu_{k}, \sigma_{k}) = \log \mathcal{N}\!\left(\mathbf{x}_{i}|\mu_{k},\sigma_{k}\right)$

```{r}
#' @param X data (n x D)
#' @param .mu centre parameters (K x D)
#' @param .ln.sig log standard dev (K x D)
log.lik <- function(X, .mu, .ln.sig){
    .ln.sig <- torch_clamp(.ln.sig,
                           log(0.01),
                           log(100.))

    .rs <- torch_square(X - .mu)
    .sig <- torch_exp(.ln.sig)
    llik <- - .rs/.sig/2.0 - .ln.sig/2.0

    torch_sum(llik, dim=2) # sum over dim
}
```

How to estimate the posterior?

$$p(Z_{ik}| \mathbf{x}_{i}, \mu, \sigma)
= \frac{\exp\{\log p(\mathbf{x}_{i}|\mu_{k},\sigma_{k})\}}{\sum_{k'} \exp\{\log p(\mathbf{x}_{i}|\mu_{k'},\sigma_{k'})\}}$$

```{r}
#' sample latent Z ~ the posterior
take.estep <- function() {
    llik.mat <- torch_zeros(nrow(X), K)
    for(k in 1:K){
        mu.k <- mu[k, , drop = FALSE]
        ln.sig.k <- ln.sig[k, , drop = FALSE]
        llik.mat[, k] <- log.lik(X, mu.k, ln.sig.k)
    }
    zz <- distr_categorical(logits = llik.mat)
    zz$sample()
}
```

*Remark*: We can stochastically sample $Z_{ik}=1$ with the posterior probability.

## EM algorithm of GMM: M-step

Maximization step to optimize model parameters

Let this expected lower-bound (ELBO)

$$\mathcal{L}(\mathbf{x}_{i}; \{\mu_{k}\}, \{\sigma_{k}\})
=
\sum_{i=1}^{n}\sum_{k=1}^{K} Z_{ik} \log p(\mathbf{x}_{i}| Z_{ik}, \mu_{k}, \sigma_{k})$$

Given $Z$, what are the unknown? We can take gradient steps (e.g., `torch`)

$$\mu_{k}^{(t)} \gets \mu_{k}^{(t-1)} + \rho \nabla_{\mu_{k}} \sum_{i} \mathcal{L}(\mathbf{x}_{i})$$

$$\sigma_{k}^{(t)} \gets \sigma_{k}^{(t-1)} + \rho \nabla_{\sigma_{k}} \sum_{i} \mathcal{L}(\mathbf{x}_{i})$$

```{r}
## `torch` implementation
opt <- optim_adam(list(mu, ln.sig), lr=.15)
take.mstep <- function(Z){
  opt$zero_grad()
  llik.1 <- log.lik(X, mu[1, , drop= FALSE],
                    ln.sig[1, , drop=FALSE])
  llik <- llik.1 * Z[, 1]
  for(k in 2:ncol(Z)){
      llik.k <- log.lik(X, mu[k, , drop=FALSE],
                        ln.sig[k, , drop=FALSE])
      llik <- llik + llik.k * Z[, k]
  }
  loss <- torch_mean(- llik)
  loss$backward()
  . <- opt$step()
  return(torch_sum(llik)$item())
}
```

*Remark*: We have an analytical solution for $\mu$ and $\sigma$ in this example.

```{r echo = FALSE}
##################################
## actually taking the EM steps ##
##################################

llik.trace <- c()
Z.trace <- data.table()

mu.proj <- as.matrix(mu) %*% .svd$v
mu.grad.proj <- as.matrix(torch_zeros_like(mu)) %*% .svd$v

mu.trace <- cbind(as.dt(mu.proj),
                  grad=as.dt(mu.grad.proj),
                  k=1:nrow(mu.proj),
                  t=0)

.grad <- torch_zeros_like(ln.sig)
ln.sig.trace <- cbind(as.dt(ln.sig),
                      grad=as.dt(.grad),
                      k = 1:nrow(mu),
                      t=0)

for(tt in 1:200){
    latent.assign <- take.estep()
    z <- nnf_one_hot(latent.assign)
    llik <- take.mstep(z)

    z.dt <- data.table(k=as.integer(latent.assign), i=1:nrow(X), t=tt)

    Z.trace <- rbind(Z.trace, z.dt)
    llik.trace <- c(llik.trace, llik)

    mu.proj <- as.matrix(mu) %*% .svd$v
    mu.grad.proj <- as.matrix(mu$grad) %*% .svd$v

    mu.dt <- cbind(as.dt(mu.proj),
                   grad=as.dt(mu.grad.proj),
                   k=1:nrow(mu.proj),
                   t=tt)

    mu.trace <- rbind(mu.trace, mu.dt)

    ln.sig.dt <- cbind(as.dt(ln.sig),
                       grad=as.dt(ln.sig$grad),
                       k=1:nrow(mu),
                       t=tt)
    ln.sig.trace <- rbind(ln.sig.trace, ln.sig.dt)
}
```

## Alternate E- and M-step until convergence

:::::: {.columns}
::: {.column width=.5}

```{r echo=TRUE, eval=FALSE}
for(tt in 1:100){
    latent.assign <- take.estep()
    z <- nnf_one_hot(latent.assign)
    llik <- take.mstep(z)
}
```

Find the details here:

```
https://github.com/STAT540-UBC/lectures
```

:::
::: {.column width=.5}

```{r echo = FALSE, fig.width=2.2, fig.height=1.5}
.dt <- as.dt(llik.trace, "llik") %>%
    mutate(iter = 1:n())

ggplot(.dt, aes(iter, llik)) +
    ylab("log-likelihood") +
    geom_smooth(se=FALSE, colour = "red", lty = 2, size=.5) +
    geom_point(pch=21, stroke = .5, size = .5, fill = "white")

```

:::
::::::

## 

```{r echo = FALSE}
xx <- as.matrix(X) %*% .svd$v ## show data projected in 2d

show.mu.trace <- function(.mu, t.max){

    ret <- ggplot(as.dt(xx), aes(V1, V2)) +
        geom_point(colour="gray", alpha=.8, size=.5) +
        theme(axis.title = element_blank())

    .mu <- .mu[t <= t.max]

    .dt <- .mu %>%
        mutate(k = factor(k, 1:K))

    .aes <- aes(V1, V2,
                fill=k,
                alpha = `t`)

    .dt.2 <- .dt[`t` == t.max] %>%
        mutate(k = factor(k, 1:K))

    .aes.2 <- aes(V1, V2,
                  colour=k,
                  xend = V1 - sign(grad.V1) * pmin(1, abs(10 * grad.V1)),
                  yend = V2 - sign(grad.V2) * pmin(1, abs(10 * grad.V2)))

    .arr <- arrow(length=unit(".2", "lines"))

    ret +        
        ggtitle(str_c("M-step Iter = ", t.max)) +
        geom_point(.aes, data=.dt, stroke = 0, pch=21, size=2) +
        geom_segment(.aes.2, size=.5, data=.dt.2, arrow=.arr) +
        scale_alpha_continuous(range=c(0, 1), guide="none") +
        scale_fill_brewer(palette = "Dark2", guide="none") +
        scale_colour_brewer(palette = "Dark2", guide="none")
}

show.z.trace <- function(.trace, t.max) {

    .dt <-
        .trace[`t` == t.max] %>%
        left_join(cbind(as.dt(xx), i=1:nrow(xx)), by="i") %>%
        mutate(`k`=factor(`k`, 1:K))

    ggplot(.dt, aes(V1, V2, fill=k)) +
        ggtitle(str_c("E-step Iter = ", t.max)) +
        geom_point(pch=21, alpha=.8, stroke=.2) +
        theme(axis.title = element_blank())+
        scale_fill_brewer(palette = "Dark2", guide = "none")
}
```

```{r fig.width = 5.5, fig.height = 2.5, only.plot = 1}
tt <- 1
show.z.trace(Z.trace, tt) | show.mu.trace(mu.trace, tt)
```

```{r fig.width = 5.5, fig.height = 2.5, only.plot = 2}
tt <- 2
show.z.trace(Z.trace, tt) | show.mu.trace(mu.trace, tt)
```

```{r fig.width = 5.5, fig.height = 2.5, only.plot = 3}
tt <- 3
show.z.trace(Z.trace, tt) | show.mu.trace(mu.trace, tt)
```

```{r fig.width = 5.5, fig.height = 2.5, only.plot = 4}
tt <- 4
show.z.trace(Z.trace, tt) | show.mu.trace(mu.trace, tt)
```

```{r fig.width = 5.5, fig.height = 2.5, only.plot = 5}
tt <- 5
show.z.trace(Z.trace, tt) | show.mu.trace(mu.trace, tt)
```

```{r fig.width = 5.5, fig.height = 2.5, only.plot = 6}
tt <- 10
show.z.trace(Z.trace, tt) | show.mu.trace(mu.trace, tt)
```

```{r fig.width = 5.5, fig.height = 2.5, only.plot = 7}
tt <- 20
show.z.trace(Z.trace, tt) | show.mu.trace(mu.trace, tt)
```

```{r fig.width = 5.5, fig.height = 2.5, only.plot = 8}
tt <- 25
show.z.trace(Z.trace, tt) | show.mu.trace(mu.trace, tt)
```

```{r fig.width = 5.5, fig.height = 2.5, only.plot = 9}
tt <- 30
show.z.trace(Z.trace, tt) | show.mu.trace(mu.trace, tt)
```

```{r fig.width = 5.5, fig.height = 2.5, only.plot = 10}
tt <- 40
show.z.trace(Z.trace, tt) | show.mu.trace(mu.trace, tt)
```

```{r fig.width = 5.5, fig.height = 2.5, only.plot = 11}
tt <- 100
show.z.trace(Z.trace, tt) | show.mu.trace(mu.trace, tt)
```

```{r fig.width = 5.5, fig.height = 2.5, only.plot = 12}
tt <- 200
show.z.trace(Z.trace, tt) | show.mu.trace(mu.trace, tt)
```

```{r fig.width = 5.5, fig.height = 2.5, only.plot = 13}
tt <- 299
show.z.trace(Z.trace, tt) | show.mu.trace(mu.trace, tt)
```

* Arrows: stochastic gradient $\nabla \mu$ (initially $\mu_{k}=0$ for all $k$)

* Colour: latent membership


# VAE: supervise approaches to unsupervised learning


## Recap: How can we learn patterns in unsupervised learning from single cell count data?

\centerline{\includegraphics[width=.75\linewidth]{img/scRNA_ETM/Matrix_Factorization.pdf}}

* Goal: Find the factor-specific gene dictionary $\beta$ and hidden factor loading $H$.

## How to "encode" high-dim data to low-dim space?

:::::: {.columns}
::: {.column width=.5}

\only<1>{\centerline{\includegraphics[width=.85\linewidth]{img/scRNA_ETM/encoder_1.pdf}}}

\only<2>{\centerline{\includegraphics[width=\linewidth]{img/scRNA_ETM/encoder_deep1.pdf}}}

:::
::: {.column width=.5}

* Take one cell as a long vector $\mathbf{x}_{i}$

* Apply an encoding function $f(\mathbf{x}_{i})$

* Neural network architectures capture relationships between variables

\onslide<2>{
$$h_{i}^{(l)} \gets f(\sum_{j} \underset{\textsf{\color{magenta} connections}}{W_{ij}} h_{j}^{(l-1)} + \underset{\textsf{\color{teal} bias}}{b_{i}})$$
}

* no connection within each layer

:::
::::::

##

\Large

Unsupervised learning of a good encoder model is fundamentally challenging because ...

\vfill

\onslide<2>{we don't have good encoding ``examples'' beforehand.}

## How can we "supervise" how well we encoded?

\centerline{\includegraphics[width=.6\linewidth]{img/scRNA_ETM/autoencoder_with_nodes.pdf}}

## "Supervise" unsupervised learning by self reconstruction

\centerline{\includegraphics[width=.5\linewidth]{img/scRNA_ETM/autoencoder.pdf}}


## Deep autoencoder first proposed in computer vision

:::::: {.columns}
::: {.column width=.6}

\centerline{\includegraphics[width=.9\linewidth]{img/scRNA_ETM/Autoencoder_DL.pdf}}

:::
::: {.column width=.4}

* [Hinton & Salakhutdinov, Science (2006)](https://www.science.org/doi/10.1126/science.1127647)

* One of the first "deep learning" idea

* Learning the latent representation of an image helps subsequent classification tasks.

* To train a deep autoencoder architecture, they pretrained the model layer by layer.

:::
::::::


## Let's build a simple autoencoder model using `torch`

```{r eval=F, echo = T}
###############################################################
## R wrapper library for `libtorch` (C++ engine for PyTorch) ##
###############################################################

library(torch)

############################################
## We can use GPU (NVIDIA Cuda), optional ##
############################################

GPU <- torch_device("cuda")
```

* See this online book if you are interested in building your own deep learning model: **Deep Learning and Scientific Computing with R torch**

\small

* [https://skeydan.github.io/Deep-Learning-and-Scientific-Computing-with-R-torch/](https://skeydan.github.io/Deep-Learning-and-Scientific-Computing-with-R-torch/)

```{r read_full_data}
library(torch)
GPU <- torch_device("cuda")

read.full.data <- function(.data, .dev = GPU){

    .info <- rcpp_mmutil_info(.data$mtx)
    .idx <- rcpp_mmutil_read_index(.data$idx)
    .mtx <- .data$mtx
    .out <- mmutilR::rcpp_mmutil_read_columns_sparse(.mtx, .idx, 1:.info$max.col)

    .inds <- rbind(.out$col, .out$row) # column is row & vice versa
    .dims <- c(.out$max.col, .out$max.row) # dimensionality

    xx <- torch_sparse_coo_tensor(indices = .inds,
                                  values = .out$val,
                                  size = .dims,
                                  dtype = torch_float(),
                                  device = .dev)

    return(xx$to_dense()$to(device=.dev))
}

raw.data <- fileset.list("../data/GSE85241/GSE85241")
x.torch <- read.full.data(raw.data)
```

## Why do we use some Deep Learning library?

\large

* Built-in functions for scientific computing

    * `log`, `log1p`, `exp`, `softmax`, `sigmoid`, `softplus`

* Faster computation both in CPU and GPU (in general)

* **No need to work on differentiation** by hands

* Trouble shooting by ML community

## Encoder: high-dim data $\to$ low-dim latent space

```{r echo = T}
build.linear.encoder <-
    nn_module(
        ## How do we want to build this module ##
        classname = "linear encoder",
        initialize = function(d.in, K){
            self$K <- K # number of hidden factors
            self$func.z <- nn_linear(d.in, K) # a linear neural net layer
            self$bn <- nn_batch_norm1d(d.in) # batch norm
        },
        ## Define how do we propate infomration ##
        forward = function(x.b){
            self$get.latent(x.b)
        },
        get.latent = function(x.b){
            x.b <- torch_log1p(x.b) # log1p transformation
            x.b <- self$bn(x.b) # to expedite training
            self$func.z(x.b) # take linear combinations
        })
```

## Check if this encoder module works okay

```{r echo = T}
lin.encoder <- build.linear.encoder(ncol(x.torch), 7)$to(dev = GPU)
lin.encoder

##########################################
## Test using the first five cells/data ##
##########################################

lin.encoder(x.torch[1:5, ])
```

## Decoder: low-dim latent space $\to$ high-dim data

```{r echo=T}
build.linear.decoder <-
    nn_module(
        classname = "linear decoder",
        ## Define how we want to build this module
        initialize = function(n.out, K) {
            self$K <- K
            self$func.logX <- nn_linear(K, n.out) # latent dim -> data dim
        },
        ## Define how do get back high-dim data
        forward = function(z.b){
            .num <- self$func.logX(z.b)
            .denom <- torch_logsumexp(.num, dim = -1, keepdim = T)
            return(.num - .denom)
        },
        ## Helper function
        get.weight = function(){
            self$func.logX$weight
        })
```

## Check flow from the encoder to decoder

```{r echo = T}
lin.decoder <- build.linear.decoder(ncol(x.torch), K=7)$to(device=GPU)

x.input <- x.torch[1:5, ]
z.b <- lin.encoder(x.input)

#############################################
## reconstruction of x based on the latent ##
#############################################
logx.recon <- lin.decoder(z.b)
logx.recon[, 1:7]
```

* Note: the reconstructed data matrix is in logarithm scale

## What's really going on in terms of functions?

For each sample/cell $i$,

* Encoder:

$$\mathbf{z}_{i} \gets \mathsf{normalize}(\log(\mathbf{x}_{i} + 1)) \cdot \mathbf{W}^{(z)} + \mathbf{b}^{(z)}$$

* Decoder:

$$\hat{\mathbf{x}}_{i} \gets \mathbf{z}_{i} \cdot \mathbf{W}^{(x)} + \mathbf{b}^{(x)}$$


## Goal: to make the reconstructed data $\approx$ the input

* Gene expression frequency:
$$\rho_{ig} = \frac{\exp( \widehat{{\log}X}_{ig} )}{\sum_{g'} \exp( \widehat{{\log}X}_{ig'} )}$$

* Multinomial log-likelihood:
$$L_{i} \overset{\mathsf{def}}{=} \sum_{g=1}^{p} X_{ig} \log \rho_{ig}$$

\only<1>{$$\log \rho_{ig} = \widehat{{\log}X}_{ig} - \log \sum_{g'} \exp(\widehat{{\log}X}_{ig'})$$}

\only<2>{$$L_{i} = \sum_{g=1}^{p} X_{ig} \left[ \widehat{\log X}_{ig} - \mathsf{logSumExp}(\widehat{\log \mathbf{x}}_{i}) \right]$$}

## Goal: to make the reconstructed data $\approx$ the input

:::::: {.columns}
::: {.column width=.6}

```{r echo = T}
multinom.llik <- function(x.input, logx.recon){
    torch_sum(x.input * logx.recon, dim = -1)
}

multinom.llik(x.input, logx.recon)
```

:::
::: {.column width=.4}

* We need to maximize this with respect to all the parameters in the encoder and decoder layers.

* Maximizing log-likelihood $\iff$ minimizing *negative* log-likelihood

:::
::::::

## `Torch` provides a convenient way to "differentiate"

```{r echo = T, size="large"}
loss <- -torch_mean(multinom.llik(x.input, logx.recon))
loss
loss$backward()
```

## Put the encoder and decoder together (so that gradients flow)

```{r echo = T}
build.linear.autoenc <-
    nn_module(
        classname = "linear autoencoder",
        initialize = function(d.data, K){
            self$enc <- build.linear.encoder(d.data, K)
            self$dec <- build.linear.decoder(d.data, K)
        },
        forward = function(x){
            z <- self$enc(x)
            x.hat <- self$dec(z)
            .loss <- - multinom.llik(x, x.hat)
            list(loss = .loss)
        })
```

```{r}
linear.autoenc <- build.linear.autoenc(ncol(x.torch), K=12)
linear.autoenc$to(device=GPU)
```

## A bit more formal definition

Minimize the total loss

$$L \overset{\textsf{def}}{=} \sum_{i=1}^{n}\mathsf{Loss}(\mathbf{x}_{i}, \widehat{\mathbf{x}}_{i})$$

where

$$\hat{\mathbf{x}}_{i} = \mathsf{Decoder}(\mathsf{Encoder}(\mathbf{x}_{i}; \theta^{(\textsf{enc})}); \theta^{(\textsf{dec})})$$

with respect to the parameters $\theta$.

## Update by stochastic gradient steps

Take gradient (direction to minimize the loss) for each parameter:

$$\nabla L \equiv \left( \frac{\partial L}{\partial \theta_{1}}, \frac{\partial L}{\partial \theta_{2}}, \ldots \right)$$

Update parameters:

$$\theta^{(t)} \gets \underset{\textsf{\color{blue} the previous}}{\theta^{(t-1)}} - \underset{\textsf{\color{teal} learning rate}}{\rho_{t}} \underbrace{\nabla L}_{\textsf{\color{magenta} gradient at $t-1$}}$$

## Update by stochastic gradient steps

```{r echo = T}
## register parameters to ADAM optimizer
.params <- linear.autoenc$parameters
adam <- optim_adam(.params, lr = 1e-2)
adam$zero_grad()
```

:::::: {.columns}
::: {.column width=.5}

```{r echo = T}
x.b <- x.torch[1:3, ]      # Select minibatch data
out <- linear.autoenc(x.b) # data -> latent -> recon.
out$loss                   # loss evalulated on X.b
```

:::
::: {.column width=.5}

1. Take minibatch $X^{(b)}$

2. Recon. $\hat{X}^{(b)} = \mathsf{Dec}(\mathsf{Enc}(X^{(b)}))$

3. Evaluate $\mathsf{Loss}(X^{(b)}, \hat{X}^{(b)})$

:::
::::::

## Update by stochastic gradient steps

:::::: {.columns}
::: {.column width=.6}

```{r echo = T}
loss.b <- torch_sum(out$loss) #
loss.b$backward()             # numerically
                              # differentiate

## Before we take one SGD step
head(adam$param_groups[[1]]$params$enc.func.z.bias, 2)
adam$step() -> .
## After we take one SGD step
head(adam$param_groups[[1]]$params$enc.func.z.bias, 2)
```

:::
::: {.column width=.4}

\onslide<2->{$\star$ Aggregate training loss across samples in this minibatch:

$\sum_{i \in \mathsf{minibatch}(b)} \mathsf{loss}(\mathbf{x}_{i}, \hat{\mathbf{x}_{i}})$}

${}$

\onslide<3->{$\star$ Take gradient with respect to encoder and decoder parameters}

${}$

\onslide<4>{$\star$ Update the parameters by taking one (stochastic) gradient descent step

$\theta^{(t)} \gets \theta^{(t-1)} + \rho \nabla L$}

:::
::::::



## Training algorithm: Repeat SGD steps many times

\large

* For many epochs

    1. Sample mini batch data

	2. Evaluate loss function $L(\mathbf{x}_{i}, \hat{\mathbf{x}}_{i})$

	3. Compute gradient $\nabla_{\theta} L$

	4. Update parameters by SGD

* Report encoding results

```{r include = F}
train.model <- function(xx, model, rds.file,
                        max.epoch = 200,
                        batch.size = 128,
                        learning.rate = 1e-3,
                        record = 10){

    GPU <- torch_device("cuda")
    CPU <- torch_device("cpu")

    adam <- optim_adam(model$parameters, lr = learning.rate)

    loss.vec <- c()

    ntot <- nrow(xx)
    nbatch <- ceiling(ntot /batch.size)
    nn <- batch.size * nbatch

    hh.list <- list()
    hh <- matrix(NA, nrow = ntot, ncol = model$enc$K)

    ww.list <- list()
    ww <- matrix(NA, nrow = ncol(xx), ncol = model$enc$K)

    for(tt in seq(0, max.epoch)){
        .shuffle <- sample(ntot, nn)
        .loss.tot <- 0

        for(b in seq(0, nbatch-1)){
            .idx <- seq(1 + b * batch.size, (b+1) * batch.size)
            .idx <- .shuffle[.idx]
            xx.b <- xx[.idx, ]

            model$train()
            adam$zero_grad()
            out <- model(xx.b)
            loss.b <- torch_sum(out$loss)$to(device=CPU)$item()
            torch_mean(out$loss)$backward()
            adam$step() -> .
            .loss.tot <- .loss.tot + loss.b

            cat(tt, " ", loss.b, "\r", file = stderr())
            flush(stderr())
        }

        loss.vec <- c(loss.vec, .loss.tot/nn)

        if(tt %% record == 0){
            model$eval()

            for(b in seq(0, nbatch-1)){
                .idx <- seq(1 + b * batch.size, min(ntot, (b+1) * batch.size))
                xx.b <- xx[.idx, ]
                hh.b <- model$enc$get.latent(xx.b)$to(device=CPU)
                hh[.idx, ] <- as.matrix(hh.b)
            }

            ww <- as.matrix(model$dec$get.weight()$to(device=CPU))
            hh.list <- c(hh.list, list(hh))
            ww.list <- c(ww.list, list(ww))
        }
    }
    saveRDS(list(loss = loss.vec, latent = hh.list, weights = ww.list), file = rds.file)
}
```

```{r run_linear_autoencoder}
.file <- fig.dir %&% "/linear_autoencodeer.rds"
if.needed(.file, {
    train.model(x.torch, linear.autoenc, .file, max.epoch = 1000, record=25)
})
```

## Results: SGD minimized the loss function

```{r}
.result <- readRDS(.file)
```

```{r fig.width=5, fig.height = 3}
nepoch <- length(.result$loss)
.df <- data.frame(epoch = seq(0, nepoch - 1), loss = .result$loss)
ggplot(.df, aes(epoch, loss)) +
    geom_point(stroke = 0)
```

## Hidden representations (factor $\times$ cell)

```{r fig.width=5, fig.height=2.5, only.plot="1"}
tt <- 1
.temp <- apply(.result$latent[[tt]], 2, scale)
.matshow(.sort.matrix(t(.temp))) + ggtitle("epoch = " %&% (1 + (tt - 1) * 25))
```

```{r fig.width=5, fig.height=2.5, only.plot="2"}
tt <- 2
.temp <- apply(.result$latent[[tt]], 2, scale)
.matshow(.sort.matrix(t(.temp))) + ggtitle("epoch = " %&% (1 + (tt - 1) * 25))
```

```{r fig.width=5, fig.height=2.5, only.plot="3"}
tt <- 10
.temp <- apply(.result$latent[[tt]], 2, scale)
.matshow(.sort.matrix(t(.temp))) + ggtitle("epoch = " %&% (1 + (tt - 1) * 25))
```

```{r fig.width=5, fig.height=2.5, only.plot="4"}
tt <- 20
.temp <- apply(.result$latent[[tt]], 2, scale)
.matshow(.sort.matrix(t(.temp))) + ggtitle("epoch = " %&% (1 + (tt - 1) * 25))
```

```{r fig.width=5, fig.height=2.5, only.plot="5"}
tt <- 30
.temp <- apply(.result$latent[[tt]], 2, scale)
.matshow(.sort.matrix(t(.temp))) + ggtitle("epoch = " %&% (1 + (tt - 1) * 25))
```

```{r fig.width=5, fig.height=2.5, only.plot="6"}
tt <- 40
.temp <- apply(.result$latent[[tt]], 2, scale)
.matshow(.sort.matrix(t(.temp))) + ggtitle("epoch = " %&% (1 + (tt - 1) * 25))
```

## Latent dimensions estimated by the encoder model

```{r}
tt <- 41
.df <- as.data.frame(.result$latent[[tt]])
```

```{r fig.width=6, fig.height=2.2, only.plot="1"}
p1 <- ggplot(.df, aes(V1, V2)) + ggrastr::rasterise(geom_point(stroke=0, alpha=.5, size=.5),dpi=300)
p2 <- ggplot(.df, aes(V3, V4)) + ggrastr::rasterise(geom_point(stroke=0, alpha=.5, size=.5),dpi=300)
p3 <- ggplot(.df, aes(V5, V6)) + ggrastr::rasterise(geom_point(stroke=0, alpha=.5, size=.5),dpi=300)
p1 | p2 | p3
```

```{r fig.width=6, fig.height=2.2, only.plot="2"}
p1 <- ggplot(.df, aes(V7, V8)) + ggrastr::rasterise(geom_point(stroke=0, alpha=.5, size=.5),dpi=300)
p2 <- ggplot(.df, aes(V9, V10)) + ggrastr::rasterise(geom_point(stroke=0, alpha=.5, size=.5),dpi=300)
p3 <- ggplot(.df, aes(V11, V12)) + ggrastr::rasterise(geom_point(stroke=0, alpha=.5, size=.5),dpi=300)
p1 | p2 | p3
```

## Variational autoencoder (VAE)

:::::: {.columns}
::: {.column width=.5}

\centerline{A classical autoencoder:}

${}$

\centerline{\includegraphics[width=\linewidth]{img/scRNA_ETM/autoencoder.pdf}}

:::
::: {.column width=.5}

\centerline{Variational autoencoder:}

${}$

\centerline{\includegraphics[width=\linewidth]{img/scRNA_ETM/VAE.pdf}}

:::
::::::

\vfill

* Define relationships between variables (auto generative process)

* Usually, the decoder side captures our scientific hypothesis

<!-- * We can use an "auto-diff" algorithm (e.g., Facebook `torch` or Google `tensorflow`) to calculate gradients for the model parameters to optimize. -->

## VAE approximates Bayesian inference

:::::: {.columns}
::: {.column width=.5}

\centerline{\includegraphics[width=\linewidth]{img/scRNA_ETM/VAE_paper.pdf}}

:::
::: {.column width=.5}

\large

Prior:

$$Z \sim p(Z| \psi)$$

Data likelihood:

$$X \sim p(X|Z, \theta)$$

Variational Encoder:

$$q(Z|X) = \mathsf{NN}(X)$$

:::
::::::

## SCVI: deep generative model for scRNA-seq

\vfill

\centerline{\includegraphics[height=.65\textheight]{img/scRNA_ETM/scVI.pdf}}

\vfill

Generative model: zero-inflated negative binomial distribution

\vfill
\tiny

Lopez, .., Jordan, Yosef, \emph{Nature Methods} (2018)

## A new encoder as a posterior inference machine

\centerline{\includegraphics[width=.45\linewidth]{img/scRNA_ETM/VAE.pdf}}

We need to define functions (neural networks) that maps from the data vector $\mathbf{x}$ to

* the mean of latent embedding: $\mu$

* the log variance of latent embedding: $\sigma$

* $z_{ig} \gets \mu_{ig} + \sigma_{ig} \epsilon,\quad \epsilon \sim \mathcal{N}\!\left(0, 1\right)$

## A new encoder as a posterior inference machine

```{r echo = T}
build.vae.encoder <- nn_module(
        classname = "vae encoder",
        initialize = function(d.in, K){
            self$K <- K                                        # number of hidden vars.
            self$z.mean <- nn_linear(d.in, K)                  # mean function
            self$z.logvar <- nn_linear(d.in, K)                # log variance function
            self$bn <- nn_batch_norm1d(d.in)                   # batch norm
        },
        forward = function(x.b){
            x.b <- self$normalize(x.b)
            mm <- self$z.mean(x.b)                             # mean evaluated
            lv <- torch_clamp(self$z.logvar(x.b), -4.0, 4.0)   # log-var evaluated
            z <- mm + torch_randn_like(lv) * torch_exp(lv/ 2.) # stochastic z
            list(z = z, z.mean = mm, z.logvar = lv)
        },
        normalize = function(x.b){                             # normalization
            x.b <- torch_log1p(x.b)                            # log1p transformation
            self$bn(x.b)                                       # to expedite training
        }, ## helper function
        get.latent = function(x.b){ self$z.mean(self$normalize(x.b)) })
```

## We will have two types of loss functions

:::::: {.columns}
::: {.column width=.5}

Data log likelihood (a generative model)

$$\log \prod_{i=1}^{n} \underset{\textsf{\color{teal} multinomial}}{p(\mathbf{x}_{i}|\mathbf{z}_{i})}$$

```{r echo = T}
multinom.llik <- function(x.input, logx.recon){
    torch_sum(x.input * logx.recon, dim = -1)
}
```

:::
::: {.column width=.5}

Divergence between prior and posterior

$$D_{\mathsf{KL}}\left(\underset{\textsf{\color{teal} posterior}}{q(\mathbf{z})} \middle\| \underset{\textsf{\color{magenta} prior}}{p(\mathbf{z})}\right)$$

```{r echo = T}
kl.loss <- function(.mean, .lnvar) {
    -0.5 * torch_sum(1. + .lnvar
                     - torch_pow(.mean, 2.)
                     - torch_exp(.lnvar),
                     dim = -1)
}
```

:::
::::::

:::::: {.columns}
::: {.column width=.5}

* The log-likelihood is the same as before

* KL loss will work like regularization

:::
::: {.column width=.5}

* We assume both $q$ and $p$ follows Gaussian distribution

:::
::::::

## A complete definition of our VAE model

```{r echo = T}
build.vae <-
    nn_module(
        classname = "variational autoencoder",
        initialize = function(d.data, K){
            self$enc <- build.vae.encoder(d.data, K)    # encoder model
            self$dec <- build.linear.decoder(d.data, K) # decoder model
        },
        forward = function(x){
            .enc <- self$enc(x)                         #
            x.hat <- self$dec(.enc$z)                   # reconstruction
            .llik <- multinom.llik(x, x.hat)            # data likelihood
            .kl <- kl.loss(.enc$z.mean, .enc$z.logvar)  # KL divergence
            .loss <- .kl - .llik                        # combined loss
            list(loss = .loss, kl=.kl)
        }
    )
```

* What do you want to change?

## VAE: Check flow from the encoder to decoder

:::::: {.columns}
::: {.column width=.5}

```{r echo = T}
vae <- build.vae(ncol(x.torch), K=12)
vae$to(device = GPU)

x.input <- x.torch[1:5, ]
out <- vae(x.input)
out$loss
```

:::
::: {.column width=.5}

```{r echo = T}
#############################################
## reconstruction of x based on the latent ##
#############################################
z.b <- vae$enc(x.input)
logx.recon <- vae$dec(z.b$z)
logx.recon[, 1:5]
```

:::
::::::

* Note: the reconstructed data matrix is in logarithm scale

```{r}
.file <- fig.dir %&% "/vae.rds"
if.needed(.file, {
    train.model(x.torch, vae, .file, max.epoch = 1000, record=21)
})
```

## Results: SGD minimized the VAE loss function

```{r}
.result <- readRDS(.file)
```

```{r fig.width=5, fig.height = 3}
nepoch <- length(.result$loss)
.df <- data.frame(epoch = seq(0, nepoch - 1), loss = .result$loss)
ggplot(.df, aes(epoch, loss)) +
    geom_point(stroke = 0)
```

## Latent dimensions in the VAE model

```{r}
tt <- 41
.df <- as.data.frame(.result$latent[[tt]])
```

```{r fig.width=6, fig.height=2.2, only.plot="1"}
p1 <- ggplot(.df, aes(V1, V2)) + ggrastr::rasterise(geom_point(stroke=0, alpha=.5, size=.5),dpi=300)
p2 <- ggplot(.df, aes(V3, V4)) + ggrastr::rasterise(geom_point(stroke=0, alpha=.5, size=.5),dpi=300)
p3 <- ggplot(.df, aes(V5, V6)) + ggrastr::rasterise(geom_point(stroke=0, alpha=.5, size=.5),dpi=300)
p1 | p2 | p3
```

```{r fig.width=6, fig.height=2.2, only.plot="2"}
p1 <- ggplot(.df, aes(V7, V8)) + ggrastr::rasterise(geom_point(stroke=0, alpha=.5, size=.5),dpi=300)
p2 <- ggplot(.df, aes(V9, V10)) + ggrastr::rasterise(geom_point(stroke=0, alpha=.5, size=.5),dpi=300)
p3 <- ggplot(.df, aes(V11, V12)) + ggrastr::rasterise(geom_point(stroke=0, alpha=.5, size=.5),dpi=300)
p1 | p2 | p3
```

## Hidden representations (factor $\times$ cell)

```{r fig.width=5, fig.height=2.5, only.plot="1"}
tt <- 1
.temp <- apply(.result$latent[[tt]], 2, scale)
.matshow(.sort.matrix(t(.temp))) + ggtitle("epoch = " %&% (1 + (tt - 1) * 25))
```

```{r fig.width=5, fig.height=2.5, only.plot="2"}
tt <- 2
.temp <- apply(.result$latent[[tt]], 2, scale)
.matshow(.sort.matrix(t(.temp))) + ggtitle("epoch = " %&% (1 + (tt - 1) * 25))
```

```{r fig.width=5, fig.height=2.5, only.plot="3"}
tt <- 10
.temp <- apply(.result$latent[[tt]], 2, scale)
.matshow(.sort.matrix(t(.temp))) + ggtitle("epoch = " %&% (1 + (tt - 1) * 25))
```

```{r fig.width=5, fig.height=2.5, only.plot="4"}
tt <- 20
.temp <- apply(.result$latent[[tt]], 2, scale)
.matshow(.sort.matrix(t(.temp))) + ggtitle("epoch = " %&% (1 + (tt - 1) * 25))
```

```{r fig.width=5, fig.height=2.5, only.plot="5"}
tt <- 30
.temp <- apply(.result$latent[[tt]], 2, scale)
.matshow(.sort.matrix(t(.temp))) + ggtitle("epoch = " %&% (1 + (tt - 1) * 25))
```

```{r fig.width=5, fig.height=2.5, only.plot="6"}
tt <- 40
.temp <- apply(.result$latent[[tt]], 2, scale)
.matshow(.sort.matrix(t(.temp))) + ggtitle("epoch = " %&% (1 + (tt - 1) * 25))
```

## Annotate factors to cell types by enrichment (`fgsea`)

```{r run_gsea_beta}
.beta <- apply(.result$weights[[tt]], 2, scale)
.rows <- fread(raw.data$row, header=F)
.rows[, c("gene","chr") := tstrsplit(`V1`, split="[_]+", keep=1:2)]
rownames(.beta) <- .rows$gene

.db <- read.panglao("Pancreas")
.genes <- unique(unlist(.db))
.dt <- sort.beta(.beta, genes.selected = .genes)
.gsea <- run.beta.gsea(.dt, .db)
.show <- select.gsea.beta(.gsea, .dt, sd.cutoff = 1e-4)
```

```{r}
p1 <-
    .gg.plot(.show$beta, aes(row, variable, fill=`value`)) +
    theme(axis.text.x = element_blank()) +
    theme(axis.ticks.x = element_blank()) +
    theme(legend.position = "none") +
    ggrastr::rasterise(geom_tile(), dpi=300) +
    ylab("VAE factors") + xlab("genes") +
    scale_fill_distiller(direction=-1, palette="RdBu")

p2 <-
    .gg.plot(.show$leading.edges, aes(row, ct)) +
    theme(axis.text.x = element_blank()) +
    theme(axis.ticks.x = element_blank()) +
    theme(panel.grid = element_blank()) +
    geom_tile(linewidth=0) +
    ylab("cell types")

p3 <-
    .gg.plot(.show$gsea, aes(variable, ct, fill=-log10(pmax(padj, 1e-4)))) +
    xlab("NMF factors") + ylab("") +
    scale_y_discrete(position = "right") +
    geom_tile(linewidth=.1, color="black") +
    theme(legend.key.width=unit(.2, "lines")) +
    theme(legend.text = element_text(size=4)) +
    scale_fill_distiller("Padj", palette = "YlGnBu", direction=1, labels=function(x) num.sci(10^(-x)))
```

```{r fig.width=6, fig.height=3, only.plot="1"}
print(p1/p2)
```

```{r fig.width=6, fig.height=3, only.plot="2"}
C <- length(unique(.show$gsea$ct))
K <- length(unique(.show$beta$variable))
p0 <- ggplot() + theme_void()
wrap_plots(p1, p0, p2, p3, nrow = 2, heights=c(K,C), widths=c(4,1))
```

##

\large

Can we improve model interpretation?

\begin{enumerate}
\item<2-> The decoder part is open to modelling in many different ways.
\item<3> Can we define latent factors by gene expression frequencies?
\end{enumerate}

## Single-cell Embedded Topic Model

\centerline{\includegraphics[width=\linewidth]{img/scRNA_ETM/scETM.pdf}}

\vfill

Zhao .. Li, Nature Comm. (2021)

## Document topic modelling

\centerline{\includegraphics[width=.8\linewidth]{img/scRNA_ETM/BleiTopicLDA_0.pdf}}

\vfill
\tiny
\flushright
Slide credit: David Blei

## Word frequencies define topics in documents

\centerline{\includegraphics[width=.8\linewidth]{img/scRNA_ETM/BleiTopicLDA_1.pdf}}

\vfill
\tiny
\flushright
Slide credit: David Blei

## Multinomial topic model for scRNA-seq data

Can we simply model scRNA-seq counts by multinomial distribution?

\only<1>{
\centerline{\includegraphics[height=.4\textheight]{img/scRNA_ETM/ETM.pdf}}
}

\vfill

\only<2>{
Likelihood model:

$$\mathcal{L} = \prod_{i=1}^{n}\prod_{g=1}^{\textsf{genes}} \left(\sum_{k} H_{ik} \beta_{kg}\right)^{X_{ij}}$$
}

\only<3>{
Likelihood model:

$$\mathcal{L} = \prod_{i=1}^{n}\prod_{g=1}^{\textsf{genes}} \left(\underset{\textsf{\color{magenta} a gene $g$'s probability in a cell $i$} \equiv \rho_{ig}}{\sum_{k} H_{ik} \beta_{kg}}\right)^{X_{ij}}$$}

\normalsize

* $X_{ig}$: gene expression of a gene $g$ in a single cell $i$

* $H_{ik}$: latent topic proportion of a cell $i$ to a topic $k$

* $\beta_{kg}$: topic $k$-specific gene probability


## Topic modelling for single-cell data

\vfill

\centerline{\includegraphics[height=.4\textheight]{img/scRNA_ETM/ETM.pdf}}

\vfill

:::::: {.columns}
::: {.column width=.5}

Probability of gene $g$ in  a cell $i$:

$$\rho_{ig} = \sum_{k \in \textsf{topics}} H_{ik} \beta_{kg}$$

:::
::: {.column width=.45}

By **not** normalizing the probability of each cell, we do not worry about modelling sequencing depths.

:::
::::::


## Document topic modelling vs. single-cell ETM

```{r echo = FALSE, results="asis"}
.dt <- data.table(
    variables = c("$D$",
                  "$d$",
                  "$N_{d}$",
                  "$j$",
                  "$K$",
                  "$k$",
                  "$V$",
                  "$v$",
                  "$W_{dj}^{v}$",
                  "$X_{dv}$"),
    `in document topic model` = c("Total number of documents (corpus)",
                    "Document index",
                    "Number of words in a document $d$",
                    "Word index, $j \\in [N_{d}]$",
                    "Total number of topics",
                    "Topic index, $k \\in [K]$",
                    "Size of vocabulary",
                    "Vocabulary index $v \\in [V]$",
                    "Indicator for a word to vocabulary $\\in \\{0, 1\\}$",
                    "Vocabulary $v$ occurrence in a document $d$"),
    `in single cell ETM` = c("Total number of cells",
                   "Cell index",
                   "Number of read counts in a cell $d$",
                   "Read index",
                   "Total number of cell type topics",
                   "Cell topic index",
                   "Total number of genes",
                   "Gene index",
                   "Indicator for a read to a gene $\\in \\{0, 1\\}$",
                   "Gene expression of a gene $v$ in a cell $d$ $\\in [0, N_{d}]$"))

knitr::kable(.dt)
```


##

```{r echo = FALSE, results = "asis"}
.dt.param <- data.table(
    variables = c("$Z_{dj}^{k}$",
                  "$H_{dk}$",
                  "$\\beta_{kv}$"),
    `in document topic model` = c("Indicator for assigning a word to a topic $k$",
                         "Hidden state $k$ of a document $d$",
                         "topic $k$-specific vocabulary $v$ frequency"),
    `in single cell ETM` = c("Indicator for assigning a read to a topic $k$",
                         "Hidden state $k$ of a cell $d$",
                         "topic $k$-specific, a gene $v$'s exression"))

knitr::kable(.dt.param)
```

* In Latent Dirichlet Allocation: $\sum_{k=1}^{K} H_{dk} = 1$ and $H_{dk} > 0$, and $\mathbf{h}_{d} \sim \mathsf{Dirichlet}(\alpha/K, \ldots, \alpha/K)$ _a priori_. Approximately, we have $\hat{H}_{dk} = \sum_{j}^{N_{d}} Z_{dj}^{k} / N_{d}$.

* In Embedded Topic model, $H_{dk}$ with the simplex constraints; $H_{dk} = \exp(\delta_{dk}) / \sum_{k'} \exp(\delta_{dk'})$ where $\delta_{dk} \sim \mathcal{N}\!\left(0,1\right)$ _a priori_.

* Additional constraints: $\beta_{kv} > 0$ and $\sum_{v} \beta_{kv} = 1$, meaning that only a handful of vocabulary $v$ contribute to a topic $k$.

## Let's modify the decoder part

```{r echo = T}
build.etm.decoder <-
    nn_module(classname = "ETM decoder",
        initialize = function(n.out, K, jitter = 1e-2) {
            self$lbeta <- nn_parameter(torch_randn(K, n.out) * jitter)
            self$beta <- nn_log_softmax(2) # topic x variant (softmax for each topic)
            self$hid <- nn_log_softmax(2)  # sample x topic (softmax for each sample)
        },
        ## Define how do get back high-dim data
        forward = function(z.b, eps = 1e-8){
            .beta <- self$get.weight()
            h.b <- self$hid(z.b)
            torch_log(torch_mm(torch_exp(h.b), torch_exp(.beta)) + eps)
        },
        ## Helper function
        get.weight = function(){
            self$beta(self$lbeta)
        })
```

## ETM: putting the encoder and decoder together

```{r echo = T}
build.etm <-
    nn_module(
        classname = "embedded topic model",
        initialize = function(d.data, K){
            self$enc <- build.vae.encoder(d.data, K)
            self$dec <- build.etm.decoder(d.data, K)
        },
        forward = function(x){
            .enc <- self$enc(x)
            x.hat <- self$dec(.enc$z)
            .llik <- multinom.llik(x, x.hat)
            .kl <- kl.loss(.enc$z.mean, .enc$z.logvar)
            .loss <- .kl - .llik
            list(loss = .loss, kl = .kl, latent = .enc$z.mean)
        })
```

## ETM: Check flow from the encoder to decoder

:::::: {.columns}
::: {.column width=.5}

```{r echo = T}
etm <- build.etm(ncol(x.torch), K=12)
etm$to(device = GPU)

x.input <- x.torch[1:5, ]
out <- etm(x.input)

out <- vae(x.input)
out$loss
```

:::
::: {.column width=.5}

```{r echo = T}
#############################################
## reconstruction of x based on the latent ##
#############################################
z.b <- etm$enc(x.input)
logx.recon <- etm$dec(z.b$z)
logx.recon[, 1:5]
```

:::
::::::

* Note: the reconstructed data matrix is in logarithm scale

## Results: SGD minimized the loss function

```{r}
.file <- fig.dir %&% "/etm.rds"
if.needed(.file, {
    train.model(x.torch, etm, .file, max.epoch = 2000, record=50)
})
```

```{r}
.result <- readRDS(.file)
```

```{r fig.width=5, fig.height = 3}
nepoch <- length(.result$loss)
.df <- data.frame(epoch = seq(0, nepoch - 1), loss = .result$loss)
ggplot(.df, aes(epoch, loss)) +
    geom_point(stroke = 0)
```

```{r}
show.struct <- function(.zz){
    .hh <- t(apply(t(.zz), 2, function(z) {
        pr <- exp(z - max(z))
        pr/sum(pr)
        }))
    .K <- ncol(.zz)
    .dt <- reshape2::melt(.hh) %>%
        mutate(row = Var2, col = Var1, weight=value) %>%
        col.order(.ro=rev(1:.K), ret.tab=TRUE) %>%
        as.data.table()

    .gg.plot(.dt, aes(`col`, `weight`, fill=factor(`Var2`, 1:.K))) +
        ylab("topic proportion") +
        xlab("cells") +
        theme(legend.position = "top") +
        theme(axis.text.x = element_blank()) +
        theme(axis.ticks.x = element_blank()) +
        geom_bar(stat="identity", linewidth=0) +
        scale_fill_brewer("latent topic", palette = "Paired")
}
```


## Latent dimensions in the ETM model

```{r}
tt <- 41
.df <- as.data.frame(.result$latent[[tt]])
```

```{r fig.width=6, fig.height=2.2, only.plot="1"}
p1 <- ggplot(.df, aes(V1, V2)) + ggrastr::rasterise(geom_point(stroke=0, alpha=.5, size=.5),dpi=300)
p2 <- ggplot(.df, aes(V3, V4)) + ggrastr::rasterise(geom_point(stroke=0, alpha=.5, size=.5),dpi=300)
p3 <- ggplot(.df, aes(V5, V6)) + ggrastr::rasterise(geom_point(stroke=0, alpha=.5, size=.5),dpi=300)
p1 | p2 | p3
```

```{r fig.width=6, fig.height=2.2, only.plot="2"}
p1 <- ggplot(.df, aes(V7, V8)) + ggrastr::rasterise(geom_point(stroke=0, alpha=.5, size=.5),dpi=300)
p2 <- ggplot(.df, aes(V9, V10)) + ggrastr::rasterise(geom_point(stroke=0, alpha=.5, size=.5),dpi=300)
p3 <- ggplot(.df, aes(V11, V12)) + ggrastr::rasterise(geom_point(stroke=0, alpha=.5, size=.5),dpi=300)
p1 | p2 | p3
```

## In ETM, the hidden dimensions are not independent

\large

$$H_{ik} = \frac{\exp(Z_{ik})}{\sum_{k'} \exp(Z_{ik'})}$$


## Hidden representations (mixing proportions)

```{r fig.width=5, fig.height=2.5, only.plot="1"}
tt <- 1
.temp <- .result$latent[[tt]]
show.struct(.temp) + ggtitle("epoch = " %&% (1 + (tt - 1) * 50))
```

```{r fig.width=5, fig.height=2.5, only.plot="2"}
tt <- 2
.temp <- .result$latent[[tt]]
show.struct(.temp) + ggtitle("epoch = " %&% (1 + (tt - 1) * 50))
```

```{r fig.width=5, fig.height=2.5, only.plot="3"}
tt <- 5
.temp <- .result$latent[[tt]]
show.struct(.temp) + ggtitle("epoch = " %&% (1 + (tt - 1) * 50))
```

```{r fig.width=5, fig.height=2.5, only.plot="4"}
tt <- 10
.temp <- .result$latent[[tt]]
show.struct(.temp) + ggtitle("epoch = " %&% (1 + (tt - 1) * 50))
```

```{r fig.width=5, fig.height=2.5, only.plot="5"}
tt <- 20
.temp <- .result$latent[[tt]]
show.struct(.temp) + ggtitle("epoch = " %&% (1 + (tt - 1) * 50))
```

```{r fig.width=5, fig.height=2.5, only.plot="6"}
tt <- 30
.temp <- .result$latent[[tt]]
show.struct(.temp) + ggtitle("epoch = " %&% (1 + (tt - 1) * 50))
```

```{r fig.width=5, fig.height=2.5, only.plot="7"}
tt <- 41
.temp <- .result$latent[[tt]]
show.struct(.temp) + ggtitle("epoch = " %&% (1 + (tt - 1) * 50))
```

## Annotate factors to cell types by enrichment (`fgsea`)

```{r run_gsea_etm_beta}
.beta <- exp(t(.result$weights[[tt]]))
.rows <- fread(raw.data$row, header=F)
.rows[, c("gene","chr") := tstrsplit(`V1`, split="[_]+", keep=1:2)]
rownames(.beta) <- .rows$gene

.db <- read.panglao("Pancreas")
.genes <- unique(unlist(.db))
.dt <- sort.beta(.beta, genes.selected = .genes)
.gsea <- run.beta.gsea(.dt, .db)
.show <- select.gsea.beta(.gsea, .dt, sd.cutoff = 1e-4)
```

```{r}
p1 <-
    .gg.plot(.show$beta, aes(row, variable, fill=pmin(`value`, 1e-2))) +
    theme(axis.text.x = element_blank()) +
    theme(axis.ticks.x = element_blank()) +
    theme(legend.position = "none") +
    ggrastr::rasterise(geom_tile(), dpi=300) +
    ylab("ETM topics") + xlab("genes") +
    scale_fill_distiller(direction=1, palette="PuRd", trans="sqrt")

p2 <-
    .gg.plot(.show$leading.edges, aes(row, ct)) +
    theme(axis.text.x = element_blank()) +
    theme(axis.ticks.x = element_blank()) +
    theme(panel.grid = element_blank()) +
    geom_tile(linewidth=0) +
    ylab("cell types")

p3 <-
    .gg.plot(.show$gsea, aes(variable, ct, fill=-log10(pmax(padj, 1e-4)))) +
    xlab("NMF factors") + ylab("") +
    scale_y_discrete(position = "right") +
    geom_tile(linewidth=.1, color="black") +
    theme(legend.key.width=unit(.2, "lines")) +
    theme(legend.text = element_text(size=4)) +
    scale_fill_distiller("Padj", palette = "YlGnBu", direction=1, labels=function(x) num.sci(10^(-x)))
```

```{r fig.width=6, fig.height=3, only.plot="1"}
print(p1/p2)
```

```{r fig.width=6, fig.height=3, only.plot="2"}
C <- length(unique(.show$gsea$ct))
K <- length(unique(.show$beta$variable))
p0 <- ggplot() + theme_void()
wrap_plots(p1, p0, p2, p3, nrow = 2, heights=c(K,C), widths=c(4,1))
```

## Discussions on latent topic modelling

* Most cells predominantly belong to one topic (one colour). Why?

* If we model cells as a mixture of cell topics, we can capture doublets or triplets.

* The underlying generative model assumes no sequencing depth! This can help avoid batch-specific differences in practice.

* VAE offers a flexible framework with which our scientific hypothesis can be formulated in a probabilistic language (`torch`).

* Potentially, this purely-unsupervised learning framework can be combined with supervised, semi-supervised learning models.

