---
title: "Regulatory Genomics"
author: "Yongjin Park"
date: "`r format(Sys.time(), '%d %B %Y')`"
classoption: "aspectratio=169"
fontsize: 12pt
bibliography: regulatory-genomics.bib
output:
    beamer_presentation:
        theme: Madrid
        keep_md: true
        keep_tex: true
        latex_engine: xelatex
        slide_level: 2
header-includes:
  - \usepackage{cancel}
  - \usepackage{tikz}
  - \definecolor{UBCblue}{rgb}{0.04706, 0.13725, 0.26667}
  - \usecolortheme[named=UBCblue]{structure}
  - \setbeamertemplate{frametitle}{\color{UBCblue}\bfseries\insertframetitle\par\vskip-6pt\hrulefill}
  - \setbeamercolor{title in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{section in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{author in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{date in head/foot}{bg=white,fg=gray}
  - \setbeamertemplate{page number in head/foot}{}
  - \setbeamertemplate{frame numbering}[none]
  - \setbeamercolor{alerted text}{bg=yellow}
  - \AtBeginSection[]{\begin{frame}\frametitle{Today's lecture}{\Large\tableofcontents[currentsection]}\end{frame}}
  - |
    \makeatletter
    \def\ps@titlepage{%
      \setbeamertemplate{footline}{}
    }
    \addtobeamertemplate{title page}{\thispagestyle{titlepage}}{}
    \makeatother
---


```{r setup, include=FALSE}
source("Setup.R")
fig.dir <- "../Fig/regulatory-genomics/"
dir.create(fig.dir, showWarnings=FALSE)
setup.env(fig.dir)
dir.create("../data", showWarnings=FALSE)

library(data.table)
library(tidyverse)
library(patchwork)
library(matrixTests)

theme_set(theme_classic() + theme(title = element_text(size=10)))
```

## Roadmap of STAT/BIOF/GSAT540 2026

_data generation_ followed by _methodology_

\vfill

\large

:::::: {.columns}
::: {.column width=.45}

3. Formulating biological questions

5. RNA-seq data

7. Experimental design

8. Single-cell genomics

9. Matrix factorization

:::
::: {.column width=.55}

10. GWAS

12. \textbf{\color{magenta} Regulatory genomics} (Today)

13. Regression (for reg. genomics)

14. Biological networks

15. Clustering

:::
::::::

\normalsize

# Technologies for measuring regulatory activities

## Chromatin Immunoprecipitation + Sequencing


**Goal**: Identify where proteins (e.g., TFs) bind to DNA genome-wide

**Experimental Protocols**:

1. Cross-link proteins to DNA
2. Fragment chromatin
3. Immunoprecipitate with antibody
4. Sequence DNA fragments
5. Map reads to genome
6. **Call peaks**

<!-- \begin{center} -->
<!-- \includegraphics[width=0.9\textwidth]{img/chipseq-protein.pdf} -->
<!-- \end{center} -->

## How ChIP-seq works

:::::: {.columns}
::: {.column width=.23}

\includegraphics[width=\textwidth]{img/chipseq-protein.pdf}

:::
::: {.column width=.23}

\onslide<2->{
\includegraphics[width=\textwidth]{img/chipseq-histone.pdf}
}

:::
::: {.column width=.23}

\onslide<3->{
\includegraphics[width=\textwidth]{img/chipseq-dnase.pdf}
}

:::
::: {.column width=.2}

* Binding

* Sequencing

$${}$$

\flushright

@Furey2012-ir

:::
::::::


## How ChIP-seq works

\only<1>{
\centerline{\includegraphics[height=.78\textheight]{img/chipseq-park-2009.pdf}}
}

\only<2>{
\centerline{\includegraphics[height=.78\textheight]{img/chipseq-park-2009-2.pdf}}
}

\flushright
\scriptsize
[@Park2009-bn]

## Peak calling: signal vs. background

:::::: {.columns}
::: {.column width=.5}

**Key idea**: Find genomic regions with more reads than expected by chance

- Divide genome into bins (e.g., 200 bp)
- Count reads in each bin
- Compare ChIP vs Input (control)
- Statistical test: Is enrichment significant?

:::
::: {.column width=.5}

**Common approaches**:

- **MACS2**: Poisson model, local background
- **normR**: Binomial mixture model
- **DiffBind**: Differential binding analysis

**Multiple testing**: FDR correction across all bins!

:::
::::::

## ChIP-seq Peak Calling Exercise {.allowframebreaks}

```{r include=F}
## Install Bioconductor packages (run once)
## BiocManager::install(c("normr", "GenomicRanges", "rtracklayer",
##                        "AnnotationHub", "Gviz"))
library(normr)
library(GenomicRanges)
set.seed(42)
```

```{r chipseq-data, echo=TRUE, cache=FALSE, message=FALSE}
library(AnnotationHub)
library(rtracklayer)

data_dir <- "../data/chipseq/"
dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)

chr <- "chr21" # Focus on chr21 to keep data manageable

# Download ENCODE CTCF ChIP peaks
peaks_file <- paste0(data_dir, "ctcf_gm12878_peaks.rds")
if.needed(peaks_file, {
    cat("Downloading ENCODE CTCF peaks...\n")
    ah <- AnnotationHub()
    query_peaks <- query(ah, c("CTCF", "GM12878", "narrowPeak", "ENCODE"))
    chip_peaks_all <- query_peaks[[1]]
    chip_peaks <- chip_peaks_all[seqnames(chip_peaks_all) == chr]
    saveRDS(chip_peaks, file = peaks_file)
})
chip_peaks <- readRDS(peaks_file)
```

```{r chipseq-data-2, echo=TRUE, cache=FALSE}
# Download ENCODE DNase-seq peaks (open chromatin for comparison)
dnase_file <- paste0(data_dir, "dnase_gm12878_peaks.rds")
if (!file.exists(dnase_file)) {
    cat("Downloading ENCODE DNase-seq data...\n")
    ah <- AnnotationHub()
    query_dnase <- query(ah, c("DNase", "GM12878", "narrowPeak", "ENCODE"))
    if (length(query_dnase) > 0) {
        dnase_peaks_all <- query_dnase[[1]]
        dnase_peaks <- dnase_peaks_all[seqnames(dnase_peaks_all) == chr]
        saveRDS(dnase_peaks, file = dnase_file)
        cat(sprintf("Saved %d DNase peaks\n", length(dnase_peaks)))
    } else {
        cat("No DNase data found\n")
        saveRDS(GRanges(), file = dnase_file)
    }
}
dnase_peaks <- readRDS(dnase_file)
```

<!-- ```{r chipseq-data-3, echo=TRUE, cache=FALSE} -->
<!-- cat(sprintf("Loaded %d DNase peaks on %s\n", length(dnase_peaks), chr)) -->
<!-- cat(sprintf("Loaded %d CTCF peaks on %s\n", length(chip_peaks), chr)) -->
<!-- ``` -->

## Peak Calling with `normr` {.allowframebreaks}

:::::: {.columns}
::: {.column width=.65}

```{r chipseq-peaks, echo=TRUE, eval=FALSE}
## Create genomic bins
binsize <- 1000
bins <- GRanges(chr, IRanges(start = seq(1, 48e6, binsize),
                             width = binsize))

## Count reads in bins
chip_counts <- countOverlaps(bins, chip_reads)
dnase_counts <- countOverlaps(bins, dnase_reads)

## Call peaks using normR
fit <- normr::enrichR(treatment = chip_counts,
                      control = dnase_counts,
                      genome = bins,
                      verbose = FALSE)

## Get significant peaks (FDR < 0.05)
peaks_called <- normr::getRanges(fit)
peaks_sig <- peaks_called[peaks_called$qvalue < 0.05]
```

:::
::: {.column width=.3}


* Foreground (CTCF) vs. background (control, DNase)

* Enrichment-based Binomial Tests

* Input: piled signals

* Output: genomic regions (BED)

* Narrow vs. broad peaks



:::
::::::



```{r include=F}
peak_widths <- width(chip_peaks)
peak_signals <- mcols(chip_peaks)$signalValue

## cat(sprintf("ENCODE Peak Statistics:\n"))
## cat(sprintf("  Number of peaks: %d\n", length(chip_peaks)))
## cat(sprintf("  Median width: %d bp\n", median(peak_widths)))
## cat(sprintf("  Mean width: %d bp\n", round(mean(peak_widths))))
## cat(sprintf("  Median signal: %.2f\n", median(peak_signals, na.rm=TRUE)))

# Create simulated coverage based on peaks
set.seed(42)
coverage_file <- paste0(data_dir, "ctcf_coverage.rds")

if.needed(coverage_file, {
    # Generate ChIP reads from peaks
    chip_reads_list <- lapply(1:length(chip_peaks), function(i) {
        peak <- chip_peaks[i]
        signal <- mcols(peak)$signalValue
        n_reads <- round(signal * 10)
        if (n_reads == 0) return(GRanges())
        peak_center <- start(peak) + width(peak)/2
        read_pos <- rnorm(n_reads, mean = peak_center, sd = width(peak)/4)
        read_pos <- round(pmax(start(peak), pmin(end(peak), read_pos)))
        GRanges(chr, IRanges(read_pos, width = 36))
    })
    chip_reads <- do.call(c, chip_reads_list)

    # Generate DNase reads (open chromatin)
    if (length(dnase_peaks) > 0) {
        dnase_reads_list <- lapply(1:length(dnase_peaks), function(i) {
            peak <- dnase_peaks[i]
            signal <- mcols(peak)$signalValue
            n_reads <- round(signal * 3)
            if (n_reads == 0) return(GRanges())
            peak_center <- start(peak) + width(peak)/2
            read_pos <- rnorm(n_reads, mean = peak_center, sd = width(peak)/3)
            read_pos <- round(pmax(start(peak), pmin(end(peak), read_pos)))
            GRanges(chr, IRanges(read_pos, width = 36))
        })
        dnase_reads <- do.call(c, dnase_reads_list)
    } else {
        dnase_reads <- GRanges()
    }

    saveRDS(list(chip = chip_reads, dnase = dnase_reads), file = coverage_file)
})

coverage_data <- readRDS(coverage_file)
chip_reads <- coverage_data$chip
dnase_reads <- coverage_data$dnase
cat(sprintf("  Simulated %d CTCF ChIP reads\n", length(chip_reads)))
cat(sprintf("  Simulated %d DNase reads\n", length(dnase_reads)))

# Download ChromHMM chromatin state segmentation
chromhmm_file <- paste0(data_dir, "chromhmm_gm12878.rds")

if.needed(chromhmm_file, {
    cat("Downloading ChromHMM data from ENCODE...\n")
    url <- "http://hgdownload.soe.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeBroadHmm/wgEncodeBroadHmmGm12878HMM.bed.gz"
    temp_file <- tempfile(fileext = ".bed.gz")
    download.file(url, temp_file, quiet = TRUE)
    chromhmm_all <- import(temp_file, format = "bed")
    chromhmm_chr21 <- chromhmm_all[seqnames(chromhmm_all) == chr]
    saveRDS(chromhmm_chr21, file = chromhmm_file)
})

chromhmm <- readRDS(chromhmm_file)
cat(sprintf("Loaded %d chromatin states on %s\n", length(chromhmm), chr))
```

## Visualizing CTCF Binding Sites (using `Gviz`)

```{r chipseq-viz, echo=FALSE, cache=FALSE, fig.width=6, fig.height=3, message=FALSE, warning=FALSE}
library(Gviz)

# Find a region with multiple strong peaks
peaks_sorted <- chip_peaks[order(mcols(chip_peaks)$signalValue, decreasing=TRUE)]
top_peak <- peaks_sorted[3]
region_start <- max(1, start(top_peak) - 100000)
region_end <- start(top_peak) + 100000

# Get peaks in this region
region_peaks <- chip_peaks[start(chip_peaks) >= region_start &
                           end(chip_peaks) <= region_end]

# Bin reads for both CTCF ChIP and DNase
bin_size <- 200
bins <- seq(region_start, region_end, by = bin_size)

# CTCF ChIP signal
region_chip <- chip_reads[start(chip_reads) >= region_start &
                          end(chip_reads) <= region_end]
chip_counts <- countOverlaps(GRanges(chr, IRanges(bins, width = bin_size)), region_chip)
chip_gr <- GRanges(chr, IRanges(bins, width = bin_size), score = chip_counts)

# DNase signal (exaggerated 3x for visibility)
region_dnase <- dnase_reads[start(dnase_reads) >= region_start &
                            end(dnase_reads) <= region_end]
dnase_counts <- countOverlaps(GRanges(chr, IRanges(bins, width = bin_size)), region_dnase)
dnase_gr <- GRanges(chr, IRanges(bins, width = bin_size), score = dnase_counts * 3)

# Create genome browser tracks
axis_track <- GenomeAxisTrack()

chip_track <- DataTrack(chip_gr, genome = "hg19", type = "histogram",
                        chromosome = chr, name = "CTCF",
                        fill = "darkblue", col.histogram = "darkblue")

dnase_track <- DataTrack(dnase_gr, genome = "hg19", type = "histogram",
                         chromosome = chr, name = "DNase (3x)",
                         fill = "orange", col.histogram = "orange")

peak_track <- AnnotationTrack(region_peaks, name = "CTCF Peaks",
                              fill = "red", col = "red", stacking = "dense")

## # ChromHMM chromatin state track
## region_chromhmm <- chromhmm[start(chromhmm) >= region_start &
##                             end(chromhmm) <= region_end]

## # Define colors for chromatin states (ENCODE standard colors)
## state_colors <- c("1_Active_Promoter" = "red",
##                   "2_Weak_Promoter" = "orange",
##                   "3_Poised_Promoter" = "purple",
##                   "4_Strong_Enhancer" = "yellow",
##                   "5_Strong_Enhancer" = "yellow",
##                   "6_Weak_Enhancer" = "lightyellow",
##                   "7_Weak_Enhancer" = "lightyellow",
##                   "8_Insulator" = "green",
##                   "9_Txn_Transition" = "darkgreen",
##                   "10_Txn_Elongation" = "darkgreen",
##                   "11_Weak_Txn" = "lightgreen",
##                   "12_Repressed" = "gray",
##                   "13_Heterochrom/lo" = "lightgray",
##                   "14_Repetitive/CNV" = "lightgray",
##                   "15_Repetitive/CNV" = "lightgray")

## chromhmm_track <- AnnotationTrack(region_chromhmm,
##                                   name = "ChromHMM",
##                                   fill = state_colors[mcols(region_chromhmm)$name],
##                                   col = state_colors[mcols(region_chromhmm)$name],
##                                   stacking = "dense")

# Plot with CTCF, DNase, peaks
plotTracks(list(axis_track, chip_track, dnase_track, peak_track),
           chromosome = chr, from = region_start, to = region_end,
           main = sprintf("CTCF ChIP-seq on %s:%s-%s",
                          chr,
                          format(region_start, big.mark=","),
                          format(region_end, big.mark=",")),
           background.title = "white",
           col.axis = "black",
           col.title = "black",
           fontsize = 8,
           cex.title = 0.6,
           cex.axis = 1,
           cex.main = 0.7)
```


## Key Concepts from ChIP-seq Peak Calling

- **Biological**: ChIP-seq identifies protein-DNA binding sites

- **Statistical** models:
  - Poisson/Binomial models for count data
  - Local background estimation
  - Multiple testing correction (FDR)

- **Computational** pipeline:
  - Millions of bins $\to$ massive multiple testing problem
  - Need efficient algorithms (`MACS2`, `normr`)

- **Quality control**:
  - Peak width distribution (sharp vs broad)
  - Fraction of reads in peaks (FRiP score)
  - Replicate concordance


## ChromHMM: Emission Probabilities

```{r chromhmm-emissions, echo=FALSE, fig.width=5, fig.height=3, message=FALSE, warning=FALSE}
library(reshape2)
library(ggtext)

# ChromHMM 15-state model emission probabilities
# Each state is defined by presence/absence of histone marks
emissions_file <- paste0(data_dir, "chromhmm_emissions.rds")

if.needed(emissions_file, {
    # Simplified emission matrix (15 states x key histone marks)
    # Based on ENCODE Broad ChromHMM model
    marks <- c("H3K4me3", "H3K4me1", "H3K36me3", "H3K27me3", "H3K9me3", "H3K27ac")
    states <- c("Active_Promoter", "Weak_Promoter", "Poised_Promoter",
                "Strong_Enhancer", "Weak_Enhancer",
                "Insulator", "Transcribed", "Repressed", "Heterochromatin")

    # Emission probabilities (simplified for teaching)
    emissions <- matrix(c(
        # H3K4me3, H3K4me1, H3K36me3, H3K27me3, H3K9me3, H3K27ac
        1.0, 0.3, 0.2, 0.0, 0.0, 1.0,  # Active Promoter
        0.8, 0.4, 0.1, 0.0, 0.0, 0.5,  # Weak Promoter
        0.7, 0.3, 0.0, 0.8, 0.0, 0.2,  # Poised Promoter
        0.2, 1.0, 0.1, 0.0, 0.0, 1.0,  # Strong Enhancer
        0.1, 0.7, 0.1, 0.0, 0.0, 0.5,  # Weak Enhancer
        0.5, 0.5, 0.1, 0.0, 0.0, 0.3,  # Insulator
        0.1, 0.1, 1.0, 0.0, 0.0, 0.2,  # Transcribed
        0.0, 0.0, 0.0, 1.0, 0.1, 0.0,  # Repressed
        0.0, 0.0, 0.0, 0.1, 1.0, 0.0   # Heterochromatin
    ), nrow = 9, ncol = 6, byrow = TRUE)

    rownames(emissions) <- states
    colnames(emissions) <- marks
    saveRDS(emissions, file = emissions_file)
})

emissions <- readRDS(emissions_file)
state_order <- rev(rownames(emissions))

# ChromHMM colors
state_colors <- c(
    "Active_Promoter" = "#FF0000", "Weak_Promoter" = "#FF6969",
    "Poised_Promoter" = "#CF0BC6", "Strong_Enhancer" = "#FFFF00",
    "Weak_Enhancer" = "#FFFF99", "Insulator" = "#0ABEFE",
    "Transcribed" = "#00B050", "Repressed" = "#808080",
    "Heterochromatin" = "#C0C0C0"
)

# Prepare data with faceting
emissions_long <- melt(emissions)
colnames(emissions_long) <- c("State", "Mark", "Value")
emissions_long$State <- factor(emissions_long$State, levels = state_order)
emissions_long$Panel <- "Emission Probabilities"

# Add color panel
color_data <- data.frame(
    State = factor(state_order, levels = state_order),
    Mark = "State",
    Value = 1,
    Panel = "State"
)

plot_data <- rbind(emissions_long, color_data)
plot_data$Panel <- factor(plot_data$Panel,
                          levels = c("State", "Emission Probabilities"))

ggplot(plot_data, aes(x = Mark, y = State)) +
    geom_tile(data = subset(plot_data, Panel == "State"),
              aes(fill = State), show.legend = FALSE, linewidth=.1, colour="gray20") +
    geom_tile(data = subset(plot_data, Panel == "Emission Probabilities"),
              aes(alpha = Value), fill = "red", linewidth=.1, colour="gray20") +
    facet_grid(. ~ Panel, scales = "free_x", space = "free_x") +
    scale_fill_manual(values = state_colors) +
    scale_alpha_continuous(range = c(0, 1), name = "Probability") +
    labs(x = "", y = "") +
    theme_minimal() +
    theme(axis.text.x = element_text(size = 8, angle = 90, hjust = 1, vjust = 0.5),
          axis.text.y = element_text(size = 7),
          axis.title = element_text(size = 9),
          plot.title = element_text(size = 10),
          legend.position = "right",
          strip.text = element_text(size = 8, face = "bold"),
          panel.grid = element_blank(),
          panel.spacing = unit(0.5, "lines"))
```

## ChromHMM: Integrating Multiple Histone Marks

**How it works:**

- Input: Multiple histone modification ChIP-seq signals
- Model: Hidden Markov Model learns chromatin states
- Each state defined by emission probabilities (previous slide)
- Output: Genome segmented into functional states

**In our genome browser:**

- ChromHMM track shows where each state occurs
- CTCF peaks often coincide with specific states (insulators, enhancers)



## Visualizing CTCF Binding Sites (using `Gviz`)

```{r chipseq-viz-chromhmm, echo=FALSE, cache=FALSE, fig.width=6, fig.height=3, message=FALSE, warning=FALSE}
library(Gviz)

# Find a region with multiple strong peaks
peaks_sorted <- chip_peaks[order(mcols(chip_peaks)$signalValue, decreasing=TRUE)]
top_peak <- peaks_sorted[3]
region_start <- max(1, start(top_peak) - 100000)
region_end <- start(top_peak) + 100000

# Get peaks in this region
region_peaks <- chip_peaks[start(chip_peaks) >= region_start &
                           end(chip_peaks) <= region_end]

# Bin reads for both CTCF ChIP and DNase
bin_size <- 200
bins <- seq(region_start, region_end, by = bin_size)

# CTCF ChIP signal
region_chip <- chip_reads[start(chip_reads) >= region_start &
                          end(chip_reads) <= region_end]
chip_counts <- countOverlaps(GRanges(chr, IRanges(bins, width = bin_size)), region_chip)
chip_gr <- GRanges(chr, IRanges(bins, width = bin_size), score = chip_counts)

# DNase signal (exaggerated 3x for visibility)
region_dnase <- dnase_reads[start(dnase_reads) >= region_start &
                            end(dnase_reads) <= region_end]
dnase_counts <- countOverlaps(GRanges(chr, IRanges(bins, width = bin_size)), region_dnase)
dnase_gr <- GRanges(chr, IRanges(bins, width = bin_size), score = dnase_counts * 3)

# Create genome browser tracks
axis_track <- GenomeAxisTrack()

chip_track <- DataTrack(chip_gr, genome = "hg19", type = "histogram",
                        chromosome = chr, name = "CTCF",
                        fill = "darkblue", col.histogram = "darkblue")

dnase_track <- DataTrack(dnase_gr, genome = "hg19", type = "histogram",
                         chromosome = chr, name = "DNase (3x)",
                         fill = "orange", col.histogram = "orange")

# ChromHMM chromatin state track
region_chromhmm <- chromhmm[start(chromhmm) >= region_start &
                            end(chromhmm) <= region_end]

# Define colors for chromatin states (ENCODE standard colors)
state_colors <- c("1_Active_Promoter" = "red",
                  "2_Weak_Promoter" = "orange",
                  "3_Poised_Promoter" = "purple",
                  "4_Strong_Enhancer" = "yellow",
                  "5_Strong_Enhancer" = "yellow",
                  "6_Weak_Enhancer" = "lightyellow",
                  "7_Weak_Enhancer" = "lightyellow",
                  "8_Insulator" = "green",
                  "9_Txn_Transition" = "darkgreen",
                  "10_Txn_Elongation" = "darkgreen",
                  "11_Weak_Txn" = "lightgreen",
                  "12_Repressed" = "gray",
                  "13_Heterochrom/lo" = "lightgray",
                  "14_Repetitive/CNV" = "lightgray",
                  "15_Repetitive/CNV" = "lightgray")

chromhmm_track <- AnnotationTrack(region_chromhmm,
                                  name = "ChromHMM",
                                  fill = state_colors[mcols(region_chromhmm)$name],
                                  col = state_colors[mcols(region_chromhmm)$name],
                                  stacking = "dense")

# Plot with CTCF, DNase, peaks
plotTracks(list(axis_track, chip_track, dnase_track, chromhmm_track),
           chromosome = chr, from = region_start, to = region_end,
           main = sprintf("CTCF ChIP-seq on %s:%s-%s",
                          chr,
                          format(region_start, big.mark=","),
                          format(region_end, big.mark=",")),
           background.title = "white",
           col.axis = "black",
           col.title = "black",
           fontsize = 8,
           cex.title = 0.6,
           cex.axis = 1,
           cex.main = 0.7)
```





## Assay for Transposase-Accessible Chromatin (ATAC)

:::::: {.columns}
::: {.column width=.5}

**Goal**: Map open chromatin regions genome-wide

**Key insight**: \alert{No antibody needed!}

**Protocol**:

1. Tn5 transposase + DNA
2. Cuts accessible chromatin preferentially
3. Adds sequencing adapters
4. Sequence DNA fragments
5. Map reads to genome
6. **Call peaks**

:::
::: {.column width=.5}

**Advantages**:

- Fast protocol (~3 hours)
- Small cell numbers (500-50K cells)
- No antibodies required
- Genome-wide view

:::
::::::



## ATAC-seq vs. ChIP-seq

:::::: {.columns}
::: {.column width=.5}

**ATAC-seq**:

- Measures \alert{chromatin accessibility}
- All open regions (enhancers, promoters, etc.)
- No protein specificity
- No antibody needed
- Cheaper, faster

**Use cases**:

- Map all regulatory elements
- Compare cell states
- Find active enhancers
- TF footprinting

:::
::: {.column width=.5}

**ChIP-seq**:

- Measures \alert{protein-DNA binding}
- Specific protein of interest
- Requires good antibody
- More expensive

**Use cases**:

- Find TF binding sites
- Map histone modifications
- Validate ATAC peaks
- Mechanistic studies

:::
::::::

## ATAC-seq produces regular fragment sizes

\only<1>{\centerline{\includegraphics[height=.75\textheight]{img/atac_seq_buenrostro-tn5.pdf}}
\textbf{Tn5}: {\color{teal} \textbf{cuts}} open DNA and {\color{magenta}\textbf{insert}} adapter sequence.}

\only<2>{\centerline{\includegraphics[height=.8\textheight]{img/atac_seq_buenrostro-2.pdf}}}

\only<3>{\vfill\centerline{\includegraphics[width=.8\textwidth]{img/atac_seq_buenrostro-3.pdf}}}

\only<4>{\vfill\centerline{\includegraphics[width=.8\textwidth]{img/atac_seq_buenrostro-4.pdf}}
\textbf{ATAC}-seq enriches known gene regulatory elements in genome}

\vfill

\flushright
\scriptsize
[@Buenrostro2013-qd]


## Differential Accessibility Analysis {.smaller}

```{r atacseq-demo, echo=TRUE, eval=FALSE}
# After peak calling, create count matrix
# Rows = genomic bins, Columns = samples
count_matrix <- matrix(...)  # Read counts per bin per sample

colData <- data.frame( # Set up experimental design
    condition = factor(c(rep("Control", 3), rep("Treatment", 3)))
)

dds <- DESeqDataSetFromMatrix(countData = count_matrix, # Run differential accessibility
                               colData = colData,       #  (same as RNA-seq!)
                               design = ~ condition)
dds <- DESeq(dds)
results <- results(dds, contrast = c("condition", "Treatment", "Control"))

sig_regions <- results[results$padj < 0.05, ] # Get significant regions (FDR < 0.05)
```

**Key point**: Same statistical framework as RNA-seq (DESeq2)!


## ATAC-seq Applications

:::::: {.columns}
::: {.column width=.4}

**1. Identify regulatory elements**

- Promoters (at TSS)
- Enhancers (distal peaks)
- Silencers
- Insulators

**2. Compare cell types/states**

- Development
- Differentiation
- Disease vs healthy

:::
::: {.column width=.55}

**3. TF footprinting**

- Bound TF → protected DNA
- Dip in signal at TF motif
- Infer active TFs

**4. Integration**

- ATAC + RNA-seq → **link enhancers to genes**
- ATAC + ChIP-seq → TF binding
- ATAC + GWAS → **functional variants**

:::
::::::

##






# DNA Methylation (credit: Keegan Korthauer)

## What is DNA Methylation?

:::::: {.columns}
::: {.column width=.5}

* **5-methylcytosine (5mC)**: a methyl group added to cytosine at CpG dinucleotides

* **Heritable**: copied faithfully during DNA replication (by DNMT1)

* **Regulatory role**:
  * CpG islands at promoters → gene silencing when methylated
  * Enhancer activity correlates with hypomethylation
  * Imprinting, X-inactivation

:::
::: {.column width=.48}

\includegraphics[width=\textwidth]{img/5mC.pdf}

\scriptsize\textit{epigentek.com}

\vspace{0.2em}

\includegraphics[width=\textwidth]{img/copymeth.pdf}

\scriptsize\textit{slideplayer.com}

:::
::::::

## DNA Methylation and Gene Regulation

:::::: {.columns}
::: {.column width=.58}
\includegraphics[width=\textwidth]{img/genereg.pdf}
:::
::: {.column width=.4}
\includegraphics[width=\textwidth]{img/5thbase.pdf}
:::
::::::

## Measuring DNA Methylation

:::::: {.columns}
::: {.column width=.48}

\centering
\includegraphics[width=\textwidth]{img/bisulfite-conversion.pdf}

\scriptsize Bisulfite conversion (\textit{diagenode.com})

:::
::: {.column width=.5}

\centering
\includegraphics[width=\textwidth]{img/methassays.pdf}

\scriptsize [@Laird2010-xv]

:::
::::::

## Bisulfite Sequencing

\centerline{\includegraphics[width=.9\textwidth]{img/methylation-bisulfite.pdf}}

## Beta ($\beta$) values

* Combine two measurements for each CpG probe, measure level of methylation:

$$\beta = \frac{\text{methylated intensity}}{(\text{methylated intensity} + \text{unmethylated intensity})}$$

* Range from 0 to 1

* Easily interpretable as a **methylation proportion**

* Best used for **visualization**

## M-values

* Transformed $\beta$ values:

$$M= \log_2\!\left(\frac{\text{methylated intensity}}{\text{unmethylated intensity}}\right) = \log_2\!\left(\frac{\beta}{1-\beta}\right)$$

* The *logit* transformation — maps a proportion onto the real line

* Alleviates heteroscedasticity of $\beta$ values
  (recall: $\text{Binomial}(n,p)$ variance $= np(1-p)$)

* Best used for **statistical testing**

\vfill
\scriptsize [@Du2010-bm]

## Comparing $\beta$ and M values

```{r meth-density, echo=FALSE, cache=FALSE, fig.width=5.5, fig.height=2.8, message=FALSE, warning=FALSE}
library(minfi)
library(RColorBrewer)

mSet <- readRDS("../data/methylation/mSet.rds")
targets <- readRDS("../data/methylation/targets.rds")

bVals <- getBeta(mSet)
mVals <- getM(mSet)

par(mfrow=c(1,2), cex=0.55, mar=c(4,4,2,1))

densityPlot(bVals,
            sampGroups = targets$Sample_Group,
            main = expression(beta~values),
            legend = FALSE,
            xlab = expression(beta~values))
legend("top",
       legend = levels(factor(targets$Sample_Group)),
       text.col = brewer.pal(8, "Dark2"),
       cex = 0.8)

densityPlot(mVals,
            sampGroups = targets$Sample_Group,
            main = "M-values",
            legend = FALSE,
            xlab = "M values")
legend("topleft",
       legend = levels(factor(targets$Sample_Group)),
       text.col = brewer.pal(8, "Dark2"),
       cex = 0.8)
```

## Methylation Array Analysis Workflow

* Analysis strategy: **linear regression on M-values** (limma)

* Key R packages:
  * `minfi` — preprocessing and normalization of array data
  * `limma` — differentially methylated cytosines (DMC)
  * `dmrseq` / `bumphunter` — differentially methylated **regions** (DMR)

* Example dataset: sorted T cells [@Zhang2013-kd]
  * Naive vs. activated naive T cells
  * [GEO: GSE49667](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE49667)

## Differentially Methylated Cytosines (DMC)

```{r meth-dmc, echo=TRUE, cache=FALSE, message=FALSE, warning=FALSE, size="tiny"}
library(limma)
cellType  <- factor(targets$Sample_Group)
individual <- factor(targets$Sample_Source)
design <- model.matrix(~ cellType + individual, data = targets)

## Fit one linear model per CpG probe
mfit <- lmFit(mVals, design)
cfit <- eBayes(mfit)

## Top differentially methylated CpGs (naive vs. act_naive)
topTable(cfit, coef = "cellTypenaive", number = 5)
```

## DMCs vs DMRs

* Individual CpG tests → **differentially methylated cytosines (DMCs)**

* Methylation of nearby CpGs is **correlated** and regulated in tandem

* Grouping nearby significant CpGs → **differentially methylated regions (DMRs)**

\vfill

\centering
\includegraphics[width=.9\textwidth]{img/dmc1.pdf}

## DMCs vs DMRs

:::::: {.columns}
::: {.column width=.49}
\centering
\includegraphics[width=\textwidth]{img/dmc2.pdf}

\scriptsize CpG 1
:::
::: {.column width=.49}
\centering
\includegraphics[width=\textwidth]{img/dmc3.pdf}

\scriptsize CpG 2
:::
::::::

## FDR at the Region Level

:::::: {.columns}
::: {.column width=.55}
\includegraphics[width=\textwidth]{img/fdr2.pdf}

\vspace{0.3em}

\includegraphics[width=\textwidth]{img/fdr3.pdf}
:::
::: {.column width=.43}

\vspace{1em}

* Grouping significant CpGs post-hoc **inflates FDR**

$$FDR_{CpG} = \frac{2}{10} = 0.20$$

$$FDR_{DMR} = \frac{1}{2} = 0.50$$

* Need to model region-level signal directly

\vfill
\scriptsize `BSmooth` [@Hansen2012-bw], `DSS` [@Park2016-rq]

:::
::::::

## Accurate DMR Inference: `dmrseq`

:::::: {.columns}
::: {.column width=.75}
\includegraphics[width=\textwidth]{img/2stage.pdf}
:::
::: {.column width=.2}

\vspace{1em}

* Model correlated methylation signal over region

* Permutation for accurate FDR control

\vspace{0.5em}
\flushright
\includegraphics[width=.5\textwidth]{img/dmrseqhex.pdf}

\vfill
\scriptsize [@Korthauer2019-dm]

:::
::::::

## `dmrseq` Output

\centering
\includegraphics[width=.85\textwidth]{img/exdmr.pdf}

\vfill

* *De novo* regions with FDR estimate
* Ranked by effect size of methylation difference
* Can adjust for covariates

## Summary: DNA Methylation Analysis

* **Data**: Illumina arrays → per-CpG $\beta$ values (proportions)

* **Model**: linear regression on logit-transformed M-values
  * Same `limma` framework as gene expression microarrays
  * Additive models, interactions, paired designs all apply

* **Output**:
  * **DMC** analysis: per-probe test (limma)
  * **DMR** analysis: region-level modelling (dmrseq/bumphunter)

* **Back-transform** estimates to $\beta$ scale for biological interpretation:
$$\hat{\beta} = \frac{2^{\hat{M}}}{1 + 2^{\hat{M}}}$$

# Multiomic Data Integration with Regulatory Genomics 

## From DMRs to Biological Meaning

We found differentially methylated regions — now what?

\vfill

\alert{Key questions}:

- Which **genes** are near these regions?
- Which **biological processes** are affected?

\vfill

**Workflow**:

$$\text{DMRs} \xrightarrow{\text{proximity}} \text{genes} \xrightarrow{\text{enrichment}} \text{pathways / GO terms}$$

\vfill

This generalises to \textbf{any} set of genomic regions: ATAC peaks, ChIP peaks, GWAS loci

## Standard Enrichment: A Problem

**Naive approach**: map each region to its nearest gene, collect gene list, run hypergeometric test

\vfill

\alert{Why this fails}:

- Gene-dense regions contribute many genes → \textbf{inflated hits}
- Distal enhancers (>5 kb) are silently ignored
- Background is "all genes", not "all genomic space"

\vfill

Example: a peak between a 50-gene cluster and an intergenic desert → the cluster dominates

## Regions → Enrichment Directly 

`GREAT` [@McLean2010-ou]. Each gene gets a **regulatory domain**:

\vfill

\textbf{Basal domain}: $\pm$ 5 kb around TSS (always kept)

\textbf{Extended domain}: expands up to 1 Mb, until the next gene's basal boundary

\vfill

\vspace{0.5em}
\begin{tikzpicture}[scale=0.72, every node/.style={font=\tiny}]
  % Chromosome bar
  \draw[thick] (0,0) -- (9,0);
  % Gene A body
  \draw[fill=gray!40, draw=gray] (1.5,-.15) rectangle (3.5,.15);
  \node at (2.5, 0) {Gene A};
  \draw[->, thick] (2.5,.15) -- (2.5,.45) node[above] {TSS$_A$};
  % Gene B body
  \draw[fill=gray!40, draw=gray] (5.8,-.15) rectangle (8,.15);
  \node at (6.9, 0) {Gene B};
  \draw[->, thick] (5.8,.15) -- (5.8,.45) node[above] {TSS$_B$};
  % Basal A
  \draw[thick, blue, |-|] (1.8,-.6) -- (3.2,-.6);
  \node[blue] at (2.5,-.85) {basal A};
  % Basal B
  \draw[thick, blue, |-|] (5.1,-.6) -- (6.5,-.6);
  \node[blue] at (5.8,-.85) {basal B};
  % Extended A
  \draw[thick, teal, |-|] (0.2,-1.3) -- (5.1,-1.3);
  \node[teal] at (2.65,-1.55) {extended A};
  % Extended B
  \draw[thick, teal, |-|] (5.1,-1.3) -- (8.8,-1.3);
  \node[teal] at (6.95,-1.55) {extended B};
  % Peak
  \draw[fill=red!70, draw=red] (3.6,-1.15) rectangle (3.9,-.45);
  \node[red] at (3.75,-.25) {peak};
  \draw[->, red, dashed] (3.75,-.45) -- (3.75,-1.15);
  \node[red] at (5.2,-1.9) {$\rightarrow$ linked to Gene A};
\end{tikzpicture}

\vfill

A region can be linked to **multiple** genes; background = all genomic space

## GREAT: Why It Works Better

| | Standard | GREAT |
|---|---|---|
| Input | gene list | genomic regions |
| Background | all genes | all genome |
| Long-range | ignored | up to 1 Mb |
| Gene density bias | yes | corrected |

\vfill

Useful for: ATAC peaks, ChIP peaks, DMRs, GWAS loci

## GREAT Step 1: Prepare Input Regions

```{r great-input, echo=TRUE, cache=FALSE, message=FALSE}
library(rGREAT)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)

## chip_peaks loaded earlier — keep standard chromosomes
std_chrs <- paste0("chr", c(1:22, "X", "Y"))
ctcf_gr  <- GRanges(as.character(seqnames(chip_peaks)),
                    IRanges(start(chip_peaks), end(chip_peaks)))
ctcf_gr  <- ctcf_gr[as.character(seqnames(ctcf_gr)) %in% std_chrs]
cat(length(ctcf_gr), "CTCF peaks ready\n")
```

* Regions = CTCF ChIP-seq peaks (GM12878, whole genome)
* Same syntax for DMRs, ATAC peaks, or GWAS loci

## GREAT Step 2: Run Enrichment

```{r great-run, echo=TRUE, message=FALSE}
great_file <- "../data/great_ctcf_gobp.rds"
if.needed(great_file, {
    great_res <- great(ctcf_gr,
                       gene_sets  = "GO:BP",
                       tss_source = "TxDb.Hsapiens.UCSC.hg19.knownGene",
                       verbose    = FALSE)
    saveRDS(great_res, great_file)
})
great_res <- readRDS(great_file)
```

* `gene_sets`: `"GO:BP"`, `"GO:MF"`, `"KEGG"`, `"MSigDB:H"`, …
* `tss_source`: genome-specific gene model
* Returns a `GreatObject` (region–gene links + enrichment)

## GREAT Step 3: Inspect Results

```{r great-table, echo=TRUE, cache=FALSE}
tb <- getEnrichmentTable(great_res)
cols <- c("description", "observed_region_hits",
          "fold_enrichment", "p_adjust")
head(tb[order(tb$p_adjust), cols], 6)
```

* `fold_enrichment`: observed / expected overlap, `p_adjust`: BH-corrected p-value

## GREAT Step 4: Visualize Top Terms

```{r great-plot, echo=FALSE, cache=FALSE, fig.width=5.5, fig.height=2.8, message=FALSE}
top <- head(tb[tb$p_adjust < 0.05, ][order(tb$p_adjust[tb$p_adjust < 0.05]), ], 15)
top$description <- factor(top$description,
                          levels = rev(top$description))
ggplot(top, aes(x = description,
                y = -log10(p_adjust),
                fill = fold_enrichment)) +
    geom_bar(stat = "identity") +
    geom_hline(yintercept = -log10(0.05),
               linetype = "dashed", colour = "red") +
    coord_flip() +
    scale_fill_gradient(low = "steelblue", high = "darkred",
                        name = "Fold\nenrichment") +
    labs(x = "", y = expression(-log[10](p[adj])),
         title = "Top GO:BP terms — CTCF peaks") +
    theme(axis.text.y = element_text(size = 7),
          legend.key.size = unit(0.4, "cm"))
```

## GWAS Variants Are Mostly Non-Coding

Disease GWAS identifies thousands of associated variants

\vfill

**The puzzle**: >90% of hits fall outside protein-coding exons

\vfill

\alert{Key questions}:

- Where do these variants **act**?
- Which **cell types** are disease-relevant?

\vfill

**Strategy**: overlap GWAS loci with cell-type-specific epigenetic annotations

## GWAS → Epigenetic States: The Approach

1. Collect GWAS summary statistics (SNPs with $P < 5 \times 10^{-8}$)

2. Download cell-type-specific chromatin annotations
   (e.g., Roadmap Epigenomics: 127 cell/tissue types)

3. For each cell type: test whether GWAS SNPs are **enriched** in active enhancers

4. The most enriched cell type = likely disease-relevant tissue

\vfill

\textbf{Statistical test}: Fisher's exact test (or LD-score regression)

## GWAS Step 1: Fetch RA Loci from GWAS Catalog

```{r gwas-fetch, echo=TRUE, cache=FALSE, message=FALSE}
library(gwasrapidd)
ra_file <- "../data/ra_gwas_hits.rds"
if (!file.exists(ra_file)) {
    assoc    <- get_associations(study_id = "GCST90132222")
    variants <- get_variants(study_id    = "GCST90132222")
    ra <- merge(assoc@risk_alleles[, c("association_id","variant_id")],
                variants@variants[, c("variant_id","chromosome_name",
                                      "chromosome_position")],
                by = "variant_id")
    ra$CHR <- paste0("chr", ra$chromosome_name)
    saveRDS(ra, ra_file)
}
ra <- readRDS(ra_file)
cat(nrow(ra), "genome-wide significant RA loci (GCST90132222)\n")
```

* Study: 22,350 European ancestry samples [@Okada2014-oa]
* Coordinates: hg19 / GRCh37

## GWAS Step 2: Build Locus Windows

```{r gwas-loci, echo=TRUE, cache=FALSE, message=FALSE}
## Expand each index SNP ±50 kb (covers typical LD block)
sig_gr <- GRanges(ra$CHR,
                  IRanges(ra$chromosome_position - 50000,
                          ra$chromosome_position + 50000))
sig_gr <- sig_gr[start(sig_gr) > 0]

## Chromosome-proportional random background (hg19)
set.seed(42)
chr_sizes <- seqlengths(BSgenome.Hsapiens.UCSC.hg19::
                            BSgenome.Hsapiens.UCSC.hg19)[paste0("chr",1:22)]
chr_s <- sample(names(chr_sizes), 5e4, replace = TRUE,
                prob = chr_sizes / sum(chr_sizes))
pos_s <- mapply(function(ch) sample(chr_sizes[ch] - 1e5, 1) + 50000, chr_s)
bg_gr <- GRanges(chr_s, IRanges(pos_s - 50000, pos_s + 50000))
cat(length(sig_gr), "locus windows,", length(bg_gr), "background windows\n")
```

## GWAS Step 3: Load Roadmap H3K27ac Enhancers

```{r gwas-enh, echo=TRUE, cache=FALSE, message=FALSE}
enh_dir    <- "../data/roadmap/"
enh_labels <- c(CD4_Naive  = "AH43615", CD4_Memory = "AH43684",
                CD19_Bcell = "AH44900", CD14_Mono  = "AH44890",
                Adult_Liver= "AH43458", Brain      = "AH43501")

## Load from locally cached RDS files (downloaded once via AnnotationHub)
enhancers <- lapply(names(enh_labels), function(ct)
    readRDS(paste0(enh_dir, ct, "_H3K27ac.rds")))
names(enhancers) <- names(enh_labels)
immune_cells <- c("CD4_Naive","CD4_Memory","CD19_Bcell","CD14_Mono")
sapply(enhancers, length)   # peaks per cell type
```

* H3K27ac = active enhancer / promoter mark
* Roadmap Epigenomics: 127 reference human epigenomes [@Kundaje2015-cy]

## GWAS Step 4: Fisher's Exact Test per Cell Type

```{r gwas-fisher, echo=TRUE, message=FALSE}
enr_file <- "../data/gwas_ra_enrichment.rds"
if.needed(enr_file, {
    enr <- do.call(rbind, lapply(names(enhancers), function(ct) {
        enh <- enhancers[[ct]]
        a <- sum(countOverlaps(sig_gr, enh) > 0)
        b <- length(sig_gr) - a
        c <- sum(countOverlaps(bg_gr,  enh) > 0)
        d <- length(bg_gr)  - c
        ft <- fisher.test(matrix(c(a, b, c, d), 2))
        data.frame(cell_type = ct, OR = ft$estimate,
                   pval = ft$p.value, frac = a / length(sig_gr))
    }))
    enr$fdr      <- p.adjust(enr$pval, "BH")
    enr$category <- ifelse(enr$cell_type %in% immune_cells, "Immune", "Non-immune")
    saveRDS(enr, enr_file)
})
enr <- readRDS(enr_file)
enr$cell_type <- factor(enr$cell_type, levels = enr$cell_type[order(enr$OR)])
```

## GWAS Enrichment: Odds Ratio

```{r gwas-plot1, echo=FALSE, cache=FALSE, fig.width=5.5, fig.height=2.8, message=FALSE}
ggplot(enr, aes(cell_type, OR, fill = category)) +
    geom_bar(stat = "identity", alpha = .8) +
    geom_hline(yintercept = 1, linetype = "dashed", colour = "red") +
    geom_text(aes(label = ifelse(fdr < .05, "*", "")),
              vjust = -.3, size = 5, colour = "red") +
    coord_flip() +
    scale_fill_manual(values = c(Immune = "#d73027", `Non-immune` = "#4575b4")) +
    labs(x = NULL, y = "Odds Ratio",
         title = "RA loci enriched in immune cell enhancers",
         subtitle = "* FDR < 0.05    |    Roadmap H3K27ac peaks") +
    theme(legend.position = "bottom")
```

## GWAS Enrichment: Statistical Significance

```{r gwas-plot2, echo=FALSE, cache=FALSE, fig.width=5.5, fig.height=2.8, message=FALSE}
enr$nlp <- -log10(enr$pval)
ggplot(enr, aes(cell_type, nlp, fill = category)) +
    geom_bar(stat = "identity", alpha = .8) +
    geom_hline(yintercept = -log10(.05 / nrow(enr)),
               linetype = "dashed", colour = "red") +
    coord_flip() +
    scale_fill_manual(values = c(Immune = "#d73027", `Non-immune` = "#4575b4")) +
    labs(x = NULL, y = expression(-log[10](p)),
         title = "Statistical significance of enrichment",
         subtitle = "dashed = Bonferroni threshold") +
    theme(legend.position = "bottom")
```

## GWAS Enrichment: Overlap Fraction

```{r gwas-plot3, echo=FALSE, cache=FALSE, fig.width=5.5, fig.height=2.8, message=FALSE}
ggplot(enr, aes(cell_type, frac * 100, fill = category)) +
    geom_bar(stat = "identity", alpha = .8) +
    coord_flip() +
    scale_fill_manual(values = c(Immune = "#d73027", `Non-immune` = "#4575b4")) +
    labs(x = NULL, y = "% RA loci overlapping enhancers",
         title = "Fraction of RA loci in cell-type enhancers") +
    theme(legend.position = "bottom")
```

## Limitation: Fisher's Test Ignores LD

**The problem**: nearby SNPs are correlated through **linkage disequilibrium (LD)**

\vfill

Two significant SNPs 20 kb apart may represent \alert{one causal signal}, not two

$\Rightarrow$ overlapping the same enhancer window gets double-counted

\vfill

**Consequence**: Fisher's test \alert{inflates enrichment} in gene-dense, LD-rich regions

\vfill

**Better approaches**:

* **Stratified LD score regression** (S-LDSC) — partitions heritability by annotation, models LD [@Finucane2015-qr]
* **GARFIELD** — permutation matching on MAF, distance to TSS, LD
* **GoShifter** — circular permutation within LD blocks

## Real-World GWAS–Epigenome Findings (examples)

\large

\textbf{Rheumatoid arthritis} $\to$ CD4$^+$ T cells \hfill \textit{adaptive immunity}

\vspace{0.4em}
\textbf{IBD} $\to$ gut immune cells \hfill \textit{mucosal inflammation}

\vspace{0.4em}
\textbf{Type 2 diabetes} $\to$ pancreatic islets \hfill \textit{$\beta$-cell dysfunction}

\vspace{0.4em}
\textbf{Schizophrenia} $\to$ neurons \hfill \textit{brain circuit defects}

\vspace{0.4em}
\textbf{Height} $\to$ chondrocytes \hfill \textit{cartilage growth plates}

\normalsize
\vfill
\flushright\scriptsize [@Kundaje2015-cy]

## Why GWAS–Epigenome Integration Matters

**Identify disease-relevant cell types**

- Guides which cells to study experimentally
- 90% of hits are non-coding — context is everything

\vfill

**Mechanistic insight**

- Variant disrupts an enhancer → changes TF binding → alters gene expression

\vfill

**Therapeutic direction**

- Target the right cell type, not just the right gene

\vfill

**Links to scRNA-seq**

- Fine-map to specific cell states within a tissue





##

## Reference {.allowframebreaks}
