---
title: "Polygenic prediction and sparse regression analysis"
author: "Yongjin Park, UBC Path + Stat, BC Cancer"
date: "`r format(Sys.time(), '%d %B %Y')`"
mainfont: Roboto
classoption: "aspectratio=169"
fontsize: 12pt
output:
    powerpoint_presentation:
        reference_doc: "_template.pptx"
    html_document:
        self_contained: true
    beamer_presentation:
        theme: Madrid
        keep_md: true
        keep_tex: true
        latex_engine: xelatex
        slide_level: 2
header-includes:
  - \definecolor{UBCblue}{rgb}{0.04706, 0.13725, 0.26667}
  - \setbeamertemplate{frametitle}{\color{UBCblue}\bfseries\insertframetitle\par\vskip-6pt\hrulefill}
  - \setbeamercolor{title in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{section in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{author in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{date in head/foot}{bg=white,fg=gray}
  - \setbeamertemplate{page number in head/foot}{}
  - \setbeamertemplate{frame numbering}[none]
  - \setbeamercolor{alerted text}{bg=yellow}
  - \AtBeginSection[]{\begin{frame}\frametitle{Today's lecture}{\Large\tableofcontents[currentsection]}\end{frame}}
  - |
    \makeatletter
    \def\ps@titlepage{%
      \setbeamertemplate{footline}{}
    }
    \addtobeamertemplate{title page}{\thispagestyle{titlepage}}{}
    \makeatother
    \include{toc}
---


```{r setup, include=FALSE}
setwd("~/work/teaching/stat540/lectures/lect14-gwas-pgs/")
library(data.table)
library(tidyverse)
library(patchwork)
library(matrixTests)
source("Setup.R")
fig.dir <- "../Fig/"
dir.create(fig.dir, showWarnings=FALSE)
setup.env(fig.dir)
dir.create("../data", showWarnings=FALSE)
theme_set(theme_classic() +
          theme(title = element_text(size=10),
                legend.background = element_blank())
          )
```

# What is GWAS (review and pick up the leftover)?

<!-- ## Today's lecture: GWAS (continued) and PGS -->
<!-- * Extension of basic GWAS methods -->
<!--     * Linear Mixed Effect model to handle population structures -->
<!-- 	* Meta analysis across multiple studies -->
<!-- * Polygenic risk prediction -->

## Using the 1000 Genomes data via `bigsnpr`

```{r eval=F, echo=T, size="large"}
library(bigsnpr)
download_1000G("../data/genotype/")
```

\vfill

```{r download_1000g_data, echo=F}
dir.create("../data/genotype/", recursive=TRUE, showWarnings=FALSE)

library(bigsnpr)
.bed.file <- "../data/genotype/1000G_phase3_common_norel.bed"
if.needed(.bed.file, {
    download_1000G("../data/genotype/")
})

.bk.file <- "../data/genotype/1000G_phase3_common_norel.rds"
if.needed(.bk.file, {
    BED <- snp_readBed(.bed.file)
})
```

```{r echo=T, size="large"}
data <- snp_attach(.bk.file)
str(data, max.level=1, strict.width = "cut")
```

```{r include = F}
pop.info.file <- "../data/genotype/1000G_phase3_common_norel.fam2"
pop.info <- fread(pop.info.file)
```

```{r include=F}
## population structure
W <- 
    pop.info %>% 
    mutate(one = 1) %>%
    dcast(`sample.ID` ~ `Super Population`, value.var = "one", fill = 0) %>%
    (function(x) left_join(data$fam[, "sample.ID", drop = F], x)) %>% 
    column_to_rownames("sample.ID") %>%
    as.matrix()

## Take first 10000 variants
X <- data$genotypes[, 1:10000]
rownames(X) <- data$fam$sample.ID
colnames(X) <- data$map$physical.pos[1:10000]

#' @param X genotype matrix
#' @param W population structure
#' @param h2 heritability
#' @param pve.w population PVe
#' @param n.causal
#' @param n.traits
simulate.pgs <- function(X, W, h2, pve.w, n.causal, n.traits = 1) {
    .rnorm <- function(d1, d2) matrix(rnorm(d1*d2), d1, d2)
    causal.snp <- sample(ncol(X), n.causal)
    xx.causal <- apply(X[, causal.snp, drop=FALSE], 2, scale)
    xx.causal[is.na(xx.causal)] <- 0

    n.ind <- nrow(X)
    y.true <- apply(xx.causal %*% .rnorm(n.causal, n.traits), 2, scale)
    if(pve.w > 0){
        y.pop <- apply(W %*% .rnorm(ncol(W), n.traits), 2, scale)
    }else{
        y.pop <- 0
    }
    y.err <- apply(.rnorm(n.ind, n.traits), 2, scale)

    y.obs <- y.true * sqrt(h2) + y.pop * sqrt(pve.w)
    y.obs <- y.obs + y.err * sqrt(1 - h2 - pve.w)

    prop <- 1/(1+ exp(-y.obs))
    y.cc <- rbinom(length(prop), 1, prop)

    list(y.q = y.obs, y.cc = y.cc, x = X, w = W, causal = causal.snp)
}
```

```{r}
set.seed(1)
.sim <- simulate.pgs(X, W, h2=.3, pve.w=0, n.causal=3)
```

```{r}
Y <- .sim$y.cc
```

## Notations

\large

* $X$: ${n} \times {p}$ genotype matrix of ${n}$ samples/individuals and ${p}$ variants

* $Y$: ${n} \times {1}$ phenotype vector of ${n}$ samples/individuals


## Let's do GWAS to find the disease-associated variants

A variant-by-variant association t-test for a variant $j$:

\begin{eqnarray*}
H_{0}:
&\,& 
\mathbb{E}\!\left[X_{ij}|Y_{i}=1\right] = \mathbb{E}\!\left[X_{ij}|Y_{i}=0\right] \\
& & \textsf{vs.} \\
\, H_{1}:&\,& \mathbb{E}\!\left[X_{ij}|Y_{i}=1\right] \neq \mathbb{E}\!\left[X_{ij}|Y_{i}=0\right]
\end{eqnarray*}

* The average genotype is the same between the case and control under the null.

## For one variant's test

```{r echo=T, size="large"}
j <- 1
t.test(X[Y == 0, j], X[Y == 1, j])
```

## Can we calculate GWAS statistics more efficiently?

```{r echo=T, size="large"}
## library(matrixTests)
.gwas <- col_t_welch(X[Y == 0, ], X[Y == 1, ])
```

* For each pair of columns in the two matrices, the function performs t-test and summarize all the results into a list of vectors.

```{r echo=T}
names(.gwas)
```


## Human genetic variation recap

\centerline{\includegraphics[height=.7\textheight]{img/altshuler_daly_lander_GWAS.pdf}}

\vfill

Highly recommend [Altshuler, Daly, Lander, _Science_ (2008)](https://www.science.org/doi/10.1126/science.1156409)

## Human genetic variation recap

\centerline{\includegraphics[height=.8\textheight]{img/altshuler_daly_lander_LD.pdf}\vspace{-16pt}}

Highly recommend [Altshuler, Daly, Lander, _Science_ (2008)](https://www.science.org/doi/10.1126/science.1156409)

## Manhattan plot: a summary of all the GWAS p-values

```{r fig.width=5, fig.height=3}
par(pch=19, col="gray40", cex=.5)
plot(colnames(X), -log10(.gwas$pvalue), xlab = "genomic location", ylab = "-log10 p-value")
```

## The covariance of $X$ between the variants (columns)

```{r}
xx <- data$genotypes[1:400, 1:700]
```

:::::: {.columns}
::: {.column width=.5}

```{r fig.width=2.8, fig.height=2.5}
.matshow(pmin(pmax(cov(xx), -.5), .5), .lab=0, .axis.lab=0, .lw=0)
```

:::
::: {.column width=.5}


For a genotype matrix $X$ ${(n{\times}p)}$,



::: {.block}
### Linkage disequilibrium matrix

A ${p\times{p}}$ matrix:

$$\hat{R} = \frac{1}{n}X^{\top}X$$

given each column $\mathbf{x}_{j}$ standardized (mean 0, SD 1).

:::

\centerline{$R_{ij} = \frac{1}{n} \sum_{r=1}^{n} X_{ri} X_{rj}$}

:::
::::::

## The covariance of $X$ between the individuals (rows)

```{r}
xx <- data$genotypes[1:500, 1:700]
```

:::::: {.columns}
::: {.column width=.5}

```{r fig.width=2.8, fig.height=2.5}
.matshow(pmin(cov(t(xx)), .7), .lab=0, .axis.lab=0, .lw=0)
```

:::
::: {.column width=.5}

For a genotype matrix $X$ ${(n{\times}p)}$


::: {.block}
### Genetic relatedness matrix

An ${n\times{n}}$ matrix:

$$\hat{K} = \frac{1}{p}XX^{\top}$$

given each row $\mathbf{x}_{i}$ standardized (mean 0, SD 1).

:::

\centerline{$K_{ij} = \frac{1}{p} \sum_{g=1}^{p} X_{ig} X_{jg}$}

:::
::::::



## What would happen if there were AG-specific biases?

\Large

$$g\left( \underset{\textsf{phenotype}}{Y} \right)
\sim \underbrace{\sum_{k \in \textsf{causal variants}} X_{k} \beta_{k}}_{\textsf{true genetic effect}} +
\underbrace{\sum_{g} \mathbf{U}_{g} \gamma_{g}}_{\textsf{ancestry group factors}} + \epsilon$$


```{r}
set.seed(1)
.sim.biased <- simulate.pgs(X, W, h2=.3, pve.w=.3, n.causal=3)
Y <- .sim.biased$y.cc
```

$${}$$

Do GWAS as usual by variant-by-variant t-test:

```{r echo=T, size="large"}
.gwas.biased <- col_t_welch(X[Y == 0, ], X[Y == 1, ])
```

## Okay... do we have that many GWA-significant variants?

```{r fig.width=6, fig.height=2.8, only.plot="1"}
par(pch=19, col="gray40", cex=.5)
hist(.gwas.biased$pvalue, 50, main="", xlab="p-value")
```

```{r fig.width=6, fig.height=2.8, only.plot="2"}
par(pch=19, col="gray40", cex=.5)
plot(colnames(X), -log10(.gwas.biased$pvalue), xlab = "genomic location", ylab = "-log10 p-value")
abline(h=-log10(0.05/ncol(X)), lty=2, col=4, lwd=2)
points(colnames(X)[.sim.biased$causal], -log10(.gwas.biased$pvalue)[.sim.biased$causal], col=2, cex=2, lwd=2, pch=1)
legend("topleft",
       c("causal variant","Bonferroni"),
       col=c(2, 4),
       pch=c(1, NA), lwd=1.5,
       lty=c(0, 2),
       cex=2)
```

## How do we know there is an unknown group structure?

:::::: {.columns}
::: {.column width=.4}

::: {.block}
### GWAS assumption: 

- Only a small portion of variants are significantly associated. Why?

- A majority of p-values under the null will follow Uniform distribution.

:::

- What do we expect from Uniform p-value?

:::
::: {.column width=.6}

```{r}
nn <- length(.gwas.biased$pvalue)
p.expected <- seq(1, nn)/(nn+1)
```

```{r fig.width=3, fig.height=2.25, onslide.plot="2"}
par(cex=.5)
hist(p.expected)
```

```{r echo=T, eval=F, size="large"}
nn <- length(.gwas$pvalue)
p.expected <- seq(1, nn)/(nn+1)
```

:::
::::::

## However, when we draw quantile-quantile

```{r fig.width=5, fig.height=3}
par(cex=.5)
qqplot(-log10(p.expected), -log10(.gwas.biased$pvalue),
       main = "qqplot(GWAS expected, GWAS observed)",
       xlab = "-log10 P-value expected",
       ylab = "-log10 P-value estimated")
abline(a=0, b=1, col=2)
```


## Can we quickly fix this? A little bit...

```{r}
Y <- .sim.biased$y.cc
```

```{r echo = T, size = "large"}
.svd <- rsvd::rsvd(X, 10)
Y.adj <- lm(Y ~ .svd$u)$residual
.gwas.adj <- col_cor_pearson(X, Y.adj)
```

\large

Assuming

$$g\left( \underset{\textsf{phenotype}}{Y} \right)
\sim \underbrace{X_{k} \beta_{k}}_{\textsf{\color{teal} true genetic effect}} +
\underbrace{\sum_{g} \mathbf{U}_{g} \gamma_{g}}_{\textsf{\color{magenta} estimate this by SVD}} + \epsilon$$

<!-- ## Fix pop structures by regressing out -->
<!-- Use `qqplot(pvalue1, pvalue2)` -->
<!-- ```{r fig.width=5, fig.height=3} -->
<!-- par(cex=.5) -->
<!-- qqplot(-log10(p.expected), -log10(.gwas.adj$pvalue), -->
<!--        main = "GWAS", -->
<!--        xlab = "-log10 P-value expected", -->
<!--        ylab = "-log10 P-value estimated") -->
<!-- abline(a=0, b=1, col=2) -->
<!-- ``` -->


## Simply regressing out population PCs can reduce biases

```{r fig.width=6, fig.height=3, only.plot="1"}
par(pch=19, col="gray40", cex=.5)
plot(colnames(X), -log10(.gwas.biased$pvalue), xlab = "genomic location", ylab = "-log10 p-value", main = "before adjustement")
abline(h=-log10(0.05/ncol(X)), lty=2, col=4)
points(colnames(X)[.sim.biased$causal], -log10(.gwas.biased$pvalue)[.sim.biased$causal], col=2, cex=2, lwd=2, pch=1)
legend("topleft",
       c("causal variant","Bonferroni"),
       col=c(2, 4),
       pch=c(1, NA), lwd=1.5,
       lty=c(0, 2),
       cex=2)
```

```{r fig.width=5, fig.height=3, only.plot="2"}
par(cex=.5)
qqplot(-log10(p.expected), -log10(.gwas.adj$pvalue),
       xlab = "-log10 P-value expected",
       ylab = "-log10 P-value estimated",
       main = "QQ after adjustment")
abline(a=0, b=1, col=2)
```

```{r fig.width=6, fig.height=3, only.plot="3"}
par(pch=19, col="gray40", cex=.5)
plot(colnames(X), -log10(.gwas.adj$pvalue), xlab = "genomic location", ylab = "-log10 p-value", main = "after adjustment")
abline(h=-log10(0.05/ncol(X)), lty=2, col=4)
points(colnames(X)[.sim.biased$causal], -log10(.gwas.adj$pvalue)[.sim.biased$causal], col=2, cex=2, lwd=2, pch=1)
legend("topright",
       c("causal variant","Bonferroni"),
       col=c(2, 4),
       pch=c(1, NA), lwd=1.5,
       lty=c(0, 2),
       cex=2)
```

## Linear Mixed Effect model: a better way to adjust pop struct

:::::: {.columns}
::: {.column width=.48}

\large

A linear regression model:


$$\mathbf{y} = \mathbf{x}_{j} \underset{\textsf{\color{blue} a fixed genetic effect}}{\beta_{j}} + \boldsymbol{\epsilon}$$


What are we missing? Can we assume homo-scedasticity, i.e., 
$$\boldsymbol{\epsilon} \overset{?}{\sim} \mathcal{N}\!\left(\mathbf{0}, \sigma^{2}I\right)$$

:::
::: {.column width=.48}

\onslide<2->{

\large

A linear model with a random effect:

\large

$$\mathbf{y} = \mathbf{x}_{j} \underset{\textsf{\color{gray} fixed}}{\beta_{j}} + \underset{\textsf{\color{magenta} random effect}}{\mathbf{u}} + \boldsymbol{\epsilon}$$

{\emph Note}: Now, we assume:

$$\boldsymbol{\epsilon} \sim \mathcal{N}\!\left(\mathbf{0}, \sigma^{2}I\right)$$

}

:::
::::::


## A linear model with population-driven random effects - 2

- Unwanted random effects to \textbf{\color{magenta} remove}: $n \times 1$ vector $\mathbf{u}$

- \textbf{\color{blue} Goal}: the fixed genetic effect of a $j$.

\large

$$\mathbf{y} = \mathbf{x}_{j} \underset{\textsf{\color{blue} goal}}{\beta_{j}} + \underset{\textsf{\color{magenta} remove}}{\mathbf{u}} + \boldsymbol{\epsilon}$$

\begin{enumerate}
\item<2-> Note that $\mathbf{u}$ shouldn't be tied to a particular variant (by definition)
\item<3-> Also, the covariation of $\mathbf{u}$ is primarily driven by relatedness among individuals, not the variants. 
\end{enumerate}

\onslide<3->{
\centerline{$\mathbf{u} \sim \mathcal{N}\!\left(\mathbf{0}, \tau^{2} K\right),\quad K \approx \frac{1}{n} XX^{\top}$}
}


## LMM to estimate unbiased GWAS, removing random $\mathbf{u}$

A hierarchical model:

\begin{eqnarray}
\mathbf{y}|X,\beta,\mathbf{u},\sigma &\sim& \mathcal{N}\!\left(X\boldsymbol{\beta} + \mathbf{u}, \sigma^{2} I\right) \\
\mathbf{u}|\tau,K &\sim& \mathcal{N}\!\left(\mathbf{0}, \tau^{2} K\right)
\end{eqnarray}

If we integrate out $\mathbf{u}$,

$$\mathbf{y}|X,\beta \sim \mathcal{N}\!\left(\mathbf{y} \, \middle| \, X \boldsymbol{\beta}, \underbrace{\tau^{2} K}_{\textsf{\color{magenta} genetic-relatedness matrix}} + \underbrace{\sigma^{2} I}_{\textsf{\color{teal} irreducible}}\right)$$


## How did we handle high-dimensionality?


## A key question in LMM: What covariance matrix?


\only<1>{If there were many types of random effects,}

\centerline{$\mathbf{y} = \underset{\textsf{\color{magenta} fixed}}{X \boldsymbol{\beta}} + \underset{\textsf{\color{teal} random effects}}{\mathbf{u} + \mathbf{w} + \ldots} + \underset{\textsf{\color{gray} unknown}}{\epsilon}$}

\onslide<2->{
We would need to many covariance matrices:
\centerline{
$\mathbf{y}|\cdot \sim \mathcal{N}\!\left(X\boldsymbol{\beta},\,\sigma^{2}(I + \underbrace{\delta_{u} K_{u}  + \delta_{w} K_{w} + \ldots}_{\textsf{\color{teal} random effects}})\right)$}
}

\onslide<3>{
If we only care about variance decomposition $\beta_{j} \sim \mathcal{N}\!\left(0,\tau\right)$:
\centerline{
$\mathbf{y} \sim \mathcal{N}\!\left(\mathbf{0},\, \sigma^{2}\left(\underset{\textsf{\color{magenta} observed genetic}}{\frac{\sigma^{2}_{\textsf{genetic}}}{n} XX^{\top}} + I + \underbrace{\delta_{u} K_{u}  + \delta_{w} K_{w} + \ldots}_{\textsf{\color{teal} random effects}}\right)\right)$}
}

## Should we worry about "over-fitting" in LMM?

\large 

An equivalent question for PCA-based confounder adjustment: 

\vfill

> How many PCs to adjust in GWAS?

> Can we include a candidate SNP in the GRM $K$ matrix?

\vfill

\only<2>{
For each chromosome $c \in \{1,\ldots, 22, \textsf{X}, \textsf{Y}\}$, build a leave-one-chromosome-out (LOCO) kinship matrix, say $K_{-c}$:

$$\mathcal{N}\!\left(\mathbf{y} | \mathbf{x}_{j} \beta_{j}, \sigma^{2} (\delta K_{-c} + I)\right)$$

\vfill
Yang, {\it et al.}, \emph{Nature Genetics} (2014); Tucker, Price, Berger, \emph{Genetics} (2014)
}

## GWAS summary statistics within each pop

```{r message=F, warning=F, include=F}
Y <- .sim$y.cc
pops <- unique(pop.info$`Super Population`)
gwas.dt <- data.table()
for(pp in pops){
    .subset <- pop.info$`Super Population` == pp
    .gwas.p <- col_t_welch(X[which(Y == 0 & .subset), ],
                           X[which(Y == 1 & .subset), ])
    .gwas.dt.p <- setDT(.gwas.p)[, .(pvalue, mean.diff, stderr)] %>%
        mutate(physical.pos = as.integer(colnames(X)), pop = pp)
    gwas.dt <- rbind(gwas.dt, .gwas.dt.p)
}
```

```{r fig.width=6, fig.height=3, only.plot="1"}
ggplot(gwas.dt[pop == "EUR"], aes(physical.pos, -log10(pvalue))) +
    ggrastr::rasterise(geom_point(pch=19, size=.5, col="gray"), dpi=300) +
    geom_hline(yintercept = -log10(0.05/ncol(X)), col=2, lty=2) +
    xlab("genomic location") +
    ggtitle("GWAS in EUR (N=" %&% sum(pop.info$`Super Population`=="EUR") %&% ")")
```

```{r fig.width=6, fig.height=3, only.plot="2"}
ggplot(gwas.dt[sample(.N)],
       aes(physical.pos, -log10(pvalue), color=pop)) +
    ggrastr::rasterise(geom_point(pch=19, size=.5), dpi=300) +
    scale_color_brewer(palette = "Paired") +
    geom_hline(yintercept = -log10(0.05/ncol(X)), col=2, lty=2) +
    theme(legend.position = c(1,1)) +
    theme(legend.justification = c(1,1)) +
    theme(legend.title = element_blank()) +
    xlab("genomic location")
```


## Meta analysis: How can we combine multiple studies?

For each population/study $k$:

* $\hat\beta_{jk}$: effect size for a variant $j$ on a study $k$

* $s_{jk} = \mathsf{SE}(\beta_{jk})$: standard error for  a variant $j$ on a study $k$

* $w_{jk} = 1/s_{jk}^{2}$: inverse variance

::: {.block}
### Meta analysis by inverse variance weighting

$$\hat\beta_{j} = 
\frac{\sum_{k} w_{k} \hat\beta_{jk}}
{\sum_{k} w_{k}},
\quad\quad
\mathsf{SE}(\hat\beta_{j}) = 
\frac{1}
{\sqrt{ \sum_{k} w_{k} }}$$

:::

## Meta analysis can boost statistical power

```{r echo = T, eval = F, size="large"}
gwas.dt[, w := 1/(`stderr`^2 + 1e-16)]

meta.dt <-
    gwas.dt[!is.na(w),
            .(`mean` = sum(`mean.diff` * w)/sum(w),
              `se` = 1/sqrt(sum(w))),
            by = .(physical.pos)]

meta.dt[, `z` := `mean`/`se`]
meta.dt[, `pvalue` := 2*pnorm(abs(`z`), lower.tail=F)]
```

```{r}
do.meta <- function(.dt, var.min = 1e-16){
    .dt[, w := 1/(`stderr`^2 + var.min)]
    meta.dt <-
        .dt[!is.na(w),
                .(`mean` = sum(`mean.diff` * w)/sum(w),
                  `se` = 1/sqrt(sum(w))),
                by = .(physical.pos)]
    meta.dt[, `z` := `mean`/`se`]
    meta.dt[, `pvalue` := 2*pnorm(abs(`z`), lower.tail=F)]
    return(meta.dt)
}
```

##

```{r fig.width=6, fig.height=3, only.plot="1"}
.show.gwas.pop <- function(pp){
    meta.dt <- do.meta(gwas.dt[pop %in% pp])
    .causal <- meta.dt[physical.pos %in% colnames(X)[.sim$causal]]
    ggplot(meta.dt, aes(physical.pos, -log10(pvalue))) +
        ggrastr::rasterise(geom_point(pch=19, size=.5, color="gray40"), dpi=300) +
        geom_hline(yintercept = -log10(0.05/ncol(X)), col=2, lty=2) +
        ggtitle(paste0(pp, collapse = " + ")) +
        geom_point(data = .causal, pch=21, col=2, size=2)
}
.show.gwas.pop("EUR")
```

```{r fig.width=6, fig.height=3, only.plot="2"}
.show.gwas.pop(c("EUR", "EAS"))
```

```{r fig.width=6, fig.height=3, only.plot="3"}
.show.gwas.pop(c("EUR", "EAS", "AMR"))
```

```{r fig.width=6, fig.height=3, only.plot="4"}
.show.gwas.pop(c("EUR", "EAS", "AMR", "SAS"))
```

```{r fig.width=6, fig.height=3, only.plot="5"}
.show.gwas.pop(c("EUR", "EAS", "AMR", "SAS", "AFR"))
```


## What's next after GWAS?

\Large

- GWAS until 2010s: heavy focuses on **mapping**

    - GWAS map: genetic variants $\to$ a phenotype

    - Stringent genome-wide p-value cutoff

    - Study design, meta analysis

- NHGRI-EBI GWAS Catalog: [`https://www.ebi.ac.uk/gwas/`](https://www.ebi.ac.uk/gwas/)

##

\huge

Let's take a look at the GWAS Catalog:

[`https://www.ebi.ac.uk/gwas/`](https://www.ebi.ac.uk/gwas/)


## What's next after GWAS?

\Large

- GWAS since 2010s: more emphases on **prediction**

	- Can we turn GWAS results to a prediction model?

	- Can we understand the mechanisms?

    - Machine learning, data integration, causal inference


# What is a polygenic score model?

<!-- ## Today's lecture: GWAS (continued) and PGS -->
<!-- * Extension of basic GWAS methods -->
<!--     * Linear Mixed Effect model to handle population structures -->
<!-- 	* Meta analysis across multiple studies -->
<!-- * Polygenic risk prediction -->
<!--     * What is a polygenic risk score model? -->
<!--     * Why does it matter? -->
<!-- 	* How can we estimate PGS reliably? -->


## Polygenic score models predict disease prevalence based on genome

- Poly + genic = many genes (units of hereditary components)

\vfill

::: {.block}
### A (linear) polygenic score (PGS)

$$Y_{i} = \sum_{j} X_{ij} \beta_{j}$$

for an individual $i$.

:::

\onslide<2>{Knowing $\beta$'s, we can predict:
$${\color{teal}Y^{\star}}_{i} \gets \sum_{j} {\color{magenta} X_{ij}^{\star}} \hat{\boldsymbol{\beta}}_{j}$$}



## Example: PGS for breast cancer occurrence

:::::: {.columns}
::: {.column width=.7}

\includegraphics[width=\linewidth]{img/PGS_Yang_Pharoah_NR_Cancer_2023_BRCA.pdf}

:::
::: {.column width=.3}

\normalsize

* Strong heritability (proportion of disease risk variation explained by genetics): 35% - 80%

* Prediction more accurate than life style and other environmental factors

* More powerful if combined with rare risk factors

:::
::::::

\vfill

\tiny

Yang .. Pharoah, Nature Review Genetics (2023)

## Example: PGS for coronary artery disorder $\to$ prevention

\centerline{\includegraphics[width=.8\linewidth]{img/PGS_Khera_Kathiresan_NatGen_2018_CAD.pdf}}

* Top .5% of PGS values $\to$ five fold increase of CAD

\vfill

\tiny

Khera .. Kathiresan, Nature Genetics (2018)

## Example: PGS stratifies individuals' disease onset \& risk

\centerline{\includegraphics[width=.8\linewidth]{img/PGS_Mars_Ripatti_Nature_Medicine_2020_1.pdf}}

\vfill

* PRS $\approx$ PGS.

* We can partition cohorts based on PGS profiles.

\tiny

Mars .. Ripatti, Nature Medicine (2020)

## A toy example: GWAS for "Obsessive `ggplot` Disorder"^[just an illustration purpose]

```{r include=F}
#' @param X genotype matrix
#' @param W population structure
#' @param h2 heritability
#' @param pve.w population PVe
#' @param n.causal
#' @param n.traits
simulate.pgs <- function(X, W, h2, pve.w, n.causal, n.traits = 1) {
    .rnorm <- function(d1, d2) matrix(rnorm(d1*d2), d1, d2)
    causal.snp <- sample(ncol(X), n.causal)
    xx.causal <- apply(X[, causal.snp, drop=FALSE], 2, scale)
    xx.causal[is.na(xx.causal)] <- 0

    n.ind <- nrow(X)
    y.true <- apply(xx.causal %*% .rnorm(n.causal, n.traits), 2, scale)

    if(pve.w > 0){
        y.pop <- apply(W %*% .rnorm(ncol(W), n.traits), 2, scale)
    }else{
        y.pop <- 0
    }
    y.err <- apply(.rnorm(n.ind, n.traits), 2, scale)

    y.obs <- y.true * sqrt(h2) + y.pop * sqrt(pve.w)
    y.obs <- y.obs + y.err * sqrt(1 - h2 - pve.w)

    prop <- 1/(1+ exp(-y.obs))
    y.cc <- rbinom(length(prop), 1, prop)

    list(y.q = y.obs, y.cc = y.cc, pgs = y.true,
         x = X, w = W, causal = causal.snp)
}
```

```{r}
## Take first chromosome
chr1 <- which(data$map$chromosome==1)
X.tot <- data$genotypes[, chr1]
rownames(X.tot) <- data$fam$sample.ID
colnames(X.tot) <- data$map$physical.pos[chr1]

## population structure
W.tot <-
    pop.info %>%
    mutate(one = 1) %>%
    dcast(`sample.ID` ~ `Super Population`, value.var = "one", fill = 0) %>%
    (function(x) left_join(data$fam[, "sample.ID", drop = F], x)) %>%
    column_to_rownames("sample.ID") %>%
    as.matrix()
```

```{r}
set.seed(1)
.sim <- simulate.pgs(X.tot, W.tot, h2=.3, pve.w=0, n.causal=7)
.train <- sample(nrow(X.tot), 1500)
Y <- .sim$y.cc[.train]
X <- .sim$x[.train, , drop = F]
```

A genotype matrix $X$ ($X_{ij} \in \{0,1,2\}$):

```{r fig.width=6, fig.height=2.3}
.matshow(X[1:30, 1:80], .lw=.01, .lab=1.5)
```

## In GWAS, we can have case vs. control phenotype

\large

$Y_{i} = 1$ if case vs. $Y_{i} = 0$ if control

For our *OGD* GWAS:

```{r}
y <- .sim$y.cc[.train]
```

:::::: {.columns}
::: {.column width=.2}

:::
::: {.column width=.8}


```{r echo=T, size="large"}
table(y)
```

:::
::::::


- E.g., Cancer vs. non-cancer, schizophrenia vs. no mental disorder

- The fundamental question is, "Which individuals/subjects can be truly labelled control/wild type?"

## The under the hood, there are quantitative "risk" scores

```{r fig.width=5, fig.height=2}
ggplot(data.table(y=as.vector(.sim$y.q)), aes(y)) +
    geom_histogram() +
    xlab("phenotypic variation")
```

- E.g., height, BMI, parents' age of death, how man pack of cigarettes, etc.

## Side note: Almost all complex traits are polygenic

Polygenic effects of common SNPs $\to$ polymorphism in phenotypes

\vfill

> A polygenic trait is a characteristic, such as height or skin color, that is influenced by two or more genes. **Because multiple genes are involved, polygenic traits do not follow the patterns of Mendelian inheritance.** Many polygenic traits are also influenced by the environment and are called multifactorial.

\vfill

https://www.genome.gov/genetics-glossary/Polygenic-Trait


## A polygenic score to predict disease prevalence

```{r}
.dt <- data.table(y = as.vector(.sim$y.q),
                  disease = factor(.sim$y.cc, c(0, 1), c("without OGD","with OGD")))
```

```{r fig.width=5.5, fig.height=2.5, only.plot="1"}
ggplot(.dt[disease == "without OGD"], aes(y)) +
    xlab("polygenic risk score (log)") +
    geom_density(alpha = .35, fill = "#000099") +
    geom_vline(xintercept = 0, lty = 2, col = 1) +
    ggtitle("polygenic score of healthy individuals")
```

```{r fig.width=5.5, fig.height=2.5, only.plot="2"}
ggplot(.dt[disease == "with OGD"], aes(y)) +
    xlab("polygenic risk score (log)") +
    geom_density(alpha = .35, fill = "red") +
    geom_vline(xintercept = 0, lty = 2, col = 1) +
    ggtitle("polygenic score of OGD individuals")
```

```{r fig.width=5.5, fig.height=2.8, only.plot="3"}
ggplot(.dt, aes(y, fill = disease, group = disease)) +
    xlab("polygenic risk score (log)") +
    geom_density(alpha = .35) +
    geom_vline(xintercept = 0, lty = 2, col = 1) +
    theme(legend.position = c(0,1),
          legend.justification = c(0,1)) +
    scale_fill_brewer(palette="Set1", direction=-1)
```

## Polygenic score $\propto$ the odds of the case vs. control in GWAS

Log-odds ratio:

$$g(y) = \log\frac{p(\textsf{disease}|Y > y)}{p(\textsf{no disease}|Y > y)}$$

\vfill

```{r fig.width=5, fig.height=1.5, onslide.plot="2"}
count.cc <- function(.tab, .cutoff){
    .tab[y > .cutoff,
        .(.N),
        by = .(disease)] %>%
        mutate(y = .cutoff)
}

n.tot <-
    .dt[, .(ntot = .N), by = .(disease)]

n.cc <-
    lapply(seq(-1.5, 1.5, by = .25), count.cc, .tab = .dt) %>%
    do.call(what = rbind) %>%
    left_join(n.tot) %>%
    mutate(pr = `N`/`ntot`) %>%
    dcast(y ~ disease, value.var = "pr")

.aes <- aes(y, log10(`with OGD`) - log10(`without OGD`))
ggplot(n.cc, .aes) +
    geom_line() +
    geom_point(pch=21, fill = "white") +
    xlab("polygenic score cutoff")+
    scale_y_continuous("odds ratio \nOGD vs. no-OGD",
                       labels = function(x)round(10^x, 2))
```

## A PGS model to explain disease prevalence with genetics

Log-odds ratio:

$$g(y) = \log\frac{p(\textsf{disease}|Y > y)}{p(\textsf{no disease}|Y > y)}$$

\vfill

**Goal**: Estimate a function to predict this log-odds ratio.

\only<1>{$$g(y_{i}) \sim \beta_{1} X_{i1} + \beta_{2} X_{i2} \cdots + \epsilon$$}

\only<2>{$$g(y_{i}) \sim \underbrace{\beta_{1} X_{i1} + \beta_{2} X_{i2} \cdots }_{\textsf{\color{teal}\bf genetic effects}} + \epsilon$$}


## PGS estimation is a linear modelling with ${p{\gg}n}$

\only<1>{$$Y_{i} \sim X_{i1} \beta_{1} + X_{i2} \beta_{2} + \cdots$$}

\only<2>{$$\left(
\begin{array}{l}
Y_{1}\\
Y_{2}\\
\vdots\\
Y_{n}
\end{array}
\right)
\sim
\left(
\begin{array}{l}
X_{11}\\
X_{21}\\
\vdots\\
X_{n1}
\end{array}
\right)
\beta_{1} +
\left(
\begin{array}{l}
X_{12}\\
X_{22}\\
\vdots\\
X_{n2}
\end{array}
\right)
\beta_{2} +
\cdots$$}

\onslide<3-5>{$$\left(
\begin{array}{l}
Y_{1}\\
Y_{2}\\
\vdots\\
Y_{n}
\end{array}
\right)
\sim
\left(
\begin{array}{l}
X_{11}\\
X_{21}\\
\vdots\\
X_{n1}
\end{array}
\right)
\beta_{1} +
\left(
\begin{array}{l}
X_{12}\\
X_{22}\\
\vdots\\
X_{n2}
\end{array}
\right)
\beta_{2} +
\cdots +
\left(
\begin{array}{l}
X_{1p}\\
X_{2p}\\
\vdots\\
X_{np}
\end{array}
\right)
\beta_{p}
$$
}

\onslide<4-5>{
$$\mathbf{y} \sim \mathbf{x}_{1} \beta_{1} + \mathbf{x}_{2} \beta_{2} + \cdots + \mathbf{x}_{p} \beta_{p}$$

$$p{\gg}n$$
}

\only<5>{\large
$n \approx 10^{4}$
but
$p \approx 10^{6}$ for many large-scale GWAS}

## $p{\gg}n$: Why can't we just fit a model?

```{r echo = T, eval = F, size = "large"}
## lm.out <- lm(Y ~ X)
```

\large

\begin{itemize}[<+-|alert@+>]
\item $p \gg n$: need to estimate $\beta_{1},\ldots,\beta_{p}$
\item How many samples? How many unknowns?
\item Is it computationally feasible?
\end{itemize}

## $p{\gg}n$: Variant-by-variant GWAS is not a bad idea

```{r echo=T, size="large"}
.gwas <- col_t_welch(X[Y == 0, ], X[Y == 1, ]) # univar.
```

```{r onslide.plot="2", fig.width=6, fig.height=2.2}
par(pch=19, col="gray40", cex=.5)
.df <- data.table(x=colnames(X), y=-log10(.gwas$pvalue))
.df.c <- .df[.sim$causal]
ggplot(.df, aes(as.integer(x),y)) +
    ggrastr::rasterise(geom_point(stroke=0, color="gray"),dpi=300) +
    geom_point(data=.df.c, col=2, pch=21, stroke=1.5, size=2) +
    geom_hline(yintercept = -log10(0.05/ncol(X)), lty=2, col=2) +
    theme(legend.position = c(0,1)) +
    theme(legend.justification = c(0,1)) +
    xlab("genomic locations (chromosome 1)") +
    ylab("-log10 p-value")
```

## A linear model including only these "GWAS" variants

:::::: {.columns}
::: {.column width=.45}


```{r echo = T, size="large"}
X.gwas <- X[, p.adjust(.gwas$pvalue) < .05]
head(X.gwas, 3)
```

:::
::: {.column width=.45}

```{r}
Y <- .sim$y.q[.train]
```

```{r echo = T, size="large"}
lm.out <- lm(Y ~ X.gwas)
```

:::
::::::

:::::: {.columns}
::: {.column width=.2}

:::
::: {.column width=.8}

```{r echo=T, size="large"}
anova(lm.out)
```

:::
::::::

##

:::::: {.columns}
::: {.column width=.5}

```{r fig.width=3, fig.height=2.8, onslide.plot="1-2"}
.df <- data.frame(y.hat = lm.out$fitted.values, y = Y)
ggplot(.df, aes(y.hat, y)) +
    geom_jitter(stroke=0, color="gray") +
    ggpubr::stat_regline_equation(size=4) +
    ggtitle("Train (N = " %&% nrow(X) %&% ")") +
    geom_smooth(method="lm", se=F, col=2)
```
:::
::: {.column width=.5}

```{r fig.width=3, fig.height=2.8, onslide.plot="2"}
x.test <- cbind(1, .sim$x[- .train, p.adjust(.gwas$pvalue) < .05])
y.test <- .sim$y.q[- .train]
y.hat.test <- x.test %*% as.matrix(coefficients(lm.out), ncol=1)

.df <- data.frame(y.hat = y.hat.test, y = y.test)
ggplot(.df, aes(y.hat, y)) +
    geom_jitter(stroke=0, color="gray") +
    ggpubr::stat_regline_equation(size=4) +
    ggtitle("Test (N = " %&% nrow(x.test) %&% ")") +
    geom_smooth(method="lm", se=F, col=2)
```

:::
::::::


## Do we need all of these GWAS variants?

```{r fig.width=5.5, fig.height=2.5}
.dt <-
    data.table(X.gwas, Y) %>%
    melt(id.vars = "Y")

ggplot(.dt, aes(factor(value, c(0,1,2)), Y, fill=as.factor(value))) +
    xlab("genotype") + ylab("logit phenotype") +
    facet_grid(.~variable) +
    geom_boxplot(show.legend = F) +
    scale_fill_brewer("",palette = "Paired")
```

## Just one of the GWAS variants works fine

```{r}
lm.out <- lm(Y ~ X.gwas[,1])
```

:::::: {.columns}
::: {.column width=.5}

```{r fig.width=3, fig.height=2.5}
.df <- data.frame(y.hat = lm.out$fitted.values, y = Y)
ggplot(.df, aes(y.hat, y)) +
    xlab("prediction") + ylab("training data") +
    geom_jitter(stroke=0, size=1, color="gray") +
    ggpubr::stat_regline_equation(size=4) +
    ggtitle("Train (N = " %&% nrow(X) %&% ")") +
    geom_smooth(method="lm", se=F, col=2)
```
:::
::: {.column width=.5}

```{r fig.width=3, fig.height=3}
x.test <- cbind(1, .sim$x[- .train, p.adjust(.gwas$pvalue) < .05][, 1])
y.test <- .sim$y.q[- .train]
y.hat.test <- x.test %*% as.matrix(coefficients(lm.out), ncol=1)

.df <- data.frame(y.hat = y.hat.test, y = y.test)
ggplot(.df, aes(y.hat, y)) +
    xlab("prediction") + ylab("testing data") +
    geom_jitter(stroke=0, size=1, color="gray") +
    ggpubr::stat_regline_equation(size=4) +
    ggtitle("Test (N = " %&% nrow(x.test) %&% ")") +
    geom_smooth(method="lm", se=F, col=2)
```

:::
::::::

## Simply add more variables?

```{r}
.dt.gwas <-
    data.table(pv = .gwas$pvalue, beta = -.gwas$mean.diff) %>%
    mutate(j = 1:n()) %>%
    mutate(x = as.integer(colnames(.sim$x))) %>% 
    arrange(pv)

pred.gwas <- function(ntop){
    xx <- .sim$x[- .train, .dt.gwas[1:ntop, ]$j, drop = F]
    data.table(y.hat = as.vector(xx %*% .dt.gwas[1:ntop, ]$beta),
               y = .sim$y.cc[- .train],
               y.q = .sim$y.q[- .train])
}

plot.pgs.gwas <- function(ntop){
    .df <- pred.gwas(ntop)
    ggplot(.df, aes(y.hat, y.q)) +
        xlab("prediction") + ylab("testing data") +
        geom_jitter(stroke=0, size=1, color="gray") +
        ggpubr::stat_cor(size=4) +
        ggtitle("Top " %&% ntop %&% ifelse(ntop>1," variants"," variant")) +
        geom_smooth(method="lm", se=F, col=2)
}

cor.pgs.gwas <- function(ntop){
    .df <- pred.gwas(ntop)
    return(cor(.df$y.q, .df$y.hat, method = "spearman"))
}
```

Let's do some experiments:

\large

1. Pick top *K* variants (ordering by GWAS, $p_{(1)} < p_{(2)} < \ldots$)

2. Take GWAS effect sizes $\beta$ (the mean difference between case vs. control)

3. Predict by a linear combination of these effects:

$$\sum_{j=1}^{K} X_{i(j)} \beta_{(j)}$$

## PGS is a variable selection problem

:::::: {.columns}
::: {.column width=.5}

```{r fig.width=3, fig.height=2.5, only.plot="1"}
plot.pgs.gwas(1)
```
```{r fig.width=3, fig.height=2.5, only.plot="2"}
plot.pgs.gwas(10)
```
```{r fig.width=3, fig.height=2.5, only.plot="3"}
plot.pgs.gwas(20)
```
```{r fig.width=3, fig.height=2.5, only.plot="4"}
plot.pgs.gwas(30)
```
```{r fig.width=3, fig.height=2.5, only.plot="5"}
plot.pgs.gwas(50)
```

:::
::: {.column width=.5}

```{r fig.width=3, fig.height=2.5, onslide.plot="5-"}
nn.vec <- c(1:75)
.dt <- data.table(p = nn.vec,
                  r = sapply(nn.vec, cor.pgs.gwas))

ggplot(.dt, aes(`p`, `r`)) +
    geom_line() +
    ylab("Spearman correlation") +
    xlab("Number of variables in PGS")
```

:::
::::::

# What are the statistical challenges in PGS estimation?

## GWAS statistics only roughly tag causal variants

\centerline{\includegraphics[width=.8\linewidth]{img/direct_indirect_associations.pdf}}

\vfill

\tiny

Balding *et al.* Nature Review Genetics (2006)

## Covariance between variants obfuscates GWAS interpretation

:::::: {.columns}
::: {.column width=.3}

* Linkage disequilibrium (LD) block

* The result of recombination events throughout generations

:::
::: {.column width=.7}

\includegraphics[height=.7\textheight]{img/altshuler_daly_lander_LD.pdf}

:::
::::::


## A typical setting of PGS training

::: {.block}
### PGS training

* Input:

    1. GWAS summary statistics (p-values, univariate effect sizes)

	2. $n{\times}p$ genotype matrix $X$, where $X_{ij}\in\{0,1,2\}$

	3. $n{\times}1$ phenotype vector $\mathbf{y}$

* Output:

    - *Polygenic* effect sizes $\hat{\beta}_{1},\ldots,\hat{\beta}_{p}$

* Objective:

    - We want to predict unseen $Y^{\star}_{i}$ by $f(X^{\star}; \hat{\beta})$ as accurately as possible

	- We want $\beta_{j} = 0$ as many as possible; only some $\beta_{j}{\neq}0$.

:::


## Common strategies to deal with LD structures

:::::: {.columns}
::: {.column width=.3}

Strategy 1. Just use them all

:::
::: {.column width=.3}

Strategy 2. Pruning/clumping

:::
::: {.column width=.3}

Strategy 3. **Fine-mapping**

:::
::::::

\vfill

:::::: {.columns}
::: {.column width=.3}
\includegraphics[width=\linewidth]{img/GWAS_LD_strategy_useall.pdf}
:::
::: {.column width=.3}
\includegraphics[width=\linewidth]{img/GWAS_LD_strategy_pruning.pdf}
:::
::: {.column width=.3}
\includegraphics[width=\linewidth]{img/GWAS_LD_strategy_fm.pdf}
:::
::::::


## True causal variables explain a large fraction of variation

```{r echo = TRUE}
.lm <- lm(.sim$y.q ~ .sim$x[, .sim$causal, drop = FALSE] - 1)
```

```{r fig.width = 4, fig.height = 2.8}
.dt <- data.table(y = as.numeric(.sim$y.q), y.hat = fitted(.lm))
.aes <- aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~"))
ggplot(.dt, aes(y.hat, y)) +
    geom_point(color = "gray45", stroke=0, size=1) +
    geom_abline(slope = 1, color = 2, lty = 2) +
    ggpubr::stat_regline_equation(.aes, size=4) +
    xlab("predicted by causal variants") +
    ylab("true polygenic score")
```


# Statistical fine-mapping to handle LD structures

## Classical statistics does not really help

\large

- Classical variable selection by univariate (one-by-one) tests will not work for a $p \gg n$ regression problem


- Especially if we have col-linearity in the design matrix $X$

- also know as Linkage Disequilibrium (LD) in GWAS/PGS


## Variable selection in high-dim. matrix ($n \ll p$)

Regression = projecting $\mathbf{y}$ on to columns $\{\mathbf{x}_{j}: j \in[p]\}$,
$$
\left(\begin{array}{l}
y_{1}\\
y_{2}\\
\vdots\\
y_{n}
\end{array}\right)
=
\beta_{1} \left(\begin{array}{l}
X_{11}\\
X_{21}\\
\vdots\\
X_{n1}
\end{array}\right) +
\cdots
\beta_{p} \left(\begin{array}{l}
X_{1p}\\
X_{2p}\\
\vdots\\
X_{np}
\end{array}\right).
$$
Variable selection = column selection.

\vfill

\begin{itemize}
\item <1-> Intuitive idea : choose the best combination of variables. $\to 2^{p}$ choices.
\item <2-> Alternative idea : make as many $\beta_{j}$'s nearly zero values.
\item <3-> What prior does: penalize $|\beta_{j}| > 0$ so that only the strong
  enough variables take non-zero values.
\end{itemize}

## Lasso, a linear regression with Laplace prior (L1)

Prior distribution
$$
p(\boldsymbol{\theta}) = \textsf{Laplace}(\boldsymbol{\theta}| \lambda) \propto \exp\left(-\lambda\|\boldsymbol{\theta}\|_{1}\right)
$$
where
\large
$$\|\boldsymbol{\theta}\|_{1} = \sum_{j=1}^{p} |\theta_{j}|,\,\textsf{\color{blue}L1-norm}.$$

\normalsize
Maximize
$$
\ln p(\mathbf{y}|X,\boldsymbol{\theta}) + \ln p(\boldsymbol{\theta}|\lambda)
= - \frac{1}{2\sigma^{2}} \sum_{i=1}^{n} (y_{i} - \mathbf{x}_{i} \boldsymbol{\theta})^{2}
- \lambda \|\boldsymbol{\theta}\|_{1}
$$

Minimize $L_{1}$-regularized error
$$
\sum_{i=1}^{n} (y_{i} - \mathbf{x}_{i} \boldsymbol{\theta})^{2}
+ \lambda \|\boldsymbol{\theta}\|_{1}
$$

\vfill

(Tibshirani, 1996)

## `glmnet` solves this regularized optimization problem

Goal (by variable-by-variable updates):\centerline{$\min_{\boldsymbol{\beta}} \quad
\overbrace{(\mathbf{y} - X\boldsymbol{\beta})^{\top}(\mathbf{y} - X\boldsymbol{\beta})}^{\textsf{\color{blue} RSS}} + \underbrace{\lambda \alpha \|\boldsymbol{\beta}\|_{1}}_{\textsf{\color{red} variable selection}} + \underbrace{\lambda (1 - \alpha) \|\boldsymbol{\beta}\|_{2}}_{\textsf{\color{magenta} shrinkage}}$}

\onslide<2->{
For each $\beta_{j}$,
}

\only<2>{$$
\hat{\beta}_{j}^{\textsf{glmnet}} \gets
\frac{S\left(
\sum_{i=1}^{n} X_{ij} (y_{i} - \hat{y}_{i}^{(-j)}),
\lambda\alpha
\right)}
{ \sum_{i=1}^{n} X_{ij}^{2} +\lambda (1- \alpha) }
$$
\tiny
Friedman {\it et al.}, Regularization Paths for Generalized Linear Models via Coordinate Descent (2010)
}

\only<3>{$$
\hat{\beta}_{j} \gets
\frac{\overset{\textsf{\color{red} threshold}}{S}
\left(
\sum_{i=1}^{n} X_{ij} \overbrace{(y_{i} - y_{i}^{(-j)})}^{\textsf{\color{red} residual w/o the variable } \beta_{j} },
\lambda\alpha
\right)}
{ \sum_{i=1}^{n} X_{ij}^{2} + \underbrace{\lambda (1- \alpha)}_{\textsf{\color{magenta} shrinkage}}}
$$
where $S(z, \tau)$ will set it to zero if $|z| < \tau$.
}


## Running `glmnet` for polygenic risk prediction of OGD

```{r}
Y <- .sim$y.q[.train]
X <- .sim$x[.train, , drop = F]
```

```{r echo=TRUE}
glm.cv.out <- glmnet::cv.glmnet(X, Y, nfolds=5, alpha=1)
lambda.cv <- glm.cv.out$lambda.min
glm.out <- glmnet::glmnet(x=X, y=Y, lambda = lambda.cv, alpha=1)
```

```{r}
lasso.dt <-
    data.table(beta.lasso = as.numeric(glm.out$beta)) %>%
    mutate(j = 1:n()) %>%
    left_join(.dt.gwas) %>%
    mutate(l = as.integer(j %in% .sim$causal)) %>%
    arrange(desc(abs(beta.lasso))) %>%
    mutate(lasso = 1:n()) %>%
    as.data.table
```

:::::: {.columns}
::: {.column width=.3}

```{r echo = T}
head(glm.out$beta)
```

:::
::: {.column width=.65}

```{r fig.width=3, fig.height=2, only.plot="2"}
ggplot(lasso.dt, aes(lasso, beta.lasso, fill=as.factor(l))) +
    theme(legend.position = c(1,1)) +
    theme(legend.justification = c(1,1)) +
    xlab("variables (ranked by glmnet)") + ylab("coefficients") +
    geom_point(data=head(lasso.dt, 100), pch=21) +
    geom_hline(yintercept = 0, color="black") +
    scale_x_sqrt() +
    scale_fill_manual("causal", values=c("gray", "red"))
```

:::
::::::


## Sum of Single Effect (SuSiE) regression

```{r run_susie, size="large", echo = T}
library(susieR)
susie.out <- susie(X, Y)
```

\vfill

\large
$$\mathbf{y} = \sum_{l=1}^{L} \underbrace{ \sum_{j} \mathbf{x}_{j} 
    \overset{\textsf{\color{magenta}probabilistic selection}}{\alpha_{j}^{(l)}} 
	\overset{\textsf{\color{teal} single variant effect}}{\beta_{j}^{(l)}} }_{\textsf{layer-by-layer}}  + \boldsymbol{\epsilon}$$

\normalsize
where $\sum_{j=1}^{p} \alpha_{j}^{(l)} = 1$ for each layer $l$.

\vfill

\small

Wang .. Stephens, _Journal of the Royal Statistical Society_ (2020)


## Ideas behind SuSiE

\large

1. Estimating univariate effects is easy (GWAS)

2. Many variants can show similar effects (LD)

3. Let's weight variants probabilistically

4. Regress out the probabilistically reweighted effects

5. Repeat 1-4.

## SuSiE identifies top causal variants

```{r echo = FALSE}
susie.dt <- 
    data.table(pip = susie_get_pip(susie.out)) %>%
    mutate(j = 1:n()) %>% 
    left_join(.dt.gwas) %>%
    mutate(l = as.integer(j %in% .sim$causal)) %>%
    arrange(desc(pip)) %>%
    mutate(susie = 1:n())
```

```{r fig.width=3, fig.height=2.8, echo = FALSE}
ggplot(susie.dt, aes(susie, pip, fill=as.factor(l))) +
    theme(legend.position = c(1,1)) +
    theme(legend.justification = c(1,1)) +
    xlab("variables (ranked by SuSiE)") +
    ylab("posterior probability") +
    geom_point(data=head(susie.dt, 100), pch=21) +
    geom_hline(yintercept = 0, color="black") +
    scale_x_sqrt() +
    scale_fill_manual("causal", values=c("gray", "red"))
```

## SuSiE can avoid the col-linearity (LD) problem

```{r fig.width=6, fig.height=2, echo = FALSE, only.plot="1"}
logit <- function(x) log(x) - log(1-x)
sigmoid <- function(x) 1/(1+exp(-x))

.aes <- aes(x, logit(pmin(pmax(pip,1e-2),1-1e-2)), colour=as.factor(l))

p1 <-
    ggplot(susie.dt[order(`pip`)], .aes) +
    theme(legend.position = c(1,1)) +
    theme(legend.justification = c(1,1)) +
    xlab("variables (along genomic axis)") +
    scale_y_continuous("posterior probability", labels=function(x) round(sigmoid(x),2)) +
    geom_vline(data=susie.dt[pip > .5], aes(xintercept=x), lty=2, size=.5, color=3) +
    ggrastr::rasterise(geom_point(stroke=0, size=1), dpi=300) +
    geom_hline(yintercept = 0, color="black", lty = 2, size = .2) +
    scale_colour_manual("causal", values=c("gray", "red"))
p1
```

```{r fig.width=6, fig.height=3, echo = FALSE, only.plot="2"}
p2 <-
    ggplot(susie.dt[order(`l`)], aes(x, -log10(pv), colour=as.factor(l))) +
    geom_vline(data=susie.dt[pip > .5], aes(xintercept=x), lty=2, size=.5, color=3) +
    ggrastr::rasterise(geom_point(stroke = 0, size=1), dpi=300) +
    scale_colour_manual("causal", values=c("gray", "red"), guide=FALSE) +
    xlab("variables (along genomic axis)")
p1/p2
```

```{r fig.width=6, fig.height=3.5, echo = FALSE, only.plot="3"}
susie_plot(susie.out, y = "PIP", add_bar = TRUE, add_legend = TRUE)
```

```{r fig.width=6, fig.height=3.5, echo = FALSE, only.plot="4"}
susie_plot(susie.out, y = "log10PIP", add_legend = TRUE)
```

## SuSiE+PGS accurately predict the held-out data

```{r}
x.test <- .sim$x[- .train, , drop = F]
y.test <- .sim$y.q[- .train]
```

```{r echo = T, tidy=T}
y.hat <- predict.susie(susie.out, x.test)
```

```{r fig.width = 4, fig.height = 2.5}
.dt <- data.table(y = y.test, y.hat = y.hat)
.aes <- aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~"))
ggplot(.dt, aes(y.hat, y)) +
    geom_point(color = "gray45", stroke=0, size=1) +
    geom_abline(slope = 1, color = 2, lty = 2) +
    ggpubr::stat_regline_equation(.aes, size=4) +
    xlab("predicted by causal variants") +
    ylab("true polygenic score")
```

## A general workflow of PGS estimation

\only<1>{\centerline{\includegraphics[height=.7\textheight]{img/PGS_Choi_Protocol_1.pdf}}}

\only<2>{\centerline{\includegraphics[height=.7\textheight]{img/PGS_Choi_Protocol_2.pdf}}}

\vfill

\small

Choi .. O'Reilly, Nature Protocol, (2020)


## A proposed workflow of PGS estimation

\large

1. Take GWAS summary statistics and target population genotype $X$

2. Run SuSiE^[SuSiE can run with summary statistics only] chromosome by chromosome to select variables

3. Predict PGS per chromosome and aggregate them

<!------------->
<!-- Wrap up -->
<!------------->

## PGS Catalog: No need to train from scratch

\large

`https://www.pgscatalog.org/`

\vfill

\centerline{\includegraphics[height=.6\textheight]{img/PGS_catalog.png}}

## Advanced topics that we might consider next lecture

\large

- Summary-based GWAS, post-GWAS analysis

- Mendelian Randomization and other causal inference techniques

