---
title: "Multiple Hypothesis Testing (in DEG, GWAS)"
author: "Yongjin Park"
date: "12 February 2026"
classoption: "aspectratio=169"
fontsize: 12pt
output:
    beamer_presentation:
        theme: Madrid
        keep_md: true
        keep_tex: true
        latex_engine: xelatex
        slide_level: 2
header-includes:
  - \usepackage{cancel}
  - \usepackage{tikz}
  - \usetikzlibrary{positioning,fit}
  - \definecolor{UBCblue}{rgb}{0.04706, 0.13725, 0.26667}
  - \usecolortheme[named=UBCblue]{structure}
  - \setbeamertemplate{frametitle}{\color{UBCblue}\bfseries\insertframetitle\par\vskip-6pt\hrulefill}
  - \setbeamercolor{title in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{section in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{author in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{date in head/foot}{bg=white,fg=gray}
  - \setbeamertemplate{page number in head/foot}{}
  - \setbeamertemplate{frame numbering}[none]
  - \setbeamercolor{alerted text}{bg=yellow}
  - \AtBeginSection[]{\begin{frame}\frametitle{Today's lecture}{\Large\tableofcontents[currentsection]}\end{frame}}
  - |
    \makeatletter
    \def\ps@titlepage{%
      \setbeamertemplate{footline}{}
    }
    \addtobeamertemplate{title page}{\thispagestyle{titlepage}}{}
    \makeatother
    \include{toc}
---


```{r setup, include=FALSE}
source("Setup.R")
fig.dir <- "../Fig/multiple-hypothesis/"
dir.create(fig.dir, showWarnings=FALSE)
setup.env(fig.dir)
dir.create("../data", showWarnings=FALSE)

library(data.table)
library(tidyverse)
library(patchwork)
library(matrixTests)

theme_set(theme_classic() + theme(title = element_text(size=10)))
```


## An example: Genome-wide Association Studies

\large

::: {.block}
### GWAS

* Input:

    - A genotype matrix $X$ ${(n{\times}p)}$, where $X_{ij} \in \{0, 1, 2\}$

    - A phenotype vector $\mathbf{y}$ ${(n{\times}1)}$

* Optimization:

\centerline{$\mathbf{y} = \mathbf{x}_{j} \beta_{j} + \epsilon,\, \textsf{for each }j$}

* Output:

    - Estimate an effect size (coefficient) $\beta_{j}$ ${(j \in 1,\ldots, p)}$

:::


* Testing $H_{0}: \beta_{j} = 0$ vs. $H_{1}: \beta_{j} \neq 0$.

\normalsize



```{r download_1000g_data, include=F}
dir.create("../data/genotype/", recursive=TRUE, showWarnings=FALSE)

library(bigsnpr)
.bed.file <- "../data/genotype/1000G_phase3_common_norel.bed"
if.needed(.bed.file, {
    download_1000G("../data/genotype/")
})

.bk.file <- "../data/genotype/1000G_phase3_common_norel.rds"
if.needed(.bk.file, {
    BED <- snp_readBed(.bed.file)
})
```

```{r include=F}
.fam.file <- "../data/genotype/1000G_phase3_common_norel.fam"
.bim.file <- "../data/genotype/1000G_phase3_common_norel.bim"
data <- snp_attach(.bk.file)
```

```{r include = F}
pop.info.file <- "../data/genotype/1000G_phase3_common_norel.fam2"
pop.info <- fread(pop.info.file)
```

## GWAS to map locations associated with case vs. control tests

```{r include=F}
## population structure
W <-
    pop.info %>%
    mutate(one = 1) %>%
    dcast(`sample.ID` ~ `Super Population`, value.var = "one", fill = 0) %>%
    (function(x) left_join(data$fam[, "sample.ID", drop = F], x)) %>%
    column_to_rownames("sample.ID") %>%
    as.matrix()

## Take first 10000 variants
X <- data$genotypes[, 1:10000]
rownames(X) <- data$fam$sample.ID
colnames(X) <- data$map$physical.pos[1:10000]

#' @param X genotype matrix
#' @param W population structure
#' @param h2 heritability
#' @param pve.w population PVe
#' @param n.causal
#' @param n.traits
simulate.pgs <- function(X, W, h2, pve.w, n.causal, n.traits = 1) {
    .rnorm <- function(d1, d2) matrix(rnorm(d1*d2), d1, d2)
    causal.snp <- sample(ncol(X), n.causal)
    xx.causal <- apply(X[, causal.snp, drop=FALSE], 2, scale)
    xx.causal[is.na(xx.causal)] <- 0

    n.ind <- nrow(X)
    y.true <- apply(xx.causal %*% .rnorm(n.causal, n.traits), 2, scale)
    if(pve.w > 0){
        y.pop <- apply(W %*% .rnorm(ncol(W), n.traits), 2, scale)
    }else{
        y.pop <- 0
    }
    y.err <- apply(.rnorm(n.ind, n.traits), 2, scale)

    y.obs <- y.true * sqrt(h2) + y.pop * sqrt(pve.w)
    y.obs <- y.obs + y.err * sqrt(1 - h2 - pve.w)

    prop <- 1/(1+ exp(-y.obs))
    y.cc <- rbinom(length(prop), 1, prop)

    list(y.q = y.obs, y.cc = y.cc, x = X, w = W, causal = causal.snp)
}
```

```{r}
.sim.file <- "../data/genotype/gwas_simulation.rds"
if.needed(.sim.file, {
    set.seed(1)
    .sim <- simulate.pgs(X, W, h2=.3, pve.w=0, n.causal=3)
    saveRDS(.sim, .sim.file)
})
.sim <- readRDS(.sim.file)
```

:::::: {.columns}
::: {.column width=.5}

A genotype matrix $X$ ($X_{ij} \in \{0,1,2\}$):

```{r fig.width=2.5, fig.height=2.35}
X <- .sim$x
.matshow(X[1:30,1:30], .lw=.1, .lab=2)
```

:::
::: {.column width=.5}

A vector of phenotypes $Y$ ($Y_{i} = 1$ if case vs. $Y_{i} = 0$ if control)

```{r}
Y <- .sim$y.cc
```

```{r echo=T, size="large"}
table(Y)
```

$${}$$

Goal: Find locations (columns of $X$) where a genotype $\mathbf{x}_{j}$ is associated with a phenotype $\mathbf{y}$

:::
::::::

## GWAS by variant-by-variant association tests

A variant-by-variant association t-test for a variant $j$:

$$H_{0}:\, \mathbb{E}\!\left[X_{ij}|Y_{i}=1\right] = \mathbb{E}\!\left[X_{ij}|Y_{i}=0\right]\quad\textsf{vs.}$$

$$\quad \mathbb{E}\!\left[X_{ij}|Y_{i}=1\right] \neq \mathbb{E}\!\left[X_{ij}|Y_{i}=0\right]$$

* The average genotype is the same between the case and control under the null.

```{r echo=T, size="large"}
## library(matrixTests)
.gwas.file <- "../data/genotype/gwas_results.rds"
if.needed(.gwas.file, {
    .gwas <- col_t_welch(X[Y == 0, ], X[Y == 1, ])
    saveRDS(.gwas, .gwas.file)
})
.gwas <- readRDS(.gwas.file)
```

# Hypothesis testing in terms of two errors

## Recap: Steps of hypothesis testing

\large

1. **Choose a test statistic** and formulate hypotheses
   - Null hypothesis $H_0$: no effect / no difference

2. **Define the rejection region**
   - Set significance level $\alpha$ (e.g., 0.05)
   - Determine critical values or thresholds

3. **Compute the test statistic** from observed data

4. **Make a decision**
   - If test statistic falls in rejection region (or p-value < $\alpha$): reject $H_0$

\normalsize


## What does type I error mean for a single test?

```{r fig.width=5.5, fig.height=2.9}
## null and alternative distributions
x <- seq(-4, 6, length.out = 500)
alpha <- 0.05
crit <- qnorm(1 - alpha/2)

df.null <- data.table(x = x, density = dnorm(x), dist = "Null (H0 true)")
df.alt <- data.table(x = x, density = dnorm(x, mean = 2), dist = "Alternative (H1 true)")
df.plot <- rbind(df.null, df.alt)

ggplot(df.plot, aes(x, density, color = dist, linetype = dist)) +
    geom_line(linewidth = 1) +
    geom_area(data = df.null[x > crit],
              aes(x, density), fill = "red", alpha = .3, inherit.aes = FALSE) +
    geom_area(data = df.null[x < -crit],
              aes(x, density), fill = "red", alpha = .3, inherit.aes = FALSE) +
    geom_vline(xintercept = c(-crit, crit), lty=1, color = "red", linewidth = .6) +
    annotate("text", x = 3.2, y = .32, label = "Type I\nerror (alpha)",
             color = "red", fontface = "bold", size = 3) +
    scale_color_manual(values = c("darkgreen", "blue")) +
    scale_linetype_manual(values = c("dashed", "solid")) +
    labs(x = "Test statistic", y = "Density",
         title = "Type I error: rejecting H0 when H0 is true",
         color = NULL, linetype = NULL) +
    theme(legend.position = "top")
```

## Type I and Type II errors (`lect04`)

|                    | $H_0$ True         | $H_0$ False          |
|--------------------|--------------------|----------------------|
| **Reject $H_0$**   | Type I Error ($\alpha$) | Correct (Power)      |
| **Fail to reject** | Correct            | Type II Error ($\beta$) |

\vfill

* **Type I Error** (false positive): Reject $H_0$ when it is actually true

* **Type II Error** (false negative): Fail to reject $H_0$ when it is actually false

## Statistical power: $1 - \beta$

\large

**Power** = Probability of correctly rejecting $H_0$ when it is false

$$\text{Power} = 1 - \beta = P(\text{reject } H_0 \mid H_1 \text{ is true})$$

\vfill

Power depends on:

1. **Significance level** $\alpha$ (Type I error rate)
2. **Effect size** $\delta$ (how different $H_1$ is from $H_0$)
3. **Sample size** $n$
4. **Variability** $\sigma$

\normalsize

## Visualizing power

```{r power-viz, fig.width=5.5, fig.height=2.9}
# Null and alternative distributions
x <- seq(-4, 6, length.out = 500)
alpha <- 0.05
effect.size <- 2  # Mean difference in standard deviation units
crit <- qnorm(1 - alpha)

df.null <- data.table(x = x, density = dnorm(x), dist = "Null (H0)")
df.alt <- data.table(x = x, density = dnorm(x, mean = effect.size), dist = "Alternative (H1)")

# Type II error region (fail to reject when H1 is true)
beta.region <- df.alt[x < crit]
power.region <- df.alt[x >= crit]

ggplot() +
    geom_line(data = rbind(df.null, df.alt),
              aes(x, density, color = dist, linetype = dist), linewidth = 1) +
    geom_area(data = power.region, aes(x, density),
              fill = "green", alpha = 0.3) +
    geom_area(data = beta.region, aes(x, density),
              fill = "orange", alpha = 0.3) +
    geom_vline(xintercept = crit, lty = 2, color = "red", linewidth = 0.6) +
    annotate("text", x = crit + 1.5, y = 0.32,
             label = sprintf("Power = %.2f", 1 - pnorm(crit, mean = effect.size)),
             color = "darkgreen", fontface = "bold", size = 4) +
    annotate("text", x = crit - 1.5, y = 0.12,
             label = sprintf("β = %.2f", pnorm(crit, mean = effect.size)),
             color = "darkorange", fontface = "bold", size = 4) +
    scale_color_manual(values = c("darkgreen", "blue")) +
    scale_linetype_manual(values = c("dashed", "solid")) +
    labs(x = "Test statistic", y = "Density",
         title = sprintf("Power analysis: alpha = %.2f, effect size = %.1f SD", alpha, effect.size),
         color = NULL, linetype = NULL) +
    theme(legend.position = "top")
```

## Why statistical power changes with $n$

\large

**Key insight**: As $n \uparrow$, the standard error $\downarrow$

$$\text{SE} = \frac{\sigma}{\sqrt{n}}$$

\vfill

* **Both** null and alternative distributions have the same SE
* Larger $n$ -> **both distributions become narrower**
* Fixed effect size $\delta$ -> distributions **separate more** (relative to SE)
* Less overlap -> easier to distinguish $H_0$ from $H_1$
* Result: **higher power**

\normalsize

## Power increases with sample size: visualization

```{r power-sample-size-n10, fig.width=5.5, fig.height=2.8, only.plot="1"}
delta <- 1.5  # True effect size
alpha <- 0.05

# Function to create plot for a given sample size
plot_power_n <- function(n) {
    sigma <- 10  # Population standard deviation (make it larger for clearer visualization)
    se <- sigma / sqrt(n)  # Standard error (same for both H0 and H1)
    crit <- qnorm(1 - alpha) * se
    power <- 1 - pnorm(crit, mean = delta, sd = se)
    beta <- pnorm(crit, mean = delta, sd = se)

    # Create x range to show both null (at 0) and alternative (at delta) ±4 SE
    x.min <- min(0 - 4*se, delta - 4*se)
    x.max <- max(0 + 4*se, delta + 4*se)
    x.vals <- seq(x.min, x.max, length.out = 500)

    df.null <- data.table(x = x.vals, density = dnorm(x.vals, mean = 0, sd = se),
                          dist = "Null (H0)")
    df.alt <- data.table(x = x.vals, density = dnorm(x.vals, mean = delta, sd = se),
                         dist = "Alternative (H1)")

    # Type I error (alpha), Type II error (beta), and power regions
    alpha.region <- df.null[x >= crit]
    beta.region <- df.alt[x < crit]
    power.region <- df.alt[x >= crit]

    ggplot() +
        geom_line(data = rbind(df.null, df.alt),
                  aes(x, density, color = dist, linetype = dist), linewidth = 1) +
        geom_area(data = alpha.region, aes(x, density),
                  fill = "red", alpha = 0.25) +
        geom_area(data = power.region, aes(x, density),
                  fill = "green", alpha = 0.3) +
        geom_area(data = beta.region, aes(x, density),
                  fill = "orange", alpha = 0.3) +
        geom_vline(xintercept = 0, lty = 2, linewidth = .1) +        
        geom_vline(xintercept = crit, color = "red", linewidth = 0.6) +
        # Mark the SE with a horizontal segment at the base
        annotate("segment", x = 0, xend = se,
                 y = -max(df.null$density) * 0.05, yend = -max(df.null$density) * 0.05,
                 color = "blue", linewidth = 1.5,
                 arrow = arrow(ends = "both", length = unit(0.15, "cm"))) +
        annotate("text", x = se/2, y = 0.01, vjust = 0,
                 label = sprintf("SE = %.2f", se),
                 color = "blue", fontface = "bold", size = 3.5) +
        annotate("text", x = crit + 1.5*se, y = max(df.null$density) * 0.4,
                 label = sprintf("alpha = %.2f", alpha),
                 color = "red", fontface = "bold", size = 3) +
        annotate("text", x = max(crit + 0.5*se, delta + 0.5*se),
                 y = max(df.alt$density) * 0.7,
                 label = sprintf("Power = %.2f", power),
                 color = "darkgreen", fontface = "bold", size = 3.5) +
        annotate("text", x = min(crit - 0.5*se, delta - 0.5*se),
                 y = max(df.alt$density) * 0.3,
                 label = sprintf("beta = %.2f", beta),
                 color = "darkorange", fontface = "bold", size = 3.5) +
        scale_color_manual(values = c("darkgreen", "blue")) +
        scale_linetype_manual(values = c("dashed", "solid")) +
        labs(x = "Test statistic", y = "Density",
             title = sprintf("n = %d", n),
             color = NULL, linetype = NULL) +
        theme(legend.position = "top")
}

# n = 10
plot_power_n(10)
```

```{r power-sample-size-n30, fig.width=5.5, fig.height=2.8, only.plot="2"}
# n = 30
plot_power_n(30)
```

```{r power-sample-size-n100, fig.width=5.5, fig.height=2.8, only.plot="3"}
# n = 100
plot_power_n(100)
```

```{r power-sample-size-n500, fig.width=5.5, fig.height=2.8, only.plot="4"}
# n = 500
plot_power_n(500)
```

## What do we need to perform power calculation?

:::::: {.columns}
::: {.column width=.5}

1. **Significance level** ($\alpha$)
   - Type I error rate we're willing to accept

2. **Effect size** ($\delta$ or Cohen's $d$)
   - Can estimate from pilot data or literature

3. **Sample size** ($n$)
   - Number of observations per group

:::
::: {.column width=.5}

4. **Standard deviation** ($\sigma$)
   - Estimate from prior experiments

5. **Power** ($1-\beta$)
   - Higher power = more confident in detecting true effects

:::
::::::

\vfill

**Typical questions:**

* Given $\alpha$, $\delta$, $\sigma$, and desired power -> find $n$
* Given $\alpha$, $n$, $\sigma$, $\delta$ -> find power



## Power calculation in R (e.g., t-test)

**Question**: How much power do we have with the sample size $n$?

```{r echo=T, eval=F, size="large"}
power.t.test(
    delta = 0.5,      # Effect size (mean diff)
    sd = 1,           # Standard deviation
    n = 50,           # Sample size per group
    sig.level = 0.05, # Rejection region (desired Type I)
    type = "two.sample"
)
```

**Answer**: roughly 0.7 power

<!-- **What affects power?** -->
<!-- * increase Sample size -> increase Power -->
<!-- * increase Effect size -> increase Power -->
<!-- * increase $\alpha$ -> increase Power (but more Type I errors!) -->
<!-- * decrease Variability -> increase Power -->

<!-- ::: -->
<!-- :::::: -->

## Sample size calculation

**Question**: How many samples do we need to achieve 80% power?

```{r echo=T, eval=F, size="large"}
# Calculate required sample size for 80% power
power.t.test(
    delta = 0.5,       # Effect size...
    sd = 1,            # Standard deviation
    power = 0.8,       # Desired power
    sig.level = 0.05,
    type = "two.sample"
)
```

**Answer**: $\approx$ 64 samples per group for the effect size $\delta=0.5$

## Power curves: sample size vs. power

```{r power-curves, fig.width=5, fig.height=2.9}
# Calculate power for different sample sizes and effect sizes
sample.sizes <- seq(10, 200, by = 5)
effect.sizes <- c(0.2, 0.5, 0.8)  # Small, medium, large

power.data <- expand.grid(
    n = sample.sizes,
    delta = effect.sizes
) %>%
    rowwise() %>%
    mutate(
        power = power.t.test(delta = delta, sd = 1, n = n,
                            sig.level = 0.05, type = "two.sample")$power,
        effect = factor(delta, levels = c(0.2, 0.5, 0.8),
                       labels = c("Small (delta=0.2)", "Medium (delta=0.5)", "Large (delta=0.8)"))
    ) %>%
    as.data.table()

ggplot(power.data, aes(x = n, y = power, color = effect)) +
    geom_line(linewidth = 1) +
    geom_hline(yintercept = 0.8, lty = 2, color = "gray40") +
    annotate("text", x = 180, y = 0.82, label = "80% power",
             color = "gray40", size = 3) +
    scale_color_manual(values = c("red3", "orange2", "green4")) +
    labs(x = "Sample size per group", y = "Statistical power",
         title = "Power curves for two-sample t-test (alpha = 0.05)",
         color = "Effect size") +
    ylim(0, 1) +
    theme(legend.position = "top")
```

# Controlling Family-Wise Error Rate 

## What if we have more than one hypothesis?

```{r type-i-ii-tradeoff, fig.width=4.5, fig.height=2.9}
library(mvtnorm)

## Grid
xs <- seq(-3.5, 6, length.out = 200)
ys <- seq(-3.5, 6, length.out = 200)
grid <- expand.grid(x = xs, y = ys)

## Two 2D Gaussians: null at origin, alternative shifted
mu0 <- c(0, 0)
mu1 <- c(2.5, 2.5)
Sigma <- matrix(c(1, 0.3, 0.3, 1), 2, 2)

grid$d.null <- dmvnorm(as.matrix(grid[, 1:2]), mean = mu0, sigma = Sigma)
grid$d.alt  <- dmvnorm(as.matrix(grid[, 1:2]), mean = mu1, sigma = Sigma)

## Decision boundary: circle of radius r around null center
r.thresh <- qnorm(1 - 0.025)
grid$dist.null <- sqrt(grid$x^2 + grid$y^2)
grid$reject <- grid$dist.null > r.thresh

dt.grid <- as.data.table(grid)

dt.grid[, region := fcase(
    reject & d.null > 0.001, "Type I: reject true null",
    !reject & d.alt > 0.001, "Type II: miss true alt",
    default = NA_character_
)]

## Simulate points from both distributions for hex density
set.seed(42)
n.pts <- 8000
pts.null <- as.data.table(rmvnorm(n.pts, mean = mu0, sigma = Sigma))
pts.alt  <- as.data.table(rmvnorm(n.pts, mean = mu1, sigma = Sigma))
setnames(pts.null, c("x", "y"))
setnames(pts.alt, c("x", "y"))
pts.null[, dist := "Null"]
pts.alt[, dist := "Alt"]

## Main 2D hex density plot
p.2d.main <- ggplot() +
    geom_hex(data = pts.null, aes(x = x, y = y), fill = "gray50", alpha = 0.5, bins = 30) +
    geom_hex(data = pts.alt, aes(x = x, y = y), fill = "steelblue", alpha = 0.5, bins = 30) +
    # Add vertical and horizontal lines for alpha = 0.05 threshold
    geom_vline(xintercept = c(-r.thresh, r.thresh), lty = 2, colour = "red", linewidth = 0.5, alpha = 0.6) +
    geom_hline(yintercept = c(-r.thresh, r.thresh), lty = 2, colour = "red", linewidth = 0.5, alpha = 0.6) +
    annotate("path",
             x = r.thresh * cos(seq(0, 2*pi, length.out = 200)),
             y = r.thresh * sin(seq(0, 2*pi, length.out = 200)),
             lty = 2, colour = "red", linewidth = 0.7) +
    annotate("text", x = 0, y = -0.3, label = "H0", size = 3, fontface = "bold", colour = "red") +
    annotate("text", x = 3.5, y = 3.2, label = "H1", size = 3, fontface = "bold", colour = "steelblue") +
    annotate("text", x = -2.8, y = 2.8, label = "reject", size = 2, colour = "red") +
    annotate("text", x = r.thresh, y = -3, label = "alpha=0.05", size = 2, colour = "red", angle = 90, vjust = -0.3) +
    annotate("text", x = -3, y = r.thresh, label = "alpha=0.05", size = 2, colour = "red", vjust = -0.3) +
    labs(x = "Hypothesis 1", y = "Hypothesis 2") +
    coord_cartesian(xlim = range(xs), ylim = range(ys)) +
    theme(legend.position = "none",
          plot.margin = margin(0, 0, 0, 0))

## Top marginal: density along x (Hypothesis 1)
dt.marg.x <- data.table(
    x = xs,
    null = dnorm(xs, mean = mu0[1], sd = sqrt(Sigma[1,1])),
    alt  = dnorm(xs, mean = mu1[1], sd = sqrt(Sigma[1,1]))
) %>% melt(id.vars = "x", variable.name = "dist", value.name = "density")

p.top <- ggplot(dt.marg.x, aes(x = x, y = density, colour = dist)) +
    geom_line(linewidth = 0.7) +
    geom_vline(xintercept = c(-r.thresh, r.thresh), lty = 2, colour = "red", linewidth = 0.4, alpha = 0.6) +
    scale_colour_manual(values = c(null = "gray30", alt = "steelblue")) +
    labs(y = NULL, title = "Hypothesis 1") +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          axis.title.x = element_blank(), axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "none",
          plot.margin = margin(0, 0, 0, 0))

## Right marginal: same structure as top, but coord_flip
dt.marg.y <- data.table(
    x = ys,
    null = dnorm(ys, mean = mu0[2], sd = sqrt(Sigma[2,2])),
    alt  = dnorm(ys, mean = mu1[2], sd = sqrt(Sigma[2,2]))
) %>% melt(id.vars = "x", variable.name = "dist", value.name = "density")

p.right <- ggplot(dt.marg.y, aes(x = x, y = density, colour = dist)) +
    geom_line(linewidth = 0.7) +
    geom_vline(xintercept = c(-r.thresh, r.thresh), lty = 2, colour = "red", linewidth = 0.4, alpha = 0.6) +
    scale_colour_manual(values = c(null = "gray30", alt = "steelblue")) +
    coord_flip() +
    labs(y = NULL, title = "Hyp 2") +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          axis.title.x = element_blank(), axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "none",
          plot.margin = margin(0, 0, 0, 0))

## Compose with patchwork
p.top + plot_spacer() + p.2d.main + p.right +
    plot_layout(ncol = 2, nrow = 2,
                widths = c(4, 1), heights = c(1, 4))
```

<!-- Marginals show the familiar 1D null vs alt for each hypothesis; the 2D -->
<!-- view shows where errors overlap. -->


## What if we decided to reject either one at $\alpha<0.05$

```{r fig.width=5, fig.height=2.9, only.plot="1"}
ggplot() +
    annotate("rect", xmin = -Inf, xmax = -r.thresh, ymin = -Inf, ymax = Inf,
             fill = "red", alpha = 0.15) +
    annotate("rect", xmin = r.thresh, xmax = Inf, ymin = -Inf, ymax = Inf,
             fill = "red", alpha = 0.15) +
    annotate("rect", xmin = -r.thresh, xmax = r.thresh, ymin = -Inf, ymax = -r.thresh,
             fill = "red", alpha = 0.15) +
    annotate("rect", xmin = -r.thresh, xmax = r.thresh, ymin = r.thresh, ymax = Inf,
             fill = "red", alpha = 0.15) +
    geom_hex(data = pts.null, aes(x = x, y = y), fill = "gray50", alpha = 0.5, bins = 30) +
    geom_vline(xintercept = c(-r.thresh, r.thresh), colour = "red", linewidth = 1) +
    geom_hline(yintercept = c(-r.thresh, r.thresh), colour = "red", linewidth = 1) +
    annotate("text", x = 0, y = 0, label = "H0", size = 4, fontface = "bold", colour = "blue") +
    labs(x = "Hypothesis 1", y = "Hypothesis 2",
         title = "Reject EITHER hypothesis at alpha=0.05") +
    coord_cartesian(xlim = range(xs), ylim = range(ys)) +
    theme(legend.position = "none")
```

```{r fig.width=5, fig.height=2.9, only.plot="2"}
ggplot() +
    annotate("rect", xmin = -Inf, xmax = -r.thresh, ymin = -Inf, ymax = Inf,
             fill = "red", alpha = 0.15) +
    annotate("rect", xmin = r.thresh, xmax = Inf, ymin = -Inf, ymax = Inf,
             fill = "red", alpha = 0.15) +
    annotate("rect", xmin = -r.thresh, xmax = r.thresh, ymin = -Inf, ymax = -r.thresh,
             fill = "red", alpha = 0.15) +
    annotate("rect", xmin = -r.thresh, xmax = r.thresh, ymin = r.thresh, ymax = Inf,
             fill = "red", alpha = 0.15) +
    geom_hex(data = pts.null, aes(x = x, y = y), fill = "gray50", alpha = 0.5, bins = 30) +
    geom_hex(data = pts.alt, aes(x = x, y = y), fill = "steelblue", alpha = 0.4, bins = 30) +
    geom_vline(xintercept = c(-r.thresh, r.thresh), colour = "red", linewidth = 1) +
    geom_hline(yintercept = c(-r.thresh, r.thresh), colour = "red", linewidth = 1) +
    annotate("text", x = 0, y = 0, label = "H0", size = 4, fontface = "bold", colour = "blue") +
    annotate("text", x = mu1[1], y = mu1[2], label = "H1", size = 4, fontface = "bold", colour = "steelblue") +
    labs(x = "Hypothesis 1", y = "Hypothesis 2",
         title = "Reject EITHER hypothesis at alpha=0.05") +
    coord_cartesian(xlim = range(xs), ylim = range(ys)) +
    theme(legend.position = "none")
```


## What if we test each hypothesis $\alpha < 0.05$?

```{r fig.width=5, fig.height=2.9, only.plot="1"}
ggplot() +
    annotate("rect", xmin = -Inf, xmax = -r.thresh, ymin = r.thresh, ymax = Inf,
             fill = "darkred", alpha = 0.2) +
    annotate("rect", xmin = r.thresh, xmax = Inf, ymin = r.thresh, ymax = Inf,
             fill = "darkred", alpha = 0.2) +
    annotate("rect", xmin = -Inf, xmax = -r.thresh, ymin = -Inf, ymax = -r.thresh,
             fill = "darkred", alpha = 0.2) +
    annotate("rect", xmin = r.thresh, xmax = Inf, ymin = -Inf, ymax = -r.thresh,
             fill = "darkred", alpha = 0.2) +
    geom_hex(data = pts.null, aes(x = x, y = y), fill = "gray50", alpha = 0.5, bins = 30) +
    geom_vline(xintercept = c(-r.thresh, r.thresh), colour = "red", linewidth = 1) +
    geom_hline(yintercept = c(-r.thresh, r.thresh), colour = "red", linewidth = 1) +
    annotate("text", x = -2.8, y = 3, label = "Reject\nboth",
             size = 2.5, colour = "darkred", fontface = "bold") +
    annotate("text", x = 2.8, y = 3, label = "Reject\nboth",
             size = 2.5, colour = "darkred", fontface = "bold") +
    annotate("text", x = -2.8, y = -3, label = "Reject\nboth",
             size = 2.5, colour = "darkred", fontface = "bold") +
    annotate("text", x = 2.8, y = -3, label = "Reject\nboth",
             size = 2.5, colour = "darkred", fontface = "bold") +
    annotate("text", x = 0, y = 0, label = "H0", size = 4, fontface = "bold", colour = "blue") +
    labs(x = "Hypothesis 1", y = "Hypothesis 2",
         title = "Reject BOTH hypotheses") +
    coord_cartesian(xlim = range(xs), ylim = range(ys)) +
    theme(legend.position = "none")
```

```{r fig.width=5, fig.height=2.9, only.plot="2"}
ggplot() +
    annotate("rect", xmin = -Inf, xmax = -r.thresh, ymin = r.thresh, ymax = Inf,
             fill = "darkred", alpha = 0.2) +
    annotate("rect", xmin = r.thresh, xmax = Inf, ymin = r.thresh, ymax = Inf,
             fill = "darkred", alpha = 0.2) +
    annotate("rect", xmin = -Inf, xmax = -r.thresh, ymin = -Inf, ymax = -r.thresh,
             fill = "darkred", alpha = 0.2) +
    annotate("rect", xmin = r.thresh, xmax = Inf, ymin = -Inf, ymax = -r.thresh,
             fill = "darkred", alpha = 0.2) +
    geom_hex(data = pts.null, aes(x = x, y = y), fill = "gray50", alpha = 0.5, bins = 30) +
    geom_hex(data = pts.alt, aes(x = x, y = y), fill = "steelblue", alpha = 0.4, bins = 30) +
    geom_vline(xintercept = c(-r.thresh, r.thresh), colour = "red", linewidth = 1) +
    geom_hline(yintercept = c(-r.thresh, r.thresh), colour = "red", linewidth = 1) +
    annotate("text", x = -2.8, y = 3, label = "Reject\nboth",
             size = 2.5, colour = "darkred", fontface = "bold") +
    annotate("text", x = 2.8, y = 3, label = "Reject\nboth",
             size = 2.5, colour = "darkred", fontface = "bold") +
    annotate("text", x = -2.8, y = -3, label = "Reject\nboth",
             size = 2.5, colour = "darkred", fontface = "bold") +
    annotate("text", x = 2.8, y = -3, label = "Reject\nboth",
             size = 2.5, colour = "darkred", fontface = "bold") +
    annotate("text", x = 0, y = 0, label = "H0", size = 4, fontface = "bold", colour = "blue") +
    annotate("text", x = mu1[1], y = mu1[2], label = "H1", size = 4, fontface = "bold", colour = "steelblue") +
    labs(x = "Hypothesis 1", y = "Hypothesis 2",
         title = "Reject BOTH hypotheses") +
    coord_cartesian(xlim = range(xs), ylim = range(ys)) +
    theme(legend.position = "none")
```

## Wait, what was the definition of p-value?

\large

**P-value**: Probability of observing a test statistic as extreme or more extreme than what we observed, **assuming the null hypothesis is true**

$$P(\text{data as extreme or more} \mid H_0 \text{ is true})$$

For a single test at $\alpha = 0.05$:

- $P(\text{Type I error}) = P(\text{reject} H_0 \mid H_0\text{ is true}) < \alpha = 0.05$

## What about two p-values?

**Union rule** (probability of A or B):
$$P(A \cup B) = P(A) + P(B) - P(A \cap B) < P(A) + P(B)$$

**Independence** (probability of A and B when independent):
$$P(A \cap B) = P(A) \times P(B)$$

When we want to "control" two hypothesis testing....

$$P(\text{reject at least one true } H_0 \mid H_{0}) \leq \alpha + \alpha$$

**What will be the $\alpha$ value that is guaranteed to control both?**

## Manhattan plot to show all the p-values...

:::::: {.columns}
::: {.column width=.5}

```{r fig.width=3, fig.height=2.5}
par(pch=19, col="gray40", cex=.5)
plot(colnames(X), -log10(.gwas$pvalue), xlab = "genomic location", ylab = "-log10 p-value")
points(colnames(X)[.sim$causal], -log10(.gwas$pvalue)[.sim$causal], col=2, cex=1, lwd=2, pch=1)
legend("topleft", "causal variant", col=2, pch=1, lwd=1.5, lty=0, cex=2)
```

:::
::: {.column width=.5}

```{r fig.width=3, fig.height=2.5}
par(pch=19, col="gray40", cex=.5)
hist(.gwas$pvalue, 50, main="", xlab="p-value")
```

:::
::::::

We have $p = 10{,}000$ p-values. How do we decide which variants are significant?

## The problem: we tested 10,000 hypotheses

```{r}
causal.variants <- .sim$causal
gwas.p <- .gwas$pvalue
```

\large

* We performed $10{,}000$ simultaneous tests.

* At $\alpha = 0.05$, how many false positives do we expect **by chance alone**?

$$\mathbb{E}[\text{\# false positives}] = 10{,}000 \times 0.05 = 500$$

* We need a principled way to control errors when testing many hypotheses at once.

\normalsize


## Why does this matter for 10,000 tests?

```{r fig.width=5.5, fig.height=2.7}
## Simulate 10000 p-values under the null
set.seed(42)
n.tests <- 10000
null.pvals <- runif(n.tests)  # all null

dt.pvals <- data.table(pvalue = null.pvals,
                        significant = null.pvals < 0.05)

ggplot(dt.pvals, aes(pvalue, fill = significant)) +
    geom_histogram(bins = 50, color = "white", linewidth = .2) +
    scale_fill_manual(values = c("gray60", "red"),
                      labels = c("Not rejected", "False positive (Type I)")) +
    geom_vline(xintercept = 0.05, lty = 2, color = "red", linewidth = .8) +
    annotate("text", x = 0.07, y = 350,
             label = paste0(sum(dt.pvals$significant), " false positives\nout of ",
                            n.tests, " null tests"),
             color = "red", fontface = "bold", size = 3.5,
             hjust = 0, vjust = 1) +
    labs(x = "p-value", y = "Count",
         title = "All 10,000 tests are null — yet we get ~500 'significant' results",
         fill = NULL) +
    theme(legend.position = "top")
```


## Conversely.... The dimensionality works unfavourably...

Each test at $\alpha = 0.05$: 5% chance of a false rejection if $H_0$ is true.

| Variants ($m$) | $P(\ge 1\text{ false rejection})$ | |
|:--------------:|:---------------------------------:|:--|
| 1              | $1 - 0.95^{1} = 0.05$            | Just $\alpha$ |
| 2              | $1 - 0.95^{2} = 0.098$           | Nearly doubled |
| 3              | $1 - 0.95^{3} = 0.143$           | Nearly tripled |
| 10             | $1 - 0.95^{10} = 0.40$           | 40% chance! |
| 100            | $1 - 0.95^{100} = 0.994$         | Almost certain |
| 10,000         | $1 - 0.95^{10000} \approx 1$     | Guaranteed |

\vfill

## Family-wise error rate (FWER)

:::::: {.columns}
::: {.column width=.5}

* $P(\text{incorrect} \mid H_0) = \alpha$; let $\alpha = 0.05$

* $P(\text{correct} \mid H_0) = 1 - \alpha = 0.95$

\vfill

* For $m$ **independent** tests, all under $H_0$:

$$P(\text{all correct}) = (1-\alpha)^{m} = 0.95^{m}$$

* **Family-wise error rate**:

$$\textsf{FWER} = P(\text{at least one error}) = 1 - 0.95^{m}$$

:::
::: {.column width=.5}

```{r fwer-curve, fig.width=2.5, fig.height=2.9}
data.table(m = 1:200) %>%
    mutate(fwer = 1 - 0.95^m) %>%
    ggplot(aes(x = m, y = fwer)) +
    geom_line(linewidth = 1.2, colour = "dodgerblue") +
    labs(x = "m (number of tests)", y = "FWER") +
    geom_hline(yintercept = 1, lty = 2, colour = "gray50") +
    annotate("text", x = 100, y = 0.5,
             label = "FWER = 1 - 0.95^m", size = 4, colour = "dodgerblue")
```

:::
::::::

## Bonferroni correction: controlling the FWER

<!-- We want: $P(\text{at least one error in } m \text{ tests}) < \alpha$ -->

\large

By the **union bound**:

$$P(\text{at least one error}) \le \sum_{i=1}^{m} P(\text{error on test } i)$$

Set each test's threshold to $\alpha_{\textsf{Bon}} = \frac{\alpha}{m}$:

$$\sum_{i=1}^{m} \alpha_{\textsf{Bon}} = m \cdot \frac{\alpha}{m} = \alpha$$

\normalsize


## Bonferroni correction: controlling the FWER

\large

* Reject $H_{0,i}$ if $p_i < \alpha / m$, or equivalently, 

* Adjust $p_{\textsf{Bon},i} = \min(m \cdot p_i,\; 1)$ and threshold at $\alpha$.


\normalsize


## Bonferroni on our GWAS example

```{r bonferroni-gwas, fig.width=5.5, fig.height=2.9}
m <- length(gwas.p)
bon.threshold <- 0.05 / m

dt.bon <- data.table(
    pos = as.numeric(colnames(X)),
    pvalue = gwas.p,
    causal = seq_along(gwas.p) %in% causal.variants
) %>% mutate(sig.bon = pvalue < bon.threshold)

p1 <- ggplot(dt.bon, aes(x = pos, y = -log10(pvalue), colour = sig.bon)) +
    geom_point(size = 0.8, alpha = 0.6) +
    geom_hline(yintercept = -log10(bon.threshold), lty = 2, colour = "red") +
    scale_colour_manual(values = c("gray50", "red")) +
    annotate("text", x = max(dt.bon$pos) * 0.7, y = -log10(bon.threshold) + 0.3,
             label = paste0("Bonferroni: alpha/m = ", signif(bon.threshold, 2)),
             colour = "red", size = 3) +
    labs(x = "Genomic location", y = "-log10(p-value)",
         title = paste0("Bonferroni: ", sum(dt.bon$sig.bon), " rejections out of ", m, " tests")) +
    theme(legend.position = "none")

p1
```

<!-- Very conservative --- may miss true signals! -->


## What if we remove correlated variants (LD clumping)?

```{r ld-clumping, include=F}
library(bigsnpr)

.ld.file <- "../data/genotype/ld_clumping_results.rds"
if.needed(.ld.file, {
    # Create a temporary FBM.code256 for genotype data (0, 1, 2)
    tmp.file <- tempfile(fileext = ".rds")
    X.fbm <- FBM.code256(nrow(X), ncol(X), code = c(0, 1, 2, rep(NA, 253)),
                          backingfile = tmp.file)
    X.fbm[] <- X

    # LD clumping using bigsnpr
    ind.keep <- snp_clumping(X.fbm,
                             infos.chr = rep(1, ncol(X)),
                             infos.pos = as.numeric(colnames(X)),
                             thr.r2 = 0.2,
                             S = -log10(gwas.p))  # Use -log10(p) as priority score

    # Clean up temporary file
    unlink(paste0(tmp.file, c("", ".bk")))

    saveRDS(ind.keep, .ld.file)
})
ind.keep <- readRDS(.ld.file)

## cat(sprintf("Original variants: %d\n", ncol(X)))
## cat(sprintf("After LD clumping (r² < 0.2): %d\n", length(ind.keep)))
```

**After LD clumping, FWER methods work much better!**

* **LD clumping** removes correlated variants, keeping only representative independent SNPs

	* It still takes some threshold parameter...

* Independent tests $\Rightarrow$ FWER and FDR are better aligned

* But: we lose power by discarding many variants

## FWER control with LD clumping

```{r bonferroni-clumped, fig.width=5.5, fig.height=2.9}
gwas.p.clumped <- gwas.p[ind.keep]
m.clumped <- length(gwas.p.clumped)
bon.threshold.clumped <- 0.05 / m.clumped

dt.bon.clumped <- data.table(
    pos = as.numeric(colnames(X)[ind.keep]),
    pvalue = gwas.p.clumped,
    causal = ind.keep %in% causal.variants
) %>% mutate(sig.bon = pvalue < bon.threshold.clumped)

ggplot(dt.bon.clumped, aes(x = pos, y = -log10(pvalue), colour = sig.bon)) +
    geom_point(size = 1, alpha = 0.7) +
    geom_hline(yintercept = -log10(bon.threshold.clumped), lty = 2, colour = "red") +
    scale_colour_manual(values = c("gray50", "red")) +
    annotate("text", x = max(dt.bon.clumped$pos) * 0.7, y = -log10(bon.threshold.clumped) + 0.3,
             label = paste0("Bonferroni: alpha/m = ", signif(bon.threshold.clumped, 2)),
             colour = "red", size = 3) +
    labs(x = "Genomic location", y = "-log10(p-value)",
         title = paste0("After LD clumping: ", sum(dt.bon.clumped$sig.bon),
                       " rejections out of ", m.clumped, " tests")) +
    theme(legend.position = "none")
```

<!-- \scriptsize -->
<!-- ```{r} -->
<!-- causal.in.clumped <- which(ind.keep %in% causal.variants) -->
<!-- padj.clumped <- p.adjust(gwas.p.clumped, method = "bonferroni") -->
<!-- false.disc.clumped <- setdiff(which(padj.clumped < 0.05), causal.in.clumped) -->
<!-- empirical.FDR.clumped <- length(false.disc.clumped) / max(sum(padj.clumped < 0.05), 1) -->
<!-- cat(sprintf("Bonferroni rejections: %d\n", sum(padj.clumped < 0.05))) -->
<!-- cat(sprintf("False discoveries: %d\n", length(false.disc.clumped))) -->
<!-- cat(sprintf("Empirical FDR: %.3f\n", empirical.FDR.clumped)) -->
<!-- ``` -->
<!-- \normalsize -->

## Holm--Bonferroni: a less conservative FWER control

* **Holm's step-down** procedure uses decreasing multipliers:

$$p_{\textsf{Holm}(1)} = m \cdot p_{(1)}, \quad p_{\textsf{Holm}(2)} = (m-1) \cdot p_{(2)}, \quad \ldots$$

* Still controls FWER $\le \alpha$, but is **uniformly more powerful** than Bonferroni

* In R: `p.adjust(p, method = "holm")`

## Bonferroni: uniform strict threshold

```{r fig.width=5.5, fig.height=2.9, only.plot="1"}
alpha <- 0.05
m <- 2

bonf.thresh <- qnorm(1 - alpha/(2*m))
uncorr.thresh <- qnorm(1 - alpha/2)

ggplot() +
    geom_hex(data = pts.null, aes(x = x, y = y), fill = "gray50", alpha = 0.5, bins = 30) +
    geom_vline(xintercept = c(-bonf.thresh, bonf.thresh), color = "red", linewidth = 1) +
    geom_hline(yintercept = c(-bonf.thresh, bonf.thresh), color = "red", linewidth = 1) +
    annotate("text", x = 0, y = 0, label = "H0", size = 3, fontface = "bold", color = "blue") +
    annotate("text", x = 5, y = bonf.thresh + 0.3, label = "alpha/m", color = "red", size = 3) +
    labs(x = "Hypothesis 1", y = "Hypothesis 2",
         title = "Bonferroni: test each at alpha/m") +
    coord_cartesian(xlim = range(xs), ylim = range(ys)) +
    theme(legend.position = "none")
```

```{r fig.width=5.5, fig.height=2.9, only.plot="2"}
ggplot() +
    annotate("rect", xmin = -Inf, xmax = -bonf.thresh, ymin = bonf.thresh, ymax = Inf,
             fill = "red", alpha = 0.2) +
    annotate("rect", xmin = bonf.thresh, xmax = Inf, ymin = bonf.thresh, ymax = Inf,
             fill = "red", alpha = 0.2) +
    annotate("rect", xmin = -Inf, xmax = -bonf.thresh, ymin = -Inf, ymax = -bonf.thresh,
             fill = "red", alpha = 0.2) +
    annotate("rect", xmin = bonf.thresh, xmax = Inf, ymin = -Inf, ymax = -bonf.thresh,
             fill = "red", alpha = 0.2) +
    geom_hex(data = pts.null, aes(x = x, y = y), fill = "gray50", alpha = 0.5, bins = 30) +
    geom_vline(xintercept = c(-bonf.thresh, bonf.thresh), color = "red", linewidth = 1) +
    geom_hline(yintercept = c(-bonf.thresh, bonf.thresh), color = "red", linewidth = 1) +
    annotate("text", x = 0, y = 0, label = "H0", size = 3, fontface = "bold", color = "blue") +
    labs(x = "Hypothesis 1", y = "Hypothesis 2",
         title = "Bonferroni rejection regions") +
    coord_cartesian(xlim = range(xs), ylim = range(ys)) +
    theme(legend.position = "none")
```

## Holm: sequential testing, one axis at a time

```{r fig.width=5.5, fig.height=2.9, only.plot="1"}
ggplot() +
    geom_hex(data = pts.null, aes(x = x, y = y), fill = "gray50", alpha = 0.5, bins = 30) +
    ## geom_vline(xintercept = c(-bonf.thresh, bonf.thresh), color = "red", linewidth = 0.5, lty = 2, alpha = 0.5) +
    ## geom_hline(yintercept = c(-bonf.thresh, bonf.thresh), color = "red", linewidth = 0.5, lty = 2, alpha = 0.5) +
    geom_vline(xintercept = c(-bonf.thresh, bonf.thresh), color = "darkgreen", linewidth = 1) +
    annotate("text", x = 0, y = 0, label = "H0", size = 3, fontface = "bold", color = "darkgray") +
    annotate("text", x = bonf.thresh + 0.5, y = 4, label = "Step 1: test H1\nat alpha/m", color = "darkgreen", size = 3) +
    labs(x = "Hypothesis 1", y = "Hypothesis 2",
         title = "Holm Step 1: test H1 (smallest p) at alpha/m") +
    coord_cartesian(xlim = range(xs), ylim = range(ys)) +
    theme(legend.position = "none")
```

```{r fig.width=5.5, fig.height=2.9, only.plot="2"}
ggplot() +
    annotate("rect", xmin = -Inf, xmax = -bonf.thresh, ymin = -Inf, ymax = Inf,
             fill = "darkgreen", alpha = 0.2) +
    annotate("rect", xmin = bonf.thresh, xmax = Inf, ymin = -Inf, ymax = Inf,
             fill = "darkgreen", alpha = 0.2) +
    geom_hex(data = pts.null, aes(x = x, y = y), fill = "gray50", alpha = 0.5, bins = 30) +
    ## geom_vline(xintercept = c(-bonf.thresh, bonf.thresh), color = "red", linewidth = 0.5, lty = 2, alpha = 0.5) +
    ## geom_hline(yintercept = c(-bonf.thresh, bonf.thresh), color = "red", linewidth = 0.5, lty = 2, alpha = 0.5) +
    geom_vline(xintercept = c(-bonf.thresh, bonf.thresh), color = "darkgreen", linewidth = 1) +
    annotate("text", x = 0, y = 0, label = "H0", size = 3, fontface = "bold", color = "darkgray") +
    annotate("text", x = 3.5, y = 4, label = "Reject H1", color = "darkgreen", size = 5, fontface="bold") +
    labs(x = "Hypothesis 1", y = "Hypothesis 2",
         title = "If H1 rejects, proceed to test H2") +
    coord_cartesian(xlim = range(xs), ylim = range(ys)) +
    theme(legend.position = "none")
```

```{r fig.width=5.5, fig.height=2.9, only.plot="3"}
ggplot() +
    annotate("rect", xmin = -Inf, xmax = -bonf.thresh, ymin = -Inf, ymax = Inf,
             fill = "darkgreen", alpha = 0.2) +
    annotate("rect", xmin = bonf.thresh, xmax = Inf, ymin = -Inf, ymax = Inf,
             fill = "darkgreen", alpha = 0.2) +
    geom_hex(data = pts.null, aes(x = x, y = y), fill = "gray50", alpha = 0.5, bins = 30) +
    ## geom_vline(xintercept = c(-bonf.thresh, bonf.thresh), color = "red", linewidth = 0.5, lty = 2, alpha = 0.5) +
    ## geom_hline(yintercept = c(-bonf.thresh, bonf.thresh), color = "red", linewidth = 0.5, lty = 2, alpha = 0.5) +
    geom_vline(xintercept = c(-bonf.thresh, bonf.thresh), color = "darkgreen", linewidth = 1) +
    geom_hline(yintercept = c(-uncorr.thresh, uncorr.thresh), color = "steelblue", linewidth = 1) +
    annotate("text", x = 0, y = 0, label = "H0", size = 3, fontface = "bold", color = "darkgray") +
    annotate("text", x = 0, y = uncorr.thresh + 0.5, label = "Step 2: test H2\nat alpha/(m-1)",
             color = "steelblue", size = 5, fontface = "bold", vjust=0) +
    labs(x = "Hypothesis 1", y = "Hypothesis 2",
         title = "Holm Step 2: test H2 at relaxed threshold alpha/(m-1)") +
    coord_cartesian(xlim = range(xs), ylim = range(ys)) +
    theme(legend.position = "none")
```

```{r fig.width=5.5, fig.height=2.9, only.plot="4"}
ggplot() +
    annotate("rect", xmin = -Inf, xmax = -bonf.thresh, ymin = -Inf, ymax = Inf,
             fill = "darkgreen", alpha = 0.2) +
    annotate("rect", xmin = bonf.thresh, xmax = Inf, ymin = -Inf, ymax = Inf,
             fill = "darkgreen", alpha = 0.2) +
    annotate("rect", xmin = -bonf.thresh, xmax = bonf.thresh, ymin = -Inf, ymax = -uncorr.thresh,
             fill = "steelblue", alpha = 0.25) +
    annotate("rect", xmin = -bonf.thresh, xmax = bonf.thresh, ymin = uncorr.thresh, ymax = Inf,
             fill = "steelblue", alpha = 0.25) +
    geom_hex(data = pts.null, aes(x = x, y = y), fill = "gray50", alpha = 0.5, bins = 30) +
    geom_vline(xintercept = c(-bonf.thresh, bonf.thresh), color = "darkgreen", linewidth = 1) +
    geom_hline(yintercept = c(-uncorr.thresh, uncorr.thresh), color = "steelblue", linewidth = 1) +
    geom_vline(xintercept = c(-bonf.thresh, bonf.thresh), color = "red", linewidth = 0.75, lty = 2, alpha = 0.5) +
    geom_hline(yintercept = c(-bonf.thresh, bonf.thresh), color = "red", linewidth = 0.75, lty = 2, alpha = 0.5) +
    annotate("text", x = 0, y = 0, label = "H0", size = 3, fontface = "bold", color = "darkgray") +
    labs(x = "Hypothesis 1", y = "Hypothesis 2",
         title = "Holm: sequential rejection regions (axis-by-axis)") +
    coord_cartesian(xlim = range(xs), ylim = range(ys)) +
    theme(legend.position = "none")
```


## Why is FWER impractical for genomics?

\large

* So much about making **one single error** out of 20,000 tests?

\vfill

* One easy way to guarantee no Type I errors: **reject nothing!**

	* But...

\vfill

::: {.alertblock}
### The trade-off
Being overly strict about Type I error $\Rightarrow$ greater Type II error.
:::

\vfill

* **Key idea**: in high-throughput biology, it is acceptable to make some errors, as long as we have a *comparatively* large number of true positives.

\normalsize


## Pop-up quiz

\huge

Why do we use ${\color{magenta} 5 \times 10^{-8}}$ as a GWAS significance threshold?

\normalsize

# False Discovery Rate

## The multiple testing table --- not the probability

\large

|                  | Not rejected | Rejected | Total   |
|------------------|:-------------|:---------|:--------|
| **$H_0$ true**   | $U$          | $\mathbf{\color{magenta}V}$      | $m_0$   |
| **$H_0$ false**  | $T$          | $S$      | $m_1$   |
| **Total**        | $m - R$      | $\mathbf{\color{teal}R}$      | $m$     |

\vfill

* $R$ = total rejections (observed)
* $V$ = false positives (unobserved)
* $S$ = true positives (unobserved)

\normalsize

## FDR vs. FPR vs. FWER

\large

* **False Discovery Rate (FDR)**:

\only<1>{
$$\textsf{\color{magenta}FDR} = \mathbb{E}\!\left[\frac{V}{R}\right] = \mathbb{E}\!\left[\frac{\text{\#false discoveries}}{\#\text{rejections}}\right] \quad \text{(define as 0 when } R = 0\text{)}$$
}

* **False Positive Rate (FPR)**:

\only<2>{
$$\textsf{FPR} = \mathbb{E}\!\left[\frac{V}{m_0}\right] = \mathbb{E}\!\left[\frac{\text{\#false discoveries}}{\text{true null}}\right]$$
}

* **Family-Wise Error Rate (FWER)**:

\only<3>{
$$\textsf{FWER} = P(V \ge 1) = P(\textsf{at least 1 false discovery})$$
}

\normalsize

<!-- ## Does Bonferroni control empirical FDR? -->
<!-- <\!-- \scriptsize -\-> -->
<!-- <\!-- ```{r} -\-> -->
<!-- <\!-- padj <- p.adjust(gwas.p, method = "bonferroni") -\-> -->
<!-- <\!-- false.discoveries <- setdiff(which(padj < 0.05), causal.variants) -\-> -->
<!-- <\!-- empirical.FDR <- length(false.discoveries) / max(sum(padj < 0.05), 1) -\-> -->
<!-- <\!-- cat(sprintf("Bonferroni rejections: %d\n", sum(padj < 0.05))) -\-> -->
<!-- <\!-- cat(sprintf("False discoveries: %d\n", length(false.discoveries))) -\-> -->
<!-- <\!-- cat(sprintf("Empirical FDR: %.3f\n", empirical.FDR)) -\-> -->
<!-- <\!-- ``` -\-> -->
<!-- <\!-- \normalsize -\-> -->
<!-- **Bonferroni controls FWER, not FDR!** -->
<!-- * Even with Bonferroni adjustment, empirical FPR can be high -->
<!-- * Why? **Linkage disequilibrium (LD)**: nearby variants are correlated -->
<!-- * FWER guarantees $P(\text{any false rejection}) \le \alpha$, not $E[\text{proportion of false rejections}] \le \alpha$ -->



## Benjamini--Hochberg (BH) procedure

\large

Control FDR at level $q$ (e.g. $q = 0.05$):

\vfill

1. Order the raw p-values: $p_{(1)} \le p_{(2)} \le \cdots \le p_{(m)}$

2. Find the largest rank $j$ such that $p_{(j)} \le \frac{j \cdot q}{m}$

3. Reject all hypotheses with rank $\le j$

\vfill

::: {.block}
### Adjusted p-values
$$p^{\textsf{BH}}_{(i)} = \min_{j \ge i}\!\left(\frac{m}{j}\, p_{(j)}\right), \quad \text{capped at } 1$$
:::

\normalsize

## BH worked example ($m = 10$, $q = 0.05$)

\small

| Rank $j$ | P-value  | BH threshold $\frac{jq}{m}$ | Adjusted p-value $p^{\textsf{BH}}$ | Reject? |
|:---------:|:--------:|:----------------------------:|:-----------------------------------:|:-------:|
| 1         | 0.0008   | 0.005                        | 0.008                               | $\checkmark$ |
| 2         | 0.009    | 0.010                        | 0.045                               | $\checkmark$ |
| 3         | 0.127    | 0.015                        | 0.423                               |         |
| 4         | 0.205    | 0.020                        | 0.513                               |         |
| 5         | 0.396    | 0.025                        | 0.750                               |         |
| 6         | 0.450    | 0.030                        | 0.750                               |         |
| 7         | 0.641    | 0.035                        | 0.916                               |         |
| 8         | 0.781    | 0.040                        | 0.976                               |         |
| 9         | 0.900    | 0.045                        | 0.993                               |         |
| 10        | 0.993    | 0.050                        | 0.993                               |         |

Bonferroni ($\alpha/m = 0.005$) would only reject rank 1.


## DEG: expression $Y \sim$ cancer stage in TCGA BRCA

```{r include=F}
library(recount3)
library(data.table)
library(tidyverse)
library(patchwork)
library(matrixTests)
library(ggpubr)
library(ggrepel)
library(limma)
library(edgeR)
library(DESeq2)

theme_set(theme_minimal())
```

```{r download-data, include = F}
data.dir <- "../data/recount3"
dir.create(data.dir, recursive = TRUE, showWarnings = FALSE)
.file <- file.path(data.dir, "tcga_brca_rse.rds")

if.needed(.file, {
    tcga.proj <- subset(available_projects(),
                        file_source == "tcga")
    brca.info <- tcga.proj[tcga.proj$project == "BRCA", ]
    rse.brca <- create_rse(brca.info)
    saveRDS(rse.brca, .file)
})
rse <- readRDS(.file)
```

:::::: {.columns}
::: {.column width=.45}

\begin{center}
\begin{tikzpicture}[
    node distance=1.5cm,
    observed/.style={circle, draw, fill=gray!15, minimum size=1.3cm}
]
\node[observed] (age) {Age};
\node[observed, below=1.2cm of age] (meno) {Meno};
\node[observed, below=1.2cm of meno] (stage) {\textbf{Stage}};
\node[observed, right=2.5cm of meno] (expr) {$Y_{g}$};

\draw[->, thick] (age) -- node[above, sloped] {$\beta_1$} (expr);
\draw[->, thick] (meno) -- node[above] {$\beta_2$} (expr);
\draw[->, thick, blue] (stage) -- node[below, sloped] {\textcolor{blue}{$\beta_3$}} (expr);
\end{tikzpicture}
\end{center}

:::
::: {.column width=.55}

**Linear model for gene $g$:**

$$Y_{ig} = \beta_{1g} \cdot \text{Age}_i + \beta_{2g} \cdot \text{Meno}_i + \textcolor{blue}{\beta_{3g} \cdot \text{Stage}_i} + \epsilon_{ig}$$

where $\epsilon_{ig} \sim N(0, \sigma^2_g)$

\vspace{0.3cm}

**Null hypothesis for gene $g$:**

$$H_{0g}: \textcolor{blue}{\beta_{3g} = 0}$$

\vspace{0.2cm}

Testing $m = 20{,}000$ genes $\Rightarrow$ $m = 20{,}000$ hypotheses!

:::
::::::


## Run limma-voom analysis

```{r prepare-data, include=F}
.file <- file.path(data.dir, "brca_prepared_data.rds")

if.needed(.file, {
    # Extract expression data
    raw.counts <- assay(rse, "raw_counts")

    # Get clinical variables from colData
    stage <- rse$tcga.gdc_cases.diagnoses.tumor_stage
    age <- rse$tcga.gdc_cases.diagnoses.age_at_diagnosis
    menopause <- rse$tcga.xml_menopause_status

    # Simplify to early (I/II) vs late (III/IV) stage
    stage_simple <- case_when(
        grepl("stage i$|stage ia|stage ib|stage ii", stage, ignore.case = TRUE) ~ "Early",
        grepl("stage iii|stage iv", stage, ignore.case = TRUE) ~ "Late",
        TRUE ~ NA_character_
    )

    # Simplify menopause status
    meno_simple <- case_when(
        grepl("Pre", menopause, ignore.case = TRUE) ~ "Pre",
        grepl("Post|Peri", menopause, ignore.case = TRUE) ~ "Post",
        TRUE ~ NA_character_
    )

    # Keep only samples with complete data
    keep <- !is.na(stage_simple) & !is.na(age) & !is.na(meno_simple)
    raw.counts <- raw.counts[, keep]
    stage_simple <- factor(stage_simple[keep])
    age <- as.numeric(age[keep])
    meno_simple <- factor(meno_simple[keep])

    # Filter low-count genes
    keep.genes <- rowSums(raw.counts > 1) >= 10
    Y <- raw.counts[keep.genes, ]

    # Save prepared data
    saveRDS(list(Y = Y, stage_simple = stage_simple,
                 age = age, meno_simple = meno_simple), .file)
})

# Load prepared data
prepared <- readRDS(.file)
Y <- prepared$Y
stage_simple <- prepared$stage_simple
age <- prepared$age
meno_simple <- prepared$meno_simple

cat(sprintf("Analyzing %d genes across %d samples\n", nrow(Y), ncol(Y)))
cat(sprintf("Early stage: %d, Late stage: %d\n",
    sum(stage_simple == "Early"), sum(stage_simple == "Late")))
```


```{r limma-voom-analysis}
.file <- file.path(data.dir, "brca_limma_voom_results.rds")

if.needed(.file, {
    # Create design matrix with covariates
    design <- model.matrix(~ age + meno_simple + stage_simple)
    colnames(design)[ncol(design)] <- "Late_vs_Early"

    # Create DGEList object and normalize
    dge <- DGEList(counts = Y)
    dge <- calcNormFactors(dge)

    # Apply voom transformation
    voom_result <- voom(dge, design = design, plot = FALSE)

    # Fit linear model with voom weights
    fit_voom <- lmFit(voom_result, design)

    # Apply empirical Bayes moderation
    fit_voom <- eBayes(fit_voom)

    # Get all results
    results <- topTable(fit_voom, coef = "Late_vs_Early", number = Inf)

    # Save results
    saveRDS(list(fit_voom = fit_voom, results = results), .file)
})

# Load results
limma_results <- readRDS(.file)
fit_voom <- limma_results$fit_voom
results <- limma_results$results
```

```{r show-top-genes}
topTable(fit_voom, coef = "Late_vs_Early", number = 5)[,
    c("logFC", "AveExpr", "t", "P.Value", "adj.P.Val")]
```

\large

How did we get the "adjusted" p-value?

\normalsize


## Volcano plot: visualizing effect size and significance


```{r volcano-plot, fig.width=3.5, fig.height=2.9}
# Prepare data for volcano plot
volcano.dt <- data.table(
    logFC = results$logFC,
    log10p = -log10(results$P.Value),
    P.Value = results$P.Value,
    gene = rownames(results)
)

# Define significance thresholds (raw p-value)
volcano.dt[, significant := ifelse(P.Value < 0.05 & abs(logFC) > 1,
                                   "Significant", "Not significant")]

ggplot(volcano.dt, aes(x = logFC, y = log10p, color = significant)) +
    geom_point(alpha = 0.4, size = 0.8) +
    scale_color_manual(values = c("gray60", "red3")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "blue") +
    labs(x = "log2 Fold Change", y = "-log10(p-value)",
         title = "Volcano plot: Late vs Early stage") +
    theme(legend.position = "bottom")
```

## Benjamini-Hochberg procedure visualized

```{r fig.width=5.5, fig.height=2.9, only.plot="1"}
q <- 0.05
m <- nrow(results)
sorted.pvals <- sort(results$P.Value)
ranks <- 1:m
bh.threshold <- ranks * q / m

bh.dt <- data.table(
    rank = ranks,
    pvalue = sorted.pvals,
    bh.threshold = bh.threshold,
    significant = sorted.pvals <= bh.threshold
)

max.sig.rank <- max(which(bh.dt$significant), 0)

zoom.max <- min(200, m)
ggplot(bh.dt[rank <= zoom.max], aes(x = rank)) +
    geom_line(aes(y = pvalue), color = "black", linewidth = 1) +
    geom_line(aes(y = bh.threshold), color = "red", linewidth = 1) +
    geom_vline(xintercept = max.sig.rank, lty = 2, color = "blue", linewidth = 0.5) +
    annotate("text", x = max.sig.rank, y = max(bh.dt[rank <= zoom.max]$pvalue) * 0.9,
             label = paste0("k = ", max.sig.rank), color = "blue", size = 3.5, hjust = -0.1) +
    annotate("text", x = zoom.max * 0.6, y = bh.threshold[as.integer(zoom.max * 0.6)] * 1.5,
             label = "BH line: i*q/m", color = "red", size = 3.5) +
    annotate("text", x = zoom.max * 0.6, y = sorted.pvals[as.integer(zoom.max * 0.6)] * 0.7,
             label = "Sorted p-values", color = "black", size = 3.5) +
    labs(x = "Rank (i)", y = "p-value",
         title = paste0("Reject ", max.sig.rank, " hypotheses at FDR < ", q)) +
    theme(legend.position = "none")
```

```{r fig.width=5.5, fig.height=2.9, only.plot="2"}
adj.pvals <- p.adjust(results$P.Value, method = "BH")
comparison.dt <- data.table(
    pvalue = results$P.Value,
    adj.pvalue = adj.pvals,
    significant = adj.pvals < q
)

top.genes <- order(results$P.Value)[1:min(200, nrow(results))]
comparison.dt.zoom <- comparison.dt[top.genes]

ggplot(comparison.dt.zoom, aes(x = pvalue, y = adj.pvalue, color = significant)) +
    geom_point(size = 1.2, alpha = 0.6) +
    geom_abline(slope = 1, intercept = 0, lty = 2, color = "gray50") +
    geom_hline(yintercept = q, lty = 2, color = "red") +
    scale_color_manual(values = c("gray60", "red3")) +
    annotate("text", x = max(comparison.dt.zoom$pvalue) * 0.7, y = q + 0.003,
             label = "FDR = 0.05", color = "red", size = 3.5) +
    labs(x = "Raw p-value", y = "BH adjusted p-value",
         title = paste0(sum(comparison.dt$significant), " genes significant at FDR < ", q)) +
    theme(legend.position = "none")
```

## Recap: what are we controlling?

\large

**FWER** (Bonferroni, Holm): Controls **probability** of making **any** false discovery

\centerline{$\text{FWER} = P(V \ge 1) \le \alpha$}

\vfill

**FDR** (Benjamini-Hochberg): Controls **expectation** of **proportion** of false discoveries

\centerline{$\text{FDR} = E\left[\frac{V}{R}\right] \le q$}

\vfill

::: {.alertblock}
### Key difference
BH controls the **expected** FDR, not the probability. In any single experiment, the actual FDR can be higher or lower than q.
:::

\normalsize

# FWER, FDR, then what?

## Well-behaved p-values: uniform under the null

\large

All multiple testing methods assume that p-values under $H_0$ are **uniformly distributed**.

\normalsize

```{r pval-uniform, fig.width=5.5, fig.height=2.5}
set.seed(23)
p.null <- replicate(5000, t.test(rnorm(50))$p.value)

p1 <- ggplot(data.table(p = p.null), aes(x = p)) +
    geom_histogram(bins = 30, fill = "steelblue", colour = "white", linewidth = 0.2) +
    labs(x = "p-value", y = "Count", title = "All null: uniform distribution")

set.seed(23)
p.mixed <- c(
    replicate(4000, t.test(rnorm(50))$p.value),
    replicate(1000, t.test(rnorm(50, mean = 1.5))$p.value)
)

p2 <- ggplot(data.table(p = p.mixed), aes(x = p)) +
    geom_histogram(bins = 30, fill = "steelblue", colour = "white", linewidth = 0.2) +
    labs(x = "p-value", y = "Count", title = "Mixed: near 0 = true signals")

p1 + p2
```

## BH adjusted p-value distribution

```{r fig.width=5.5, fig.height=2.9}
p.mixed.adj <- p.adjust(p.mixed, method = "BH")

p3 <- ggplot(data.table(p = p.mixed.adj), aes(x = p)) +
    geom_histogram(bins = 30, fill = "coral", colour = "white", linewidth = 0.2) +
    labs(x = "BH adjusted p-value", y = "Count",
         title = "After BH adjustment")

p2 + p3
```




## Poorly behaved p-values

\large

When p-values **misbehave**, FDR estimates become invalid.

\vfill

Common causes:

* Model misspecification (wrong distributional assumptions)
* Uncontrolled variation or batch effects
* Selection bias / data filtering after testing

Great reference: [http://varianceexplained.org/statistics/interpreting-pvalue-histogram/](http://varianceexplained.org/statistics/interpreting-pvalue-histogram/)




## What can we do to make sure that p-values

Possible solutions:

* **Nonparametric tests** (but limited flexibility for covariates)

* **Permutation testing**: shuffle group labels to construct an empirical null distribution; compare observed statistics to the permutation distribution

    - Limitation: computationally intensive; not ideal for very small sample sizes



## What can we estimate from data?

:::::: {.columns}
::: {.column width=.55}

```{r fig.width=3, fig.height=2, only.plot="1"}
p.2d.main
```

```{r fig.width=3, fig.height=2, only.plot="2"}
midpoint <- (mu0 + mu1) / 2
p.2d.main +
    geom_abline(intercept = sum(midpoint), slope = -1,
                color = "darkgreen", linewidth = 1, lty = 1)
```

:::
::: {.column width=.4}

* Can we estimate the distribution under $H_{0}$?

* Can we estimate the distribution under $H_{1}$?

* Do we need to estimate what's going on under $H_{1}$?

\onslide<2>{

$${}$$

Can we come up with an \textit{optimal} boundary?

}

:::
::::::

## Bayesian view of multiple testing

\large

**Mixture model framework**: Each hypothesis is either

:::::: {.columns}
::: {.column width=.45}

- **Null** ($H_0$) with prior probability $\pi_0$

- **Alternative** ($H_1$) with probability $1-\pi_0$

:::
::: {.column width=.45}

\normalsize

```{r fig.width=2, fig.height=1.5}
library(ggplot2)
df <- data.frame(
    Status = c("Null", "Alternative"),
    Probability = c(0.6, 0.4)
)
ggplot(df, aes(x = "", y = Probability, fill = Status)) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar(theta = "y") +
    scale_fill_manual(values = c("Alternative" = "steelblue", "Null" = "coral")) +
    theme_void() +
    theme(legend.position = "bottom", legend.text = element_text(size = 8))
```

:::
::::::

\vfill

**Bayesian interpretation** (note what's in the condition)  [@Storey2003-yd]: 

$$q(p_i) \approx P(H_0 \mid p \le p_i)$$


## Bayesian view of multiple testing

\large

We want to control: $P(H_0 \mid \text{reject}) = P(H_0 \mid p \le t)$

**Bayes rule**:

$$P(H_0 \mid p \le t) = \frac{P(p \le t \mid H_0) \cdot P(H_0)}{P(p \le t)}$$

\normalsize

* It is called "positive" FDR because of the denominator.

* What is the first term of the numerator?

 [@Storey2003-yd]

## Storey's q-value

\large

**Key insight**: BH assumes **all** hypotheses are null ($\pi_0 = 1$)

But in reality, many hypotheses are **truly alternative** (non-null)

\vfill

**Storey's q-value**: Estimates $\pi_0$ (proportion of true nulls) from data

:::::: {.columns}
::: {.column width=.45}

$$q(p_i) = \min_{j:p_j \ge p_i} \frac{\color{magenta}{\hat{\boldsymbol{\pi}}_0 \cdot \mathbf{m}} \cdot p_j}{j}$$

:::
::: {.column width=.45}

\normalsize

Compare to BH: 

$$\textsf{BH}(p_i) = \min_{j:p_j \ge p_i} \frac{m \cdot p_j}{j}$$

\normalsize

:::
::::::

\normalsize

\vfill

When $\hat{\pi}_0 < 1$: q-value is **less conservative** than BH

\normalsize

## Estimating $\pi_0$ from p-value histogram

```{r fig.width=5.5, fig.height=2.9}
library(qvalue)

pvals <- results$P.Value
qobj <- qvalue(pvals)

ggplot(data.table(p = pvals), aes(x = p)) +
    geom_histogram(bins = 50, fill = "steelblue", color = "white", linewidth = 0.2) +
    geom_hline(yintercept = length(pvals) * qobj$pi0 / 50,
               color = "red", linewidth = 1, lty = 2) +
    annotate("text", x = 0.7, y = length(pvals) * qobj$pi0 / 50 + 20,
             label = paste0("Estimated pi0 = ", round(qobj$pi0, 3)),
             color = "red", size = 4, vjust=0) +
    labs(x = "p-value", y = "Count",
         title = "P-value histogram: red line estimates null density")
```

## Comparing BH vs. q-value

```{r fig.width=5.5, fig.height=2.9}
qobj.pi05 <- qvalue(pvals, pi0 = 0.5)

comparison <- data.table(
    pvalue = pvals,
    BH = p.adjust(pvals, method = "BH"),
    qvalue = qobj$qvalues,
    qvalue.pi05 = qobj.pi05$qvalues
) %>% arrange(pvalue) %>% mutate(rank = 1:n())

top.genes <- 1:min(200, nrow(comparison))
comparison.zoom <- comparison[top.genes]

qvalue_label <- paste0("q-value (pi0=", round(qobj$pi0, 2), ")")

ggplot(comparison.zoom, aes(x = rank)) +
    geom_line(aes(y = BH, color = "BH (pi0=1)"), linewidth = 1) +
    geom_line(aes(y = qvalue, color = qvalue_label), linewidth = 1) +
    geom_line(aes(y = qvalue.pi05, color = "q-value (pi0=0.5)"), linewidth = 1, lty = 2) +
    scale_color_manual(values = setNames(c("coral", "steelblue", "darkgreen"),
                                         c("BH (pi0=1)", qvalue_label, "q-value (pi0=0.5)"))) +
    geom_hline(yintercept = 0.05, lty = 2, color = "red", linewidth = 0.5) +
    annotate("text", x = max(comparison.zoom$rank) * 0.8, y = 0.055,
             label = "FDR = 0.05", color = "red", size = 2.5) +
    labs(x = "Rank (by p-value)", y = "Adjusted p-value",
         title = "BH vs q-value with different pi0",
         color = "Method") +
    theme(legend.position = "top", legend.text = element_text(size = 8))
```

## Effect of different $\pi_0$ estimates

```{r fig.width=5.5, fig.height=2.9}
pi0.values <- c(0.3, 0.5, 0.7, qobj$pi0, 1.0)
qval.comparison <- data.table()

for(pi0 in pi0.values) {
    qobj.pi0 <- qvalue(pvals, pi0 = pi0)
    qval.comparison <- rbind(qval.comparison,
        data.table(
            pvalue = pvals,
            qvalue = qobj.pi0$qvalues,
            pi0 = paste0("pi0 = ", round(pi0, 3))
        )
    )
}

top.genes.idx <- order(pvals)[1:min(200, length(pvals))]
qval.zoom <- qval.comparison[pvalue %in% pvals[top.genes.idx]]

ggplot(qval.zoom, aes(x = pvalue, y = qvalue, color = pi0)) +
    geom_line(linewidth = 0.8) +
    geom_abline(slope = 1, intercept = 0, lty = 2, color = "gray50") +
    scale_color_manual(values = c("darkred", "orange", "gold", "steelblue", "gray50")) +
    labs(x = "Raw p-value", y = "q-value",
         title = "Effect of pi0 on q-values",
         color = "pi0 estimate") +
    theme(legend.position = "right")
```

:::
::::::

## When to use Storey's q-value vs. BH?

:::::: {.columns}
::: {.column width=.5}

**BH procedure:**

* Conservative (assumes $\pi_0 = 1$)
* No assumptions about null proportion
* Valid (if independent)
* Recommended as default

:::
::: {.column width=.5}

**Storey's q-value:**

* Less conservative when $\pi_0 < 1$
* Estimates $\pi_0$ from data
* More powerful when many true signals
* Risk: $\pi_0$ estimation can be unstable with few tests

:::
::::::

\vfill

**Rule of thumb**: Use BH by default; consider q-value when you have $>1000$ tests with many expected signals

# Beyond global FDR: gene-by-gene localized control

## Local false discovery rate (Efron 2001)

\large

**FDR**: Expected proportion of false discoveries **among all rejections**

**Local fdr**: Probability that **a specific test** is null

$$\text{lfdr}(z_i) = P(H_0 \mid z_i) = \frac{\pi_0 f_0(z_i)}{f(z_i)}$$

\vfill

:::::: {.columns}
::: {.column width=.48}

**Numerator**: $\pi_0 f_0(z_i)$

- Null contribution at $z_i$

:::
::: {.column width=.48}

**Denominator**: $f(z_i)$

- Total density at $z_i$

:::
::::::

\vfill

\normalsize

Can estimate from empirical distribution of z-scores

## Local fdr: DEG example

```{r fig.width=5.5, fig.height=3.2}
library(locfdr)

## Extract effect sizes and standard errors from results
betahat <- results$logFC
sebetahat <- results$logFC / results$t

## Compute z-scores
z_scores <- betahat / sebetahat

## Fit local fdr with built-in visualization
## plot=1 shows histogram with null and alternative densities
lfdr_res <- locfdr(z_scores, plot = 1, nulltype = 1)
```

## q-value vs. local fdr

:::::: {.columns}
::: {.column width=.48}

\normalsize

**q-value (Storey)**

$$q(p_i) = \min_{j:p_j \ge p_i} \frac{\pi_0 \cdot m \cdot p_j}{j}$$

\vfill

- Based on **p-values**
- Ranks tests by p-value
- Minimum FDR at threshold $p_i$
- **Tail probability** interpretation

:::
::: {.column width=.48}

\normalsize

**Local fdr (Efron)**

$$\text{lfdr}(z_i) = \frac{\pi_0 f_0(z_i)}{f(z_i)}$$

\vfill

- Based on **test statistics** (z-scores)
- Density ratio at specific $z_i$
- Posterior P(null $\mid$ $z_i$)
- **Point-wise** interpretation

:::
::::::

\vfill

\normalsize

Both use $\pi_0$, but q-value uses **cumulative** p-values; local fdr uses **density** at each point

## What FDR doesn't tell us

\large

FDR controls the **expected proportion** of false discoveries

But for each gene, we want to know:

* Is this gene **truly differentially expressed**?
* What is the **direction** of the effect?
* What is the **magnitude** of the effect?

\vfill

::: {.alertblock}
### The problem
A small p-value tells us there's an effect, but not its direction or size with certainty
:::

\normalsize

## Local False Sign Rate (lfsr)

\large

**Local fsr**: Probability that the **sign** of an effect is wrong

$$\text{lfsr} = P(\text{sign}(\beta) \neq \text{sign}(\hat{\beta}) \mid \text{data})$$

\vfill

**Key idea**: Account for uncertainty in **both**:

* Whether there's an effect ($\beta \neq 0$)
* The direction of the effect (sign of $\beta$)

\normalsize

## Adaptive Shrinkage (ashr)

\large

**ashr** (Stephens 2017): Empirical Bayes approach

\vfill

**Key features:**

1. Estimates effect size distribution from data
2. Shrinks estimates toward zero (like in ridge regression)
3. Provides posterior probabilities for sign and magnitude
4. Computes local false sign rate (lfsr)

\vfill

More conservative than FDR for **directional claims**

\normalsize

## ashr: Estimating null vs alternative distributions

```{r fig.width=5.5, fig.height=2.8}
library(ashr)

## Fit ashr model using effect sizes and standard errors
betahat <- results$logFC
sebetahat <- results$logFC / results$t
ash_res <- ash(betahat, sebetahat, mixcompdist = "normal")

## Extract fitted mixture components
fitted_g <- ash_res$fitted_g

## Use qvalue's pi0 estimate (more interpretable than ashr's)
## qvalue estimates pi0 from p-value distribution
pi0_for_plot <- qobj$pi0

## Create data for plotting
z_scores <- betahat / sebetahat
density_data <- data.table(z = z_scores)

## Calculate histogram for scaling
hist_data <- hist(z_scores[abs(z_scores) <= 10], breaks = 100, plot = FALSE)
scale_factor <- max(hist_data$density) / max(dnorm(0))

## Pre-compute labels
label_all_null <- "If all null (pi0=1)"
label_null_component <- paste0("Null component (pi0=", round(pi0_for_plot, 2), ")")

ggplot(density_data, aes(x = z)) +
    geom_histogram(aes(y = after_stat(density)), bins = 100,
                   fill = "steelblue", alpha = 0.3, color = "black", linewidth = 0.2) +
    stat_function(fun = function(x) dnorm(x) * scale_factor,
                  aes(color = label_all_null), linewidth = 1, lty = 2) +
    stat_function(fun = function(x) pi0_for_plot * dnorm(x) * scale_factor,
                  aes(color = label_null_component), linewidth = 1.2) +
    scale_color_manual(values = setNames(c("gray50", "coral"),
                                         c(label_all_null, label_null_component))) +
    labs(x = "Z-score (effect size / SE)",
         y = "Density",
         title = "Observed vs null distribution",
         color = "") +
    xlim(-10, 10) +
    theme(legend.position = "top")
```

## ashr: Different alternative distributions

```{r fig.width=5.5, fig.height=2.8}
## Fit ashr with different mixture distributions
ash_normal <- ash(betahat, sebetahat, mixcompdist = "normal")
ash_uniform <- ash(betahat, sebetahat, mixcompdist = "uniform")
ash_halfuniform <- ash(betahat, sebetahat, mixcompdist = "halfuniform")

## Compare lfsr for top genes
lfsr_comparison <- data.table(
    rank = 1:1000,
    Normal = sort(ash_normal$result$lfsr)[1:1000],
    Uniform = sort(ash_uniform$result$lfsr)[1:1000],
    HalfUniform = sort(ash_halfuniform$result$lfsr)[1:1000]
) %>% melt(id.vars = "rank", variable.name = "Method", value.name = "lfsr")

ggplot(lfsr_comparison, aes(x = rank, y = lfsr, color = Method)) +
    geom_line(linewidth = 0.6) +
    geom_hline(yintercept = 0.05, lty = 2, color = "red", linewidth = 0.5) +
    scale_color_manual(values = c("Normal" = "steelblue",
                                   "Uniform" = "darkgreen",
                                   "HalfUniform" = "coral"),
                       labels = c("Normal mixture",
                                  "Uniform mixture",
                                  "Half-uniform (one-sided)")) +
    labs(x = "Rank (by lfsr)",
         y = "Local false sign rate",
         title = "Effect of alternative distribution assumption",
         color = "Alternative",
         caption = paste0("Note: qvalue estimates pi0=", round(qobj$pi0, 2),
                         " from p-values; ashr can give different estimates from effect sizes")) +
    theme(legend.position = "top", legend.text = element_text(size = 8),
          plot.caption = element_text(size = 6, hjust = 0))
```

## Why shrinkage?

\large

**Problem**: Raw effect size estimates are **noisy**

:::::: {.columns}
::: {.column width=.48}

**Genes with large SE:**
- Less precise measurements
- Estimates unreliable
- May be spuriously large

:::
::: {.column width=.48}

**Genes with small SE:**
- More precise measurements
- Estimates more reliable
- Should be trusted more

:::
::::::

\vfill

**Solution**: **Shrink** uncertain estimates toward zero

- Small/noisy effects → shrunk more
- Large/precise effects → preserved
- **Borrows strength** across all genes

\normalsize

## ashr in action: Effect size shrinkage

```{r fig.width=5.5, fig.height=2.8}
## ash_res already fitted above

## Compare original vs shrunken effect sizes
shrinkage_df <- data.table(
    original = betahat,
    shrunken = ash_res$result$PosteriorMean,
    lfsr = ash_res$result$lfsr
)

## Plot top genes (by abs original effect)
top_idx <- order(abs(betahat), decreasing = TRUE)[1:3000]
shrinkage_df_top <- shrinkage_df[top_idx]

ggplot(shrinkage_df_top, aes(x = original, y = shrunken)) +
    geom_point(aes(color = lfsr < 0.05), alpha = 0.3, size = 0.8) +
    geom_abline(slope = 1, intercept = 0, lty = 2, color = "red") +
    scale_color_manual(values = c("FALSE" = "gray", "TRUE" = "steelblue"),
                       labels = c("lfsr >= 0.05", "lfsr < 0.05")) +
    labs(x = "Original log fold-change",
         y = "Shrunken log fold-change",
         title = "Effect size shrinkage by ashr",
         color = "Significance",
         caption = "Points below diagonal: shrunk toward zero (less confident). Points near diagonal: preserved (more confident).") +
    theme(legend.position = "top", plot.caption = element_text(size = 6, hjust = 0))
```

## ashr: local false sign rate vs q-value

```{r fig.width=5.5, fig.height=2.8}
## Compare lfsr with q-values
comparison_ash <- data.table(
    lfsr = ash_res$result$lfsr,
    qvalue = qobj$qvalues,
    abs_effect = abs(betahat)
) %>% arrange(lfsr)

## Zoom to top 100 for clarity
comparison_ash_top <- comparison_ash[1:100]

ggplot(comparison_ash_top, aes(x = 1:nrow(comparison_ash_top))) +
    geom_line(aes(y = lfsr, color = "lfsr"), linewidth = 0.6) +
    geom_line(aes(y = qvalue, color = "q-value"), linewidth = 0.6) +
    geom_hline(yintercept = 0.05, lty = 2, color = "red", linewidth = 0.5) +
    scale_color_manual(values = c("lfsr" = "darkgreen", "q-value" = "steelblue")) +
    labs(x = "Rank (by lfsr)",
         y = "Error rate",
         title = "Local false sign rate vs q-value",
         color = "Method") +
    theme(legend.position = "top")
```

## Summary: Multiple testing methods

:::::: {.columns}
::: {.column width=.48}

\normalsize

**Control FWER** (strict):

- **Bonferroni**: $\alpha/m$
  - Simple, conservative
  - Use for small $m$ or independent tests

- **Holm**: Sequential Bonferroni
  - More powerful than Bonferroni
  - Still controls FWER

:::
::: {.column width=.48}

\normalsize

**Control FDR** (flexible):

- **Benjamini-Hochberg**: Assumes $\pi_0=1$
  - Standard FDR control
  - More discoveries than FWER

- **Storey q-value**: Estimates $\pi_0$
  - More powerful when $\pi_0 < 1$
  - Uses p-values only

- **ashr**: Uses effect sizes + SE
  - Accounts for measurement precision
  - `lfsr` for directions

:::
::::::

