---
title: "Clustering and Differential Analysis"
author: "Yongjin Park"
date: "`r format(Sys.time(), '%d %B %Y')`"
classoption: "aspectratio=169"
fontsize: 12pt
bibliography: network.bib
output:
    beamer_presentation:
        theme: Madrid
        keep_md: true
        keep_tex: true
        latex_engine: pdflatex
        slide_level: 2
header-includes:
  - \usepackage{cancel}
  - \usepackage{tikz}
  - \usetikzlibrary{positioning,fit,shapes,calc}
  - \definecolor{UBCblue}{rgb}{0.04706, 0.13725, 0.26667}
  - \usecolortheme[named=UBCblue]{structure}
  - \setbeamertemplate{frametitle}{\color{UBCblue}\bfseries\insertframetitle\par\vskip-6pt\hrulefill}
  - \setbeamercolor{title in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{section in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{author in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{date in head/foot}{bg=white,fg=gray}
  - \setbeamertemplate{page number in head/foot}{}
  - \setbeamertemplate{frame numbering}[none]
  - \setbeamercolor{alerted text}{bg=yellow}
  - \AtBeginSection[]{\begin{frame}\frametitle{Today's lecture}{\Large\tableofcontents[currentsection]}\end{frame}}
  - |
    \makeatletter
    \def\ps@titlepage{%
      \setbeamertemplate{footline}{}
    }
    \addtobeamertemplate{title page}{\thispagestyle{titlepage}}{}
    \makeatother
---


```{r setup, include=FALSE}
source("Setup.R")
fig.dir <- "../Fig/clustering-differential-analysis/"
dir.create(fig.dir, showWarnings=FALSE)
setup.env(fig.dir)
dir.create("../data", showWarnings=FALSE)
dir.create("../data/network", showWarnings=FALSE)

library(data.table)
library(tidyverse)
library(patchwork)
library(matrixTests)
library(igraph)

theme_set(theme_minimal())

## --- Network helper functions (from lect14) ---

.read.online <- function(.src,
                         .dir="../data/network/",
                         .zip = str_c(.dir,basename(.src)),
                         .tgt = str_replace(.zip, "zip$", "gml")){
    if.needed(.tgt, {
        .gml <- basename(.tgt)
        download.file(.src, .zip)
        unzip(.zip, files=.gml, exdir = .dir)
    })
    return(igraph::read_graph(.tgt, format="gml"))
}

find.nice.layout <- function(G, layout.file) {
    if(!file.exists(layout.file)){
        layout.dt <- layout_nicely(G) %>% as.data.table()
        saveRDS(layout.dt, layout.file)
    }
    readRDS(layout.file)
}

apply.layout.dt <- function(G, .name){
    .out.file <- str_c(fig.dir,"/",.name,".rds")
    .out <- find.nice.layout(G, .out.file) %>%
        dplyr::mutate(v = 1:dplyr::n())
    as_long_data_frame(G) %>%
        as.data.table %>%
        merge(.out, by.x= "from", by.y="v") %>%
        merge(.out, by.x= "to", by.y="v", suffixes=c(".from",".to"))
}

plot.layout.dt <- function(.dt, .palette="Paired", .palette.e = "Set2",
                           .esize = .5, .vsize = 5, .add.lab=TRUE){
    if(!("k.from" %in% colnames(.dt))) .dt[, k.from := 1]
    if(!("k.to" %in% colnames(.dt))) .dt[, k.to := 1]
    .rename <- function(xx) { colnames(xx) <- c("v","V1","V2","k"); xx }
    .dt.v <- rbind(.dt[, .(`from`, `V1.from`, `V2.from`, `k.from`)] %>% .rename(),
                   .dt[, .(`to`, `V1.to`, `V2.to`, `k.to`)] %>% .rename()) %>%
        dplyr::mutate(k = as.factor(k)) %>% unique
    .aes.e <- aes(V1.from, V2.from, xend=V1.to, yend=V2.to)
    if("k.edge" %in% colnames(.dt)) {
        .aes.e <- aes(V1.from, V2.from, xend=V1.to, yend=V2.to, colour=as.factor(k.edge))
    }
    plt <- ggplot() + theme_void()
    if("k.edge" %in% colnames(.dt)) {
        plt <- plt + geom_segment(.aes.e, data = .dt, size = .esize) +
            scale_colour_brewer(palette=.palette.e, na.value="grey", guide="none")
    } else {
        plt <- plt + geom_segment(.aes.e, data = .dt, size = .esize, colour="gray")
    }
    plt <- plt +
        geom_point(aes(V1, V2, fill=k), data = .dt.v, pch=21, stroke=.1, size = .vsize) +
        scale_fill_brewer(palette = .palette, guide="none")
    if(.add.lab){ plt <- plt + geom_text(aes(V1, V2, label=v), data = .dt.v, size = 3) }
    return(plt)
}
```


## Roadmap: From biology to methodology

:::::: {.columns}
::: {.column width=.45}

**14: Biological Networks**

* Protein-protein, genetic interactions

* Cell-cell interaction networks

* Network as a tool: propagation

:::
::: {.column width=.45}

**15: Clustering**

* Graph-based clustering

* Single-cell example

* When clustering goes wrong

:::
::::::

# Graph-based clustering

## Last time: networks encode relationships

* We learned how to build and visualize networks

* Protein-protein interactions, genetic interactions, cell-cell communication

* Today: how do we find **groups** within a network?

## Recap: `igraph` in R

```{r echo=TRUE, eval=FALSE}
library(igraph)
```

```{r echo=TRUE}
karate <- .read.online("http://www-personal.umich.edu/~mejn/netdata/karate.zip")
karate
```

* $n = |V|$: number of vertices, $m = |E|$: number of edges

* $d_v$: degree of vertex $v$ (how many neighbours?)





## A classic example: Karate club [@Zachary1977-rm]

:::::: {.columns}
::: {.column width=.4}

* 34 members, edges = friendships

* The club split into two factions after a dispute

* Can we recover this split from the network alone?

* Who are the "influencers"?

	* **1**: the instructor ("Mr. Hi")

	* **34**: the president (John A.)

:::
::: {.column width=.55}

```{r echo=FALSE, fig.width=3, fig.height=2.5}
.karate.dt <- apply.layout.dt(karate, "karate")
plot.layout.dt(.karate.dt)
```

:::
::::::

## Idea: split the network by cutting edges

:::::: {.columns}
::: {.column width=.4}


* Consider two groups — find a set of edges to cut so the graph splits into two disconnected components

* **Min-cut**: the fewest edges to remove

$${}$$

A trivial solution?

:::
::: {.column width=.55}

```{r trivial-mincut, echo=FALSE, fig.width=3, fig.height=2.5}
.dt <- apply.layout.dt(karate, "karate")
.dt[, k.from := 1]
.dt[, k.to := 1]
.dt[`from` == 12, k.from := 2]
.dt[`to` == 12, k.to := 2]
plot.layout.dt(.dt)
```

:::
::::::



## Split $G$ by the fewest edges between two parties

```{r echo=TRUE, size="large"}
.cuts <-
    as.directed(karate) %>%
    st_min_cuts(source = "1", target = c("33","34"))
.cuts$value / 2  # number of edges to cut
```

## Min-cut partitions from source to target

```{r mincut-partitions, echo=FALSE, fig.width=6, fig.height=1.5}
.sets <- .cuts[["partition1s"]] %>% lapply(as.character)

plots <- lapply(seq_along(.sets)[1:min(3, length(.sets))], function(i) {
    .dt <- apply.layout.dt(karate, "karate")
    .dt[, k.from := ifelse(`from` %in% .sets[[i]], 2, 1)]
    .dt[, k.to := ifelse(`to` %in% .sets[[i]], 2, 1)]
    plot.layout.dt(.dt, .vsize=2, .add.lab=FALSE) +
        ggtitle(paste0("Partition ", i, " (", length(.sets[[i]]), " members)")) +
        theme(title=element_text(size=6))
})
wrap_plots(plots, ncol=3)
```

* Multiple solutions with the same min-cut = `r .cuts$value/2` edges

* But min-cut requires knowing **source and target** — can we find groups without labels?

## The network as an adjacency matrix $A$

:::::: {.columns}
::: {.column width=.5}

```{r fig.width=2.8, fig.height=2.8}
A <- as_adjacency_matrix(karate)
.matshow(A, .scale=FALSE, .lab = 0)
```

:::
::: {.column width=.5}

$${}$$

* Row/column: a member

* Non-zero element: friendship

* Can you see group structures?

$$A_{ij} = 
\left\{
\begin{array}{ll}
1 & i,j \text{ connected}\\
0 & \text{otherwise}
\end{array}
\right.$$

:::
::::::

## Stochastic block model: a generative model of $A$?

* An adjacency matrix $A$ can be **data**

$$A_{ij} \sim \text{Bernoulli}\left(\sum_{k,k'}\theta_{kk'} Z_{ik} Z_{jk}\right)$$

where
$$Z_{ik} =
\left\{
\begin{array}{ll}
1 & i \in \text{ cluster/block/membership } k\\
0 & \text{otherwise}
\end{array}
\right.$$
and
$$\theta_{kk'} = P(A=1|\text{between communities } k \text{ and } k')$$


* **Goal**: factorize $A \approx Z^{\top} \Theta Z$

## Fitting the stochastic block model

```{r echo=TRUE}
library(hsblock)
.out <- fit.hsblock(A, "bernoulli", vbiter=100,
                    inner.iter=10, verbose=FALSE)
Z <- as.matrix(.out$Z)
```

:::::: {.columns}
::: {.column width=.45}

```{r echo=TRUE, size="scriptsize"}
n <- ncol(Z)
Edges <- 0.5 * Z %*% A %*% t(Z)
.ones <- matrix(1,n,n)
diag(.ones) <- 0
Tot <- Z %*% .ones %*% t(Z)/2
Theta <- Edges / Tot
```

:::
::: {.column width=.45}

* **Z**: `r nrow(Z)` clusters × `r ncol(Z)` vertices
* $\Theta$: edge probabilities

:::
::::::

## Matrix factorization view: $A \approx Z^{\top} \Theta Z$

```{r echo=FALSE, fig.width=6, fig.height=2.2}
.p.eq <- ggplot() + theme_void() + geom_text(aes(x=1,y=1,label="≈"), size=10)
.p.data <- .matshow(A, .scale=FALSE, .lab = 0)
.p.z <- .matshow(Z, .scale=FALSE, .lab = 0)
.p.zt <- .matshow(t(Z), .scale=FALSE, .lab = 0)
.p.theta <- .matshow(Theta, .scale=TRUE, .lab = 0)
.p0 <- ggplot() + theme_void()

wrap_plots(.p.data, .p0, .p.eq, .p0, .p.zt, .p0,
           wrap_plots(.p.theta, .p0, ncol=1, heights=c(1,9)),
           .p0,
           wrap_plots(.p.z, .p0, ncol=1, heights=c(1,9)),
           nrow = 1,
           widths=c(4, .01, .5, .05, .5, .05, .5, .05, 4))
```

## Reordering by cluster reveals the block structure

```{r echo=FALSE, fig.width=6, fig.height=2}
oo <- order(apply(Z,2,which.max), decreasing=TRUE)

.p.data <- .matshow(A[oo,oo], .scale=FALSE, .lab = 0)
.p.z <- .matshow(Z[,oo], .scale=FALSE, .lab = 0)
.p.zt <- .matshow(t(Z[,oo]), .scale=FALSE, .lab = 0)

wrap_plots(.p.data, .p0, .p.eq, .p0, .p.zt, .p0,
           wrap_plots(.p.theta, .p0, ncol=1, heights=c(1,9)),
           .p0,
           wrap_plots(.p.z, .p0, ncol=1, heights=c(1,9)),
           nrow = 1,
           widths=c(4, .01, .5, .05, .5, .05, .5, .05, 4))
```

* Sorting vertices by cluster membership makes the block structure visible
* Each block shows edge density within/between groups

## How do we know clustering is good enough?

:::::: {.columns}
::: {.column width=.4}

\includegraphics[width=\textwidth]{img/newman_modularity-1.pdf}

$${}$$

* cluster $\approx$ densely connected subgraph

* but how dense?

:::
::: {.column width=.4}

\includegraphics[width=\textwidth]{img/newman_modularity-2.pdf}

It's not always as perfect as this...

:::
::::::

## In a dendrogram of clusters, where should we cut?

:::::: {.columns}
::: {.column width=.4}

${}$

\includegraphics[width=\textwidth]{img/newman_modularity-4.pdf}

${}$

This is perhaps what happened after the split...

:::
::: {.column width=.2}

\includegraphics[height=.8\textheight]{img/newman_modularity-5.pdf}

:::
::: {.column width=.25}

$${}$$

* **What will be the proper score?**

* The underlying combinatorial optimization is one of the hardest CS problems.

:::
::::::



## What is modularity? [@Newman2004-fa]

Given a partition of vertices into groups, **modularity** $Q$ measures how many more edges fall *within* groups than expected by chance:

$$Q = \frac{1}{2m} \sum_{ij} \left[ {\color{teal} \mathbf{A}_{ij}} - {\color{magenta} \frac{d_i d_j}{2m} } \right] \delta(c_i, c_j)$$

* $A_{ij}$: adjacency matrix, $d_i$: degree of vertex $i$, $m$: total edges

* $\delta(c_i, c_j) = 1$ if vertices $i,j$ are in the same group

* $Q \in [-0.5, 1]$; higher = stronger community structure

* The null model: edges are distributed proportionally to degree

## Why $d_i d_j / 2m$?

* Each vertex $i$ has $d_i$ "stubs" (half-edges). Total stubs: $\sum_i d_i = 2m$.

* Pick a stub from $i$. It connects to one of $2m - 1 \approx 2m$ other stubs.

* Vertex $j$ has $d_j$ stubs, so:

$$Pr(\text{stub from } i \text{ lands on } j) = \frac{d_j}{2m}$$

* Vertex $i$ has $d_i$ stubs, so the expected number of edges between $i$ and $j$:

$$\mathbb{E}[A_{ij}] = \frac{d_i \, d_j}{2m}$$


## Louvain: Greedy optimization of the modularity score

:::::: {.columns}
::: {.column width=.45}

${}$

\includegraphics[width=\textwidth]{img/louvain_leiden-1.pdf}

\scriptsize

[@Blondel2008-fk]:

:::
::: {.column width=.45}

1. Each vertex starts as its own community
2. Move vertices to neighbour's community if it increases $Q$
3. Aggregate communities into super-nodes
4. Repeat until $Q$ converges

:::
::::::


## Louvain algorithm: vertices $\to$ "super" nodes

\centerline{\includegraphics[height=.75\textheight]{img/louvain_leiden-2.pdf}}

\scriptsize

[@Blondel2008-fk]

## Louvain deals with the "colours" of "super" nodes

\centerline{\includegraphics[width=.8\textwidth]{img/louvain_leiden-5.pdf}}

\vfill

**Key idea**: Keep coarsening a large graph into a smaller one

\vfill

\scriptsize

[@Traag2019-ua]

## Louvain can easily scale up to a million vertices

\centerline{\includegraphics[height=.75\textheight]{img/louvain_leiden-3-belgian-telephone.pdf}}

\vfill

\scriptsize

[@Blondel2008-fk]


## Louvain's supernode with disconnected components

:::::: {.columns}
::: {.column width=.35}

\centerline{\includegraphics[width=\textwidth]{img/louvain_leiden-4.pdf}}

:::
::: {.column width=.6}

\centerline{\includegraphics[width=\textwidth]{img/louvain_leiden-6.pdf}}

:::
::::::

\vfill

\scriptsize

[@Traag2019-ua]

## Leiden "refines" to ensure connected within a cluster

\only<1>{\includegraphics[width=.6\textwidth]{img/louvain_leiden-move-1.pdf}}

\only<2>{\includegraphics[width=.9\textwidth]{img/louvain_leiden-move-2.pdf}}

\only<3>{\includegraphics[width=.6\textwidth]{img/louvain_leiden-move-3.pdf}}

\only<4>{\includegraphics[width=.9\textwidth]{img/louvain_leiden-move-4.pdf}}

\vfill

\scriptsize

[@Traag2019-ua]

## Leiden vs. Louvain

```{r echo=TRUE, size="large"}
cl.louvain <- cluster_louvain(karate)
cl.leiden  <- cluster_leiden(karate,
    objective_function = "modularity")
```

## Leiden vs. Louvain

```{r echo=FALSE, fig.width=5.5, fig.height=2.5}
.dt <- apply.layout.dt(karate, "karate")
.dt[, k.from := membership(cl.louvain)[`from`]]
.dt[, k.to := membership(cl.louvain)[`to`]]
p1 <- plot.layout.dt(.dt, .vsize=3, .add.lab=FALSE) +
    ggtitle(paste0("Louvain Q=", round(modularity(cl.louvain),3)))

.dt[, k.from := membership(cl.leiden)[`from`]]
.dt[, k.to := membership(cl.leiden)[`to`]]
p2 <- plot.layout.dt(.dt, .vsize=3, .add.lab=FALSE) +
    ggtitle(paste0("Leiden Q=", round(modularity(karate, membership(cl.leiden)),3)))

p1 + p2 + plot_layout(ncol=2)
```

## What should we learn from Louvain/Leiden?

\large

When $n$ is large...

\vfill

* What shall we do?

* Any counter examples for Leiden?


## Side note: The resolution parameter matters

```{r echo=FALSE, fig.width=5, fig.height=3}
res.vals <- c(0.5, 1.0, 1.5, 2.0)
plots <- lapply(res.vals, function(r) {
    cl <- cluster_leiden(karate, objective_function="modularity",
                         resolution=r)
    .dt <- apply.layout.dt(karate, "karate")
    .dt[, k.from := membership(cl)[`from`]]
    .dt[, k.to := membership(cl)[`to`]]
    plot.layout.dt(.dt, .vsize=2, .add.lab=FALSE) +
        ggtitle(paste0("res=", r, " (k=", max(membership(cl)), ")")) +
        theme(title=element_text(size=6))
})
wrap_plots(plots, ncol=2)
```


<!-- * Both Louvain and Leiden have a **resolution** parameter -->

<!-- * Higher resolution $\to$ more, smaller clusters -->

<!-- * Lower resolution $\to$ fewer, larger clusters -->



# Single-cell clustering in practice

## Recall: the standard single-cell workflow

\centerline{\includegraphics[width=.8\textwidth]{img/scRNA_overview/Karchenko_1.pdf}}

\centerline{\includegraphics[width=.8\textwidth]{img/scRNA_overview/Karchenko_2.pdf}}

\vfill

\flushright
\tiny
[@Kharchenko2021-wr]



## Clustering the PDAC data

```{r load-seurat, message=FALSE, warning=FALSE}
library(Seurat)
```

```{r load-pdac, include=FALSE}
pdac.rds <- "../data/GSE155698_PDAC_Steele2020/pdac_tissue1_processed.rds"
if.needed(pdac.rds, {
    data.dir <- "../data/GSE155698_PDAC_Steele2020/PDAC_TISSUE_1/"
    pdac <- Read10X(data.dir = data.dir)
    pdac <- CreateSeuratObject(counts = pdac,
                               project = "PDAC_TISSUE_1",
                               min.cells = 3,
                               min.features = 200)
    pdac[["percent.mt"]] <- PercentageFeatureSet(pdac, pattern = "^MT-")
    pdac <- subset(pdac,
                   nFeature_RNA > 200 &
                   nFeature_RNA < 6000 &
                   percent.mt < 20)
    pdac <- pdac |>
        NormalizeData() |>
        FindVariableFeatures(nfeatures = 2000) |>
        ScaleData() |>
        RunPCA(npcs = 30) |>
        FindNeighbors(dims = 1:20) |>
        RunUMAP(dims = 1:20)
    saveRDS(pdac, pdac.rds)
})
pdac <- readRDS(pdac.rds)
```

```{r show-pdac-pipeline, eval=FALSE, echo=TRUE}
pdac <- Read10X(data.dir = data.dir)
pdac <- CreateSeuratObject(counts = pdac,
    project = "PDAC_TISSUE_1",
    min.cells = 3, min.features = 200)

pdac <- pdac |>
    NormalizeData() |>
    FindVariableFeatures(nfeatures = 2000) |>
    ScaleData() |>
    RunPCA(npcs = 30) |>
    FindNeighbors(dims = 1:20) |>
    RunUMAP(dims = 1:20)
```

* Same PDAC Tissue 1 dataset from Lecture 8 [@Steele2020-tq]

* `FindNeighbors()` builds the **KNN** and **SNN** graphs in PCA space

## Building the KNN graph

```{r pdac-knn-prep, include=FALSE}
knn.mat <- pdac@graphs$RNA_nn
snn.mat <- pdac@graphs$RNA_snn

set.seed(42)
n.sub <- 500
cells.sub <- sample(colnames(pdac), n.sub)
umap.coords <- Embeddings(pdac, "umap")[cells.sub, ]
```

:::::: {.columns}
::: {.column width=.45}

```{r pdac-knn-show, echo=TRUE, eval=FALSE, size="large"}
knn.mat <- pdac@graphs$RNA_nn
dim(knn.mat)
```

```{r pdac-knn-dim, echo=FALSE}
dim(knn.mat)
```

* `FindNeighbors(dims = 1:20)` connects each cell to its $k=20$ nearest neighbours in PCA space

* $A_{ij} = 1$ if cell $j$ is among $k$-NN of cell $i$

:::
::: {.column width=.5}

```{r pdac-knn-igraph, echo=FALSE, fig.width=2.8, fig.height=2.7}
knn.sub <- knn.mat[cells.sub, cells.sub]
G.knn <- graph_from_adjacency_matrix(knn.sub, mode = "undirected", diag = FALSE)

.dt.knn <- data.table(
    x = umap.coords[,1], y = umap.coords[,2],
    v = seq_len(n.sub))
.el <- as_edgelist(G.knn)
.edge.dt <- data.table(
    x = umap.coords[.el[,1], 1], y = umap.coords[.el[,1], 2],
    xend = umap.coords[.el[,2], 1], yend = umap.coords[.el[,2], 2])

ggplot(.edge.dt) +
    geom_segment(aes(x, y, xend=xend, yend=yend),
                 colour="grey70", linewidth=0.1, alpha=0.3) +
    geom_point(aes(x, y), data=.dt.knn,
               size=0.5, colour="steelblue") +
    labs(x = "UMAP 1", y = "UMAP 2")
```

:::
::::::

## Shared neighbours graph

:::::: {.columns}
::: {.column width=.45}

* **SNN**: weight edges by shared neighbour overlap (Jaccard index)

$$w_{ij} = \frac{|N_i \cap N_j|}{|N_i \cup N_j|}$$

* Cells sharing many neighbours get stronger connections

* This is what Leiden/Louvain clusters on

:::
::: {.column width=.5}

```{r pdac-snn-igraph, echo=FALSE, fig.width=2.8, fig.height=2.7}
snn.sub <- snn.mat[cells.sub, cells.sub]
G.snn <- graph_from_adjacency_matrix(snn.sub, mode = "undirected",
                                     weighted = TRUE, diag = FALSE)
.el.s <- as_edgelist(G.snn)
.ew <- E(G.snn)$weight
.ew <- .ew / max(.ew)

.edge.snn <- data.table(
    x = umap.coords[.el.s[,1], 1], y = umap.coords[.el.s[,1], 2],
    xend = umap.coords[.el.s[,2], 1], yend = umap.coords[.el.s[,2], 2],
    w = .ew)

ggplot(.edge.snn) +
    geom_segment(aes(x, y, xend=xend, yend=yend, alpha=w),
                 colour="grey30", linewidth=0.1) +
    scale_alpha_identity() +
    geom_point(aes(x, y), data=.dt.knn,
               size=0.5, colour="steelblue") +
    labs(x = "UMAP 1", y = "UMAP 2")
```

:::
::::::

<!-- ## From KNN to SNN: the Jaccard index -->
<!-- **Seurat** uses binary KNN ($A_{ik} \in \{0,1\}$, each cell has exactly $k$ neighbours): -->
<!-- $$w_{ij} = \frac{|N_i \cap N_j|}{|N_i \cup N_j|} = \frac{(A A^\top)_{ij}}{k + k - (A A^\top)_{ij}}$$ -->
<!-- **Weighted KNN** ($A_{ik} \geq 0$): the **weighted Jaccard** generalizes this -->
<!-- $$w_{ij} = \frac{\sum_\ell \min(A_{i\ell},\, A_{j\ell})}{\sum_\ell \max(A_{i\ell},\, A_{j\ell})}$$ -->
<!-- * When $A$ is binary: $\min = \text{AND}$, $\max = \text{OR}$ $\to$ recovers standard Jaccard -->
<!-- * E.g., BBKNN [@Polanski2020-ck] produces a weighted KNN across batches -->
<!-- This gives a **weighted adjacency** where cross-batch edges are explicitly encouraged. -->

## Clustering the PDAC data

```{r pdac-leiden, include=FALSE}
pdac.clust.rds <- "../data/GSE155698_PDAC_Steele2020/pdac_tissue1_clustered.rds"
if.needed(pdac.clust.rds, {
    pdac <- FindClusters(pdac, algorithm = 4, resolution = 0.5)
    saveRDS(pdac, pdac.clust.rds)
})
pdac <- readRDS(pdac.clust.rds)
```

:::::: {.columns}
::: {.column width=.4}

```{r pdac-leiden-show, eval=FALSE, echo=TRUE, size="normalsize"}
pdac <- FindClusters(
    pdac,
    algorithm = 4,
    resolution = 0.5)
```

* Leiden on the SNN graph

* `r length(unique(Idents(pdac)))` clusters found

* Resolution = 0.5

:::
::: {.column width=.55}

```{r pdac-umap, echo=FALSE, fig.width=3.5, fig.height=3}
DimPlot(pdac, reduction = "umap", pt.size = 0.2) +
    NoLegend() +
    labs(title = "PDAC Tissue 1: Leiden clustering") +
    theme(plot.title = element_text(size = 9))
```

:::
::::::

## Effect of the resolution parameter

```{r pdac-resolution, echo=FALSE, fig.width=5.5, fig.height=3}
pdac.res.rds <- "../data/GSE155698_PDAC_Steele2020/pdac_tissue1_resolution_plots.rds"
if.needed(pdac.res.rds, {
    res.vals <- c(0.2, 0.5, 1.0, 2.0)
    plots <- lapply(res.vals, function(r) {
        pdac <- FindClusters(pdac, algorithm = 4, resolution = r, verbose = FALSE)
        DimPlot(pdac, reduction = "umap", pt.size = 0.15) +
            NoLegend() +
            labs(title = paste0("res=", r, " (k=", length(unique(Idents(pdac))), ")")) +
            theme(plot.title = element_text(size = 7),
                  axis.title = element_text(size = 6),
                  axis.text = element_text(size = 5))
    })
    saveRDS(plots, pdac.res.rds)
})
plots <- readRDS(pdac.res.rds)
wrap_plots(plots, ncol = 2)
```

## The resolution parameter in the modularity score

The **resolution parameter** $\gamma$ scales the null model:

$$Q_\gamma = \frac{1}{2m} \sum_{ij} \left[ A_{ij} - {\color{magenta}\gamma} \frac{d_i d_j}{2m} \right] \delta(c_i, c_j)$$

* $\gamma > 1$: penalizes large clusters more $\to$ **more, smaller** clusters

* $\gamma < 1$: tolerates sparser communities $\to$ **fewer, larger** clusters

* $\gamma = 1$: standard modularity



# When clustering goes wrong





## Are clusters real or artifacts of the algorithm?

\centerline{\includegraphics[width=.7\textwidth]{img/grabski_irizarry_signficance-1.pdf}}

\tiny
\flushright

[@Grabski2023-vy]

## Clustering a single population finds spurious clusters

\includegraphics[height=.7\textheight]{img/grabski_irizarry_signficance-2.pdf}

\vfill
\tiny

Simulated data with **no true subpopulations**. PCA (left) shows one blob; Leiden (right) still finds clusters. [@Grabski2023-vy]


## Higher resolution keeps splitting --- but are the splits real?

:::::: {.columns}
::: {.column width=.6}

\includegraphics[width=.9\textwidth]{img/grabski_irizarry_signficance-3.pdf}

:::
::: {.column width=.35}

Each node = a cluster at a given resolution. Arrows show how cells redistribute as resolution increases.

\vspace{1em}

Thin arrows = small proportion of cells moving $\to$ potentially spurious splits.

:::
::::::

\vfill
\tiny

[@Grabski2023-vy]


## sc-SHC: recursively test each split for significance

\includegraphics[width=.9\textwidth]{img/grabski_irizarry_signficance-4.pdf}

\vfill
\tiny

[@Grabski2023-vy]

## Can we compute the p-value of each split?

\includegraphics[height=.7\textheight]{img/grabski_irizarry_signficance-5.pdf}

\vfill
\tiny

Top: two real populations $\to$ observed $w$ far from null ($P < 0.001$). Bottom: one population $\to$ $w$ typical under null ($P = 0.97$). [@Grabski2023-vy]


## Capillary aerocytes are a distinct subtype

\includegraphics[width=.9\textwidth]{img/grabski_irizarry_signficance-6.pdf}

\vfill
\tiny

[@Grabski2023-vy]

## Not all splits are real: capillary subtypes overlap

\includegraphics[width=.9\textwidth]{img/grabski_irizarry_signficance-7.pdf}

\vfill
\tiny

[@Grabski2023-vy]

<!-- sc-SHC demo removed: testClusters() too slow for live build -->


## Over-clustering [@DenAdel2025-ed]

**Problem**: Leiden with a high resolution parameter $\to$ too many clusters $\to$ spurious DE genes

**RECALL** (calibrated clustering with artificial variables):

1. **Augment** the expression matrix with **artificial null genes** — synthetic variables that match the data distribution but have no association with cell types

2. **Cluster** cells using both real and artificial genes

3. **Test for DE** between clusters, for both real and artificial genes

4. Since artificial genes are known nulls, their test statistics reveal the **double-dipping effect**

5. If real genes are not more significant than artificial ones $\to$ **merge** those clusters

6. Repeat until clusters are well-calibrated

Works with any clustering algorithm that has a resolution parameter.

## Default clustering over-clusters; RECALL calibrates [@DenAdel2025-ed]

\centerline{\includegraphics[height=.8\textheight]{img/artificial_var_over_cluster-1.pdf}}


## RECALL: iteratively reduce resolution until calibrated

\centerline{\includegraphics[width=\textwidth]{img/artificial_var_over_cluster-2.pdf}}


## RECALL correctly identifies true clusters

\centerline{\includegraphics[height=.7\textheight]{img/artificial_var_over_cluster-3.pdf}}

\vfill
\tiny

B--C: one population --- Seurat finds about 5, RECALL finds 1. D--E: three populations --- both recover 3. [@DenAdel2025-ed]


## RECALL outperforms alternatives on real tissue data

\centerline{\includegraphics[height=.65\textheight]{img/artificial_var_over_cluster-4.pdf}}

\vfill
\tiny

Diaphragm and limb muscle datasets. ARI and V-measure vs curated labels. [@DenAdel2025-ed]


## Seurat default over-splits; RECALL matches cell ontology

\centerline{\includegraphics[height=.7\textheight]{img/artificial_var_over_cluster-5.pdf}}

\vfill
\tiny

Limb muscle: Seurat finds 11 clusters, RECALL finds 6 matching the 6 annotated cell types. Heatmaps show cleaner marker gene blocks. [@DenAdel2025-ed]


## How does RECALL generate artificial null genes?

For each gene $j$, fit a **zero-inflated Poisson** (ZIP) to the observed counts:

$$P(X_j = 0) = \pi_{0j} + (1 - \pi_{0j})\, e^{-\lambda_j}, \quad P(X_j = x) = (1 - \pi_{0j})\, \frac{\lambda_j^x\, e^{-\lambda_j}}{x!}$$

Estimate parameters from the data:

* $\hat{\lambda}_j$ from the mean $\bar{x}_j$ and zero-fraction $r_{0j}$ via the Lambert $W$ function

* $\hat{\pi}_{0j} = 1 - \bar{x}_j / \hat{\lambda}_j$

Then sample synthetic null genes: $\tilde{x}_j \sim \text{ZIP}(\hat{\pi}_{0j},\, \hat{\lambda}_j)$

Each null gene matches the **marginal** distribution (mean, zero-inflation) of the real gene.

<!-- If real genes are not more significant than null genes $\to$ the cluster split is **spurious**. -->

## RECALL on the PDAC data

```{r recall-run, include=FALSE}
recall.rds <- "../data/GSE155698_PDAC_Steele2020/pdac_tissue1_recall.rds"
if.needed(recall.rds, {
    library(recall)
    pdac.recall <- FindClustersRecall(pdac,
                                      algorithm = "leiden",
                                      resolution_start = 0.5,
                                      dims = 1:20,
                                      cores = 10,
                                      verbose = TRUE)
    saveRDS(pdac.recall, recall.rds)
})
pdac.recall <- readRDS(recall.rds)
```

```{r recall-show, eval=FALSE, echo=TRUE, size="normalsize"}
library(recall)
pdac.recall <- FindClustersRecall(pdac,
    algorithm = "leiden",
    resolution_start = 0.5, dims = 1:20)
```

```{r recall-plot, echo=FALSE, fig.width=5.5, fig.height=2.7}
p1 <- DimPlot(pdac, reduction = "umap", pt.size = 0.15) +
    NoLegend() +
    labs(title = paste0("Leiden res=0.5 (",
         length(unique(Idents(pdac))), " clusters)")) +
    theme(plot.title = element_text(size = 7),
          axis.title = element_text(size = 6),
          axis.text = element_text(size = 5))

p2 <- DimPlot(pdac.recall, reduction = "umap", pt.size = 0.15) +
    NoLegend() +
    labs(title = paste0("RECALL (",
         length(unique(Idents(pdac.recall))), " clusters)")) +
    theme(plot.title = element_text(size = 7),
          axis.title = element_text(size = 6),
          axis.text = element_text(size = 5))

p1 + p2
```


## 

\centerline{\includegraphics[width=.8\textwidth]{img/zhang_tse_sc_diff-1.pdf}}

\vfill

* Cluster cells using gene expression $\to$ find marker genes that differ between clusters

* But clusters were *defined* to maximize gene expression differences

* This is **selection bias**: the same data used to form and evaluate the hypothesis

\vfill
\tiny
\flushright 

[@Zhang2019-ku]


## Clustering creates artificial DE even from one population

\centerline{\includegraphics[width=.6\textwidth]{img/zhang_tse_sc_diff-2.pdf}}

\vfill

500 cells from **one** distribution ($\mu = 0.5$). After $k$-means, the two clusters look different --- but the difference is an artefact. [@Zhang2019-ku]


## Standard DE tests are inflated; corrected tests are not

\centerline{\includegraphics[width=.9\textwidth]{img/zhang_tse_sc_diff-3.pdf}}

\vfill

\small
Blue = standard test, orange = truncated normal (TN) corrected test. \textsf{\color{red}Red gene names} = housekeeping genes that should not be DE. [@Zhang2019-ku]


## Clustering then testing $\to$ inflated significance

Even when there is **no true grouping**, standard DE after clustering produces many "significant" genes:

* A "double dipping" problem

* The clustering algorithm forces cells into groups

* Cells assigned to the same cluster are more similar *by construction*

* Standard tests (e.g., Wilcoxon, t-test) ignore this selection step $\to$ **inflated p-values**



## The double-dipping problem

We observe a count matrix $X$ (cells $\times$ genes), where $X_{ij} \sim \text{Poisson}(\lambda_{ij})$

**Standard pipeline** (double dipping):

* Use $X$ to estimate clusters $\hat{c}_1, \ldots, \hat{c}_n$

* Use $X$ again to test: "Is gene $j$ differentially expressed between clusters?"

* The test does not account for the fact that clusters were chosen to make genes look different

**Result**: under the null ($\lambda_{ij}$ does not depend on cell type), we still get many small p-values

\vfill
\tiny
\flushright

[@Neufeld2023-fd]

## 

\centerline{\includegraphics[width=.7\textwidth]{img/neufeld_double_dipping-1.pdf}}

## Null genes look DE after clustering: both $p < 0.001$

\centerline{\includegraphics[width=.95\textwidth]{img/neufeld_double_dipping-2.pdf}}

\vfill

Two genes drawn from the **same** Poisson distribution across all cells. After clustering (right), both appear significantly DE ($p < 0.001$). [@Neufeld2023-fd]


## Count splitting [@Neufeld2023-fd]

**Key insight**: Poisson thinning creates independent splits

If $X_{ij} \sim \text{Poisson}(\lambda_{ij})$, then for any $\epsilon \in (0,1)$:

$$X_{ij}^{\text{train}} \mid X_{ij} \sim \text{Binomial}(X_{ij}, \epsilon), \quad X_{ij}^{\text{test}} = X_{ij} - X_{ij}^{\text{train}}$$

Then $X^{\text{train}} \perp X^{\text{test}}$ (they are **independent**!)

**The recipe**:

1. Split $X$ into $X^{\text{train}}$ and $X^{\text{test}}$ via Poisson thinning

2. Cluster using $X^{\text{train}}$ only

3. Test for DE using $X^{\text{test}}$ only

This controls Type I error because the test data is independent of the clustering.


## Count splitting controls Type I error

:::::: {.columns}
::: {.column width=.6}

\centerline{\includegraphics[height=.7\textheight]{img/neufeld_double_dipping-3.pdf}}

:::
::: {.column width=.35}

* QQ plots under null: double dipping (purple) inflates p-values

* Count splitting (orange) stays on the diagonal. 

* Count splitting across signal-to-noise ratios. 

${}$

\tiny
\flushright

[@Neufeld2023-fd]

:::
::::::

# Discussion: How should we test cell-cell communication?

## CellPhoneDB infers communication from clusters

:::::: {.columns}
::: {.column width=.4}

\centerline{\includegraphics[height=.6\textheight]{img/cellphonedb-pipeline.pdf}}

:::
::: {.column width=.6}

1. **Cluster** cells into discrete types (e.g., Leiden)

2. **Look up** known ligand-receptor (L-R) pairs from a curated database

3. For each L-R pair $(l, r)$ and each cell-type pair $(A, B)$:
   - Compute the **mean expression** of $l$ in $A$ and $r$ in $B$
   - Score $= \operatorname{mean}(\bar{x}_l^A,\, \bar{x}_r^B)$

4. **Permutation test**: randomly shuffle cell-type labels $\to$ null distribution of scores

5. Report pairs with $p < 0.05$

:::
::::::

\vfill

\footnotesize

[@Vento-Tormo2018-wn; @Armingol2021-tu; @Efremova2020-qy]

## CellPhoneDB output: significant L-R pairs

\centerline{\includegraphics[height=.7\textheight]{img/cellphonedb-output.pdf}}

\footnotesize

[@Vento-Tormo2018-wn]

## Discussion

\Large

Given what we learned today, what could go wrong with this approach?

\vspace{1em}

\normalsize

**Hint 1:** Clustering uses the same expression matrix as ...

\vspace{0.5em}

**Hint 2:** But L-R genes may not overlap with the genes that ...

## Can we fix this? Could count splitting help?

\Large

- Split counts: $X = X^{\text{train}} + X^{\text{test}}$

- Cluster on $X^{\text{train}}$, test L-R pairs on $X^{\text{test}}$

\vspace{1em}

\normalsize

But wait ...

- L-R scores need **mean expression** — do we have enough signal in $X^{\text{test}}$?

- Ligand in cluster $A$, receptor in cluster $B$ — we need counts from **different cells**, not the same cell split twice

- What if the clustering itself changes with only half the counts?


## Takeaways

**Graph-based clustering**

* Modularity optimization (Louvain/Leiden) is the workhorse of single-cell clustering

* The resolution parameter controls granularity — there is no single "correct" value

* SBMs offer a principled alternative with automatic model selection

**When clustering goes wrong**

* Clustering then testing on the same data = **double dipping** $\to$ inflated p-values

* **sc-SHC**: build significance testing into the clustering itself

* **Count splitting**: exploit Poisson thinning to create independent train/test sets

* **RECALL**: use artificial null genes to calibrate and detect over-clustering


# Appendix

## No perfect clustering algorithm

**Impossibility Theorem** [@Kleinberg2002-im]: no clustering function $f$ can simultaneously satisfy all three axioms:

1. **Scale invariance**: scaling all distances by $\alpha > 0$ doesn't change the clustering

2. **Richness**: every possible partition can be produced by some distance function

3. **Consistency**: shrinking within-cluster distances and expanding between-cluster distances doesn't change the clustering

Any algorithm must violate at least one:

\vfill

\centering
\small
\begin{tabular}{lccc}
\hline
\textbf{Method} & \textbf{Scale inv.} & \textbf{Richness} & \textbf{Consistency} \\
\hline
$k$-cluster (fixed $k$) & \checkmark & $\times$ & \checkmark \\
Distance threshold (fixed $r$) & $\times$ & \checkmark & \checkmark \\
Proportional ($\alpha \cdot d_{\max}$) & \checkmark & \checkmark & $\times$ \\
\hline
\end{tabular}

## Axiom 1: Scale invariance

\begin{center}
\begin{tikzpicture}[scale=0.9, every node/.style={circle, fill, inner sep=2pt}]

% Left cluster
\node[fill=teal] (a1) at (0,0) {};
\node[fill=teal] (a2) at (0.6,0.3) {};
\node[fill=teal] (a3) at (0.3,-0.4) {};

% Right cluster
\node[fill=magenta] (b1) at (3,0) {};
\node[fill=magenta] (b2) at (3.5,0.4) {};
\node[fill=magenta] (b3) at (3.2,-0.3) {};

\draw[gray] (a1)--(a2) node[midway,above,font=\scriptsize,fill=none] {1};
\draw[gray] (a1)--(b1) node[midway,above,font=\scriptsize,fill=none] {3};

\node[fill=none, font=\large] at (5,0) {$\xrightarrow{\times \alpha}$};

% Scaled version
\node[fill=teal] (c1) at (7,0) {};
\node[fill=teal] (c2) at (7.6,0.3) {};
\node[fill=teal] (c3) at (7.3,-0.4) {};

\node[fill=magenta] (d1) at (10,0) {};
\node[fill=magenta] (d2) at (10.5,0.4) {};
\node[fill=magenta] (d3) at (10.2,-0.3) {};

\draw[gray] (c1)--(c2) node[midway,above,font=\scriptsize,fill=none] {$\alpha$};
\draw[gray] (c1)--(d1) node[midway,above,font=\scriptsize,fill=none] {$3\alpha$};

\node[fill=none, font=\small] at (1.5,-1.5) {$f(d) = \{\text{\textcolor{teal}{A}}, \text{\textcolor{magenta}{B}}\}$};
\node[fill=none, font=\small] at (8.5,-1.5) {$f(\alpha \cdot d) = \{\text{\textcolor{teal}{A}}, \text{\textcolor{magenta}{B}}\}$};

\end{tikzpicture}
\end{center}

Multiplying all distances by $\alpha > 0$ should not change the partition.

## Axiom 2: Richness

\begin{center}
\begin{tikzpicture}[scale=0.9, every node/.style={circle, fill, inner sep=2pt}]

% Partition 1: {AB}{CD}
\node[fill=teal] (a1) at (0,0.3) {};
\node[fill=teal] (a2) at (0.7,0) {};
\node[fill=magenta] (a3) at (2.5,0.3) {};
\node[fill=magenta] (a4) at (3.2,0) {};

\draw[rounded corners, dashed, teal] (-0.4,-0.5) rectangle (1.1,0.8);
\draw[rounded corners, dashed, magenta] (2.1,-0.5) rectangle (3.6,0.8);

\node[fill=none, font=\scriptsize] at (1.6,-1) {$d_1 \to \{\text{\textcolor{teal}{AB}}, \text{\textcolor{magenta}{CD}}\}$};

% Partition 2: {AC}{BD}
\node[fill=orange] (b1) at (6,0.5) {};
\node[fill=blue] (b2) at (7.5,0) {};
\node[fill=orange] (b3) at (6,-0.3) {};
\node[fill=blue] (b4) at (7.5,0.7) {};

\draw[rounded corners, dashed, orange] (5.6,-0.7) rectangle (6.4,0.9);
\draw[rounded corners, dashed, blue] (7.1,-0.4) rectangle (7.9,1.1);

\node[fill=none, font=\scriptsize] at (6.9,-1) {$d_2 \to \{\text{\textcolor{orange}{AC}}, \text{\textcolor{blue}{BD}}\}$};

\end{tikzpicture}
\end{center}

For **any** partition of $n$ points, there exists some distance function $d$ that produces it.

## Axiom 3: Consistency

\begin{center}
\begin{tikzpicture}[scale=0.9, every node/.style={circle, fill, inner sep=2pt}]

% Before
\node[fill=teal] (a1) at (0,0) {};
\node[fill=teal] (a2) at (1,0.3) {};
\node[fill=magenta] (b1) at (3,0) {};
\node[fill=magenta] (b2) at (4,0.3) {};

\draw[teal, thick] (a1)--(a2) node[midway,above,font=\scriptsize,fill=none] {1};
\draw[gray] (a2)--(b1) node[midway,above,font=\scriptsize,fill=none] {2};
\draw[magenta, thick] (b1)--(b2) node[midway,above,font=\scriptsize,fill=none] {1};

\node[fill=none, font=\large] at (5.5,0) {$\to$};

% After: within shrinks, between grows
\node[fill=teal] (c1) at (7,0) {};
\node[fill=teal] (c2) at (7.5,0.15) {};
\node[fill=magenta] (d1) at (11,0) {};
\node[fill=magenta] (d2) at (11.5,0.15) {};

\draw[teal, thick] (c1)--(c2) node[midway,above,font=\scriptsize,fill=none] {0.5};
\draw[gray] (c2)--(d1) node[midway,above,font=\scriptsize,fill=none] {5};
\draw[magenta, thick] (d1)--(d2) node[midway,above,font=\scriptsize,fill=none] {0.5};

\node[fill=none, font=\scriptsize] at (2,-1) {$f(d) = \{\text{\textcolor{teal}{A}}, \text{\textcolor{magenta}{B}}\}$};
\node[fill=none, font=\scriptsize] at (9.2,-1) {$f(d') = \{\text{\textcolor{teal}{A}}, \text{\textcolor{magenta}{B}}\}$ ?};

\end{tikzpicture}
\end{center}

If we make clusters "tighter" and "farther apart," the same partition should remain.


## References {.allowframebreaks}


