---
title: "Single-cell RNA-seq, Technology and Quality Control"
author: "Yongjin Park"
date: "February 3, 2026"
classoption: "aspectratio=169"
fontsize: 12pt
bibliography: SingleCell.bib
output:
    beamer_presentation:
        theme: Madrid
        keep_md: true
        keep_tex: true
        latex_engine: pdflatex
        slide_level: 2
header-includes:
  - \usepackage{cancel}
  - \definecolor{UBCblue}{rgb}{0.04706, 0.13725, 0.26667}
  - \usecolortheme[named=UBCblue]{structure}
  - \setbeamertemplate{frametitle}{\color{UBCblue}\bfseries\insertframetitle\par\vskip-6pt\hrulefill}
  - \setbeamercolor{title in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{section in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{author in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{date in head/foot}{bg=white,fg=gray}
  - \setbeamertemplate{page number in head/foot}{}
  - \setbeamertemplate{frame numbering}[none]
  - \setbeamercolor{alerted text}{bg=yellow}
  - \AtBeginSection[]{\begin{frame}\frametitle{Today's lecture}{\Large\tableofcontents[currentsection]}\end{frame}}
  - |
    \makeatletter
    \def\ps@titlepage{%
      \setbeamertemplate{footline}{}
    }
    \addtobeamertemplate{title page}{\thispagestyle{titlepage}}{}
    \makeatother
    \include{toc}
---

<!--
Rules: 

1. After editing this file, run `make` to rebuild the slides: make lect08-single-cell-technology-slides.pdf

2. Use data-beans and senna (legume) Rust-based CLI tools (already installed)

3. Small data sets: we can show both Seurat and Scanpy how we can process

4. Do not mix R and python in the same slide/frame (it's confusing)

5. Prefer hdf5 backend since zarr takes too many files.

6. Prefer onslide rather than only in latex beamer

7. use \large for discussion (text-driven) slides then close it with \normalsize

8. Figure width never exceeds 6 and height never 3

9. Rasterize plots with too many dots with dpi=300 using `ggrastr`

-->


## Learning objectives

```{r setup, include=FALSE}
source("Setup.R")
fig.dir <- "../Fig/single-cell-technology/"
dir.create(fig.dir, showWarnings=FALSE)
setup.env(fig.dir)
dir.create("../data", showWarnings=FALSE)

library(data.table)
library(tidyverse)
library(patchwork)
library(matrixTests)

theme_set(theme_minimal())
```

\large

* Introduction to single-cell RNA-seq technology

* Understand single-cell RNA-seq data structure

* Introduction to analysis workflow
  * Data quality control
  * Exploratory data analysis


# Single-cell sequencing technology

## Droplet-based single-cell sequencing technology

\vfill
\centerline{\includegraphics[width=\textwidth]{img/dropseq_technology/dropseq1.pdf}}
\vfill
\flushright
\small
[@Macosko2015-vi]

## Idea 1: Capture one cell with a microbead in a droplet

\vfill
\centerline{\includegraphics[height=.7\textheight]{img/dropseq_technology/dropseq2.pdf}}
\vfill
\flushright
\small
[@Macosko2015-vi]

## Idea 2: Massively-parallel sequencing followed by cell-specific barcoding

\vfill
\centerline{\includegraphics[width=\textwidth]{img/dropseq_technology/dropseq3.pdf}}
\vfill
\flushright
\small
[@Macosko2015-vi]


## Idea 3: How do we keep track of mRNA short reads' membership to a certain droplet?

\vfill
\centerline{\includegraphics[width=.9\textwidth]{img/dropseq_technology/dropseq4.pdf}}
\vfill
\flushright
\small
[@Macosko2015-vi]

## How do we construct millions of unique barcodes? Use DNA as a hashing function!

\vfill

\only<1>{
\includegraphics[height=.5\textheight]{img/dropseq_technology/dropseq5.pdf}
}

\only<2>{
\includegraphics[height=.5\textheight]{img/dropseq_technology/dropseq6.pdf}
}

\only<3>{
\includegraphics[height=.55\textheight]{img/dropseq_technology/dropseq7.pdf}
}

\vfill

\flushright
\small
[@Macosko2015-vi]

## The lengths of barcode sequences determine data dimensionality

\vfill

:::::: {.columns}
::: {.column width=.75}

\centerline{\includegraphics[width=\linewidth]{img/dropseq_technology/dropseq8.pdf}}

:::
::: {.column width=.2}

\begin{block}{}
Technically, we can build up to a `r num.int(4^8)` $\times$ `r num.int(4^12)`, gene $\times$ cell expression matrix in one single-cell RNA-seq experiment.
\end{block}

:::
::::::

\vfill
\small
[@Macosko2015-vi]


## Droplet-based single-cell sequencing pipeline

\vfill
\centerline{\includegraphics[width=.8\textwidth]{img/dropseq_technology/dropseq9.pdf}}
\vfill
\flushright
\small
[@Macosko2015-vi]

## Discussions: What are the potential challenges?

\large

* What are the same challenges we had in bulk RNA-seq?

* What are the new challenges?

\normalsize


## Overview of single-cell data analysis

\centerline{\includegraphics[width=\textwidth]{img/scRNA_overview/Karchenko_1.pdf}}

\vfill

\flushright
\small
[@Kharchenko2021-wr]

## Overview of single-cell data analysis cont'd

\centerline{\includegraphics[width=\textwidth]{img/scRNA_overview/Karchenko_2.pdf}}

\vfill

\flushright
\small
[@Kharchenko2021-wr]


# Basic quality control

## Example: Comprehensive immune profiling in pancreatic cancer

\vfill

:::::: {.columns}
::: {.column width=.6}

Multimodal mapping of the tumor and peripheral blood immune landscape in human pancreatic cancer

**Approach:**

* **Disease model:** Pancreatic ductal adenocarcinoma (PDA)
* **Technologies:** CyTOF, \textbf{\color{magenta} single-cell RNA sequencing}, multiplex immunohistochemistry
* **Sample diversity:** Tumors, matched blood, and healthy tissues
* **Goal:** Comprehensive immune characterization

:::
::: {.column width=.35}

\begin{block}{Key discoveries}
\begin{itemize}
\item Revealed complex cellular immune networks
\item Mapped patient-specific checkpoint receptor patterns
\item Identified CD8$^+$ T cell states and transitions
\item Discovered TIGIT as potential therapeutic target
\end{itemize}
\end{block}

:::
::::::

\vfill

\flushright
\small
[@Steele2020-tq]

## Dataset overview: 41 samples

\vfill

:::::: {.columns}
::: {.column width=.5}

* 16 PDAC tissue
* 3 Adjacent normal tissue
* 16 PDAC PBMC
* 4 Healthy PBMC

\vspace{0.5cm}

**Platform:** 10x Genomics Chromium

:::
::: {.column width=.45}

\small
```
../data/
  GSE155698_PDAC_Steele2020/
```
\normalsize

**GEO:** GSE155698

:::
::::::

\vfill

## Sample directories

\small

:::::: {.columns}
::: {.column width=.32}

**PDAC Tissue (16)**
```
PDAC_TISSUE_1
PDAC_TISSUE_2
...
PDAC_TISSUE_16
```

:::
::: {.column width=.32}

**PDAC PBMC (16)**
```
PDAC_PBMC_1
PDAC_PBMC_2
...
PDAC_PBMC_16
```

:::
::: {.column width=.32}

**Controls**
```
AdjNorm_TISSUE_1
AdjNorm_TISSUE_2
AdjNorm_TISSUE_3

Healthy_PBMC_1
Healthy_PBMC_2
Healthy_PBMC_3
Healthy_PBMC_4
```

:::
::::::

\normalsize

## 10x Genomics data format

\vfill

:::::: {.columns}
::: {.column width=.5}

\small
```bash
PDAC_TISSUE_1/
|-- barcodes.tsv.gz   (9 KB)
|-- features.tsv.gz   (303 KB)
+-- matrix.mtx.gz     (17 MB)
```
\normalsize

\vspace{0.5cm}

* **`barcodes.tsv.gz`** - Cell IDs
* **`features.tsv.gz`** - Genes
* **`matrix.mtx.gz`** - Counts (sparse)

:::
::: {.column width=.45}

**Matrix Market format**

* Sparse matrix standard
* Memory efficient
* Compatible with Seurat, Scanpy

\vspace{0.5cm}

**Why sparse?** Most genes are not expressed in most cells

:::
::::::

\vfill

## Loading 10x data with Seurat (R) \hfill \includegraphics[height=1em]{img/logos/seurat_logo.jpg}

```{r load-seurat, message=FALSE, warning=FALSE}
library(Seurat)

data.dir <- "../data/GSE155698_PDAC_Steele2020/PDAC_TISSUE_1/"

pdac <- Read10X(data.dir = data.dir)
pdac <- CreateSeuratObject(counts = pdac,
                           project = "PDAC_TISSUE_1",
                           min.cells = 3,
                           min.features = 200)
pdac
```

## Seurat object structure

* `SeuratObject` class stores everything
* Genes in rows, cells in columns
* Access counts: `GetAssayData(pdac)`
* Cell metadata: `pdac@meta.data`

```{r seurat-cell-metadata, message=FALSE}
head(pdac@meta.data, 3)
```

```{r seurat-gene-metadata, message=FALSE}
head(rownames(pdac), 5)
```

## Loading 10x data with scanpy (Python) \hfill \includegraphics[height=1em]{img/logos/scanpy_logo.png}

```{python load-scanpy}
import scanpy as sc

data_dir = "../data/GSE155698_PDAC_Steele2020/PDAC_TISSUE_1/"

adata = sc.read_10x_mtx(data_dir, var_names='gene_symbols', cache=True)
adata
```

## AnnData object structure

* `AnnData` class stores everything
* Cells in rows, genes in columns (transposed from Seurat!)
* Access counts: `adata.X`
* Cell metadata: `adata.obs`, gene metadata: `adata.var`

```{python scanpy-cell-metadata}
print(adata.obs_names[:5].tolist())
```

```{python scanpy-gene-metadata}
print(adata.var_names[:5].tolist())
```

## Cell QC metrics with Seurat

```{r cell-qc-metrics, message=FALSE}
# Calculate mitochondrial gene percentage
pdac[["percent.mt"]] <- PercentageFeatureSet(pdac, pattern = "^MT-")

# View QC metrics
head(pdac@meta.data)
```

* `nCount_RNA` = total UMI counts per cell
* `nFeature_RNA` = number of genes detected
* `percent.mt` = % mitochondrial reads (high = dying cells)

## Why high mitochondrial?

\large

\begin{itemize}
\item MT genes: 13 protein-coding genes
\item But 100s--1000s of mitochondria per cell
\item MT RNA is abundant and stable
\end{itemize}

\onslide<2->{
\textbf{Proportion effect:}

Cytoplasmic RNA degrades $\to$ MT\% goes up
}

## Why does high MT implicate dying cells?

\large

\textbf{Conventional wisdom:}

Membrane breaks $\to$ cytoplasmic RNA leaks $\to$ MT RNA stays

\onslide<2->{
\textbf{But...}

\begin{itemize}
\item Why wouldn't mitochondria also be damaged?
\item Other organelles have membranes too (ER, nucleus)
\end{itemize}
}

\onslide<3->{
High MT\% correlates with low quality --- mechanism unclear
}

\small
[@Yates2025-sd]

## Seurat: QC visualization commands

```{r seurat-qc-commands, eval=FALSE}
# Violin plots
VlnPlot(pdac, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"))

# Scatter plots
FeatureScatter(pdac, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
FeatureScatter(pdac, feature1 = "nCount_RNA", feature2 = "percent.mt")
```

## Visualizing QC metrics

:::::: {.columns}
::: {.column width=.32}

```{r qc-vln-nfeature, echo=FALSE, fig.width=2, fig.height=2.5, onslide.plot="1-"}
VlnPlot(pdac, features = "nFeature_RNA", pt.size = 0, cols = "steelblue") +
    NoLegend() +
    ggtitle("nFeature_RNA") +
    theme(plot.title = element_text(size = 10)) +
    theme(axis.text.x = element_blank(), axis.title.x = element_blank())
```

:::
::: {.column width=.32}

```{r qc-vln-ncount, echo=FALSE, fig.width=2, fig.height=2.5, onslide.plot="2-"}
VlnPlot(pdac, features = "nCount_RNA", pt.size = 0, cols = "steelblue") +
    NoLegend() +
    ggtitle("nCount_RNA") +
    theme(plot.title = element_text(size = 10)) +
    theme(axis.text.x = element_blank(), axis.title.x = element_blank())
```

:::
::: {.column width=.32}

```{r qc-vln-mt, echo=FALSE, fig.width=2, fig.height=2.5, onslide.plot="3-"}
VlnPlot(pdac, features = "percent.mt", pt.size = 0, cols = "steelblue") +
    NoLegend() +
    ggtitle("percent.mt") +
    theme(plot.title = element_text(size = 10)) +
    theme(axis.text.x = element_blank(), axis.title.x = element_blank())
```

:::
::::::

## QC: Scatter plots

:::::: {.columns}
::: {.column width=.48}

```{r qc-scatter-1, echo=FALSE, fig.width=3, fig.height=2.5, onslide.plot="1-"}
FeatureScatter(pdac, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", pt.size = 0.3, cols = "steelblue") +
    NoLegend() +
    theme(plot.title = element_blank())
```

\onslide<1->{
\begin{itemize}
\item Positive correlation expected
\item Low-low $\to$ empty/dying
\end{itemize}
}

:::
::: {.column width=.48}

```{r qc-scatter-2, echo=FALSE, fig.width=3, fig.height=2.5, onslide.plot="2-"}
FeatureScatter(pdac, feature1 = "nCount_RNA", feature2 = "percent.mt", pt.size = 0.3, cols = "steelblue") +
    NoLegend() +
    theme(plot.title = element_blank())
```

\onslide<2->{
\begin{itemize}
\item High MT + low count $\to$ dying
\item High MT + high count $\to$ keep?
\end{itemize}
}

:::
::::::

## Cell QC metrics with scanpy \hfill \includegraphics[height=1em]{img/logos/scanpy_logo.png}

```{python scanpy-qc-metrics}
# Calculate QC metrics (mitochondrial genes start with "MT-")
adata.var['mt'] = adata.var_names.str.startswith('MT-')
sc.pp.calculate_qc_metrics(adata, qc_vars=['mt'], inplace=True)
adata.obs.head()[['n_genes_by_counts', 'total_counts', 'pct_counts_mt']]
```

## scanpy: QC visualization commands \hfill \includegraphics[height=1em]{img/logos/scanpy_logo.png}

```{python scanpy-qc-commands, eval=FALSE}
# Violin plots
sc.pl.violin(adata, ['n_genes_by_counts', 'total_counts', 'pct_counts_mt'])

# Scatter plots
sc.pl.scatter(adata, x='total_counts', y='n_genes_by_counts')
sc.pl.scatter(adata, x='total_counts', y='pct_counts_mt')
```

## Visualizing QC metrics (scanpy) \hfill \includegraphics[height=1em]{img/logos/scanpy_logo.png}

```{python scanpy-qc-violin, echo=FALSE, dev='png', fig.width=6, fig.height=2, results='hide'}
import matplotlib.pyplot as plt
plt.rcParams.update({'font.size': 8, 'axes.titlesize': 8, 'axes.labelsize': 7, 'xtick.labelsize': 6, 'ytick.labelsize': 6})
fig, axes = plt.subplots(1, 3, figsize=(6, 2))
sc.pl.violin(adata, 'n_genes_by_counts', ax=axes[0], show=False)
sc.pl.violin(adata, 'total_counts', ax=axes[1], show=False)
sc.pl.violin(adata, 'pct_counts_mt', ax=axes[2], show=False)
plt.tight_layout()
plt.show()
```

## QC: Scatter plots (scanpy) \hfill \includegraphics[height=1em]{img/logos/scanpy_logo.png}

```{python scanpy-qc-scatter, echo=FALSE, dev='png', fig.width=5, fig.height=2, results='hide'}
fig, axes = plt.subplots(1, 2, figsize=(5, 2))
sc.pl.scatter(adata, x='total_counts', y='n_genes_by_counts', ax=axes[0], show=False)
sc.pl.scatter(adata, x='total_counts', y='pct_counts_mt', ax=axes[1], show=False)
plt.tight_layout()
plt.show()
```

```{r reset-dev, include=FALSE}
knitr::opts_chunk$set(dev = "cairo_pdf")
```

```{bash compute-cell-stats, include=FALSE}
DATA_DIR="../data/GSE155698_PDAC_Steele2020"
H5_FILES=$(find ${DATA_DIR} -maxdepth 2 -name "*.h5" | grep -v "TISSUE_14" | grep -v "TISSUE_16" | sort)

[ -f cell_stats.parquet ] || echo ${H5_FILES} | xargs data-beans stat -s column -o cell_stats.parquet
[ -f cell_stats_mt.parquet ] || echo ${H5_FILES} | xargs data-beans stat -s column -r "_MT-" -o cell_stats_mt.parquet
[ -f gene_stats.parquet ] || echo ${H5_FILES} | xargs data-beans stat -s row -o gene_stats.parquet
```

## Overview: where are we?

\centerline{\includegraphics[width=\textwidth]{img/scRNA_overview/Karchenko_1.pdf}}

\vfill

\flushright
\small
[@Kharchenko2021-wr]

## Cell stats: total counts vs. MT%

```{r load-cell-stats, include=F}
library(arrow)
library(ggrastr)

# Load cell stats and MT stats
cell_stats <- as.data.table(read_parquet("cell_stats.parquet"))
mt_stats <- as.data.table(read_parquet("cell_stats_mt.parquet"))
gene_stats <- as.data.table(read_parquet("gene_stats.parquet"))

# Merge and compute pct_mt
merged <- merge(cell_stats, mt_stats, by = "row", suffixes = c("", "_mt"))
merged[, pct_mt := 100 * tot_mt / tot]
```

We have **`r scales::comma(nrow(cell_stats))`** cells across 38 samples.

```{r cell-stats-scatter-tot, echo=FALSE, fig.width=5, fig.height=2.8}
ggplot(merged[sample(nrow(merged), min(10000, nrow(merged)))],
       aes(x = log10(tot), y = pct_mt)) +
    rasterise(geom_point(alpha = 0.3, size = 0.5, color = "steelblue"), dpi = 300) +
    labs(x = "log10(total counts)", y = "% MT")
```


## Cell stats: # genes expressed vs. MT%

```{r cell-stats-scatter, echo=FALSE, fig.width=5, fig.height=3}
ggplot(merged[sample(nrow(merged), min(10000, nrow(merged)))],
       aes(x = log10(nnz), y = pct_mt)) +
    rasterise(geom_point(alpha = 0.3, size = 0.5, color = "steelblue"), dpi = 300) +
    labs(x = "log10(# genes expressed)", y = "% MT")
```

## Cell stats: 2D density

```{r cell-stats-density, echo=FALSE, fig.width=5, fig.height=3}
ggplot(merged, aes(x = log10(nnz), y = pct_mt)) +
    stat_density_2d(geom = "raster",
                    aes(fill = after_stat(sqrt(density))),
                    contour = FALSE) +
    scale_fill_viridis_c() +
    labs(x = "log10(# genes expressed)", y = "% MT")
```

## Mean-variance relationship: cells

```{r mean-var-cell, echo=FALSE, fig.width=5, fig.height=3}
ggplot(cell_stats[sample(nrow(cell_stats), min(10000, nrow(cell_stats)))],
       aes(x = log10(mu), y = log10(sig^2))) +
    rasterise(geom_point(alpha = 0.3, size = 0.5, color = "steelblue"), dpi = 300) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    labs(x = "log10(mean)", y = "log10(variance)")
```

## Mean-variance relationship: cells (density)

```{r mean-var-cell-density, echo=FALSE, fig.width=5, fig.height=3}
ggplot(cell_stats, aes(x = log10(mu), y = log10(sig^2))) +
    stat_density_2d(geom = "raster",
                    aes(fill = after_stat(sqrt(density))),
                    contour = FALSE) +
    scale_fill_viridis_c() +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    labs(x = "log10(mean)", y = "log10(variance)")
```

## Mean-variance relationship: genes

```{r mean-var-gene, echo=FALSE, fig.width=5, fig.height=3, dev="png", dpi=300}
ggplot(gene_stats, aes(x = log10(mu), y = log10(sig^2))) +
    geom_point(alpha = 0.3, size = 0.5, color = "steelblue") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    labs(x = "log10(mean)", y = "log10(variance)")
```

## Mean-variance relationship: genes (density)

```{r mean-var-gene-density, echo=FALSE, fig.width=5, fig.height=3, dev="png", dpi=300}
ggplot(gene_stats, aes(x = log10(mu), y = log10(sig^2))) +
    stat_density_2d(geom = "raster",
                    aes(fill = after_stat(sqrt(density))),
                    contour = FALSE) +
    scale_fill_viridis_c() +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    labs(x = "log10(mean)", y = "log10(variance)")
```

## Can we use the count data as they are?

\large

* In bulk RNA-seq, we transformed the count data to adjust size factors that vary across samples.

* In single-cell data analysis, we generally care a lot about zero values, not knowing whether they come from actual "no expression" or technical drop-out.

* Should we take some sort of `voom` transformation?

\normalsize

## Transformation methods for scRNA-seq

\vfill

\large

:::::: {.columns}
::: {.column width=.25}

**Raw counts**

$$Y$$

:::
::: {.column width=.25}

**Delta method**

$$\log(Y/s + 1)$$

:::
::: {.column width=.25}

**GLM residual**

$$\frac{Y - \mu}{\sqrt{\mu + \alpha\mu^2}}$$

:::
::: {.column width=.25}

**Latent expression**

$$Y \sim \text{Poisson}(M)$$
$$M \sim \text{logNormal}(\mu, \sigma^2)$$

:::
::::::

\vfill

\small
[@Ahlmann-Eltze2023-lq]

## Effect of transformation on PCA

\vfill

\centerline{\includegraphics[width=\textwidth]{img/scRNA_transformation/Ahlmann_Huber_2023-2.pdf}}

\vfill

\small
[@Ahlmann-Eltze2023-lq]

## Mean-variance relationship after transformation

\vfill

\centerline{\includegraphics[width=\textwidth]{img/scRNA_transformation/Ahlmann_Huber_2023-3.pdf}}

\vfill

\small
[@Ahlmann-Eltze2023-lq]

## Gene expression distribution after transformation

\vfill

\centerline{\includegraphics[width=\textwidth]{img/scRNA_transformation/Ahlmann_Huber_2023-4.pdf}}

\vfill

\small
[@Ahlmann-Eltze2023-lq]

## What would happen after cell Q/C?

```{r seurat-umap-noqc, include=FALSE, cache=TRUE}
# Process without QC filtering
pdac <- pdac |>
    NormalizeData() |>
    FindVariableFeatures(nfeatures = 2000) |>
    ScaleData() |>
    RunPCA(npcs = 30) |>
    RunUMAP(dims = 1:20)
```

:::::: {.columns}
::: {.column width=.5}

```{r eval=F, size="large"}
pdac <- pdac |>
    NormalizeData() |>
    FindVariableFeatures() |>
    ScaleData() |>
    RunPCA() |>
    RunUMAP(dims = 1:20)

DimPlot(pdac,
        reduction = "umap")
```

:::
::: {.column width=.5}

```{r seurat-umap-plot, echo=FALSE, fig.width=3, fig.height=3}
DimPlot(pdac, reduction = "umap", pt.size = 0.5) +
    NoLegend()
```

:::
::::::

## What would happen after Q/C? Let's see..

```{r umap-qc-metrics, echo=FALSE, message=FALSE, fig.width=6, fig.height=2.8}
p1 <- FeaturePlot(pdac, features = "percent.mt", pt.size = 0.3) +
    scale_color_viridis_c() +
    labs(title = "% MT")
p2 <- FeaturePlot(pdac, features = "nCount_RNA", pt.size = 0.3) +
    scale_color_viridis_c(trans = "log10") +
    labs(title = "Total counts (log10)")
p1 | p2
```


## It's useful to show high-dimensional data in 2D

\vfill

:::::: {.columns}
::: {.column width=.45}

\onslide<1->{
\includegraphics[width=.8\linewidth]{img/scRNA_data_cartoon.pdf}
}

:::
::: {.column width=.45}

\onslide<2->{
\includegraphics[width=.7\linewidth]{img/scRNA_data_cartoon_tSNE.pdf}
}

:::
::::::

\vfill

\onslide<2->{
\small
\texttt{tSNE}: t-distributed Stochastic Neighbourhood Embedding (Van der Maaten \& Hinton, 2008).
}

## Warning: Don't over interpret 2D embedding

\vfill
\centerline{\includegraphics[width=\textwidth]{img/scRNA_ETM/tSNE_over.pdf}}
\vfill

\small

[@Kobak2021-yu]

## Don't over-interpret 2D embedding results

Check out these papers:

* [Initialization is critical for preserving global data structure in both t-SNE and UMAP](https://www.nature.com/articles/s41587-020-00809-z)

* [The art of using t-SNE for single-cell transcriptomics](https://www.nature.com/articles/s41467-019-13056-x)

* [Dimensionality reduction for visualizing single-cell data using UMAP](https://www.nature.com/articles/nbt.4314)


## What if we capture more than one cell in a droplet?

\vfill
\centerline{\includegraphics[width=.8\textwidth]{img/dropseq_technology/dropseq9.pdf}}
\vfill

\flushright
\small
[@Macosko2015-vi]

## What is a doublet in single-cell data?

**Biological/technical definition:**

- One or more cells captured (usually at most two cells by chance)

- Thus, multiple cells accidentally share the same cell barcode sequence

- Not so clear in general... since we missed the chance to assign
different tags to different cells encapsulated in the same droplet.

**Statistical definition**:

- If marker genes of multiple cell types are
simultaneously expressed...

- An unvetted approach: Find ambiguous/intermediate coordinates in
PCA/tSNE/UMAP (after removing ambient cells).

## Idea: k-Nearest Neighbour classification for doublet detection

\vfill

\centerline{\includegraphics[width=.9\textwidth]{img/doublet/doublet_knn.pdf}}

\vfill
\small



## Can we create artificial doublets?

\large

A straightforward definition (used in [`DoubletFinder`](https://github.com/chris-mcginnis-ucsf/DoubletFinder)):

For each cell $i$:

* Take some other $j$ by random selection

* Create an artificial doublet

\centerline{$\tilde{\mathbf{x}} \gets \frac{1}{2}(\mathbf{x}_{i} + \mathbf{x}_{j})$}

Some thought questions:

* Doublets within the same cell type?

* Doublets between the different cell types?

## DoubletFinder on PDAC_TISSUE_1

```{r doubletfinder, include=FALSE, cache=TRUE}
library(DoubletFinder)

# Estimate doublet rate (~5 per 1000 cells)
nExp_poi <- round(.05 * ncol(pdac)^2 / 1000)

# Find optimal pK
sweep.res <- paramSweep(pdac, PCs = 1:20, sct = FALSE)
sweep.stats <- summarizeSweep(sweep.res, GT = FALSE)
bcmvn <- find.pK(sweep.stats)
pK <- as.numeric(as.character(bcmvn$pK[which.max(bcmvn$BCmetric)]))

# Run DoubletFinder
pdac <- doubletFinder(pdac, PCs = 1:20, pN = 0.25, pK = pK,
                      nExp = nExp_poi, sct = FALSE)
```

:::::: {.columns}
::: {.column width=.3}

```{r eval=F, size="normalsize"}
library(DoubletFinder)

sweep.res <- paramSweep(
    pdac, PCs = 1:20)
sweep.stats <- summarizeSweep(
    sweep.res, GT = FALSE)
bcmvn <- find.pK(sweep.stats)

pdac <- doubletFinder(
    pdac, PCs = 1:20,
    pN = 0.25, pK = pK,
    nExp = nExp_poi)
```

:::
::: {.column width=.6}

```{r doubletfinder-plot, echo=FALSE, fig.width=3.5, fig.height=3}
# Get the DoubletFinder classification column
df_col <- grep("DF.classifications", colnames(pdac@meta.data), value = TRUE)[1]
DimPlot(pdac, reduction = "umap", group.by = df_col, pt.size = 0.5,
        cols = c("Singlet" = "steelblue", "Doublet" = "red")) +
    labs(title = "DoubletFinder") +
    theme(legend.position = "bottom")
```

:::
::::::


<!--------------------->
<!-- dont edit below -->
<!--------------------->

# Data normalization across many batches

## Where are we?

\centerline{\includegraphics[width=.85\textwidth]{img/scRNA_overview/Karchenko_1.pdf}}

\centerline{\includegraphics[width=.85\textwidth]{img/scRNA_overview/Karchenko_2.pdf}}

\vfill

\flushright
\small
[@Kharchenko2021-wr]

## Why batch normalization?

\large

- Single-cell experiments are often performed across **multiple samples, batches, or technologies**

- Technical variation can cause cells to cluster by **batch** rather than **cell type**

- Without correction, biological signals are confounded with technical artifacts

- **Goal**: Remove technical variation while preserving biological heterogeneity

\normalsize

## Batch normalization aims to minimize the difference between nearest cells across different batches

\vfill

:::::: {.columns}
::: {.column width=.45}
\only<1-2>{\centerline{\includegraphics[width=.85\linewidth]{img/scRNA_batch_normalization/mutual_neighbour1.pdf}}}

:::
::: {.column width=.45}

\only<2>{\centerline{\includegraphics[width=.85\linewidth]{img/scRNA_batch_normalization/mutual_neighbour2.pdf}}}

:::
::::::

\only<3>{\centerline{\includegraphics[width=.7\linewidth]{img/scRNA_batch_normalization/mutual_neighbour3.pdf}}}

:::::: {.columns}
::: {.column width=.45}

\only<4>{
\centerline{\includegraphics[width=.85\linewidth]{img/scRNA_batch_normalization/mutual_neighbour4.pdf}}
}

\only<5>{
\centerline{\includegraphics[width=.85\linewidth]{img/scRNA_batch_normalization/mutual_neighbour5.pdf}}
}

:::
::: {.column width=.45}

\only<4-5>{
What is the gap $\Delta$ between the batches?

\centerline{$\min_{\Delta} \sum_{a,b} W_{ab} \| \mathbf{x}_{a} - \mathbf{x}_{b} - \Delta\|_{2}$}
}

\only<4>{
A key assumption:
$$0 \approx W_{ac} < W_{ab}$$
}

\only<5>{
Fixed point (local) optimal solution:
$$\Delta \gets \frac{\sum_{a,b} W_{ab}(\mathbf{x}_{a} - \mathbf{x}_{b})}{\sum_{b} W_{ab}}$$
}

:::
::::::

\vfill
\flushright
\small
[@Haghverdi2016-jp]

## A batch-balancing k-nearest neighbour graph

BBKNN method strikes balance between over- and under-normalization

\vfill

:::::: {.columns}
::: {.column width=.75}

\only<1>{\includegraphics[width=\linewidth]{img/scRNA_batch_normalization/bbknn1.pdf}}

:::
::: {.column width=.2}

\only<1>{
What kind of differences are due to inter-batch, technical discrepancy, not inter-cell-type divergence?
}

:::
::::::

\only<2>{\centerline{\includegraphics[width=.8\linewidth]{img/scRNA_batch_normalization/bbknn2.pdf}}}
\only<3>{\centerline{\includegraphics[width=.8\linewidth]{img/scRNA_batch_normalization/bbknn3.pdf}}}

\vfill
\small
[@Polanski2019-hz]



##

\vfill

:::::: {.columns}
::: {.column width=.75}

\centerline{
\includegraphics[width=\linewidth]{img/scRNA_batch_normalization/scanorama.pdf}
}

:::
::: {.column width=.2}

$${}$$

How do we integrate multiple samples/batches?

$${}$$

**Scanorama**: mutual nearest neighbourhood-based data integration

:::
::::::

\vfill
\small
[@Hie2019-fh]

## Canonical Correlation Analysis (CCA)

\large

- CCA finds linear combinations of features that are **maximally correlated** between two datasets

- Given two data matrices $X_1$ and $X_2$, find vectors $a$ and $b$ such that:
$$\text{corr}(X_1 a, X_2 b) \text{ is maximized}$$

- The projected values $X_1 a$ and $X_2 b$ are called **canonical variates**

- Multiple canonical variates can be found (orthogonal to previous ones)

\normalsize

## CCA for batch integration

\large

**Intuition**: Cells of the same type should have similar gene expression patterns, regardless of batch

\onslide<2->

**Seurat CCA integration**:

1. Run CCA on each pair of datasets to find shared correlation structure

2. Identify **anchor cells** --- pairs of cells (one from each batch) that are mutual nearest neighbours in CCA space

3. Use anchors to compute correction vectors that align the datasets

\normalsize

## Batch normalization with Seurat

```{r load-batch-helpers, include=FALSE}
source("batch_normalization.R")
```

```{r load-multiple-samples, include=FALSE, cache=TRUE}
samples <- c("PDAC_TISSUE_1", "PDAC_TISSUE_2", "PDAC_TISSUE_3")
data_dir <- "../data/GSE155698_PDAC_Steele2020"

seu_list <- load_pdac_samples(samples, data_dir)
seu_list <- preprocess_samples(seu_list)
```

```{r merge-without-integration, include=FALSE, cache=TRUE}
seu_merged <- merge_without_integration(seu_list)
```

:::::: {.columns}
::: {.column width=.5}



```{r umap-before-integration, echo=FALSE, fig.width=3, fig.height=3}
ggrastr::rasterise(
    DimPlot(seu_merged, reduction = "umap", group.by = "orig.ident",
            pt.size = 0.1, alpha = 0.3) +
        labs(title = "Merged (no integration)") +
        theme(text = element_text(size = 8),
              legend.text = element_text(size = 6),
              legend.title = element_blank(),
              legend.key.size = unit(0.3, "cm"),
              legend.margin = margin(0, 0, 0, 0),
              plot.title = element_text(size = 9)),
    dpi = 300)
```

:::
::: {.column width=.5}

```{r seurat-integration, include=FALSE, cache=TRUE}
seu_integrated <- integrate_seurat_cca(seu_list)
```



```{r umap-after-integration, echo=FALSE, fig.width=3, fig.height=3}
ggrastr::rasterise(
    DimPlot(seu_integrated, reduction = "umap", group.by = "orig.ident",
            pt.size = 0.1, alpha = 0.3) +
        labs(title = "Seurat CCA integration") +
        theme(text = element_text(size = 8),
              legend.text = element_text(size = 6),
              legend.title = element_blank(),
              legend.key.size = unit(0.3, "cm"),
              legend.margin = margin(0, 0, 0, 0),
              plot.title = element_text(size = 9)),
    dpi = 300)
```

:::
::::::

## Seurat integration workflow

```{r eval=FALSE}
# 1. Preprocess each sample separately
seu_list <- lapply(seu_list, function(seu) {
    seu |> NormalizeData() |> FindVariableFeatures()
})

# 2. Find integration anchors
features <- SelectIntegrationFeatures(seu_list)
anchors <- FindIntegrationAnchors(seu_list, anchor.features = features)

# 3. Integrate data
seu_integrated <- IntegrateData(anchors)

# 4. Standard workflow on integrated data
seu_integrated <- seu_integrated |>
    ScaleData() |> RunPCA() |> RunUMAP(dims = 1:20)
```

## Harmony algorithm

\vfill

\centerline{
\onslide<1->{\includegraphics[height=.7\textheight]{img/scRNA_batch_normalization/harmony1.pdf}}
\onslide<2->{\includegraphics[height=.7\textheight]{img/scRNA_batch_normalization/harmony2.pdf}}
\onslide<3->{\includegraphics[height=.7\textheight]{img/scRNA_batch_normalization/harmony3.pdf}}
\onslide<4->{\includegraphics[height=.7\textheight]{img/scRNA_batch_normalization/harmony4.pdf}}
}

\vfill
\small
[@Korsunsky2019-vv]

## Harmony: a fast alternative

```{r harmony-integration, include=FALSE, cache=TRUE}
library(harmony)
seu_harmony <- seu_merged |>
    RunHarmony("orig.ident") |>
    RunUMAP(reduction = "harmony", dims = 1:20)
```

:::::: {.columns}
::: {.column width=.4}

```{r eval=FALSE}
library(harmony)

seu_harmony <- seu_merged |>
    RunHarmony("orig.ident") |>
    RunUMAP(reduction = "harmony",
            dims = 1:20)
```

:::
::: {.column width=.6}

```{r umap-harmony, echo=FALSE, fig.width=3.5, fig.height=3}
ggrastr::rasterise(
    DimPlot(seu_harmony, reduction = "umap", group.by = "orig.ident",
            pt.size = 0.1, alpha = 0.3) +
        labs(title = "Harmony integration") +
        theme(text = element_text(size = 8),
              legend.text = element_text(size = 6),
              legend.title = element_blank(),
              legend.key.size = unit(0.3, "cm"),
              legend.margin = margin(0, 0, 0, 0),
              plot.title = element_text(size = 9)),
    dpi = 300)
```

:::
::::::

##

<!--------------------->
<!-- intential blank -->
<!--------------------->


## Reference {.allowframebreaks}
