---
title: "Matrix Factorization with applications in single-cell RNA-seq"
author: "Yongjin Park"
date: "`r format(Sys.time(), '%d %B %Y')`"
classoption: "aspectratio=169"
fontsize: 12pt
bibliography: singleCellFactorization.bib
output:
    beamer_presentation:
        theme: Madrid
        keep_md: true
        keep_tex: true
        latex_engine: pdflatex
        slide_level: 2
header-includes:
  - \usepackage{cancel}
  - \usepackage{tikz}
  - \usetikzlibrary{fit}
  - \definecolor{UBCblue}{rgb}{0.04706, 0.13725, 0.26667}
  - \usecolortheme[named=UBCblue]{structure}
  - \setbeamertemplate{frametitle}{\color{UBCblue}\bfseries\insertframetitle\par\vskip-6pt\hrulefill}
  - \setbeamercolor{title in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{section in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{author in head/foot}{bg=white,fg=gray}
  - \setbeamercolor{date in head/foot}{bg=white,fg=gray}
  - \setbeamertemplate{page number in head/foot}{}
  - \setbeamertemplate{frame numbering}[none]
  - \setbeamercolor{alerted text}{bg=yellow}
  - \AtBeginSection[]{\begin{frame}\frametitle{Today's lecture}{\Large\tableofcontents[currentsection]}\end{frame}}
  - |
    \makeatletter
    \def\ps@titlepage{%
      \setbeamertemplate{footline}{}
    }
    \addtobeamertemplate{title page}{\thispagestyle{titlepage}}{}
    \makeatother
    \include{toc}
---

<!--
Rules: 

1. After editing this file, run `make` to rebuild the slides: make lect08-single-cell-technology-slides.pdf

2. Use data-beans and senna (legume) Rust-based CLI tools (already installed)

3. Small data sets: we can show both Seurat and Scanpy how we can process

4. Do not mix R and python in the same slide/frame (it's confusing)

5. Prefer hdf5 backend since zarr takes too many files.

6. Prefer onslide rather than only in latex beamer

7. use \large for discussion (text-driven) slides then close it with \normalsize

8. Figure width never exceeds 6 and height never 3

9. Rasterize plots with too many dots with dpi=300 using `ggrastr`

10. use pdflatex preferrably...

-->


```{r setup, include=FALSE}
source("Setup.R")
fig.dir <- "../Fig/single-cell-factorization/"
dir.create(fig.dir, showWarnings=FALSE)
setup.env(fig.dir)
dir.create("../data", showWarnings=FALSE)

library(data.table)
library(tidyverse)
library(patchwork)
library(matrixTests)
library(ggrastr)
library(rsvd)

theme_set(theme_minimal())
```

## Roadmap of STAT/BIOF/GSAT540 2026

_data generation_ followed by _inference_ or validation

\vfill

\large

:::::: {.columns}
::: {.column width=.45}

3. Formulating biological questions

4. Hypothesis testing

5. RNA-seq data

6. Inference focusing on RNA-seq

7. Experimental design

:::
::: {.column width=.45}

8. Single-cell genomics

9. \textbf{\color{magenta} Matrix factorization} (Today)

10. Genome-wide association studies

11. Multiple hypothesis testing

:::
::::::

\normalsize


## Recap: Overview of single-cell data analysis

\centerline{\includegraphics[width=\textwidth]{img/scRNA_overview/Kharchenko_1.pdf}}

\vfill

\flushright
\small
[@Kharchenko2021-wr]

## Overview of single-cell data analysis cont'd

\centerline{\includegraphics[width=\textwidth]{img/scRNA_overview/Kharchenko_2.pdf}}

\vfill

\flushright
\small
[@Kharchenko2021-wr]


## Learning objectives

- Learn Principal Component Analysis (PCA) and Singular Value Decomposition (SVD)

- Understand matrix factorization and its applications in single-cell analysis

- Introduction to useful tools based on matrix factorization

	- `glmpca`, `rliger`, `fastTopic`, `MOFA`, ...


# What is Principal Component Analysis?


```{r child='lect09-pca-section.rmd'}
```


## Generalized Linear Model PCA to model count data

:::::: {.columns}
::: {.column width=.45}

\large

- scRNA-seq data are **counts** (discrete, non-negative)

$$\log(\mu_{ij}) = (UV^{\top})_{ij}$$

$$Y_{ij} \sim \text{Poisson}(\mu_{ij})$$

or

$$Y_{ij} \sim \text{NB}(\mu_{ij}, \phi)$$

\normalsize

:::
::: {.column width=.45}

```{r eval=FALSE, echo=TRUE}
library(glmpca)

## Run GLM-PCA on count matrix
res <- glmpca(X, L = 5, fam = "nb")

factors <- res$factors
loadings <- res$loadings
```

:::
::::::

## GLM-PCA vs standard PCA

\large

**When to use GLM-PCA?**

\normalsize

- Working with **raw counts** (not normalized)
- Want to avoid log-transformation artifacts
- More computationally intensive than standard PCA

\vspace{0.5cm}

**Reference:** [@Townes2019-rg]


# Non-negative Matrix Factorization

```{r child='lect09-nmf-section.rmd'}
```

## `rliger`: Integrative NMF for single-cell data

:::::: {.columns}
::: {.column width=.25}

\includegraphics[width=\textwidth]{img/logos/rliger_logo.png}

:::
::: {.column width=.6}

- Joint NMF across multiple datasets
- Shared factors ($W$) capture common biology
- Dataset-specific factors ($V_i$) capture batch effects
- Quantile normalization aligns factor loadings

$$X^{(d)} \approx H^{(d)} (W + V^{(d)})^{\top}$$

$H^{(d)}$: cell factor loadings for dataset $d$

$W$: shared gene factors, $V^{(d)}$: dataset-specific

:::
::::::


@Welch2019-jl; @Gao2021-vy; @Kriebel2022-fe 


## `rliger`: Running integrative NMF

```{r load-seurat-main, message=FALSE, warning=FALSE}
library(Seurat)

data.dir <- "../data/GSE155698_PDAC_Steele2020/PDAC_TISSUE_1/"

pdac <- Read10X(data.dir = data.dir)
pdac <- CreateSeuratObject(counts = pdac,
                           project = "PDAC_TISSUE_1",
                           min.cells = 3,
                           min.features = 200)

X <- as.matrix(GetAssayData(pdac, layer = "counts"))
mt.genes <- grep("^MT-", rownames(X), value = TRUE)
X <- X[!rownames(X) %in% mt.genes, ]
```


```{r liger-run, include=FALSE}
library(rliger)
liger.rds <- "../data/GSE155698_PDAC_Steele2020/liger_obj_3tissues.rds"

if.needed(liger.rds, {
  tissues <- c("PDAC_TISSUE_1", "PDAC_TISSUE_2", "PDAC_TISSUE_3")
  data.list <- lapply(tissues, function(tissue) {
    data <- Read10X(data.dir = paste0("../data/GSE155698_PDAC_Steele2020/", tissue, "/"))
    mt.genes <- grep("^MT-", rownames(data), value = TRUE)
    data[!rownames(data) %in% mt.genes, ]
  })
  names(data.list) <- tissues
  liger_obj <- createLiger(data.list)
  liger_obj <- normalize(liger_obj)
  liger_obj <- selectGenes(liger_obj)
  liger_obj <- scaleNotCenter(liger_obj)
  liger_obj <- optimizeALS(liger_obj, k = 20)
  liger_obj <- quantile_norm(liger_obj)
  liger_obj <- louvainCluster(liger_obj, resolution = 0.4)
  liger_obj <- runUMAP(liger_obj, n_neighbors = 30)
  saveRDS(liger_obj, liger.rds)
})

liger_obj <- readRDS(liger.rds)
```

```{r liger-load, eval=FALSE, echo=TRUE}
library(rliger)

## Load three PDAC tissue samples
tissues <- c("PDAC_TISSUE_1", "PDAC_TISSUE_2", "PDAC_TISSUE_3")

data.list <- lapply(tissues, function(tissue) {
  data <- Read10X(data.dir = paste0("../data/GSE155698_PDAC_Steele2020/", tissue, "/"))
  ## Remove MT-genes
  mt.genes <- grep("^MT-", rownames(data), value = TRUE)
  data[!rownames(data) %in% mt.genes, ]
})
names(data.list) <- tissues
```

## `rliger`: Create object and preprocess

```{r liger-preprocess, eval=FALSE, echo=TRUE}
## Create liger object with three datasets
liger_obj <- createLiger(data.list)

## Normalize and select variable genes
liger_obj <- normalize(liger_obj)
liger_obj <- selectGenes(liger_obj)
liger_obj <- scaleNotCenter(liger_obj)
```

## `rliger`: Run integrative NMF

```{r liger-inmf, eval=FALSE, echo=TRUE, size="Large"}
## Run integrative NMF
liger_obj <- optimizeALS(liger_obj, k = 20)
```

## `rliger`: Align datasets and cluster

```{r liger-cluster, eval=FALSE, echo=TRUE}
## Quantile normalization for alignment
liger_obj <- quantile_norm(liger_obj)

## Clustering on aligned factors
liger_obj <- louvainCluster(liger_obj, resolution = 0.4)

## UMAP visualization
liger_obj <- runUMAP(liger_obj, n_neighbors = 30)
```

## `rliger`: UMAP by dataset

```{r liger-umap-dataset, eval=TRUE, echo=TRUE, fig.width=6, fig.height=2.5}
## UMAP colored by dataset
plotDatasetDimRed(liger_obj)
```

## `rliger`: UMAP by cluster

```{r liger-umap-cluster, eval=TRUE, echo=TRUE, fig.width=6, fig.height=2.5}
## UMAP colored by cluster
plotClusterDimRed(liger_obj)
```

## `fastTopics`: Topic modeling for scRNA-seq

\large

- Topic model: each cell $F$ = **mixture of topics**
- each topic $L$ = distribution over genes

$$Y_{gj} = \sum_{k} L_{gk} F_{kj}$$

\normalsize

@Carbonetto2021-ft; @Carbonetto2023-gd

```{r fasttopic-run, include=FALSE}
library(fastTopics)

## Use the same PDAC tissue 1 count matrix (cells x genes)
gene_var <- apply(X, 1, var)
top_genes <- names(sort(gene_var, decreasing = TRUE))[1:5000]
counts <- t(X[top_genes, ])

fasttopic.rds <- "../data/GSE155698_PDAC_Steele2020/fasttopic_fit.rds"
if.needed(fasttopic.rds, {
  fit <- fit_topic_model(counts, k = 10)
  saveRDS(fit, fasttopic.rds)
})
fit <- readRDS(fasttopic.rds)

fasttopic.de.rds <- "../data/GSE155698_PDAC_Steele2020/fasttopic_de.rds"
if.needed(fasttopic.de.rds, {
  de <- de_analysis(fit, counts)
  saveRDS(de, fasttopic.de.rds)
})
de <- readRDS(fasttopic.de.rds)
```

```{r fasttopic-load, eval=FALSE, echo=TRUE}
library(fastTopics)

## Use the same PDAC tissue 1 count matrix (cells x genes)
## Select top 5000 most variable genes
gene_var <- apply(X, 1, var)
top_genes <- names(sort(gene_var, decreasing = TRUE))[1:5000]
counts <- t(X[top_genes, ])
dim(counts)
```

## `fastTopics`: Fit topic model

```{r fasttopic-fit, eval=FALSE, echo=TRUE, size="Large"}
## Fit topic model with k=10 topics
fit <- fit_topic_model(counts, k = 10)
```

## `fastTopics`: Topic proportions per cell

```{r fasttopic-L, eval=TRUE, echo=TRUE}
## L matrix: cells x topics (topic proportions)
head(round(fit$L, 3))
```

## `fastTopics`: Top genes per topic

```{r fasttopic-genes, eval=TRUE, echo=TRUE}
## F matrix: genes x topics (gene weights)
top_genes <- apply(fit$F, 2, function(x) {
  names(sort(x, decreasing = TRUE)[1:10])
})
top_genes[, 1:3]
```

## `fastTopics`: Structure plot

```{r fasttopic-structure, eval=TRUE, fig.width=6, fig.height=2.5, message=FALSE, warning=FALSE, results='hide'}
## Structure plot: stacked bar of topic proportions per cell
structure_plot(fit, topics = 1:10, n = 500)
```

## `fastTopics`: Volcano plot — genes driving each topic

\only<1>{Topic 1: PRSS1, REG3A, CTRB1 — pancreatic acinar cells}
\only<2>{Topic 2: MALAT1, RPL41, RPL10 — housekeeping/ribosomal}
\only<3>{Topic 3: FTL, SPP1, CTSD — macrophage/myeloid markers}
\only<4>{Topic 4: RARRES1, PLA2G2A, RPL18A — ductal/regenerating}
\only<5>{Topic 5: TMSB4X, GNLY, GZMA — cytotoxic T/NK cells}
\only<6>{Topic 6: RARRES1, FTH1, HLA-B — ductal/epithelial}
\only<7>{Topic 7: S100A6, RPLP1, RPS19 — epithelial}
\only<8>{Topic 8: S100A9, S100A8, LYZ — neutrophils/monocytes}
\only<9>{Topic 9: S100A6, FTH1, TMSB10 — proliferating}
\only<10>{Topic 10: MALAT1, NEAT1, HLA-A — antigen presentation/stress}

```{r fasttopic-volcano-1, eval=TRUE, fig.width=5, fig.height=2.5, message=FALSE, warning=FALSE, only.plot="1"}
volcano_plot(de, k = 1)
```

```{r fasttopic-volcano-2, eval=TRUE, fig.width=5, fig.height=2.5, message=FALSE, warning=FALSE, only.plot="2"}
volcano_plot(de, k = 2)
```

```{r fasttopic-volcano-3, eval=TRUE, fig.width=5, fig.height=2.5, message=FALSE, warning=FALSE, only.plot="3"}
volcano_plot(de, k = 3)
```

```{r fasttopic-volcano-4, eval=TRUE, fig.width=5, fig.height=2.5, message=FALSE, warning=FALSE, only.plot="4"}
volcano_plot(de, k = 4)
```

```{r fasttopic-volcano-5, eval=TRUE, fig.width=5, fig.height=2.5, message=FALSE, warning=FALSE, only.plot="5"}
volcano_plot(de, k = 5)
```

```{r fasttopic-volcano-6, eval=TRUE, fig.width=5, fig.height=2.5, message=FALSE, warning=FALSE, only.plot="6"}
volcano_plot(de, k = 6)
```

```{r fasttopic-volcano-7, eval=TRUE, fig.width=5, fig.height=2.5, message=FALSE, warning=FALSE, only.plot="7"}
volcano_plot(de, k = 7)
```

```{r fasttopic-volcano-8, eval=TRUE, fig.width=5, fig.height=2.5, message=FALSE, warning=FALSE, only.plot="8"}
volcano_plot(de, k = 8)
```

```{r fasttopic-volcano-9, eval=TRUE, fig.width=5, fig.height=2.5, message=FALSE, warning=FALSE, only.plot="9"}
volcano_plot(de, k = 9)
```

```{r fasttopic-volcano-10, eval=TRUE, fig.width=5, fig.height=2.5, message=FALSE, warning=FALSE, only.plot="10"}
volcano_plot(de, k = 10)
```


# How can we use factorization results?

## Gene set enrichment analysis with PanglaoDB

```{r gsea-run, include=FALSE}
library(fgsea)

gsea.rds <- "../data/GSE155698_PDAC_Steele2020/fasttopic_gsea.rds"
if.needed(gsea.rds, {
  panglao.db <- read.panglao("Pancreas")
  .beta <- fit$F
  rownames(.beta) <- colnames(counts)
  .genes <- unique(unlist(panglao.db))
  .dt <- sort.beta(.beta, genes.selected = .genes)
  .gsea <- run.beta.gsea(.dt, panglao.db)
  saveRDS(list(gsea = .gsea, beta = .dt, db = panglao.db), gsea.rds)
})
.gsea.result <- readRDS(gsea.rds)
.gsea <- .gsea.result$gsea
.dt <- .gsea.result$beta
panglao.db <- .gsea.result$db
```

```{r gsea-panglao, eval=FALSE, echo=TRUE}
library(fgsea)

## Read PanglaoDB cell-type marker gene sets
panglao.db <- read.panglao("Pancreas")

## Prepare gene weights from fastTopics F matrix
## Each column = topic, rows = genes
.beta <- apply(fit$F, 2, scale)
rownames(.beta) <- colnames(counts)

## Sort genes by topic assignment
.genes <- unique(unlist(panglao.db))
.dt <- sort.beta(.beta, genes.selected = .genes)

## Run fgsea per topic
.gsea <- run.beta.gsea(.dt, panglao.db)
```

## GSEA results: which topics correspond to cell types?

```{r gsea-plot, eval=TRUE, fig.width=5, fig.height=3, message=FALSE, warning=FALSE}
.show <- select.gsea.beta(.gsea, .dt, sd.cutoff = 1e-4)

## Enrichment heatmap: -log10(padj) per topic x cell type
.gg.plot(.show$gsea, aes(variable, ct, fill=-log10(pmax(padj, 1e-4)))) +
    xlab("topics") + ylab("cell types") +
    geom_tile(linewidth=.1, color="black") +
    theme(legend.key.width=unit(.2, "lines")) +
    theme(legend.text = element_text(size=4)) +
    scale_fill_distiller("Padj", palette = "YlGnBu", direction=1,
                         labels=function(x) num.sci(10^(-x)))
```

## GSEA: Gene weights across topics

```{r gsea-beta, eval=TRUE, fig.width=6, fig.height=3, message=FALSE, warning=FALSE, only.plot="1"}
p1 <-
    .gg.plot(.show$beta, aes(row, variable, fill=`value`)) +
    theme(axis.text.x = element_blank()) +
    theme(axis.ticks.x = element_blank()) +
    theme(legend.position = "none") +
    ggrastr::rasterise(geom_tile(), dpi=300) +
    ylab("topics") + xlab("genes") +
    scale_fill_distiller(direction=-1, palette="RdBu")

p2 <-
    .gg.plot(.show$leading.edges, aes(row, ct)) +
    theme(axis.text.x = element_blank()) +
    theme(axis.ticks.x = element_blank()) +
    theme(panel.grid = element_blank()) +
    geom_tile(linewidth=0) +
    ylab("cell types")

print(p1/p2)
```

```{r gsea-combined, eval=TRUE, fig.width=6, fig.height=3, message=FALSE, warning=FALSE, only.plot="2"}
p3 <-
    .gg.plot(.show$gsea, aes(variable, ct, fill=-log10(pmax(padj, 1e-4)))) +
    xlab("topics") + ylab("") +
    scale_y_discrete(position = "right") +
    geom_tile(linewidth=.1, color="black") +
    theme(legend.key.width=unit(.2, "lines")) +
    theme(legend.text = element_text(size=4)) +
    scale_fill_distiller("Padj", palette = "YlGnBu", direction=1,
                         labels=function(x) num.sci(10^(-x)))

C <- length(unique(.show$gsea$ct))
K <- length(unique(.show$beta$variable))
p0 <- ggplot() + theme_void()
wrap_plots(p1, p0, p2, p3, nrow = 2, heights=c(K,C), widths=c(4,1))
```

## `rliger`: GSEA on shared gene loadings (W matrix)

```{r liger-gsea-run, include=FALSE}
library(fgsea)

liger.gsea.rds <- "../data/GSE155698_PDAC_Steele2020/liger_gsea.rds"
if.needed(liger.gsea.rds, {
  W <- getMatrix(liger_obj, "W")
  panglao.db <- read.panglao("Pancreas")
  .beta <- apply(W, 2, scale)
  rownames(.beta) <- rownames(W)
  .genes <- unique(unlist(panglao.db))
  .dt <- sort.beta(.beta, genes.selected = .genes)
  .gsea <- run.beta.gsea(.dt, panglao.db)
  saveRDS(list(gsea = .gsea, beta = .dt, db = panglao.db), liger.gsea.rds)
})
.liger.gsea <- readRDS(liger.gsea.rds)
```

```{r liger-gsea-plot, eval=TRUE, fig.width=5, fig.height=3, message=FALSE, warning=FALSE, only.plot="1"}
.lshow <- select.gsea.beta(.liger.gsea$gsea, .liger.gsea$beta, sd.cutoff = 1e-4)

lp1 <-
    .gg.plot(.lshow$beta, aes(row, variable, fill=`value`)) +
    theme(axis.text.x = element_blank()) +
    theme(axis.ticks.x = element_blank()) +
    theme(legend.position = "none") +
    ggrastr::rasterise(geom_tile(), dpi=300) +
    ylab("iNMF factors") + xlab("genes") +
    scale_fill_distiller(direction=-1, palette="RdBu")

lp2 <-
    .gg.plot(.lshow$leading.edges, aes(row, ct)) +
    theme(axis.text.x = element_blank()) +
    theme(axis.ticks.x = element_blank()) +
    theme(panel.grid = element_blank()) +
    geom_tile(linewidth=0) +
    ylab("cell types")

print(lp1/lp2)
```

```{r liger-gsea-combined, eval=TRUE, fig.width=6, fig.height=3, message=FALSE, warning=FALSE, only.plot="2"}
lp3 <-
    .gg.plot(.lshow$gsea, aes(variable, ct, fill=-log10(pmax(padj, 1e-4)))) +
    xlab("iNMF factors") + ylab("") +
    scale_y_discrete(position = "right") +
    geom_tile(linewidth=.1, color="black") +
    theme(legend.key.width=unit(.2, "lines")) +
    theme(legend.text = element_text(size=4)) +
    scale_fill_distiller("Padj", palette = "YlGnBu", direction=1,
                         labels=function(x) num.sci(10^(-x)))

C <- length(unique(.lshow$gsea$ct))
K <- length(unique(.lshow$beta$variable))
lp0 <- ggplot() + theme_void()
wrap_plots(lp1, lp0, lp2, lp3, nrow = 2, heights=c(K,C), widths=c(4,1))
```

## Wrap-up

\large

- **PCA/SVD**: Find orthogonal axes of maximum variance; fast, good for QC
- **NMF**: Non-negative, parts-based decomposition; interpretable gene programs
- **`rliger`**: Integrative NMF across multiple datasets (batch correction)
- **`fastTopics`**: Scalable topic model; cells as mixtures of gene programs
- **GSEA on factors**: Link learned topics to known cell types (e.g., PanglaoDB + `fgsea`)

\normalsize

##

## Reference {.allowframebreaks}
