
## Let's start with multivariate linear regression

**Linear model for gene $g$:**

$$Y_{ig} = \beta_{1g} \cdot X_{i1} + \beta_{2g} \cdot X_{i2} + \beta_{3g} \cdot X_{i3} + \epsilon_{ig},\, \epsilon_{ig} \sim N(0, \sigma^2_g)$$

:::::: {.columns}
::: {.column width=.4}

\begin{center}
\input{texfrag/graphical-model-linear-regression.tex}
\end{center}

:::
::: {.column width=.45}

* $X_{ij}$: covariate $j$ for sample $i$
* $\beta_{jg}$: effect of covariate $j$ on gene $g$
* $Y_{ig}$: expr. of gene $g$ in sample $i$

:::
::::::

:::
::::::

## A design matrix = a column space view

We can rewrite the linear model as:

$$\mathbf{y}_g = \beta_{0g} \mathbf{x}_0 + \beta_{1g} \mathbf{x}_1 + \beta_{2g} \mathbf{x}_2 + \beta_{3g} \mathbf{x}_3 + \boldsymbol{\epsilon}_g$$

where $\mathbf{x}_j$ is the $j$-th column of $\mathbf{X}$

\onslide<2>{

\textbf{\color{teal} What is a linear regression?}

Finding the "best" \textbf{\color{teal} linear combination} of {\color{magenta} column vectors} to approximate $\mathbf{y}_g$

\begin{itemize}
\item Each column $\mathbf{x}_j$ represents a covariate pattern across all samples
\item The coefficients $\beta_{jg}$ tell us how much each pattern contributes
\item The fitted values $\hat{\mathbf{y}}_g = \mathbf{X}\hat{\boldsymbol{\beta}}_g$ lie in the \textbf{\color{magenta} column space} of $\mathbf{X}$
\end{itemize}

}

## Linear regression = projection onto linear space

If the whole prediction space can be made up by a linear combination of two $n$ vectors $\mathbf{x}_{1}$ and $\mathbf{x}_{2}$...

\begin{center}
\input{texfrag/column-space-projection.tex}
\end{center}

$\hat{\mathbf{y}}_g$ is the **projection** of $\mathbf{y}_g$ onto the column space of $\mathbf{X}$

## Linear regression = projection onto some space

:::::: {.columns}
::: {.column width=.4}

\begin{center}
\input{texfrag/graphical-model-linear-regression-latent.tex}
\end{center}

:::
::: {.column width=.45}

**What if our covariates are unknown?**

$$Y_{ig} = U_{i1} V_{1g} + U_{i2} V_{2g} + X_{i3} V_{3g} + \epsilon_{ig}$$

\vspace{0.3cm}

* $U_{ij}$: **latent** factor $j$ for sample $i$
* $V_{jg}$: loading of gene $g$ on factor $j$
* $Y_{ig}$: observed expression of gene $g$ in sample $i$

:::
::::::

## Many linear regression models on many genes/columns

\begin{center}
\input{texfrag/graphical-model-multi-gene.tex}
\end{center}

## Many linear regression models on many genes/columns

\begin{center}
\input{texfrag/graphical-model-matrix-factorization.tex}
\end{center}


## The goal is to find hidden patterns

:::::: {.columns}
::: {.column width=.5}


```{r echo=FALSE}
U <- .rnorm(100, 2); V <- .rnorm(100, 2)
X <- U %*% t(V) + .rnorm(100, 100)
colnames(X) <- paste0("cell", 1:ncol(X))
rownames(X) <- paste0("gene", 1:nrow(X))
```

A data matrix was generated by the following:

$$\mathbf{x}_{i} = \mathbf{u}_{i} V^{\top} + \epsilon_{i}$$

for all $i$ with $\epsilon_{i} \sim \mathcal{N}\!\left(\mathbf{0},I\right)$

```{r echo=FALSE}
X <- .sort.matrix(X)
```

```{r echo=FALSE, fig.width=2.5, fig.height=2, onslide.plot="1-"}
.matshow(X, .lab=0, .lw=0)
```

:::
::: {.column width=.5}

```{r echo=FALSE}
.svd <- rsvd::rsvd(X, k=5)
u.hat <- sweep(.svd$u, 2, .svd$d, `*`)
```

Can we reverse engineer the $U$ and $V$ matrices from the data $X$?

$$X \to U V^{\top}$$

```{r echo=FALSE, fig.width=2.7, fig.height=2.3, only.plot="2"}
p1 <- .matshow(u.hat, .lab=0, .lw=0) +
    ggtitle("features")
p.times <- ggplot() + theme_void() +
    geom_text(aes(0,0,label="X"), size=5)
p2 <- wrap_plots(.matshow(t(.svd$v), .lab=0, .lw=0),
                 ggplot(),
                 ncol=1, heights=c(1,9)) +
    ggtitle("feature loading")
wrap_plots(list(p1, p.times, p2), nrow=1, widths=c(1,2,8))
```

:::
::::::


## Let's look at the PDAC 10x Genomics data

\vfill

:::::: {.columns}
::: {.column width=.5}

\small
```bash
PDAC_TISSUE_1/
|-- barcodes.tsv.gz   (9 KB)
|-- features.tsv.gz   (303 KB)
+-- matrix.mtx.gz     (17 MB)
```
\normalsize

\vspace{0.5cm}

* **`barcodes.tsv.gz`** - Cell IDs
* **`features.tsv.gz`** - Genes
* **`matrix.mtx.gz`** - Counts (sparse)

:::
::: {.column width=.45}

**Matrix Market format**

* Sparse matrix standard
* Memory efficient
* Compatible with Seurat, Scanpy

:::
::::::

## Loading 10x data with Seurat (R) \hfill \includegraphics[height=1em]{img/logos/seurat_logo.jpg}

```{r load-seurat, message=FALSE, warning=FALSE}
library(Seurat)

data.dir <- "../data/GSE155698_PDAC_Steele2020/PDAC_TISSUE_1/"

pdac <- Read10X(data.dir = data.dir)
pdac <- CreateSeuratObject(counts = pdac,
                           project = "PDAC_TISSUE_1",
                           min.cells = 3,
                           min.features = 200)
pdac
```

Extract the count matrix for later use:

```{r}
## Extract raw counts from Seurat object for downstream analysis
X <- as.matrix(GetAssayData(pdac, layer = "counts"))

## Remove MT-genes (mitochondrial genes)
mt.genes <- grep("^MT-", rownames(X), value = TRUE)
X <- X[!rownames(X) %in% mt.genes, ]
dim(X)
```


## Let's just use top 200 most variable genes and cells

```{r}
.sd <- apply(X, 1, sd, na.rm=TRUE)
msv <- order(.sd, decreasing = TRUE)[1:200]
.sd <- apply(X, 2, sd, na.rm=TRUE)
msw <- order(.sd, decreasing = TRUE)[1:200]
```

Top 200 genes

:::::: {.columns}
::: {.column width=.25}

```{r echo=FALSE, fig.width=2, fig.height=2, only.plot="1"}
x.sub <- as.matrix(X)[msv, msw, drop = FALSE]
.matshow(x.sub, .scale=TRUE, .lab=0, .lw=0)
```

```{r echo=FALSE, fig.width=2, fig.height=2, only.plot="2"}
x.sub <- t(scale(t(log1p(x.sub))))
.matshow(pmin(x.sub, 3), .scale=TRUE, .lab=0, .lw=0)
```

:::
::: {.column width=.65}

* We will call such a high-dimensional matrix $X$ ($m \times n$)

* X has $m$=`r num.int(nrow(X))` rows (transcripts/genes/features)

* X has $n$=`r num.int(ncol(X))` columns (cells/#data points)

* The rows were log-transformed and scaled by `scale()` for visualization

* Each cell is a `r num.int(nrow(X))`-dimensional vector!

* Each gene is a `r num.int(ncol(X))`-dimensional vector...

:::
::::::


## 1-dimensional representations/summary of data matrix

:::::: {.columns}
::: {.column width=.25}

Top 200 genes

```{r echo=FALSE, fig.width=2, fig.height=2}
.matshow(pmin(x.sub, 3), .scale=TRUE, .lab=0, .lw=0)
```

:::
::: {.column width=.65}

```{r echo=FALSE, fig.width=3.8, fig.height=1.25, onslide.plot="1-"}
ggplot(as.data.table(X[msv[1],]),
         aes(x=colnames(X), y=V1)) +
    ggtitle("Most variable cell (column):") +
    ylab("expression") +
    xlab("cells") +
    theme(axis.text.y = element_text(size=4)) +
    theme(axis.text.x = element_blank()) +
    theme(axis.ticks.x = element_blank()) +
    geom_point(pch=21, size=.25)
```

```{r echo=FALSE, fig.width=3.8, fig.height=1.25, onslide.plot="2"}
ggplot(as.data.table(X[,msw[1]]),
         aes(x=1:nrow(X), y=V1)) +
    ggtitle("Most variable gene (row):") +
    ylab("expression") +
    xlab("genes") +
    theme(axis.text.y = element_text(size=4)) +
    theme(axis.text.x = element_blank()) +
    theme(axis.ticks.x = element_blank()) +
    geom_point(pch=21, size=.2, alpha=.5)
```

:::
::::::


## Can you see some common patterns?

```{r echo=FALSE}
x.sorted <- .sort.matrix(x.sub)
```

:::::: {.columns}
::: {.column width=.5}

```{r echo=FALSE, fig.width=2, fig.height=2.5}
.matshow(pmin(x.sorted, 3), .scale=TRUE, .lab=0, .lw=0) +
    ggtitle("200 genes and cells")
```

:::
::: {.column width=.5}

Common patterns will become more apparent after applying PCA.

:::
::::::

## How do we recover "common" patterns from data?

Principal Component Analysis

:::::: {.columns}
::: {.column width=.45}

::: {.block}

### [@Pearson1901-on]

* Projection [of the original data] that minimizes the projection cost between the original and projected

* The cost = mean squared distance

:::

:::
::: {.column width=.45}

::: {.block}

### [@Hotelling1933-aa]

* Orthogonal projection of data into a lower-dimensional **[principal]** sub-space,

* such that the total **variation of the projected data** is maximized

:::

:::
::::::

## Let's see how much the first eigenvector can explain

:::::: {.columns}
::: {.column width=.35}

```{r echo=FALSE, fig.width=2, fig.height=2.5, only.plot="1"}
.matshow(pmin(x.sorted, 3), .lab=0, .lw=0) +
    ggtitle("Top 200 genes")
```

```{r}
.eigen <- eigen(x.sorted %*% t(x.sorted))
U <- .eigen$vector[,1,drop=FALSE]
.lm <- lm(x.sorted ~ U -1)
```

```{r echo=FALSE, fig.width=2, fig.height=2.5, only.plot="2"}
.matshow(pmin(.lm$fitted.values,3), .lab=0, .lw=0) +
    ggtitle("eigen proj.")
```

```{r echo=FALSE, fig.width=2, fig.height=2.5, only.plot="3"}
.matshow(pmin(x.sorted, 3), .lab=0, .lw=0) +
    ggtitle("Top 200 genes")
```

:::
::: {.column width=.65}

```{r echo=FALSE, fig.width=3.2, fig.height=2.7, only.plot="2"}
p0 <- ggplot() + theme_void()
p.eq <- ggplot() + theme_void() + geom_text(aes(0,0,label="="), size=10)
p.prod <- ggplot() + theme_void() + geom_text(aes(0,0,label="x"), size=10)
p1 <- .matshow(U, .lab=0, .lw=0, .scale=TRUE) +
    ggtitle("e-vec")
p2 <- .matshow(matrix(.lm$coefficients, nrow=1),
               .lab=0, .lw=0, .scale=TRUE) +
    ggtitle("loading")
p2 <- wrap_plots(p2, p0, ncol=1, heights = c(1,13))
wrap_plots(p.eq, p0, p1, p0, p.prod, p2, nrow=1, widths=c(1, 1, 1, 1, 1, 11))
```

```{r echo=FALSE, fig.width=2, fig.height=2.5, only.plot="3"}
var.exp <- mean(apply(.lm$fitted.values, 2, var) / apply(x.sorted, 2, var))
.matshow(pmin(.lm$fitted.values,3), .lab=0, .lw=0) +
    ggtitle(round(var.exp*100) %&% "%")
```

:::
::::::


## Top two eigen-vectors

:::::: {.columns}
::: {.column width=.35}

```{r echo=FALSE, fig.width=2, fig.height=2.5, only.plot="1"}
.matshow(pmin(x.sorted, 3), .scale=TRUE, .lab=0, .lw=0) +
    ggtitle("Top 200 genes")
```

```{r}
U <- .eigen$vector[,1:2,drop=FALSE]
.lm <- lm(x.sorted ~ U -1)
```

```{r echo=FALSE, fig.width=2, fig.height=2.5, only.plot="2"}
.matshow(pmin(.lm$fitted.values,3), .lab=0, .lw=0) +
    ggtitle("eigen proj.")
```

```{r echo=FALSE, fig.width=2, fig.height=2.5, only.plot="3"}
.matshow(pmin(x.sorted, 3), .lab=0, .lw=0) +
    ggtitle("Top 200 genes")
```

:::
::: {.column width=.65}

```{r echo=FALSE, fig.width=3.2, fig.height=2.7, only.plot="2"}
p0 <- ggplot() + theme_void()
p.eq <- ggplot() + theme_void() + geom_text(aes(0,0,label="="), size=10)
p.prod <- ggplot() + theme_void() + geom_text(aes(0,0,label="x"), size=10)
p1 <- .matshow(U, .lab=0, .lw=0, .scale=TRUE) +
    ggtitle("e-vec")
p2 <- .matshow(as.matrix(.lm$coefficients),
               .lab=0, .lw=0, .scale=TRUE) +
    ggtitle("loading")
p2 <- wrap_plots(p2, p0, ncol=1, heights = c(1,13))
wrap_plots(p.eq, p0, p1, p0, p.prod, p2, nrow=1, widths=c(1, 1, 1, 1, 1, 11))
```

```{r echo=FALSE, fig.width=2, fig.height=2.5, only.plot="3"}
var.exp <- mean(apply(.lm$fitted.values, 2, var) / apply(x.sorted, 2, var))
.matshow(pmin(.lm$fitted.values,3), .lab=0, .lw=0) +
    ggtitle(round(var.exp*100) %&% "%")
```

:::
::::::

## Top three eigen-vectors

:::::: {.columns}
::: {.column width=.35}

```{r echo=FALSE, fig.width=2, fig.height=2.5, only.plot="1"}
.matshow(pmin(x.sorted, 3), .lab=0, .lw=0) +
    ggtitle("Top 200 genes and 200 cells")
```

```{r}
U <- .eigen$vector[,1:3,drop=FALSE]
.lm <- lm(x.sorted ~ U -1)
```

```{r echo=FALSE, fig.width=2, fig.height=2.5, only.plot="2"}
.matshow(pmin(.lm$fitted.values,3), .lab=0, .lw=0) +
    ggtitle("eigen proj.")
```

```{r echo=FALSE, fig.width=2, fig.height=2.5, only.plot="3"}
.matshow(pmin(x.sorted, 3), .lab=0, .lw=0) +
    ggtitle("Top 200 genes and 200 cells")
```

:::
::: {.column width=.65}

```{r echo=FALSE, fig.width=3.2, fig.height=2.7, only.plot="2"}
p0 <- ggplot() + theme_void()
p.eq <- ggplot() + theme_void() + geom_text(aes(0,0,label="="), size=10)
p.prod <- ggplot() + theme_void() + geom_text(aes(0,0,label="x"), size=10)
p1 <- .matshow(U, .lab=0, .lw=0, .scale=TRUE) +
    ggtitle("e-vec")
p2 <- .matshow(as.matrix(.lm$coefficients),
               .lab=0, .lw=0, .scale=TRUE) +
    ggtitle("loading")
p2 <- wrap_plots(p2, p0, ncol=1, heights = c(1,13))
wrap_plots(p.eq, p0, p1, p0, p.prod, p2, nrow=1, widths=c(1, 1, 1, 1, 1, 11))
```

```{r echo=FALSE, fig.width=2, fig.height=2.5, only.plot="3"}
var.exp <- mean(apply(.lm$fitted.values, 2, var) / apply(x.sorted, 2, var))
.matshow(pmin(.lm$fitted.values,3), .lab=0, .lw=0) +
    ggtitle(round(var.exp*100) %&% "%")
```

:::
::::::

##

\huge

What is Principal Component Analysis?

\large

(a more mathematical definition)

## PCA: Consider this multivariate linear model

\only<1>{
$$\left(\begin{array}{l}
X_{1}\\
X_{2}\\
\vdots \\
X_{p}\\
\end{array}
\right) =
\left(
\begin{array}{l l l}
U_{11} & \cdots & U_{1k}\\
U_{21} & \cdots & U_{2k}\\
\vdots & \vdots & \vdots \\
U_{p1} & \cdots & U_{pk}\\
\end{array}
\right)
\left(
\begin{array}{l}
V_{1}\\
\vdots\\
V_{k}
\end{array}
\right)
+
\left(
\begin{array}{l}
\epsilon_{1}\\
\epsilon_{2}\\
\vdots\\
\epsilon_{p}
\end{array}
\right)$$

or

$\mathbf{x} = U \mathbf{v} + \epsilon$}

\only<2>{For many columns, $X = U V + E$, or
$$\left(\begin{array}{l l l}
X_{11} & \cdots & X_{1n}\\
X_{21} & \cdots & X_{2n}\\
\vdots & & \vdots \\
X_{p1} & \cdots & X_{pn}\\
\end{array}
\right) =
\left(
\begin{array}{l l l}
U_{11} & \cdots & U_{1k}\\
U_{21} & \cdots & U_{2k}\\
\vdots & \vdots & \vdots \\
U_{p1} & \cdots & U_{pk}\\
\end{array}
\right)
\left(
\begin{array}{l l l}
V_{11} & \cdots & V_{1n}\\
\vdots & & \vdots \\
V_{k1} & \cdots & V_{kn} \\
\end{array}
\right)
+ \cdots$$}

* If we **knew** $U$, we would be able to solve weights $V$.

* And vice versa...

## PCA: What is a projection matrix?

Suppose we knew $V$. For each gene (row) $g$, we solve the following regression.

\begin{eqnarray*}
\mathbf{x}_{g}^{\top} &\sim& V \mathbf{u}_{g}^{\top} + \epsilon I\\
\onslide<2->{\left(\begin{array}{l}
X_{g1} \\
\vdots \\
X_{gn}
\end{array}\right)
&\sim&
\left(\begin{array}{l l l}
V_{11} & \cdots & V_{1k} \\
\vdots & \vdots & \vdots \\
V_{n1} & \cdots & V_{nk} \\
\end{array}\right)
\left(\begin{array}{l}
U_{g1} \\
\vdots \\
U_{gk}
\end{array}\right) +
\left(\begin{array}{l}
\epsilon \\
\vdots \\
\epsilon
\end{array}\right)}
\end{eqnarray*}

Least square solution:

\only<1-2>{$$\underset{\mathbf{u}_{g}}{\min} \, \|\mathbf{x}_{g}^{\top} - V \mathbf{u}_{g}^{\top}\|$$}

\only<3>{$$\hat{\mathbf{u}}_{g}^{\top} = (V^{\top}V)^{-1} V^{\top}\mathbf{x}_{g}^{\top}$$}

## PCA: What is a projection matrix?

Plugging

$$\hat{\mathbf{u}}_{g}^{\top} = (V^{\top}V)^{-1} V^{\top}\mathbf{x}_{g}^{\top}$$

into the following prediction model:

\begin{eqnarray*}
\hat{\mathbf{x}}_{g}^{\top} &=& V \hat{\mathbf{u}}_{g}^{\top} \\
\only<2>{
&=& V (V^{\top}V)^{-1} V^{\top}\mathbf{x}_{g}^{\top} \\ }
\only<3>{
\hat{\mathbf{x}}_{g} &=& \mathbf{x}_{g} \underbrace{V (V^{\top}V)^{-1} V^{\top}}_{\textsf{$n{\times}n$ projection matrix}} }
\end{eqnarray*}

## PCA: Projection can be done on the other side

A multivariate regression model for sample (column) $j$:

$$\mathbf{x}_{j} \sim U \mathbf{v}_{j} + \epsilon$$

The least-square solution for the $V$:

$$\hat{\mathbf{v}}_{j} = (U^{\top}U)^{-1} U^{\top} \mathbf{x}_{j}$$

Then we have

$$\hat{\mathbf{x}}_{j} = \underbrace{U (U^{\top}U)^{-1} U^{\top}}_{\textsf{\color{red} $p{\times}p$ projection matrix}} \mathbf{x}_{j}$$

## PCA: maximizing total variance of the projected

::: {.block}

### Constrained optimization

Given the rank-1 factorization
$$\hat{X} = \mathbf{u} \mathbf{v}^{\top}$$
and
the least square solution $\hat{\mathbf{v}}$
$$\hat{\mathbf{v}}^{\top} = \frac{\mathbf{u}^{\top}}{\mathbf{u}^{\top}\mathbf{u}} X$$

We want to maximize the variance of this projected vector $\mathbf{v}$
$$\hat{\mathbf{v}}^{\top} \hat{\mathbf{v}}
= \frac{\mathbf{u}^{\top}}{\mathbf{u}^{\top}\mathbf{u}} X
X^{\top} \frac{\mathbf{u}}{\mathbf{u}^{\top}\mathbf{u}}
= \mathbf{u}^{\top}XX^{\top}\mathbf{u}$$
where we assume $\mathbf{u}$ is a unit vector.

:::

## PCA is an eigen value problem

:::::: {.columns}
::: {.column width=.45}

::: {.block}
### PCA

Letting the covariance matrix $\hat{\Sigma} = XX^{\top}/(p-1)$, we want to find a unit vector $\mathbf{v}$ by

$$\max \mathbf{v}^{\top} \hat{\Sigma}\mathbf{v}$$

subject to $\mathbf{v}^{\top}\mathbf{v} = 1$.

:::

:::
::: {.column width=.45}

::: {.block}
### Eigen value problem

Given the covariance matrix $\hat{\Sigma}$, we can resolve an eigen-value $\lambda$ and the corresponding eigen-vector $\mathbf{v}$ such that

$$\hat{\Sigma} \mathbf{v} = \lambda \mathbf{v}$$

:::

:::
::::::


## SVD: another equivalent method for PCA

:::::: {.columns}
::: {.column width=.45}

::: {.block}
### Singular Value Decomposition

SVD identifies three matrices of $X$:

$$X = U D V^{\top}$$

where both $U$ and $V$ vectors are orthonormal,
namely,

- $U^{\top}U = I$, $\mathbf{u}_{k}^{\top}\mathbf{u}_{k}=1$ for all $k$,

- $V^{\top}V = I$, $\mathbf{v}_{k}^{\top}\mathbf{v}_{k} = 1$ for all $k$.

:::

:::
::: {.column width=.45}

::: {.block}
### Covariance by SVD

Covariance across the columns (samples)

$$X^{\top}X/p = V D^{2} V^{\top}/p$$

Covariance across the rows (genes)

$$XX^{\top}/n = U D^{2} U^{\top}/n$$

:::

*Remark*: The covariance formulas above assume data is centered (mean-subtracted).

:::
::::::

## SVD: another equivalent method for PCA

We can confirm the equivalent relations by multiplying singular vectors to the covariance matrix:

\small

\begin{eqnarray*}
\onslide<1->{
\underbrace{\frac{\left( X^{\top}X \right)}{p-1}}_{\textsf{\color{blue} sample covariance}} \mathbf{v}_{1}
&\propto&
\left(\mathbf{v}_{1}, \mathbf{v}_{2},\ldots, \mathbf{v}_{k}\right)
\left(
\begin{array}{l l l l}
D_{1}^{2} & 0 & \ldots & \ldots \\
0 & D_{2}^{2} & 0 & \ldots \\
0 & \ldots & \ddots & 0 \\
0 & \ldots & 0 & D_{k}^{2} \\
\end{array} \right)
\left(
\begin{array}{l}
\mathbf{v}_{1}^{\top}\\
\mathbf{v}_{2}^{\top}\\
\vdots\\
\mathbf{v}_{k}^{\top}
\end{array}
\right)
\mathbf{v}_{1} \\
}
\only<2>{
&\propto&
\left(\mathbf{v}_{1}, \mathbf{v}_{2},\ldots, \mathbf{v}_{k}\right)
\left(
\begin{array}{l l l l}
D_{1}^{2} & 0 & \ldots & \ldots \\
0 & D_{2}^{2} & 0 & \ldots \\
0 & \ldots & \ddots & 0 \\
0 & \ldots & 0 & D_{k}^{2} \\
\end{array} \right)
\left(
\begin{array}{l}
1 \\
0 \\
\vdots\\
0
\end{array}
\right) \\
}
\only<3>{
&\propto&
\left(\mathbf{v}_{1}, \mathbf{v}_{2},\ldots, \mathbf{v}_{k}\right)
\left(
\begin{array}{l}
D_{1}^{2} \\
0 \\
0 \\
0 \\
\end{array} \right)}
\only<4>{
&=&
\underbrace{\frac{D_{1}^{2}}{p-1}}_{\textsf{\color{red}eigenvalue}}
\underbrace{\mathbf{v}_{1}}_{\textsf{\color{red}eigenvector}}
}
\end{eqnarray*}

## Run SVD to find principal components

```{r echo=TRUE}
svd.out <- rsvd::rsvd(x.sorted, k = 7)
U <- svd.out$u; D <- diag(svd.out$d); V <- svd.out$v
```

:::::: {.columns}
::: {.column width=.25}

```{r echo=FALSE, fig.width=2, fig.height=2.3}
.matshow(pmin(x.sorted, 3), .lab=0, .lw=0) +
    ggtitle("data matrix")
```

:::
::: {.column width=.65}

```{r echo=FALSE, fig.width=3.2, fig.height=2.5}
p0 <- ggplot() + theme_void()

p1 <- .matshow(U, .lab=0, .lw=.0, .scale=TRUE) +
    ggtitle("U")

p.D <- .matshow(D, .lab=0, .lw=.0, .scale=TRUE) +
    ggtitle("D")

p.V <- .matshow(t(V), .lab=0, .lw=.0, .scale=TRUE) +
    ggtitle("transpose(V)")

p.eq <- ggplot() + theme_void() + geom_text(aes(0,0,label="="), size=6)
p.prod <- ggplot() + theme_void() + geom_text(aes(0,0,label="x"), size=5)
p.prod <- wrap_plots(p.prod, p0, ncol=1, heights=c(1,8))

p2 <- wrap_plots(p.D, p0, ncol=1, heights=c(1,8))
p3 <- wrap_plots(p.V, p0, ncol=1, heights=c(1,8))

wrap_plots(p.eq, p0, p1, p.prod, p2, p.prod, p3,
     nrow=1,
     widths=c(1, 1, 1, 1, 1, 1, 8))
```

:::
::::::


## Summary: PCA

\large

* Matrix factorization = alternating linear regressions 

* PCA is a matrix factorization

* PCA can be found by Eigen value/vector problem

* PCA can be found by Singular Value Decomposition

\normalsize

